### 2022-10-18 manu-draft.qmd
quarto manu skeleton added

---
format: 
  docx:
    reference-doc: my_styles.docx
    number-sections: true
title: "EBASE manuscript"
bibliography: refs.bib
csl: limnology-and-oceanography.csl
author: 
  - Marcus W Beck:
      email: mbeck@tbep.org
      institute: [TBEP]
      correspondence: true
  - Jill M Arriola:
      email: jva5648@psu.edu 
      institute: [PSU]
      correspondence: false
  - Maria:
      email: mxh367@psu.edu
      institute: [PSU]
      correspondence: false
  - Raymond G Najjar:
      email: rgn1@psu.edu
      institute: [PSU]
      correspondence: false

institute:
  - TBEP: Tampa Bay Estuary Program, St. Petersburg, Florida 33701 USA
  - PSU: Pennsylvania State University, University Park, Pennsylvania 16802 USA

filters: 
  - templates/scholarly-metadata.lua
  - templates/author-info-blocks.lua

execute: 
  echo: false
  warning: false
  ft.keepnext: false
  ft.align: left
---

```{r, results = 'hide'}
library(knitr)
library(here)
library(flextable)

```

## Abstract {.unnumbered}

*Key words*:

## Introduction

## Materials and Procedures

## Assessment

## Discussion

## Comments and Recommendations

## Acknowledgments {.unnumbered}

\newpage

## Figures {.unnumbered}

\newpage

## Tables {.unnumbered}

\newpage

## References {.unnumbered}

### 2022-10-18 manu-draft.qmd
working on outline

@@
       correspondence: false
-  - Maria:
+  - Maria Hermann:
       email: mxh367@psu.edu
@@
 
+* Overview of metabolism - importance
+* Existing methods: Odum/WtRegDO, BASEmetab, references in @Grace15, applications to lakes and streams
+* Applications to estuaries - challenges
+* Goals and objectives
+
 ## Materials and Procedures
 
+* EBASE theory
+* EBASE R package
+
 ## Assessment
+
+* Comparison with Odum and BASEmetab: Fwoxy Appalachicola
+* Comparison with Odum and BASemetab: Apalachicola
+* Sensitivity analyses:
+     * optimization length
+     * priors
+* Application to representative NERR sites
 
@@
 
-\newpage
+<!-- \newpage -->
 
@@
 
-\newpage
+Figure 1: Fwoxy comparisons
+
+Figure 2: Apalachicola comparisons
+
+Figure 3: Sensitivity analyses
+
+Figure 4: NERRS application
+
+<!-- \newpage  -->
 
@@
 
-\newpage
+Table 1: EBASE parameters and outputs
+
+Table 2: Summary stats of Fwoxy comparison
+
+Table 3: Summary stats of Apalachicola comparison
+
+Table 4: Key parameters form NERRS application
+
+<!-- \newpage -->
 

### 2022-12-15 R/figs.R
comparison of apa obs and dtd figure added

@@ -1,0 +1,233 @@
+library(tidyverse)
+library(lubridate)
+library(here)
+library(patchwork)
+
+load(file = url('https://github.com/fawda123/BASEmetab_script/raw/master/data/apacmp.RData'))
+
+toplo <- apacmp %>% 
+  mutate(
+    val = case_when(
+      var == 'Rt_vol' ~ -1 * val, 
+      T ~ val
+    ), 
+    var = factor(var, 
+                 levels = c('NEM', 'Pg_vol', 'Rt_vol', 'D'), 
+                 labels = c('NEM', 'P', 'R', 'D')
+    ), 
+    seas = case_when(
+      month(Date) %in% c(1:5, 10:12) ~ 'dry', 
+      T ~ 'wet'
+    )
+  ) %>% 
+  pivot_wider(names_from = 'typ', values_from = 'val')
+
+toplo1 <- toplo %>% 
+  filter(dotyp == 'observed')
+toplo1a <- toplo1 %>% 
+  select(-BASEmetab)
+toplo1b <- toplo1 %>% 
+  select(-Odum)
+
+toplo2 <- toplo %>% 
+  filter(dotyp == 'detided') 
+toplo2a <- toplo2 %>% 
+  select(-BASEmetab)
+toplo2b <- toplo2 %>% 
+  select(-Odum)
+
+thm <- theme_minimal() + 
+  theme(
+    strip.background = element_blank(), 
+    legend.position = 'top', 
+    panel.grid.minor = element_blank(), 
+    axis.text = element_text(size = 6)
+  )
+
+alph <- 0.8
+cols <- c('darkgrey', 'black')
+
+# obs do, odum v ebase
+vars <- levels(toplo1a$var)
+for(vr in vars){
+  
+  toplotmp <- toplo1a %>% 
+    filter(var == vr)
+  ind <- which(vr == vars)
+  
+  corv <- cor.test(toplotmp$Odum, toplotmp$EBASE) %>% 
+    .$estimate %>% 
+    round(., 2) %>% 
+    format(., nsmall = 2)
+  
+  toplotmp <- toplotmp %>% 
+    mutate(
+      var = paste0(var, ' (', corv, ')')
+    )
+  
+  ylb <- NULL
+  ttl <- NULL
+  if(ind == 1){
+    ylb <- 'Odum'
+    ttl <- '(a) Observed dissolved oxygen'
+  }
+  
[…]

### 2022-12-16 R/figs.R
apacmp plot uses a function now

@@ -22,212 +22,16 @@ toplo <- apacmp %>%
   ) %>% 
   pivot_wider(names_from = 'typ', values_from = 'val')
 
-toplo1 <- toplo %>% 
-  filter(dotyp == 'observed')
-toplo1a <- toplo1 %>% 
-  select(-BASEmetab)
-toplo1b <- toplo1 %>% 
-  select(-Odum)
+p1a <- apacmp_plo(toplo, xlb = NULL, ylb = 'Odum', dotyp = 'observed', addtitle = T)
+p1b <- apacmp_plo(toplo, xlb = NULL, ylb = 'BASEmetab', dotyp = 'observed', addtitle = F)
+p2a <- apacmp_plo(toplo, xlb = NULL, ylb = 'Odum', dotyp = 'detided', addtitle = T)
+p2b <- apacmp_plo(toplo, xlb = 'EBASE', ylb = 'BASEmetab', dotyp = 'detided', addtitle = F)
 
-toplo2 <- toplo %>% 
-  filter(dotyp == 'detided') 
-toplo2a <- toplo2 %>% 
-  select(-BASEmetab)
-toplo2b <- toplo2 %>% 
-  select(-Odum)
-
-thm <- theme_minimal() + 
-  theme(
-    strip.background = element_blank(), 
-    legend.position = 'top', 
-    panel.grid.minor = element_blank(), 
-    axis.text = element_text(size = 6)
-  )
-
-alph <- 0.8
-cols <- c('darkgrey', 'black')
-
-# obs do, odum v ebase
-vars <- levels(toplo1a$var)
-for(vr in vars){
-  
-  toplotmp <- toplo1a %>% 
-    filter(var == vr)
-  ind <- which(vr == vars)
-  
-  corv <- cor.test(toplotmp$Odum, toplotmp$EBASE) %>% 
-    .$estimate %>% 
-    round(., 2) %>% 
-    format(., nsmall = 2)
-  
-  toplotmp <- toplotmp %>% 
-    mutate(
-      var = paste0(var, ' (', corv, ')')
-    )
-  
-  ylb <- NULL
-  ttl <- NULL
-  if(ind == 1){
-    ylb <- 'Odum'
-    ttl <- '(a) Observed dissolved oxygen'
-  }
-  
-  lims <- range(toplotmp[, c('Odum', 'EBASE')], na.rm = T)
-  
-  ptmp <- ggplot(toplotmp, aes(x = EBASE, y = Odum, colour = seas)) + 
[…]

### 2022-12-16 R/funcs.R

@@ -1,0 +1,82 @@
+# capitalization function
+simpleCap <- function(x) {
+  s <- strsplit(x, " ")[[1]]
+  paste(toupper(substring(s, 1,1)), substring(s, 2),
+        sep="", collapse=" ")
+}
+
+# create subplots for apalachicola obs v dtd metab comparisons
+apacmp_plo <- function(dat, xlb = 'EBASE', ylb = 'Odum', dotyp = 'observed', addtitle = T, alph = 0.8, cols = c('darkgrey', 'black')){
+
+  thm <- theme_minimal() + 
+    theme(
+      strip.background = element_blank(), 
+      legend.position = 'top', 
+      panel.grid.minor = element_blank(), 
+      axis.text = element_text(size = 6)
+    )
+  
+  names(dat)[names(dat) == ylb] <- 'yval'
+  
+  subfigs <- list(
+    observed = '(a)', 
+    detided = '(b)'
+  )
+  
+  dat <- dat %>% 
+    filter(dotyp == !!dotyp) %>% 
+    select(var, seas, yval, EBASE) %>% 
+    na.omit()
+  
+  vars <- levels(dat$var)
+  for(vr in vars){
+    
+    toplotmp <- dat %>% 
+      filter(var == vr)
+    ind <- which(vr == vars)
+    
+    corv <- cor.test(toplotmp$yval, toplotmp$EBASE) %>% 
+      .$estimate %>% 
+      round(., 2) %>% 
+      format(., nsmall = 2)
+    
+    toplotmp <- toplotmp %>% 
+      mutate(
+        var = paste0(var, ' (', corv, ')')
+      )
+    
+    if(ind != 1)
+      ylb <- NULL
+    
+    ttl <- NULL
+    if(addtitle & ind == 1){
+      subfig <- subfigs[[dotyp]]
+      ttl <- paste(subfig, simpleCap(dotyp), 'dissolved oxygen')
+    }
+    
+    lims <- range(toplotmp[, c('yval', 'EBASE')], na.rm = T)
+    
+    ptmp <- ggplot(toplotmp, aes(x = EBASE, y = yval, colour = seas)) + 
+      geom_point(alpha = alph) + 
+      geom_abline(intercept = 0, slope = 1) +
+      facet_wrap(~var, ncol = 1) +
+      scale_colour_manual(values = cols) + 
+      scale_x_continuous(limits = lims) + 
+      scale_y_continuous(limits = lims) + 
[…]

### 2022-12-16 manu-draft.qmd
added some methods text about model comparisons at apa

@@
 * Comparison with Odum and BASEmetab: Fwoxy Appalachicola
-* Comparison with Odum and BASemetab: Apalachicola
+* Comparison with Odum and BASemetab: Apalachicola, @fig-apacmp
+
+Metabolic estimates from EBASE were compared to those from existing methods to benchmark and evaluate potential differences from alternative model formulations used by each approach.
+An approximate one year continuous time series from Apalachicola Bay, Florida was provided as input to each method.
+The input data were 2012 observations from February 22nd to December 19th at the Cat Point water quality monitoring station (29.7021$^{\circ}$ N, -84.8802$^{\circ}$ W) and included 15 minute observations of dissolved oxygen (mg/L), water temperature (C), salinity (psu), and sonde depth (m).
+Continuous meteorological observations from the nearby East Bay station (29.7909$^{\circ}$ N, -84.8834$^{\circ}$ W, ~10km from Cat Point) were combined with the water quality data at the same time step and included air temperature (C), wind speed (m/s), PAR (mmol/m$^2$ as total per 15 minute observation), and barometric pressure (mb).
+The combined input data were provided to each method after converting the units as required (e.g., BASEmetab requires PAR as $\mu$mol m$^{-2}$ s$^{-1}$).
+Water column depth at Cat Point is approximately 1.8 m and all metabolic estimates returned by each method were converted to volumetric rates by dividing by the depth.
+Each comparison also evaluated metabolic estimates using the observed dissolved oxygen time series and a detided time series using weighted regression [@Beck15].
+The latter comparison provided an assessment of metabolic rates that are expected to be minimally influenced by tidal advection.
+All metabolic estimates were obtained using the default setting for each of the three methods, with an exception that the prior distribution for the reaeration coefficient ($K$) in BASEmetab was informed by results from the Odum output.
+Initial assessments indicated that unstable and inaccurate results are obtained from BASEmetab using an uninformed prior distribution for $K$.
+All other prior distributions for BASEmetab and EBASE were set as the software default.
+A one-day model optimization period was used for EBASE.
+
+The common set of metabolic parameters returned by the Odum, BASEmetab, and EBASE methods are net ecosystem metabolism, gross production, respiration, and gas exchange.
+Pairwise comparisons of the daily estimates returned by each method were evaluated using simple summary statistics including Pearson correlation coefficients and linear regression fits to evaluate potential biases in each method.
+For the latter analysis, differences in intercept suggested an overall bias in magnitude, where differences in slope suggested a bias that depended on relative value of each observation (e.g., bias may differ between lower or higher values).
+Comparisons were made for the whole year and by wet (June, July, August, September) or dry (all other months) season observations.
+Results were also assessed separately for estimates based on the observed or detided disssolved oxygen time series.
+
 * Sensitivity analyses:
@@
 
+\newpage
+
+![Comparisons of daily metabolic estimates from the Odum, BASEmetab, and EBASE methods for one year of continuous monitoring data at Apalachicola Bay.
+Results for net ecosystem metabolism (NEM), gross production (P), respiration (R), and gas exchange (D) are compared using the (a) observed and (b) detided dissolved oxygen time series as input to each model.
+All values are volumetric as mmol O$_2$ m$^3$ d$^{-1}$ and are grouped by the dry or wet season.
+Lines show the 1:1 comparison and the the Pearson correlation values are shown in parentheses for each facet title.](figs/apacmp.png){#fig-apacmp fig-align="center"}
+
 <!-- \newpage  -->

### 2022-12-16 R/figs.R

@@ -3,6 +3,8 @@ library(lubridate)
 library(here)
 library(patchwork)
 
+source(here('R/funcs.R'))
+
 load(file = url('https://github.com/fawda123/BASEmetab_script/raw/master/data/apacmp.RData'))
 
 toplo <- apacmp %>% 
@@ -32,6 +34,6 @@ p <- p1a[[1]] + p1a[[2]] + p1a[[3]] + p1a[[4]] + p1b[[1]] + p1b[[2]] + p1b[[3]] 
   plot_layout(ncol = 4, guides = 'collect') & 
   theme(legend.position = 'bottom')
 
-png(here('figs/apacmp2.png'), height = 9, width = 8, family = 'serif', units = 'in', res = 500)
+png(here('figs/apacmp.png'), height = 9, width = 8, family = 'serif', units = 'in', res = 500)
 print(p)
 dev.off()

### 2022-12-16 manu-draft.qmd
working on summary table for apacmp

@@
 
+load(file = here('tabs/apacmptab.RData'))
 ```
@@
 * Comparison with Odum and BASEmetab: Fwoxy Appalachicola
-* Comparison with Odum and BASemetab: Apalachicola, @fig-apacmp
+* Comparison with Odum and BASemetab: Apalachicola, @fig-apacmpfig
 
@@
 Pairwise comparisons of the daily estimates returned by each method were evaluated using simple summary statistics including Pearson correlation coefficients and linear regression fits to evaluate potential biases in each method.
-For the latter analysis, differences in intercept suggested an overall bias in magnitude, where differences in slope suggested a bias that depended on relative value of each observation (e.g., bias may differ between lower or higher values).
+For the latter analysis, differences in intercept suggested an overall bias in magnitude, whereas the slope was further tested to determine if the estimate was significantly different than one, suggesting a bias that depended on relative value of each observation (e.g., bias may differ between lower or higher values).
 Comparisons were made for the whole year and by wet (June, July, August, September) or dry (all other months) season observations.
@@
 All values are volumetric as mmol O$_2$ m$^3$ d$^{-1}$ and are grouped by the dry or wet season.
-Lines show the 1:1 comparison and the the Pearson correlation values are shown in parentheses for each facet title.](figs/apacmp.png){#fig-apacmp fig-align="center"}
+Black lines show the 1:1 comparison, the blue line shows the linea regression fit, and the Pearson correlation values are shown in parentheses for each facet title.](figs/apacmpfig.png){#fig-apacmpfig fig-align="center"}
 
@@
 
+\newpage
+
+```{r}
+#| tbl-cap: "Table 1: "  
+# apacmptab
+```
+
 <!-- \newpage -->

### 2022-12-16 R/figs.R

@@ -34,6 +34,6 @@ p <- p1a[[1]] + p1a[[2]] + p1a[[3]] + p1a[[4]] + p1b[[1]] + p1b[[2]] + p1b[[3]] 
   plot_layout(ncol = 4, guides = 'collect') & 
   theme(legend.position = 'bottom')
 
-png(here('figs/apacmp.png'), height = 9, width = 8, family = 'serif', units = 'in', res = 500)
+png(here('figs/apacmpfig.png'), height = 9, width = 8, family = 'serif', units = 'in', res = 500)
 print(p)
 dev.off()

### 2022-12-16 R/funcs.R

@@ -1,3 +1,16 @@
+# function for formatting p-values in tables
+p_ast <- function(x){
+  
+  sig_cats <- c('**', '*', '')
+  sig_vals <- c(-Inf, 0.005, 0.05, Inf)
+  
+  out <- cut(x, breaks = sig_vals, labels = sig_cats, right = FALSE)
+  out <- as.character(out)
+  
+  return(out)
+  
+}
+
 # capitalization function
 simpleCap <- function(x) {
   s <- strsplit(x, " ")[[1]]
@@ -56,9 +69,10 @@ apacmp_plo <- function(dat, xlb = 'EBASE', ylb = 'Odum', dotyp = 'observed', add
     
     lims <- range(toplotmp[, c('yval', 'EBASE')], na.rm = T)
     
-    ptmp <- ggplot(toplotmp, aes(x = EBASE, y = yval, colour = seas)) + 
-      geom_point(alpha = alph) + 
+    ptmp <- ggplot(toplotmp, aes(x = EBASE, y = yval)) + 
+      geom_point(aes(colour = seas), alpha = alph) + 
       geom_abline(intercept = 0, slope = 1) +
+      geom_smooth(method = 'lm', se = F, colour = 'tomato1', size = 0.7) +
       facet_wrap(~var, ncol = 1) +
       scale_colour_manual(values = cols) + 
       scale_x_continuous(limits = lims) + 

### 2022-12-16 R/tabs.R

@@ -1,0 +1,85 @@
+library(flextable)
+library(tidyverse)
+library(here)
+
+source(here('R/funcs.R'))
+
+load(file = here('data/apacmp.RData'))
+
+
+tosum <- apacmp %>% 
+  mutate(
+    val = case_when(
+      var == 'Rt_vol' ~ -1 * val, 
+      T ~ val
+    ), 
+    var = factor(var, 
+                 levels = c('NEM', 'Pg_vol', 'Rt_vol', 'D'), 
+                 labels = c('NEM', 'P', 'R', 'D')
+    ), 
+    seas = case_when(
+      month(Date) %in% c(1:5, 10:12) ~ 'dry', 
+      T ~ 'wet'
+    )
+  ) %>% 
+  pivot_wider(names_from = 'typ', values_from = 'val')
+
+grd <- crossing(
+  dotyp = unique(tosum$dotyp), 
+  var = levels(tosum$var), 
+  seas = c('dry', 'wet', 'all'), 
+  comp = c('Odum v EBASE', 'BASEmetab v EBASE')
+  ) %>% 
+  mutate(
+    corv = NA, 
+    int = NA,
+    slo = NA, 
+    rse = NA
+  )
+
+for(i in 1:nrow(grd)){
+  
+  cat(i, '\t')
+  
+  tosel <- grd[i, ]
+               
+  dotyp <- grd[i, ]$dotyp
+  var <- grd[i, ]$var
+  seas <- grd[i, ]$seas
+  comp <- grd[i, ]$comp
+  
+  if(seas == 'all')
+    seas <- c('dry', 'wet')
+  
+  comp <- strsplit(comp, ' v ')[[1]]
+  
+  # filter data
+  tocmp <- tosum %>% 
+    filter(dotyp == !!dotyp) %>% 
+    filter(var == !!var) %>% 
+    filter(seas %in% !!seas) %>% 
+    rename(
+      yval = !!comp[1], 
+      xval = !!comp[2]
+    )
+  
+  # get summaries
+  corv <- cor.test(tocmp$xval, tocmp$yval)
+  corv <- paste0(round(corv$estimate, 2), p_ast(corv$p.value))
+  
+  lmmod <- lm(yval ~ xval, data = tocmp)
+  int <- paste0(round(summary(lmmod)$coefficients[1, 1], 2), p_ast(summary(lmmod)$coefficients[1, 4]))
[…]

### 2022-12-19 manu-draft.qmd
finished apa cmp table

@@
 library(here)
+library(officer)
+library(officedown)
 library(flextable)
@@
 Pairwise comparisons of the daily estimates returned by each method were evaluated using simple summary statistics including Pearson correlation coefficients and linear regression fits to evaluate potential biases in each method.
-For the latter analysis, differences in intercept suggested an overall bias in magnitude, whereas the slope was further tested to determine if the estimate was significantly different than one, suggesting a bias that depended on relative value of each observation (e.g., bias may differ between lower or higher values).
-Comparisons were made for the whole year and by wet (June, July, August, September) or dry (all other months) season observations.
-Results were also assessed separately for estimates based on the observed or detided disssolved oxygen time series.
+For the latter analysis, differences in intercept estimates less than or greater than zero suggested a bias in magnitude, whereas differences in slope estimates less than or greater than one suggested a bias that depended on relative value of each observation (e.g., bias may differ between lower or higher values).
+Comparisons of the slope and estimates values to zero and one, respectively, were based on confidence intervals from the standard errors of each estimate.
+Results were assessed separately for estimates based on the observed or detided dissolved oxygen time series.
 
@@
 
+`r run_reference(id = "apacmptab")`
+
 ```{r}
-#| tbl-cap: "Table 1: "  
-# apacmptab
+#| label: apacmptab
+#| tbl-cap: "Summary statisics of daily metabolic estimates comparing Odum and BASEmetab to EBASE for net ecosystem metabolism (NEM), gross production (P), respiration (R), and gas exchange (D).
+Separate comparisons are made for estimates based on observed and detided dissolved oxygen.
+Summary statistics include Pearson correlation coefficients and intercept, slope, and residual standard deviation of a linear fit between the same metabolic estimates obtained from each method.
+Significance of the p-values for the intercept and slope indicates if the estimate is different from zero or one, respectively."
+apacmptab
 ```

### 2022-12-19 R/funcs.R

@@ -47,17 +47,7 @@ apacmp_plo <- function(dat, xlb = 'EBASE', ylb = 'Odum', dotyp = 'observed', add
     toplotmp <- dat %>% 
       filter(var == vr)
     ind <- which(vr == vars)
-    
-    corv <- cor.test(toplotmp$yval, toplotmp$EBASE) %>% 
-      .$estimate %>% 
-      round(., 2) %>% 
-      format(., nsmall = 2)
-    
-    toplotmp <- toplotmp %>% 
-      mutate(
-        var = paste0(var, ' (', corv, ')')
-      )
-    
+  
     if(ind != 1)
       ylb <- NULL
     

### 2022-12-19 R/tabs.R

@@ -1,11 +1,9 @@
 library(flextable)
 library(tidyverse)
+library(lubridate)
 library(here)
 
-source(here('R/funcs.R'))
-
-load(file = here('data/apacmp.RData'))
-
+load(file = url('https://github.com/fawda123/BASEmetab_script/raw/master/data/apacmp.RData'))
 
 tosum <- apacmp %>% 
   mutate(
@@ -16,10 +14,6 @@ tosum <- apacmp %>%
     var = factor(var, 
                  levels = c('NEM', 'Pg_vol', 'Rt_vol', 'D'), 
                  labels = c('NEM', 'P', 'R', 'D')
-    ), 
-    seas = case_when(
-      month(Date) %in% c(1:5, 10:12) ~ 'dry', 
-      T ~ 'wet'
     )
   ) %>% 
   pivot_wider(names_from = 'typ', values_from = 'val')
@@ -27,13 +21,14 @@ tosum <- apacmp %>%
 grd <- crossing(
   dotyp = unique(tosum$dotyp), 
   var = levels(tosum$var), 
-  seas = c('dry', 'wet', 'all'), 
   comp = c('Odum v EBASE', 'BASEmetab v EBASE')
   ) %>% 
   mutate(
     corv = NA, 
     int = NA,
     slo = NA, 
+    slolo = NA,
+    slohi = NA,
     rse = NA
   )
 
@@ -45,11 +40,7 @@ for(i in 1:nrow(grd)){
                
   dotyp <- grd[i, ]$dotyp
   var <- grd[i, ]$var
-  seas <- grd[i, ]$seas
   comp <- grd[i, ]$comp
-  
-  if(seas == 'all')
-    seas <- c('dry', 'wet')
   
   comp <- strsplit(comp, ' v ')[[1]]
   
@@ -57,7 +48,6 @@ for(i in 1:nrow(grd)){
   tocmp <- tosum %>% 
     filter(dotyp == !!dotyp) %>% 
     filter(var == !!var) %>% 
-    filter(seas %in% !!seas) %>% 
     rename(
       yval = !!comp[1], 
       xval = !!comp[2]
@@ -66,20 +56,60 @@ for(i in 1:nrow(grd)){
   # get summaries
   corv <- cor.test(tocmp$xval, tocmp$yval)
   corv <- paste0(round(corv$estimate, 2), p_ast(corv$p.value))
-  
+
   lmmod <- lm(yval ~ xval, data = tocmp)
[…]

### 2022-12-19 manu-draft.qmd
working on sensitivity analysis text

@@
 
+### EBASE comparison with modelled observations
+
 * Comparison with Odum and BASEmetab: Fwoxy Appalachicola
-* Comparison with Odum and BASemetab: Apalachicola, @fig-apacmpfig
+
+### EBASE comparison with real observations
 
@@
 The input data were 2012 observations from February 22nd to December 19th at the Cat Point water quality monitoring station (29.7021$^{\circ}$ N, -84.8802$^{\circ}$ W) and included 15 minute observations of dissolved oxygen (mg/L), water temperature (C), salinity (psu), and sonde depth (m).
-Continuous meteorological observations from the nearby East Bay station (29.7909$^{\circ}$ N, -84.8834$^{\circ}$ W, ~10km from Cat Point) were combined with the water quality data at the same time step and included air temperature (C), wind speed (m/s), PAR (mmol/m$^2$ as total per 15 minute observation), and barometric pressure (mb).
+Continuous meteorological observations from the nearby East Bay station (29.7909$^{\circ}$ N, -84.8834$^{\circ}$ W, ~10km north of Cat Point) were combined with the water quality data at the same time step and included air temperature (C), wind speed (m/s), PAR (mmol/m$^2$ as total per 15 minute observation), and barometric pressure (mb).
 The combined input data were provided to each method after converting the units as required (e.g., BASEmetab requires PAR as $\mu$mol m$^{-2}$ s$^{-1}$).
-Water column depth at Cat Point is approximately 1.8 m and all metabolic estimates returned by each method were converted to volumetric rates by dividing by the depth.
+Water column depth at Cat Point is approximately 1.8 m and all daily metabolic estimates returned by each method were converted to volumetric rates by dividing by the depth ($\mu$mol O$_2$ m$^{-3}$ d$^{-1}$).
 Each comparison also evaluated metabolic estimates using the observed dissolved oxygen time series and a detided time series using weighted regression [@Beck15].
 The latter comparison provided an assessment of metabolic rates that are expected to be minimally influenced by tidal advection.
-All metabolic estimates were obtained using the default setting for each of the three methods, with an exception that the prior distribution for the reaeration coefficient ($K$) in BASEmetab was informed by results from the Odum output.
+All metabolic estimates were obtained using the default settings for each of the three methods, with an exception that the prior distribution for the reaeration coefficient ($K$) in BASEmetab was informed by results from the Odum output.
 Initial assessments indicated that unstable and inaccurate results are obtained from BASEmetab using an uninformed prior distribution for $K$.
@@
 
-* Sensitivity analyses:
-     * optimization length
-     * priors
-* Application to representative NERR sites
+@fig-apacmpfig and `r run_reference(id = "apacmptab")` provide a comparison of the EBASE metabolic estimates with those from the Odum and BASmetab methods.
+
+### Sensitivity analyses
+
+Several configurations of the EBASE model can influence how the metabolic rates are estimated.
+A simulated time series with known values for the $a$, $r$, and $b$ parameters that are returned by EBASE was used to determine if 1) EBASE estimates similar values, and 2) how the estimated values change with different model configurations.
+Two types of sensitivity analyses were conducted to evaluate changes in the parameters returned by EBASE, where in all cases, the parameters were compared to the known values in the simulated time series.
+First, parameter estimates were compared to the known values after changing characteristics of their prior distributions.
+Second, parameter estimates were compared using a model period of optimization for one day and one week of observations.
+Details of these analyses are as follows.
+
+The standard prior distributions for the EBASE parameters follow normal distributions with defined means and standard deviations.
+The sensitivity analysis evaluated changes in the distributions by varying the standard deviations from small to large to test the effect of more or less constrained ranges, respectively, on the priors.
+For each parameter, low, medium, and high values for the standard deviation were assessed as 10%, 100%, and 1000% of the default value.
+Each unique combination of low, medium, and high values for the standard deviation of each parameter were evaluated, creating 3$^3$ = 27 different combinations of prior distributions that were used with the simulated time series in EBASE.
+Because initial assessments indicated that all parameter estimates were especially sensitive to changes in the prior distribution for the $b$ parameter, a second sensitivity analysis evaluated additional changes in both the mean and standard deviation of the prior for the $b$ parameter.
+Changes in both the mean and standard deviation of the $b$ prior were evaluated, where the mean was varied as 50%, 100%, and 200% (the upper limit) of the default and the standard deviation was varied as before.
+This resulted in 3$^2$ = 9 unique combinations of means and standard deviations for the $b$ prior that were evaluated.
+
+The optimization period used for a time series determines how much data are used to estimate the three EBASE parameters ($a$, $r$, and $b$).
+Unlike BASEmetab, the EBASE software allows flexibility in the period of time used for optimization, with a minimum period of time being one day and a maximum period of being the length of the time series.
+For a chosen optimization period, a single estimate for $r$ and $b$ is returned by the model, whereas $n$ number of observations in each optimization period are returned for the $a$ parameter (e.g., 96 values for  24 hour time period if the samplign interval is 15 minutes).
+Short periods of observation may return unstable parameter estimates if forcing factors (e.g., wind, temperature) that affect estuarine biogeochemical rates occur at time scales longer the chosen period of optimization for EBASE.
+As such, potentially longer optimization periods may return more stable and robust estimates given that more data are used for parameter estimation.
+An optimization period of one day was compared to results using an optimization period of seven days for the simulated one-year time series.
+Each optimization period was evaluated for every unique combination of prior distributions, totaling 54 unique analyses for the assessment of changing standard deviations of all parameters and 18 unique analyses for the assessment of changing the mean and standard deviation of only the $b$ parameter.
+
+The metabolic estimates returned from each analysis were compared to the known values to determine which priors the model was most sensitive to and if the conclusions varied by optimization period.
+Estimates for the modeled dissolved oxygen, gross production, respiration ($r$), gas exchange ($D$), and $a$ were compared to those from the simulated time series using coefficient of variation ($R^2$) and root mean square error ($RMSE$) summaries of the 1:1 comparison of the estimated vs simulated values.
+For the one day optimization period, all parameter estimates returned at the time step of the input data (e.g., $a$) were daily-averaged for comparison.
+For the seven day optimization period, all parameter estimates and those from the known time series were weekly-averaged.
+Estimates for the $b$ parameter were evaluated only by visual comparison because a single value was used in the simulated time series, preventing the estimate of quantitative comparisons.
+
+### Application to represenatiave NERR sites
 
@@
 
-`r run_reference(id = "apacmptab")`
-
 ```{r}

### 2022-12-20 manu-draft.qmd
working on ebase theory

@@
 
+The Estuarine BAyesian Single station Estimation (EBASE) method provides a simple approach to quantify metabolic parameters from continuous dissolved oxygen time series.
+The estimates are based on a mass balance equation described in @Grace15 that is modified using a gas exchange calculation described in @Wanninkhof14 (@eq-ebase).
+
+$$
+\frac{\delta{C_d}}{\delta{t}} = [\,aPAR]\, - [\,r]\, - \frac{1}{H}\left[-bU_{10}^2\left(\frac{s_c}{600} \right)^{-0.5} \left(C_{Sat} - C_d \right )\right]
+$$ {#eq-ebase}
+
+The metabolic estimates are defined by the change in dissolved oxygen over the time step $\frac{\delta{C_d}}{\delta{t}}$, where gross production is provided by $aPAR$, respiration is provided by $r$, and gas exchange is provided by the remainder [as in @Wanninkhof14].
+Required data for the equation are dissolved oxygen concentration as $C_d$ (mmol/m$^3$), solar radiation as $PAR$ (W/m$^2$/s), water column depth as $H$ (m), and wind speed at 10 m height as $U$ (m/s).
+The schmidt number $s_c$  is the ratio of the kinematic viscosity of water to the molecular diffusivity of oxygen calculated as a function of water tempererature and salinity using the polynomial wit of @Wanninkhof14.
+
+Dissolved oxygen at saturation $C_{Sat}$ is calculated from the input data using @Garcia92 as a function of water temperature (C, $T$) and salinity (psu, $sal$).
+@eq-ts is temperature scaled, as input to @eq-co that calculates the solubility of dissolved oxygen per volume of seawater at the given temperature.
+The coefficients are A$_0$ = 5.80818, A$_1$ = 3.20684, A$_2$ = 4.11890, A$_3$ = 4.93845, A$_4$ = 1.01567, A$_5$ = 1.41575, B$_0$ = -7.01211e-3, B$_1$ = -7.25958e-3, B$_2$ = -7.93334e-3, B$_3$ = -5.54491e-3, and C$_0$ = -1.32412e-7.
+
+$$
+T_s = \log((298.15 - T)/(273.15 + T)
+$$ {#eq-ts}
+$$
+C_O = A_0 + A_1 \cdot T_s + A_2 \cdot T_s^2 + A_3 \cdot T_s^3 + A_4 \cdot T_s^4 + A_5 \cdot T_s^5 + sal \cdot (B_0 + B_1 \cdot T_s + B_2 \cdot T_s^2 + B_3 \cdot T_s^3) + C_0 \cdot sal^2
+$$ {#eq-co}
+
+Dissolved oxygen saturation is then calculated by multiplying $C_o$ by the density of seawater given temperature and salinity (@eq-csat).
+The parameter $\rho$ (kg/m$^3$) is seawater density at zero elevation given temperature and salinity [@Millero81].
+
+$$
+C_{sat} = \exp(C_O) \cdot \rho_{T, sal} \cdot 1e-3
+$$ {#eq-csat}
+
+The remaining three parameters $a$, $r$, and $b$ are estimated by likelihood given the observed data with the JAGS model using prior distributions shown in the model file.
+At each time step, the change in oxygen concentration between time steps is calculated from the equation using model inputs and parameter guesses, and then a finite difference approximation is used to estimate modeled oxygen concentration.
+The estimated concentration is also returned, which can be compared to observed as one measure of model performance.
+
 * EBASE theory
@@
 * Comparison with Odum and BASEmetab: Fwoxy Appalachicola
+
+### Sensitivity analyses
+
+Several configurations of the EBASE model can influence how the metabolic rates are estimated.
+A simulated time series with known values for the $a$, $r$, and $b$ parameters that are returned by EBASE was used to determine if 1) EBASE estimates similar values, and 2) how the estimated values change with different model configurations.
+Two types of sensitivity analyses were conducted to evaluate changes in the parameters returned by EBASE, where in all cases, the parameters were compared to the known values in the simulated time series.
+First, parameter estimates were compared to the known values after changing characteristics of their prior distributions.
+Second, parameter estimates were compared using a model period of optimization for one day and one week of observations.
+Details of these analyses are as follows.
+
+The standard prior distributions for the EBASE parameters follow normal distributions with defined means and standard deviations.
+The sensitivity analysis evaluated changes in the distributions by varying the standard deviations from small to large to test the effect of more or less constrained ranges, respectively, on the priors.
+For each parameter, low, medium, and high values for the standard deviation were assessed as 10%, 100%, and 1000% of the default values.
+Each unique combination of low, medium, and high values for the standard deviation of each parameter were evaluated, creating 3$^3$ = 27 different combinations of prior distributions that were used with the simulated time series in EBASE.
+Because initial assessments indicated that all parameter estimates were especially sensitive to changes in the prior distribution for the $b$ parameter, a second sensitivity analysis evaluated additional changes in both the mean and standard deviation of the prior.
+The mean of the $b$ prior was varied as 50%, 100%, and 200% (the upper limit used in the software) of the default and the standard deviation was varied as before.
+This resulted in 3$^2$ = 9 unique combinations of means and standard deviations for the $b$ prior that were evaluated.
+
+The optimization period used for a time series determines how much data are used to estimate the three EBASE parameters ($a$, $r$, and $b$).
+Unlike BASEmetab, the EBASE software allows flexibility in the period of time used for optimization, with a minimum period of time being one day and a maximum period of being the length of the time series.
+For a chosen optimization period, a single estimate for $r$ and $b$ is returned by the model, whereas $n$ number of observations equal to the number of time steps in each optimization period are returned for the $a$ parameter (e.g., 96 values for  24 hour time period if the sampling interval is 15 minutes).
+Short periods of observation may return unstable parameter estimates if forcing factors (e.g., wind, temperature) that affect estuarine biogeochemical rates occur at time scales longer the chosen period of optimization for EBASE.
+As such, potentially longer optimization periods may return more stable and robust estimates given that more data are used for parameter estimation.
+An optimization period of one day was compared to results using an optimization period of seven days for the simulated one-year time series.
+Each optimization period was evaluated for every unique combination of prior distributions, totaling 54 unique analyses for the assessment of changing standard deviations of all parameters and 18 unique analyses for the assessment of changing the mean and standard deviation of only the $b$ parameter.
+
+The metabolic estimates returned from each analysis were compared to the known values to identify priors the model was most sensitive to and if the conclusions varied by optimization period.
+Estimates for the modeled dissolved oxygen, gross production, respiration ($r$), gas exchange ($D$), and $a$ were compared to those from the simulated time series using coefficient of variation ($R^2$) and root mean square error ($RMSE$) summaries of the 1:1 comparison of the estimated vs simulated values.
+For the one day optimization period, all parameter estimates returned at the time step of the input data (i.e., modeled dissoled oxygen, $D$, and $a$) were daily-averaged for comparison.
+For the seven day optimization period, all parameter estimates and those from the known time series were weekly-averaged.
+Estimates for the $b$ parameter were evaluated only by visual comparison because a single value was used in the simulated time series, preventing the estimate of quantitative comparisons.
 
@@
 @fig-apacmpfig and `r run_reference(id = "apacmptab")` provide a comparison of the EBASE metabolic estimates with those from the Odum and BASmetab methods.
-
-### Sensitivity analyses
-
-Several configurations of the EBASE model can influence how the metabolic rates are estimated.
-A simulated time series with known values for the $a$, $r$, and $b$ parameters that are returned by EBASE was used to determine if 1) EBASE estimates similar values, and 2) how the estimated values change with different model configurations.
-Two types of sensitivity analyses were conducted to evaluate changes in the parameters returned by EBASE, where in all cases, the parameters were compared to the known values in the simulated time series.
-First, parameter estimates were compared to the known values after changing characteristics of their prior distributions.
-Second, parameter estimates were compared using a model period of optimization for one day and one week of observations.
-Details of these analyses are as follows.
-
-The standard prior distributions for the EBASE parameters follow normal distributions with defined means and standard deviations.
-The sensitivity analysis evaluated changes in the distributions by varying the standard deviations from small to large to test the effect of more or less constrained ranges, respectively, on the priors.
-For each parameter, low, medium, and high values for the standard deviation were assessed as 10%, 100%, and 1000% of the default value.
-Each unique combination of low, medium, and high values for the standard deviation of each parameter were evaluated, creating 3$^3$ = 27 different combinations of prior distributions that were used with the simulated time series in EBASE.
-Because initial assessments indicated that all parameter estimates were especially sensitive to changes in the prior distribution for the $b$ parameter, a second sensitivity analysis evaluated additional changes in both the mean and standard deviation of the prior for the $b$ parameter.
-Changes in both the mean and standard deviation of the $b$ prior were evaluated, where the mean was varied as 50%, 100%, and 200% (the upper limit) of the default and the standard deviation was varied as before.
-This resulted in 3$^2$ = 9 unique combinations of means and standard deviations for the $b$ prior that were evaluated.
-
-The optimization period used for a time series determines how much data are used to estimate the three EBASE parameters ($a$, $r$, and $b$).
-Unlike BASEmetab, the EBASE software allows flexibility in the period of time used for optimization, with a minimum period of time being one day and a maximum period of being the length of the time series.
-For a chosen optimization period, a single estimate for $r$ and $b$ is returned by the model, whereas $n$ number of observations in each optimization period are returned for the $a$ parameter (e.g., 96 values for  24 hour time period if the samplign interval is 15 minutes).
-Short periods of observation may return unstable parameter estimates if forcing factors (e.g., wind, temperature) that affect estuarine biogeochemical rates occur at time scales longer the chosen period of optimization for EBASE.
-As such, potentially longer optimization periods may return more stable and robust estimates given that more data are used for parameter estimation.
-An optimization period of one day was compared to results using an optimization period of seven days for the simulated one-year time series.
-Each optimization period was evaluated for every unique combination of prior distributions, totaling 54 unique analyses for the assessment of changing standard deviations of all parameters and 18 unique analyses for the assessment of changing the mean and standard deviation of only the $b$ parameter.
-
-The metabolic estimates returned from each analysis were compared to the known values to determine which priors the model was most sensitive to and if the conclusions varied by optimization period.
-Estimates for the modeled dissolved oxygen, gross production, respiration ($r$), gas exchange ($D$), and $a$ were compared to those from the simulated time series using coefficient of variation ($R^2$) and root mean square error ($RMSE$) summaries of the 1:1 comparison of the estimated vs simulated values.
-For the one day optimization period, all parameter estimates returned at the time step of the input data (e.g., $a$) were daily-averaged for comparison.
-For the seven day optimization period, all parameter estimates and those from the known time series were weekly-averaged.
-Estimates for the $b$ parameter were evaluated only by visual comparison because a single value was used in the simulated time series, preventing the estimate of quantitative comparisons.
 

### 2022-12-23 manu-draft.qmd
almost done with ebase theory, added prior plot figure

@@
 
+\newcommand{\auni}{\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)}
+\newcommand{\runi}{\text{mmol}/\text{m}^3/\text{d}}
+\newcommand{\buni}{\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)}
+
 ## Abstract {.unnumbered}
@@
 The Estuarine BAyesian Single station Estimation (EBASE) method provides a simple approach to quantify metabolic parameters from continuous dissolved oxygen time series.
-The estimates are based on a mass balance equation described in @Grace15 that is modified using a gas exchange calculation described in @Wanninkhof14 (@eq-ebase).
+The estimates are based on a modified version of the mass balance equation described in @Grace15.
+The original mass balance equation is:
 
 $$
-\frac{\delta{C_d}}{\delta{t}} = [\,aPAR]\, - [\,r]\, - \frac{1}{H}\left[-bU_{10}^2\left(\frac{s_c}{600} \right)^{-0.5} \left(C_{Sat} - C_d \right )\right]
-$$ {#eq-ebase}
+\frac{\delta{C_d}}{\delta{t}} = [\,aPAR^p]\, - [\,r\Theta^{T - \bar{T}}]\, - \left[-K\cdot1.0241^{T-\bar{T}}\left(C_{sat} - C_d \right )\right]
+$$ {#eq-base}
 
-The metabolic estimates are defined by the change in dissolved oxygen over the time step $\frac{\delta{C_d}}{\delta{t}}$, where gross production is provided by $aPAR$, respiration is provided by $r$, and gas exchange is provided by the remainder [as in @Wanninkhof14].
-Required data for the equation are dissolved oxygen concentration as $C_d$ (mmol/m$^3$), solar radiation as $PAR$ (W/m$^2$/s), water column depth as $H$ (m), and wind speed at 10 m height as $U$ (m/s).
-The schmidt number $s_c$  is the ratio of the kinematic viscosity of water to the molecular diffusivity of oxygen calculated as a function of water tempererature and salinity using the polynomial wit of @Wanninkhof14.
-
-Dissolved oxygen at saturation $C_{Sat}$ is calculated from the input data using @Garcia92 as a function of water temperature (C, $T$) and salinity (psu, $sal$).
+where the terms in the square brackets are production (P), respiration (R), and gas exchange (D) as a function of the change in dissolved oxygen ($\delta{C_d}$) per unit time of the input data ($\delta{t}$).
+Five unknown parameters are estimated by the model: $a$, $p$, $r$, $\Theta$, and $K$.
+The required inputs are the time series of diel oxygen ($C_d$, mmol/m$^2$), surface irradiance ($PAR$, $\mu$mol/m$^2$/s), water temperature ($T$, C), and oxygen saturation ($C_{sat}$, mmol/m$^3$).
+Dissolved oxygen at saturation $C_{sat}$ is calculated from the input data as a function of water temperature and salinity (psu) [@Garcia92].
 @eq-ts is temperature scaled, as input to @eq-co that calculates the solubility of dissolved oxygen per volume of seawater at the given temperature.
-The coefficients are A$_0$ = 5.80818, A$_1$ = 3.20684, A$_2$ = 4.11890, A$_3$ = 4.93845, A$_4$ = 1.01567, A$_5$ = 1.41575, B$_0$ = -7.01211e-3, B$_1$ = -7.25958e-3, B$_2$ = -7.93334e-3, B$_3$ = -5.54491e-3, and C$_0$ = -1.32412e-7.
+The coefficients are $A_0$ = 5.80818, $A_1$ = 3.20684, $A_2$ = 4.11890, $A_3$ = 4.93845, $A_4$ = 1.01567, $A_5$ = 1.41575, $B_0$ = -7.01211e-3, $B_1$ = -7.25958e-3, $B_2$ = -7.93334e-3, $B_3$ = -5.54491e-3, and $C_0$ = -1.32412e-7.
 
@@
 
-The remaining three parameters $a$, $r$, and $b$ are estimated by likelihood given the observed data with the JAGS model using prior distributions shown in the model file.
+The first modification to @eq-base for EBASE is use of a wind-based parameterization for gas exchange that is more appropriate for estuaries [@Wanninkhof14]:
+
+$$
+\frac{\delta{C_d}}{\delta{t}} = [\,aPAR^p]\, - [\,r\Theta^{T - \bar{T}}]\, - \frac{1}{H}\left[-bU_{10}^2\left(\frac{s_c}{600} \right)^{-0.5} \left(C_{Sat} - C_d \right )\right]
+$$ {#eq-basewan}
+
+where $U_{10}^2$ is the wind speed at 10-m height and $S_c$ is the Schmidt number defined as the ratio of the kinematic viscosity of water to the molecular diffusivity of the gas for oxygen calculated as a function of water temperature and salinity using the polynomial fit of @Wanninkhof14.
+The parameter $b$ ($\buni$) must be estimated from the data.
+The other variables are as defined as in @eq-base.
+
+The final modification to @eq-base for EBASE is a simplification to remove the $\Theta$ and $p$ parameters.
+For EBASE, the temperature dependency of respiration is removed by setting $\Theta^(T-\bar{T}$ = 1) and $p$ is also set as 1.
+This further simplifies @eq-basewan to:
+
+$$
+\frac{\delta{C_d}}{\delta{t}} = [\,aPAR]\, - [\,r]\, - \frac{1}{H}\left[-bU_{10}^2\left(\frac{s_c}{600} \right)^{-0.5} \left(C_{Sat} - C_d \right )\right]
+$$ {#eq-ebase}
+
+As for BASE, the metabolic estimates in EBASE are defined by the change in dissolved oxygen over the time step $\frac{\delta{C_d}}{\delta{t}}$, where gross production is provided by $aPAR$, respiration is provided by $r$, and gas exchange is provided by the remainder [as in @Wanninkhof14].
+Units for each of the metabolic estimates are $\runi$.
+
+The remaining three parameters $a$ (light efficiency, $\auni$), $r$ (respiration, $\runi$), and $b$ ($\buni$) are estimated by likelihood given the observed data with the JAGS model using prior distributions shown in the model file.
 At each time step, the change in oxygen concentration between time steps is calculated from the equation using model inputs and parameter guesses, and then a finite difference approximation is used to estimate modeled oxygen concentration.
-The estimated concentration is also returned, which can be compared to observed as one measure of model performance.
+The estimated dissolved oxygen concentration is also returned, which can be compared to observed as one measure of model performance.
 
-* EBASE theory
-* EBASE R package
+The metabolism estimates and their parameters returned by the Bayesian routine implemented in JAGS are affected by the prior distributions assigned to each.
+As for the BASE model, relatively uninformed prior distributions following a normal Gaussian distribution are used by default, although the priors can of course be changed based on previous knowledge of parameter specific to an ecosystem.
+The default priors were based on those that returned reasonable parameter estimates using a forward-feeding oxygen mass balance equation call FWOXY (ForWard OXYgen).
+For FWOXY, a dissolved oxygen time series is simulated by providing known parameter values in an equation similar to @eq-ebase.
+All of the physical parameters are held constant, e.g., wind speed, temperature, and salinity.
+The other required inputs are the $a$, $r$, and $b$ parameters, which are set by the user.
+As such, the dissolved oxygen output simulated by FWOXY as change per unit time is modelled by EBASE using the constant physical parameters as input.
+Reasonable prior distributions by EBASE were identified based on the ability to return the known parameters from FWOXY:
+
+$$
+a \sim N(0.2, 0.1)T(0,)
+$$ {#eq-aprior}
+
+$$
+r \sim N(20, 5)T(0,)
+$$ {#eq-rprior}
+
+$$
+b \sim N(0.251, 0.01)T(0,0.504)
+$$ {#eq-bprior}
+
+The $T$ notation is JAGS model syntax indicating the distributions are truncated to a minimum (left) and maximum (right) value defined in the parentheses.
+All prior distributions were constrained to positive values based on requirements of the model formula in @eq-ebase.
+Additionally, the prior distribution for $b$ was constrained to an upper limit of 0.504 (~twice the default mean).
+@Wanninkhof14 states that the $b$ parameter typically does not vary by more than 20% and the standard deviation of the default prior is sufficiently high to expand beyond this range.
+Initial development of EBASE also showed that using an undefined upper limit of $b$ led to unstable and unreasonable parameter estimates.
+The density distributions for the default priors in EBASE are shown in @fig-priorplot.
+
+### EBASE R package
+
+For ease of application, an R package was developed to implement the EBASE methods described above.
 
@@
 
+\newpage
+
+![The default prior distributions for the $a$, $r$, and $b$ parameters used in EBASE.
+All priors are normal distributions and truncated to positive values.
+The $b$ parameter is also truncated to a maximum of 0.504.](figs/priorplot.png){#fig-priorplot fig-align="center"}
+
 <!-- \newpage -->

### 2022-12-23 R/figs.R

@@ -2,8 +2,19 @@ library(tidyverse)
 library(lubridate)
 library(here)
 library(patchwork)
+library(EBASE)
 
 source(here('R/funcs.R'))
+
+# prior distribution plot ---------------------------------------------------------------------
+
+p <- prior_plot(n = 1e5)
+
+png(here('figs/priorplot.png'), height = 3, width = 8, family = 'serif', units = 'in', res = 500)
+print(p)
+dev.off()
+
+# Fwoxy apa comparison ------------------------------------------------------------------------
 
 load(file = url('https://github.com/fawda123/BASEmetab_script/raw/master/data/apacmp.RData'))
 

### 2023-01-03 manu-draft.qmd
minor additions to methods

@@
 The Estuarine BAyesian Single station Estimation (EBASE) method provides a simple approach to quantify metabolic parameters from continuous dissolved oxygen time series.
-The estimates are based on a modified version of the mass balance equation described in @Grace15.
-The original mass balance equation is:
+The estimates use a modified version of the BASE mass balance equation described in @Grace15.
+The original equation is:
 
@@
 
-where the terms in the square brackets are production (P), respiration (R), and gas exchange (D) as a function of the change in dissolved oxygen ($\delta{C_d}$) per unit time of the input data ($\delta{t}$).
+where the terms in the square brackets are production (P), respiration (R), and gas exchange (D), respecitively, modeled as a function of the change in dissolved oxygen ($\delta{C_d}$) per unit time of the input data ($\delta{t}$).
 Five unknown parameters are estimated by the model: $a$, $p$, $r$, $\Theta$, and $K$.
-The required inputs are the time series of diel oxygen ($C_d$, mmol/m$^2$), surface irradiance ($PAR$, $\mu$mol/m$^2$/s), water temperature ($T$, C), and oxygen saturation ($C_{sat}$, mmol/m$^3$).
+The required inputs are the time series of surface irradiance ($PAR$, $\mu$mol/m$^2$/s), water temperature ($T$, C), diel oxygen ($C_d$, mmol/m$^2$), and oxygen saturation ($C_{sat}$, mmol/m$^3$).
 Dissolved oxygen at saturation $C_{sat}$ is calculated from the input data as a function of water temperature and salinity (psu) [@Garcia92].
@@
 
-where $U_{10}^2$ is the wind speed at 10-m height and $S_c$ is the Schmidt number defined as the ratio of the kinematic viscosity of water to the molecular diffusivity of the gas for oxygen calculated as a function of water temperature and salinity using the polynomial fit of @Wanninkhof14.
+where $U_{10}^2$ is the wind speed at 10-m height, $H$ is water column depth (m, as a fixed parameter or time series from a pressure sensor) and $S_c$ is the Schmidt number defined as the ratio of the kinematic viscosity of water to the molecular diffusivity of the gas for oxygen [from water temperature and salinity using the polynomial fit of @Wanninkhof14].
 The parameter $b$ ($\buni$) must be estimated from the data.
-The other variables are as defined as in @eq-base.
+The other variables are defined as in @eq-base.
 
 The final modification to @eq-base for EBASE is a simplification to remove the $\Theta$ and $p$ parameters.
-For EBASE, the temperature dependency of respiration is removed by setting $\Theta^(T-\bar{T}$ = 1) and $p$ is also set as 1.
+For EBASE, the temperature dependency of respiration is removed by setting $\Theta^{(T-\bar{T})}$ = 1.
+Th $p$ parameter, which is dimensionless, is also set to 1 as in @Holtgrieve10.
 This further simplifies @eq-basewan to:
@@
 
-As for BASE, the metabolic estimates in EBASE are defined by the change in dissolved oxygen over the time step $\frac{\delta{C_d}}{\delta{t}}$, where gross production is provided by $aPAR$, respiration is provided by $r$, and gas exchange is provided by the remainder [as in @Wanninkhof14].
+As for BASE, the metabolic estimates in EBASE are defined by the change in dissolved oxygen over the time step $\delta{C_d}/\delta{t}$, where gross production is provided by $aPAR$, respiration is provided by $r$, and gas exchange is provided by the remainder [as in @Wanninkhof14].
 Units for each of the metabolic estimates are $\runi$.
@@
 As such, the dissolved oxygen output simulated by FWOXY as change per unit time is modelled by EBASE using the constant physical parameters as input.
-Reasonable prior distributions by EBASE were identified based on the ability to return the known parameters from FWOXY:
+Reasonable prior distributions by EBASE were identified (Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]) based on the ability to return the known parameters from FWOXY:
 
@@
 For ease of application, an R package was developed to implement the EBASE methods described above.
+A detailed web page (<https://fawda123.github.io/EBASE/>) describes the required data inputs for estimating the metabolic parameters using the core R functions with the JAGS software.
+A JAGS model file is included with the package that implements @eq-ebase, with options to supply a custom model file as needed.
+Users can also specify alternative prior distributions from those in Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior] for each parameter.
+Additional functions are available for viewing model results, including a plot of the modelled dissolved oxygen with the observed, a time series plot of the primary metabolic estimates (P, R, and D), and a time series plot of the credible intervals for the $a$, $r$, and $b$ parameters.
+All results provided herein were generated using the EBASE R package.
 

### 2023-01-04 manu-draft.qmd
adding more to methods including text from MH

@@
 The Estuarine BAyesian Single station Estimation (EBASE) method provides a simple approach to quantify metabolic parameters from continuous dissolved oxygen time series.
-The estimates use a modified version of the BASE mass balance equation described in @Grace15.
+The EBASE method is a modified version of the BASEmetab mass balance equation described in @Grace15.
 The original equation is:
@@
 
-where the terms in the square brackets are production (P), respiration (R), and gas exchange (D), respecitively, modeled as a function of the change in dissolved oxygen ($\delta{C_d}$) per unit time of the input data ($\delta{t}$).
+where the terms in the square brackets are production (P), respiration (R), and gas exchange (D), respectively, modeled as a function of the change in dissolved oxygen ($\delta{C_d}$) per unit time of the input data ($\delta{t}$).
 Five unknown parameters are estimated by the model: $a$, $p$, $r$, $\Theta$, and $K$.
 The required inputs are the time series of surface irradiance ($PAR$, $\mu$mol/m$^2$/s), water temperature ($T$, C), diel oxygen ($C_d$, mmol/m$^2$), and oxygen saturation ($C_{sat}$, mmol/m$^3$).
-Dissolved oxygen at saturation $C_{sat}$ is calculated from the input data as a function of water temperature and salinity (psu) [@Garcia92].
-@eq-ts is temperature scaled, as input to @eq-co that calculates the solubility of dissolved oxygen per volume of seawater at the given temperature.
-The coefficients are $A_0$ = 5.80818, $A_1$ = 3.20684, $A_2$ = 4.11890, $A_3$ = 4.93845, $A_4$ = 1.01567, $A_5$ = 1.41575, $B_0$ = -7.01211e-3, $B_1$ = -7.25958e-3, $B_2$ = -7.93334e-3, $B_3$ = -5.54491e-3, and $C_0$ = -1.32412e-7.
+Dissolved oxygen at saturation $C_{sat}$ is calculated from the input data as a function of water temperature, salinity, and barometric pressure.
+
+### Modification of the gas exchange parameterization
+
+The underlying idea behind most gas exchange parameterizations for metabolism estimates is that the transfer of gases of relatively low solubility, like oxygen, is proportional to the concentration gradient in a thin film of water at the air–water interface.
+It is further assumed that the water at the top of the film is saturated with respect to the atmosphere, so that the concentration gradient across the thin film can be approximated as the difference between oxygen concentration measured in the surface water layer, $C_d$, and the saturation (or equilibrium) concentration $C_{Sat}$, which is usually calculated as a function of water temperature and salinity (the negative sign is needed because positive flux is set to be from water to air):
 
 $$
-T_s = \log((298.15 - T)/(273.15 + T)
-$$ {#eq-ts}
-$$
-C_O = A_0 + A_1 \cdot T_s + A_2 \cdot T_s^2 + A_3 \cdot T_s^3 + A_4 \cdot T_s^4 + A_5 \cdot T_s^5 + sal \cdot (B_0 + B_1 \cdot T_s + B_2 \cdot T_s^2 + B_3 \cdot T_s^3) + C_0 \cdot sal^2
-$$ {#eq-co}
+D = -k_w \left(C_{sat}-C_d \right).
+$$ {#eq-gasex}
 
-Dissolved oxygen saturation is then calculated by multiplying $C_o$ by the density of seawater given temperature and salinity (@eq-csat).
-The parameter $\rho$ (kg/m$^3$) is seawater density at zero elevation given temperature and salinity [@Millero81].
+The proportionality parameter, $k_w$, usually called the gas transfer velocity because it has the dimensions of length over time, increases as the water column in the vicinity of the air–water interface gets more turbulent.
+As such, $k_w$  is usually modeled as a function of measurable variables that can potentially predict turbulent mixing of the surface water, such as wind speed, current velocity, and water depth [@Ho16].
+$k_w$ also increases with temperature because as the random motion of gas and water molecules in the thin film speeds up, the film thins out and the transfer of molecules across the film becomes easier.
+This dependence of $k_w$ on temperature is usually modeled with the Schmidt number, which combines information on the viscosity of the fluid and the diffusivity of the gas.
+The Schmidt number, in turn, is often parameterized as a gas-specific polynomial function of water temperature and salinity.
+
+Instead of one of the traditional wind-based parameterization used in oceanography [e.g., @Wanninkhof14], @Grace15 uses the following for the gas transfer:
 
 $$
-C_{sat} = \exp(C_O) \cdot \rho_{T, sal} \cdot 1e-3
-$$ {#eq-csat}
+\frac{k_w}{H} = K \cdot 1.024^{\left(T - \bar{T}\right)}
+$$ {#eq-gasexbase}
 
-The first modification to @eq-base for EBASE is use of a wind-based parameterization for gas exchange that is more appropriate for estuaries [@Wanninkhof14]:
+where $\bar{T}$ is the average water temperature over each 24-hour period in the diel oxygen time series and $K$ is the optimization parameter, often called the reaeration coefficient, which has dimensions of inverse time.
+@eq-gasexbase indicates that $k_w$ increases with temperature and depth.
+The temperature dependence is consistent with the known decrease of $k_w$ with the Schmidt number, but the depth dependence is difficult to justify.
+One expects that, for a given current speed, $k_w$ would increase as the water depth decreases because turbulence generated at the sediment–water interface by currents would more easily reach the surface [@Ho16].
+
+Based on the above differences, the first modification to @eq-base for EBASE is use of a wind-based parameterization for gas exchange that is more appropriate for estuaries [@Wanninkhof14]:
 
@@
 The parameter $b$ ($\buni$) must be estimated from the data.
-The other variables are defined as in @eq-base.
 
-The final modification to @eq-base for EBASE is a simplification to remove the $\Theta$ and $p$ parameters.
+The remaining variables in @eq-basewan are defined as in @eq-base, with the exception that $C_{sat}$ is calculated using only temperature (C) and salinity (psu), removing the need for barometric pressure.
+Methods from @Garcia92 are used as follows:
+
+$$
+C_{sat} = \exp(C_O) \cdot \rho_{T, sal} \cdot 1e-3
+$$ {#eq-csat}
+
+where $C_O$ is solubility of dissolved oxygen per volume of seawater at the given temperature and $\rho$ (kg/m$^3$) is seawater density at zero elevation given temperature and salinity [@Millero81].
+$C_O$ is calculated using a polynomial equation:
+
+$$
+C_O = A_0 + A_1 \cdot T_s + A_2 \cdot T_s^2 + A_3 \cdot T_s^3 + A_4 \cdot T_s^4 + A_5 \cdot T_s^5 + sal \cdot (B_0 + B_1 \cdot T_s + B_2 \cdot T_s^2 + B_3 \cdot T_s^3) + C_0 \cdot sal^2
+$$ {#eq-co}
+
+where the coefficients are $A_0$ = 5.80818, $A_1$ = 3.20684, $A_2$ = 4.11890, $A_3$ = 4.93845, $A_4$ = 1.01567, $A_5$ = 1.41575, $B_0$ = -7.01211e-3, $B_1$ = -7.25958e-3, $B_2$ = -7.93334e-3, $B_3$ = -5.54491e-3, and $C_0$ = -1.32412e-7.
+$T_s$ is temperature scaled
+as input to @eq-co defined as:
+
+$$
+T_s = \log((298.15 - T)/(273.15 + T)
+$$ {#eq-ts}
+
+### Modification of the production and respiration parameterization
+
+The second modification to @eq-base for EBASE is a simplification to remove the $p$ and $\Theta$ parameters from the production (P) and respiration (R) components, respectively, of the mass balance equation.
+The BASEmetab method provides options to estimate both $\Theta$ and $p$ given the data or to set as fixed values defined in the model file ($\Theta = 1.07177$, $p=1$).
 For EBASE, the temperature dependency of respiration is removed by setting $\Theta^{(T-\bar{T})}$ = 1.
-Th $p$ parameter, which is dimensionless, is also set to 1 as in @Holtgrieve10.
+The $p$ parameter, which is dimensionless, is also set to 1 as in @Holtgrieve10.
+Removing these components from the model improved model stability by creating more reasonable estimates for the other parameters.
 This further simplifies @eq-basewan to:
@@
 
-As for BASE, the metabolic estimates in EBASE are defined by the change in dissolved oxygen over the time step $\delta{C_d}/\delta{t}$, where gross production is provided by $aPAR$, respiration is provided by $r$, and gas exchange is provided by the remainder [as in @Wanninkhof14].
+As for BASEmetab, the metabolic estimates in EBASE are defined by the change in dissolved oxygen over the time step $\delta{C_d}/\delta{t}$, where gross production is provided by $aPAR$, respiration is provided by $r$, and gas exchange is provided by the remainder [as in @Wanninkhof14].
 Units for each of the metabolic estimates are $\runi$.
@@
 
+### Bayesian prior
+
 The metabolism estimates and their parameters returned by the Bayesian routine implemented in JAGS are affected by the prior distributions assigned to each.
-As for the BASE model, relatively uninformed prior distributions following a normal Gaussian distribution are used by default, although the priors can of course be changed based on previous knowledge of parameter specific to an ecosystem.
+As for the BASEmetab model, relatively uninformed prior distributions following a normal Gaussian distribution are used by default, although the priors can of course be changed based on previous knowledge of parameter specific to an ecosystem.
 The default priors were based on those that returned reasonable parameter estimates using a forward-feeding oxygen mass balance equation call FWOXY (ForWard OXYgen).
@@
 
-@fig-apacmpfig and `r run_reference(id = "apacmptab")` provide a comparison of the EBASE metabolic estimates with those from the Odum and BASmetab methods.
+@fig-apacmpfig and `r run_reference(id = "apacmptab")` provide a comparison of the EBASE metabolic estimates with those from the Odum and BASEmetab methods.
 

### 2023-01-05 manu-draft.qmd
more on methods

@@
 Five unknown parameters are estimated by the model: $a$, $p$, $r$, $\Theta$, and $K$.
-The required inputs are the time series of surface irradiance ($PAR$, $\mu$mol/m$^2$/s), water temperature ($T$, C), diel oxygen ($C_d$, mmol/m$^2$), and oxygen saturation ($C_{sat}$, mmol/m$^3$).
+The required inputs are the time series of surface irradiance ($PAR$, $\mu$mol/m$^2$/s), water temperature ($T$, C), atmospheric pressure (atm), diel oxygen ($C_d$, mmol/m$^2$), and oxygen saturation ($C_{sat}$, mmol/m$^3$).
 Dissolved oxygen at saturation $C_{sat}$ is calculated from the input data as a function of water temperature, salinity, and barometric pressure.
+Several changes, outlined and justified below, were made to @eq-base for EBASE.
 
@@
 
-Instead of one of the traditional wind-based parameterization used in oceanography [e.g., @Wanninkhof14], @Grace15 uses the following for the gas transfer:
+Instead of one of the traditional wind-based parameterization used in oceanography [e.g., @Wanninkhof14], @Grace15 uses the following for the gas transfer in BASEmetab:
 
@@
 
-where $C_O$ is solubility of dissolved oxygen per volume of seawater at the given temperature and $\rho$ (kg/m$^3$) is seawater density at zero elevation given temperature and salinity [@Millero81].
+where $C_O$ is thr solubility of dissolved oxygen per volume of seawater at the given temperature and $\rho$ (kg/m$^3$) is seawater density at zero elevation given temperature and salinity [@Millero81].
 $C_O$ is calculated using a polynomial equation:
@@
 The second modification to @eq-base for EBASE is a simplification to remove the $p$ and $\Theta$ parameters from the production (P) and respiration (R) components, respectively, of the mass balance equation.
-The BASEmetab method provides options to estimate both $\Theta$ and $p$ given the data or to set as fixed values defined in the model file ($\Theta = 1.07177$, $p=1$).
+The BASEmetab method provides options to estimate both $\Theta$ and $p$ from the data or to set as fixed values defined in the model file ($\Theta = 1.07177$, $p=1$).
 For EBASE, the temperature dependency of respiration is removed by setting $\Theta^{(T-\bar{T})}$ = 1.
 The $p$ parameter, which is dimensionless, is also set to 1 as in @Holtgrieve10.
-Removing these components from the model improved model stability by creating more reasonable estimates for the other parameters.
-This further simplifies @eq-basewan to:
+Removing these components mproved model stability by creating more reasonable estimates for the other parameters.
+This further simplifies @eq-basewan to the final form for EBASE:
 
@@
 
-As for BASEmetab, the metabolic estimates in EBASE are defined by the change in dissolved oxygen over the time step $\delta{C_d}/\delta{t}$, where gross production is provided by $aPAR$, respiration is provided by $r$, and gas exchange is provided by the remainder [as in @Wanninkhof14].
+As for BASEmetab, the metabolic estimates in EBASE are defined by the change in dissolved oxygen over the time step, $\delta{C_d}/\delta{t}$, where gross production is provided by $aPAR$, respiration is provided by $r$, and gas exchange is provided by the remainder [as in @Wanninkhof14].
 Units for each of the metabolic estimates are $\runi$.
 
-The remaining three parameters $a$ (light efficiency, $\auni$), $r$ (respiration, $\runi$), and $b$ ($\buni$) are estimated by likelihood given the observed data with the JAGS model using prior distributions shown in the model file.
+The remaining three parameters $a$ (light efficiency, $\auni$), $r$ (respiration, $\runi$), and $b$ ($\buni$) are estimated by likelihood given the observed data and prior distributions for the parameters.
+The JAGS software [@Plummer03] is used with the EBASE R package [described below, @Beck22;@RCT22] to estimate the unknown parameters using Markov Chain Monte Carlo simulations in a Bayesian framework.
 At each time step, the change in oxygen concentration between time steps is calculated from the equation using model inputs and parameter guesses, and then a finite difference approximation is used to estimate modeled oxygen concentration.
+Estimations proceed using a forward in time integration where the dissolved oxygen at each time step is informed by parameter estimates from the preceding time steps.
 The estimated dissolved oxygen concentration is also returned, which can be compared to observed as one measure of model performance.
@@
 The metabolism estimates and their parameters returned by the Bayesian routine implemented in JAGS are affected by the prior distributions assigned to each.
-As for the BASEmetab model, relatively uninformed prior distributions following a normal Gaussian distribution are used by default, although the priors can of course be changed based on previous knowledge of parameter specific to an ecosystem.
-The default priors were based on those that returned reasonable parameter estimates using a forward-feeding oxygen mass balance equation call FWOXY (ForWard OXYgen).
+As for the BASEmetab model, relatively uninformed prior distributions following a normal Gaussian distribution are used by default, although the priors can of course be changed based on previous knowledge of parameters specific to an ecosystem or as informed by other metabolic modeling approaches.
+The default priors were based on those that returned reasonable parameter estimates using a forward-feeding oxygen mass balance equation call FWOXY [ForWard OXYgen, @Arriola22].
 For FWOXY, a dissolved oxygen time series is simulated by providing known parameter values in an equation similar to @eq-ebase.
@@
 The other required inputs are the $a$, $r$, and $b$ parameters, which are set by the user.
-As such, the dissolved oxygen output simulated by FWOXY as change per unit time is modelled by EBASE using the constant physical parameters as input.
+An initial oxygen concentration is provided and the simulated time series is created by forward addition of the inputs.
+As such, the dissolved oxygen output simulated by FWOXY as change per unit time were modelled by EBASE using the constant physical parameters as input.
 Reasonable prior distributions by EBASE were identified (Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]) based on the ability to return the known parameters from FWOXY:
@@
 The $T$ notation is JAGS model syntax indicating the distributions are truncated to a minimum (left) and maximum (right) value defined in the parentheses.
-All prior distributions were constrained to positive values based on requirements of the model formula in @eq-ebase.
+All prior distributions were constrained to positive values based on known physical constraints and requirements of the model formula in @eq-ebase.
 Additionally, the prior distribution for $b$ was constrained to an upper limit of 0.504 (~twice the default mean).
@@
 
-For ease of application, an R package was developed to implement the EBASE methods described above.
-A detailed web page (<https://fawda123.github.io/EBASE/>) describes the required data inputs for estimating the metabolic parameters using the core R functions with the JAGS software.
+For ease of application, an R package was developed to implement the EBASE methods described above [@Beck22].
+A detailed web page (<https://fawda123.github.io/EBASE/>) describes the required data inputs for estimating the metabolic parameters using the core R functions that pass inputs to the JAGS software.
 A JAGS model file is included with the package that implements @eq-ebase, with options to supply a custom model file as needed.
@@
 Additional functions are available for viewing model results, including a plot of the modelled dissolved oxygen with the observed, a time series plot of the primary metabolic estimates (P, R, and D), and a time series plot of the credible intervals for the $a$, $r$, and $b$ parameters.
+An additional option also allows the user to specify the model optimization period, where the time period can vary from a minimum of one day to the maximum of the entire length of the time series.
+This differs from the BASEmetab approach where the optimization period is set at one day.
 All results provided herein were generated using the EBASE R package.
@@
 
-* Comparison with Odum and BASEmetab: Fwoxy Appalachicola
+A preliminary assessment of the ability of EBASE to produce reasonable parameter estimates for $a$, $r$, and $b$ (@eq-ebase) was conducted using a simulated time series created with FWOXY.
+These assessments evaluated if 1) EBASE estimates similar values for the known parameters, and 2) how the estimated values change with different model configurations.
+The simulated time series used inputs from an approximate one year of continuous data from Apalachicola Bay, Florida.
+The input data were 2012 observations from February 22nd to December 19th at the Cat Point water quality monitoring station (29.7021$^{\circ}$ N, -84.8802$^{\circ}$ W) and included 15 minute observations of dissolved oxygen (mg/L), water temperature (C), and salinity (psu).
+Continuous meteorological observations from the nearby East Bay station (29.7909$^{\circ}$ N, -84.8834$^{\circ}$ W, ~10km north of Cat Point) were combined with the water quality data at the same time step and included air temperature (C), wind speed (m/s), and PAR (mmol/m$^2$ as total per 15 minute observation).
+An oxygen time series was generated from Fwoxy using the inputs from Apalachicola Bay, such that the simulated dissolved oxygen was created as close as possible to the observed given the inputs.
+Known values for $a$ and $r$ parameters were returned for each time step of the simulated data, as well as all metabolic estimates for production, respiration, and gas exchange.
+The only parameter that could not be simulated by Fwoxy was $b$, which was set to 0.251.
 
-### Sensitivity analyses
+The simulated time series is shown in figure......
 
 Several configurations of the EBASE model can influence how the metabolic rates are estimated.
-A simulated time series with known values for the $a$, $r$, and $b$ parameters that are returned by EBASE was used to determine if 1) EBASE estimates similar values, and 2) how the estimated values change with different model configurations.
 Two types of sensitivity analyses were conducted to evaluate changes in the parameters returned by EBASE, where in all cases, the parameters were compared to the known values in the simulated time series.
@@
 Estimates for the modeled dissolved oxygen, gross production, respiration ($r$), gas exchange ($D$), and $a$ were compared to those from the simulated time series using coefficient of variation ($R^2$) and root mean square error ($RMSE$) summaries of the 1:1 comparison of the estimated vs simulated values.
-For the one day optimization period, all parameter estimates returned at the time step of the input data (i.e., modeled dissoled oxygen, $D$, and $a$) were daily-averaged for comparison.
+For the one day optimization period, all parameter estimates returned at the time step of the input data (i.e., modeled dissolved oxygen, $D$, and $a$) were daily-averaged for comparison.
 For the seven day optimization period, all parameter estimates and those from the known time series were weekly-averaged.
@@
 
+Results for the sensitivity analyses are shown in ......
+
 ### EBASE comparison with real observations
 
-Metabolic estimates from EBASE were compared to those from existing methods to benchmark and evaluate potential differences from alternative model formulations used by each approach.
-An approximate one year continuous time series from Apalachicola Bay, Florida was provided as input to each method.
-The input data were 2012 observations from February 22nd to December 19th at the Cat Point water quality monitoring station (29.7021$^{\circ}$ N, -84.8802$^{\circ}$ W) and included 15 minute observations of dissolved oxygen (mg/L), water temperature (C), salinity (psu), and sonde depth (m).
-Continuous meteorological observations from the nearby East Bay station (29.7909$^{\circ}$ N, -84.8834$^{\circ}$ W, ~10km north of Cat Point) were combined with the water quality data at the same time step and included air temperature (C), wind speed (m/s), PAR (mmol/m$^2$ as total per 15 minute observation), and barometric pressure (mb).
-The combined input data were provided to each method after converting the units as required (e.g., BASEmetab requires PAR as $\mu$mol m$^{-2}$ s$^{-1}$).
+Metabolic estimates from EBASE were further compared to those from existing methods to benchmark and evaluate potential differences from alternative model formulations used by each approach.
+The same one year continuous time series described above for water quality and weather data from Apalachicola Bay was used to compare metabolic estimates from EBASE, BASEmetab, and the Odum open-water method [@Odum56].
+The Odum method used an approach described in @Murrell18 created for estuarine application that used a gas-exchange approach described in @Thebault08.
+For better comparison with the other methods, the gas exchange parmaterization in the Odum method was modified to use the @Wanninkhof14 approach [as available in the WtRegDO R package, @Beck21].
+The input data were provided to each method after converting the units as required (e.g., BASEmetab requires PAR as $\mu$mol m$^{-2}$ s$^{-1}$).
 Water column depth at Cat Point is approximately 1.8 m and all daily metabolic estimates returned by each method were converted to volumetric rates by dividing by the depth ($\mu$mol O$_2$ m$^{-3}$ d$^{-1}$).
@@
 
-### Application to represenatiave NERR sites
+### Application to representative NERR sites
 

### 2023-01-06 manu-draft.qmd
simulated apa fig and cross refs added

@@
 Continuous meteorological observations from the nearby East Bay station (29.7909$^{\circ}$ N, -84.8834$^{\circ}$ W, ~10km north of Cat Point) were combined with the water quality data at the same time step and included air temperature (C), wind speed (m/s), and PAR (mmol/m$^2$ as total per 15 minute observation).
-An oxygen time series was generated from Fwoxy using the inputs from Apalachicola Bay, such that the simulated dissolved oxygen was created as close as possible to the observed given the inputs.
-Known values for $a$ and $r$ parameters were returned for each time step of the simulated data, as well as all metabolic estimates for production, respiration, and gas exchange.
+An oxygen time series was generated from Fwoxy using the inputs from Apalachicola Bay ([@fig-simapa]a), such that the simulated dissolved oxygen was created as close as possible to the observed given the inputs.
+Known values for $a$ and $r$ parameters were returned for each time step of the simulated data, as well as all metabolic estimates for production, respiration, and gas exchange ([@fig-simapa]b and c).
 The only parameter that could not be simulated by Fwoxy was $b$, which was set to 0.251.
-
-The simulated time series is shown in figure......
 
@@
 Each optimization period was evaluated for every unique combination of prior distributions, totaling 54 unique analyses for the assessment of changing standard deviations of all parameters and 18 unique analyses for the assessment of changing the mean and standard deviation of only the $b$ parameter.
-
-The metabolic estimates returned from each analysis were compared to the known values to identify priors the model was most sensitive to and if the conclusions varied by optimization period.
-Estimates for the modeled dissolved oxygen, gross production, respiration ($r$), gas exchange ($D$), and $a$ were compared to those from the simulated time series using coefficient of variation ($R^2$) and root mean square error ($RMSE$) summaries of the 1:1 comparison of the estimated vs simulated values.
-For the one day optimization period, all parameter estimates returned at the time step of the input data (i.e., modeled dissolved oxygen, $D$, and $a$) were daily-averaged for comparison.
-For the seven day optimization period, all parameter estimates and those from the known time series were weekly-averaged.
-Estimates for the $b$ parameter were evaluated only by visual comparison because a single value was used in the simulated time series, preventing the estimate of quantitative comparisons.
 
@@
 
+![The default prior distributions for the $a$, $r$, and $b$ parameters used in EBASE.
+All priors are normal distributions and truncated to positive values.
+The $b$ parameter is also truncated to a maximum of 0.504.](figs/priorplot.png){#fig-priorplot fig-align="center"}
+
+\newpage
+
+![Simulated time series using the FWOXY model for one year of continuous data at Apalachicola Bay.
+Simulated input to EBASE was the (a) dissolved oxygen time series, with results compared to the (b) simulated $a$ parameter and (c) simulatd metabolic estimates for production (P), respiration (R), and gas exchange (D).](figs/simapa.png){#fig-simapa}
+
 ![Comparisons of daily metabolic estimates from the Odum, BASEmetab, and EBASE methods for one year of continuous monitoring data at Apalachicola Bay.
@@
 \newpage
-
-![The default prior distributions for the $a$, $r$, and $b$ parameters used in EBASE.
-All priors are normal distributions and truncated to positive values.
-The $b$ parameter is also truncated to a maximum of 0.504.](figs/priorplot.png){#fig-priorplot fig-align="center"}
 

### 2023-01-06 R/figs.R

@@ -3,6 +3,7 @@ library(lubridate)
 library(here)
 library(patchwork)
 library(EBASE)
+library(readr)
 
 source(here('R/funcs.R'))
 
@@ -11,6 +12,53 @@ source(here('R/funcs.R'))
 p <- prior_plot(n = 1e5)
 
 png(here('figs/priorplot.png'), height = 3, width = 8, family = 'serif', units = 'in', res = 500)
+print(p)
+dev.off()
+
+# simulated apa data --------------------------------------------------------------------------
+
+fwdat <- read_csv(file = url('https://raw.githubusercontent.com/fawda123/BASEmetab_script/master/data/apafwoxy.csv'))
+
+fwdatcmp <- fwdat %>% 
+  mutate(
+    DateTimeStamp = dmy_hms(datet, tz = 'America/Jamaica'),
+    Date = as.Date(DateTimeStamp, tz = 'America/Jamaica'),
+    DO_obs = `oxy,mmol/m3`, 
+    a = `aparam,(mmolO2/m2/d)/(W/m2)` / `ht,m`,
+    Rt_vol = `er,mmol/m2/d` / `ht,m`,
+    Pg_vol = `gpp,mmol/m2/d` / `ht,m`,
+    D = -1 * `gasex,mmol/m2/d` / `ht,m`,
+    b = 100 * 3600 * `kw,m/s` / `wspd2,m2/s2` / (`sc,dimensionless` / 660) ^ -0.5 # (m/s)/(m2/s2) to (cm/hr) / (m2/s2)
+  ) %>% 
+  select(Date, DateTimeStamp, DO_obs, a, b, Pg_vol, Rt_vol, D)
+
+p1 <- ggplot(fwdatcmp, aes(x = DateTimeStamp, y = DO_obs)) + 
+  geom_line(linewidth = 0.2) + 
+  scale_x_datetime(breaks = "1 month", labels = date_format("%b")) +
+  theme_minimal() + 
+  labs(
+    x = NULL, 
+    y = expression(paste(O [2], ' (mmol ', m^-3, ')')),
+    title = '(a) Simulated dissolved oxygen'
+  )
+
+p2 <- ggplot(fwdatcmp, aes(x = DateTimeStamp, y = a)) + 
+  geom_line(linewidth = 0.2) + 
+  scale_x_datetime(breaks = "1 month", labels = date_format("%b")) +
+  theme_minimal() + 
+  labs(
+    x = NULL, 
[…]

### 2023-01-08 manu-draft.qmd
adding figures for apasumdat and apasumdatmean, results text for apasumdat

@@
 Users can also specify alternative prior distributions from those in Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior] for each parameter.
-Additional functions are available for viewing model results, including a plot of the modelled dissolved oxygen with the observed, a time series plot of the primary metabolic estimates (P, R, and D), and a time series plot of the credible intervals for the $a$, $r$, and $b$ parameters.
+Functions are available for viewing model results, including a plot of the modelled dissolved oxygen with the observed, a time series plot of the primary metabolic estimates (P, R, and D), and a time series plot of the credible intervals for the $a$, $r$, and $b$ parameters.
 An additional option also allows the user to specify the model optimization period, where the time period can vary from a minimum of one day to the maximum of the entire length of the time series.
 This differs from the BASEmetab approach where the optimization period is set at one day.
+Sensitivity of the EBASE method to different optimization periods is described in @sec-cmpmod.
 All results provided herein were generated using the EBASE R package.
@@
 
-### EBASE comparison with modelled observations
+### EBASE comparison with modelled observations {#sec-cmpmod}
 
@@
 Known values for $a$ and $r$ parameters were returned for each time step of the simulated data, as well as all metabolic estimates for production, respiration, and gas exchange ([@fig-simapa]b and c).
-The only parameter that could not be simulated by Fwoxy was $b$, which was set to 0.251.
+The only parameter that could not be simulated by Fwoxy was $b$, which was set to 0.251 $\buni$.
 
-Several configurations of the EBASE model can influence how the metabolic rates are estimated.
-Two types of sensitivity analyses were conducted to evaluate changes in the parameters returned by EBASE, where in all cases, the parameters were compared to the known values in the simulated time series.
-First, parameter estimates were compared to the known values after changing characteristics of their prior distributions.
-Second, parameter estimates were compared using a model period of optimization for one day and one week of observations.
+Several configurations of the EBASE model can influence how the parameters and metabolic rates are estimated.
+Two types of sensitivity analyses were conducted to evaluate changes in the results returned by EBASE, where in all cases, the results were compared to the known values in the simulated time series.
+First, results were compared to the known values after changing characteristics of their prior distributions.
+Second, results were compared using a model period of optimization for one day and one week of observations.
 Details of these analyses are as follows.
 
-The standard prior distributions for the EBASE parameters follow normal distributions with defined means and standard deviations.
+The standard prior distributions for the EBASE parameters follow normal distributions with defined means and standard deviations (Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]).
 The sensitivity analysis evaluated changes in the distributions by varying the standard deviations from small to large to test the effect of more or less constrained ranges, respectively, on the priors.
@@
 Unlike BASEmetab, the EBASE software allows flexibility in the period of time used for optimization, with a minimum period of time being one day and a maximum period of being the length of the time series.
-For a chosen optimization period, a single estimate for $r$ and $b$ is returned by the model, whereas $n$ number of observations equal to the number of time steps in each optimization period are returned for the $a$ parameter (e.g., 96 values for  24 hour time period if the sampling interval is 15 minutes).
+For a chosen optimization period, a single estimate for $r$ and $b$ is returned by the model, whereas $n$ number of observations equal to the number of time steps in each optimization period are returned for the $a$ parameter (e.g., 96 values for a 24 hour time period if the sampling interval is 15 minutes).
 Short periods of observation may return unstable parameter estimates if forcing factors (e.g., wind, temperature) that affect estuarine biogeochemical rates occur at time scales longer the chosen period of optimization for EBASE.
@@
 An optimization period of one day was compared to results using an optimization period of seven days for the simulated one-year time series.
-Each optimization period was evaluated for every unique combination of prior distributions, totaling 54 unique analyses for the assessment of changing standard deviations of all parameters and 18 unique analyses for the assessment of changing the mean and standard deviation of only the $b$ parameter.
+Each optimization period was evaluated for every unique combination of prior distributions, totaling 54 (27 times the two optimization periods) unique analyses for the assessment of changing standard deviations of all parameters and 18 (9 times the two optimization periods) unique analyses for the assessment of changing the mean and standard deviation of only the $b$ parameter.
 
-Results for the sensitivity analyses are shown in ......
+Changing the variance of each of the prior distributions affected the ability of EBASE to return the known parameters from the simulated time series.
+@fig-apasumdat shows a summary of the comparisons of the simulated data with EBASE for each unique combination of prior distributions and two optimization periods of one and seven days.
+The results are compared using R-squared ($R^2$) values from the linear fit between the simulated data and results from EBASE.
+Darker colors represent lower $R^2$ values and can be used to assess how changing the variance of the priors affects the ability of EBASE to estimate the true parameter.
+
+Several conclusions can be made from the results in @fig-apasumdat based on the patterns of the $R^2$ values.
+Overall, the model was able to estimate the $a$ parameter with reasonable precision regardless of any of the prior distributions, whereas most of the variation was observed in the $r$ and $b$ parameters.
+Changing the variance of the $a$ parameter had little effect on any of the comparisons.
+Smaller variance of the prior distribution for the $r$ parameter generally contributed to lower $R^2$ for all parameters, with the precision being worse for the 7-day optimization period (e.g., the $a$ parameter).
+Larger variance of the prior distribution for the $b$ parameter generally contributed to lower $R^2$ only for the gas exchange estimate (D), although this effect was minimized with the 7-day optimization period.
+As such, the results suggest that the default prior distributions used in EBASE may be appropriate in some cases, but increasing the prior variance for $a$ and $r$ and decreasing the variance for $b$ will likely produce more optimal solutions.
 
@@
 
+\newpage
+
+![Comparison of EBASE results with known metabolic parameters from a simulated time series.
+Results are compared for unique combination of prior distributions, where the variances are evaluated from low (L), medium (M), to high (H) for a one-day and seven-day model optimization period.
+The metabolic results for dissolved oxygen (DO), production (P), respiration (R), gas exchange (D), and the $a$ parameter are evaluated based on the $R^2$ value of the linear fit between the EBASE results and simulated values.
+The $b$ parameter is not evaluated because it is a fixed constant for the simulated data.](figs/apasumdat.png){#fig-apasumdat}
+
+\newpage
+
+![Comparison of EBASE results with known metabolic parameters from a simulated time series.
+Results are compared for unique combination of prior distributions, where the variances and means for the prior of the $b$ parameter are evaluated from low (L), medium (M), to high (H) for a one-day and seven-day model optimization period.
+The metabolic results for dissolved oxygen (DO), production (P), respiration (R), gas exchange (D), and the $a$ parameter are evaluated based on the $R^2$ value of the linear fit between the EBASE results and simulated values.](figs/apasumdatmean.png){#fig-apasumdatmean}
+
+\newpage
+
 ![Comparisons of daily metabolic estimates from the Odum, BASEmetab, and EBASE methods for one year of continuous monitoring data at Apalachicola Bay.

### 2023-01-08 R/figs.R

@@ -64,6 +64,27 @@ dev.off()
 
 # Fwoxy apa comparison ------------------------------------------------------------------------
 
+fl <- paste0(tempdir(), '/apasumdat.RData')
+download.file('https://github.com/fawda123/BASEmetab_script/raw/master/data/apasumdat.RData', destfile = fl)
+load(file = fl)
+fl <- paste0(tempdir(), '/apasumdatmean.RData')
+download.file('https://github.com/fawda123/BASEmetab_script/raw/master/data/apasumdatmean.RData', destfile = fl)
+load(file = fl)
+
+p1 <- priorcomp(apasumdat, 1)
+
+p2 <- priorcompmean(apasumdatmean, 1)
+
+png(here('figs/apasumdat.png'), height = 6, width = 6, family = 'serif', units = 'in', res = 500)
+print(p1)
+dev.off()
+
+png(here('figs/apasumdatmean.png'), height = 3, width = 6, family = 'serif', units = 'in', res = 500)
+print(p2)
+dev.off()
+
+# actual apa comparison -----------------------------------------------------------------------
+
 load(file = url('https://github.com/fawda123/BASEmetab_script/raw/master/data/apacmp.RData'))
 
 toplo <- apacmp %>% 

### 2023-01-08 R/funcs.R

@@ -84,3 +84,202 @@ apacmp_plo <- function(dat, xlb = 'EBASE', ylb = 'Odum', dotyp = 'observed', add
   return(out)
   
 }
+
+# Fwoxy apa comparison to EBASE for different priors sd only
+priorcomp <- function(dat, ind){
+  
+  met <- tibble(
+    lbs = c('r2', 'rmse', 'aved'),
+    lbspr = c('italic(R)^2', 'RMSE', 'Ave.\nDiff.'), 
+    direc = c(-1, 1, 1)
+  )
+  
+  toshw <- met$lbs[ind]
+  leglb <- met$lbspr[ind]
+  direc <- met$direc[ind]
+  
+  toplo <- dat %>% 
+    select(-amean, -rmean, -bmean) %>% 
+    unnest('ests') %>% 
+    select(ndays, asd, rsd, bsd, var, matches(toshw)) %>%
+    filter(!var %in% 'b') %>% 
+    pivot_wider(names_from = 'var', values_from = !!toshw) %>% 
+    mutate(
+      ind = sort(rep(1: (nrow(.) / 2), times = 2)), 
+      ndays = case_when(
+        ndays == 1 ~ paste(ndays, 'day'), 
+        T ~ paste(ndays, 'days')
+      )
+    )
+  
+  toplo1 <- toplo %>% 
+    select(ind, asd, rsd, bsd) %>% 
+    unique() %>% 
+    mutate(
+      def = case_when(
+        asd == 0.1 & rsd == 5 & bsd == 0.01 ~ '*', 
+        T ~ ''
+      ),
+      asd = factor(asd, labels = c('L', 'M', 'H')), 
+      rsd = factor(rsd, labels = c('L', 'M', 'H')), 
+      bsd = factor(bsd, labels = c('L', 'M', 'H')), 
+    ) %>% 
+    pivot_longer(-c('ind', 'def'), names_to = 'var', values_to = 'val') %>% 
+    mutate(
+      var = factor(var, 
+                   levels = c('asd', 'rsd', 'bsd'), 
+                   labels = c('italic(a)', 'italic(r)', 'italic(b)'))
+    ) 
+  
+  toplo2 <- toplo %>% 
+    select(-asd, -rsd, -bsd) %>%
+    pivot_longer(-c(ind, ndays), names_to = 'var', values_to = 'val') %>% 
+    mutate(
+      var = factor(var, 
[…]

### 2023-01-08 manu-draft.qmd
removed text and figures for apasumdatmean eval

@@
 Each unique combination of low, medium, and high values for the standard deviation of each parameter were evaluated, creating 3$^3$ = 27 different combinations of prior distributions that were used with the simulated time series in EBASE.
-Because initial assessments indicated that all parameter estimates were especially sensitive to changes in the prior distribution for the $b$ parameter, a second sensitivity analysis evaluated additional changes in both the mean and standard deviation of the prior.
-The mean of the $b$ prior was varied as 50%, 100%, and 200% (the upper limit used in the software) of the default and the standard deviation was varied as before.
-This resulted in 3$^2$ = 9 unique combinations of means and standard deviations for the $b$ prior that were evaluated.
 
@@
 An optimization period of one day was compared to results using an optimization period of seven days for the simulated one-year time series.
-Each optimization period was evaluated for every unique combination of prior distributions, totaling 54 (27 times the two optimization periods) unique analyses for the assessment of changing standard deviations of all parameters and 18 (9 times the two optimization periods) unique analyses for the assessment of changing the mean and standard deviation of only the $b$ parameter.
+Each optimization period was evaluated for every unique combination of prior distributions, totaling 54 (27 times the two optimization periods) unique analyses for the assessment of changing standard deviations of all parameters.
 
@@
 
-![Comparison of EBASE results with known metabolic parameters from a simulated time series.
-Results are compared for unique combination of prior distributions, where the variances and means for the prior of the $b$ parameter are evaluated from low (L), medium (M), to high (H) for a one-day and seven-day model optimization period.
-The metabolic results for dissolved oxygen (DO), production (P), respiration (R), gas exchange (D), and the $a$ parameter are evaluated based on the $R^2$ value of the linear fit between the EBASE results and simulated values.](figs/apasumdatmean.png){#fig-apasumdatmean}
-
-\newpage
-
 ![Comparisons of daily metabolic estimates from the Odum, BASEmetab, and EBASE methods for one year of continuous monitoring data at Apalachicola Bay.

### 2023-01-08 R/figs.R

@@ -67,20 +67,11 @@ dev.off()
 fl <- paste0(tempdir(), '/apasumdat.RData')
 download.file('https://github.com/fawda123/BASEmetab_script/raw/master/data/apasumdat.RData', destfile = fl)
 load(file = fl)
-fl <- paste0(tempdir(), '/apasumdatmean.RData')
-download.file('https://github.com/fawda123/BASEmetab_script/raw/master/data/apasumdatmean.RData', destfile = fl)
-load(file = fl)
 
 p1 <- priorcomp(apasumdat, 1)
 
-p2 <- priorcompmean(apasumdatmean, 1)
-
 png(here('figs/apasumdat.png'), height = 6, width = 6, family = 'serif', units = 'in', res = 500)
 print(p1)
-dev.off()
-
-png(here('figs/apasumdatmean.png'), height = 3, width = 6, family = 'serif', units = 'in', res = 500)
-print(p2)
 dev.off()
 
 # actual apa comparison -----------------------------------------------------------------------

### 2023-01-08 manu-draft.qmd
working on optex fig

@@
 Several conclusions can be made from the results in @fig-apasumdat based on the patterns of the $R^2$ values.
-Overall, the model was able to estimate the $a$ parameter with reasonable precision regardless of any of the prior distributions, whereas most of the variation was observed in the $r$ and $b$ parameters.
+Overall, the model was able to estimate the $a$ parameter with reasonable precision regardless of any of the prior distributions, whereas most of the variation was observed in the $r$ (respiration, R) and $b$ parameters.
 Changing the variance of the $a$ parameter had little effect on any of the comparisons.
@@
 As such, the results suggest that the default prior distributions used in EBASE may be appropriate in some cases, but increasing the prior variance for $a$ and $r$ and decreasing the variance for $b$ will likely produce more optimal solutions.
+The actual model results for the best performing combination of prior distributions for the one-day and 7-day optimization is shown in @fig-optex (all results are averaged at the time-step of the optimization period for comparability).
 

### 2023-01-08 R/figs.R

@@ -74,6 +74,42 @@ png(here('figs/apasumdat.png'), height = 6, width = 6, family = 'serif', units =
 print(p1)
 dev.off()
 
+# best fwoxy apa comparison -------------------------------------------------------------------
+
+fl <- paste0(tempdir(), '/apagrd.RData')
+download.file('https://github.com/fawda123/BASEmetab_script/raw/master/data/apagrd.RData', destfile = fl)
+load(file = fl)
+
+fwdat <- read_csv(file = url('https://raw.githubusercontent.com/fawda123/BASEmetab_script/master/data/apafwoxy.csv'))
+
+fwdatcmp <- fwdat %>% 
+  mutate(
+    DateTimeStamp = dmy_hms(datet, tz = 'America/Jamaica'),
+    Date = as.Date(DateTimeStamp, tz = 'America/Jamaica'),
+    DO_obs = `oxy,mmol/m3`, 
+    a = `aparam,(mmolO2/m2/d)/(W/m2)` / `ht,m`,
+    Rt_vol = `er,mmol/m2/d` / `ht,m`,
+    Pg_vol = `gpp,mmol/m2/d` / `ht,m`,
+    D = -1 * `gasex,mmol/m2/d` / `ht,m`,
+    b = 100 * 3600 * `kw,m/s` / `wspd2,m2/s2` / (`sc,dimensionless` / 660) ^ -0.5 # (m/s)/(m2/s2) to (cm/hr) / (m2/s2)
+  ) %>% 
+  select(Date, DateTimeStamp, DO_obs, a, b, Pg_vol, Rt_vol, D)
+
+p1 <- optex(apagrd, fwdatcmp, asdin = 1, rsdin = 50, bsdin = 0.001, ndaysin = 1) + 
+  labs(
+    subtitle = parse(text = paste(c('ndays=1', 'italic(a)(sd)=1', 'italic(r)(sd)=50', 'italic(b)(sd)=0.001'), collapse = ', '))
+  )
+p2 <- optex(apagrd, fwdatcmp, asdin = 0.01, rsdin = 50, bsdin = 0.001, ndaysin = 7) #+ 
+  # labs(
+  #   subtitle = parse(text = 'ndays=7,'~'italic(a)(sd)=0.01,'~'italic(r)(sd)=50,'~'italic(b)(sd)=0.001')
+  # )
+
+p <- p1 + p2 + plot_layout(ncol = 2, guides = 'collect')
+
[…]

### 2023-01-08 R/funcs.R

@@ -185,101 +185,109 @@ priorcomp <- function(dat, ind){
   
 }
 
-# Fwoxy apa comparison to EBASE for different priors, b mean and sd changes only
-priorcompmean <- function(dat, ind){
+# comparison of fwoxy and ebase results for selected prior at time step of ndays
+optex <- function(apagrd, fwdatcmp, asdin, rsdin, bsdin, ndaysin){
   
-  met <- tibble(
-    lbs = c('r2', 'rmse', 'aved'),
-    lbspr = c('italic(R)^2', 'RMSE', 'Ave.\nDiff.'), 
-    direc = c(-1, 1, 1)
+  res <- apagrd %>% 
+    filter(
+      asd == asdin & rsd == rsdin & bsd == bsdin & ndays == ndaysin
+    ) %>% 
+    pull(out) %>% 
+    .[[1]] %>% 
+    .[[1]]
+  cmp <- inner_join(fwdatcmp, res, by = c('Date', 'DateTimeStamp')) %>%
+    select(-converge, -dDO, -DO_obs.y, -rsq, -matches('lo$|hi$')) %>%
+    rename(
+      DO_mod.x = DO_obs.x,
+      DO_mod.y = DO_mod
+    ) %>%
+    pivot_longer(!all_of(c('DateTimeStamp', 'Date', 'grp')), names_to = 'var', values_to = 'val') %>%
+    separate(var, c('var', 'mod'), sep = '\\.') %>%
+    mutate(
+      mod = case_when(
+        mod == 'x' ~ 'Fwoxy',
+        mod == 'y' ~ 'EBASE'
+      )
+    ) %>%
+    pivot_wider(names_from = 'mod', values_from = 'val')
+  
+  toplo1 <- cmp %>% 
+    filter(var %in% c('Pg_vol', 'Rt_vol', 'D')) %>% 
+    group_by(grp, var) %>% 
+    summarise(
+      Fwoxy = mean(Fwoxy, na.rm = T), 
+      EBASE = mean(EBASE, na.rm = T),
+      Date  = min(Date),
+      .groups = 'drop'
+    ) %>% 
+    pivot_longer(-c(Date, grp, var), names_to = 'model', values_to = 'est') %>% 
+    mutate(
+      var = factor(var, 
+                   levels = c('Pg_vol', 'Rt_vol', 'D'), 
+                   labels = c('Pg [vol]', 'Rt [vol]', 'D')
+      )
[…]

### 2023-01-08 R/figs.R
save file

@@ -95,10 +95,10 @@ fwdatcmp <- fwdat %>%
   ) %>% 
   select(Date, DateTimeStamp, DO_obs, a, b, Pg_vol, Rt_vol, D)
 
-p1 <- optex(apagrd, fwdatcmp, asdin = 1, rsdin = 50, bsdin = 0.001, ndaysin = 1) + 
-  labs(
-    subtitle = parse(text = paste(c('ndays=1', 'italic(a)(sd)=1', 'italic(r)(sd)=50', 'italic(b)(sd)=0.001'), collapse = ', '))
-  )
+p1 <- optex(apagrd, fwdatcmp, asdin = 1, rsdin = 50, bsdin = 0.001, ndaysin = 1) #+ 
+  # labs(
+  #   subtitle = parse(text = paste(c('ndays=1', 'italic(a)(sd)=1', 'italic(r)(sd)=50', 'italic(b)(sd)=0.001'), collapse = ', '))
+  # )
 p2 <- optex(apagrd, fwdatcmp, asdin = 0.01, rsdin = 50, bsdin = 0.001, ndaysin = 7) #+ 
   # labs(
   #   subtitle = parse(text = 'ndays=7,'~'italic(a)(sd)=0.01,'~'italic(r)(sd)=50,'~'italic(b)(sd)=0.001')

### 2023-01-09 R/figs.R
optex fig done, need to add to manu

@@ -95,18 +95,15 @@ fwdatcmp <- fwdat %>%
   ) %>% 
   select(Date, DateTimeStamp, DO_obs, a, b, Pg_vol, Rt_vol, D)
 
-p1 <- optex(apagrd, fwdatcmp, asdin = 1, rsdin = 50, bsdin = 0.001, ndaysin = 1) #+ 
-  # labs(
-  #   subtitle = parse(text = paste(c('ndays=1', 'italic(a)(sd)=1', 'italic(r)(sd)=50', 'italic(b)(sd)=0.001'), collapse = ', '))
-  # )
-p2 <- optex(apagrd, fwdatcmp, asdin = 0.01, rsdin = 50, bsdin = 0.001, ndaysin = 7) #+ 
-  # labs(
-  #   subtitle = parse(text = 'ndays=7,'~'italic(a)(sd)=0.01,'~'italic(r)(sd)=50,'~'italic(b)(sd)=0.001')
-  # )
+subttl <- expression(paste('ndays = 1, ', italic(a), ' (sd) = 1, ', italic(r), ' (sd) = 50, ', italic(b), ' (sd) = 0.001'))
+p1 <- optex(apagrd, fwdatcmp, asdin = 1, rsdin = 50, bsdin = 0.001, ndaysin = 1, subttl = subttl)
 
-p <- p1 + p2 + plot_layout(ncol = 2, guides = 'collect')
+subttl <- expression(paste('ndays = 7, ', italic(a), ' (sd) = 0.01, ', italic(r), ' (sd) = 50, ', italic(b), ' (sd) = 0.001'))
+p2 <- optex(apagrd, fwdatcmp, asdin = 0.01, rsdin = 50, bsdin = 0.001, ndaysin = 7, subttl = subttl, ylbs = F)
 
-png(here('figs/optex.png'), height = 8, width = 8, family = 'serif', units = 'in', res = 500)
+p <- ((p1 + plot_layout(ncol = 1)) | (p2 + plot_layout(ncol = 1)))  + plot_layout(ncol = 2, guides = 'collect') & theme(legend.position = 'top')
+
+png(here('figs/optex.png'), height = 9, width = 8, family = 'serif', units = 'in', res = 500)
 print(p)
 dev.off()
 

### 2023-01-09 R/funcs.R

@@ -186,7 +186,7 @@ priorcomp <- function(dat, ind){
 }
 
 # comparison of fwoxy and ebase results for selected prior at time step of ndays
-optex <- function(apagrd, fwdatcmp, asdin, rsdin, bsdin, ndaysin){
+optex <- function(apagrd, fwdatcmp, asdin, rsdin, bsdin, ndaysin, subttl, ylbs = TRUE){
   
   res <- apagrd %>% 
     filter(
@@ -224,7 +224,7 @@ optex <- function(apagrd, fwdatcmp, asdin, rsdin, bsdin, ndaysin){
     mutate(
       var = factor(var, 
                    levels = c('Pg_vol', 'Rt_vol', 'D'), 
-                   labels = c('Pg [vol]', 'Rt [vol]', 'D')
+                   labels = c('P~(mmol~m^{3}~d^{-1})', 'R~(mmol~m^{3}~d^{-1})', 'D~(mmol~m^{3}~d^{-1})')
       )
     ) %>% 
     select(-grp)
@@ -232,8 +232,8 @@ optex <- function(apagrd, fwdatcmp, asdin, rsdin, bsdin, ndaysin){
   ylab <- expression(paste(O [2], ' (mmol ', m^-3, ' ', d^-1, ')'))
   
   p1 <- ggplot(toplo1, aes(x = Date, y = est, group = model, color = model)) + 
-    geom_line() +
-    geom_point() + 
+    geom_line(alpha = 1) +
+    # geom_point() + 
     facet_wrap(~var, ncol = 1, strip.position = 'left', scales = 'free_y', labeller = label_parsed) + 
     theme_minimal() + 
     theme(
@@ -241,16 +241,18 @@ optex <- function(apagrd, fwdatcmp, asdin, rsdin, bsdin, ndaysin){
       strip.background = element_blank(), 
       legend.position = 'top', 
       legend.title = element_blank(),
-      strip.text = element_text(size = rel(1))
+      strip.text = element_text(size = rel(0.7)), 
+      axis.text.x = element_blank()
     ) + 
     labs(
       x = NULL, 
-      y = ylab
+      y = NULL,
+      subtitle = subttl
     )
   
-  labs <- c('DO[mod]~(mmol~O[2]~m^{3}~d^{-1})',
-            'a~(mmol~m^{-3}~d^{-1})/(W~m^{-2})', 
-            'b~(cm~hr^{-1})/(m^{2}~s^{-2})'
[…]

### 2023-01-11 manu-draft.qmd
optex results text added

@@
 As for the BASEmetab model, relatively uninformed prior distributions following a normal Gaussian distribution are used by default, although the priors can of course be changed based on previous knowledge of parameters specific to an ecosystem or as informed by other metabolic modeling approaches.
-The default priors were based on those that returned reasonable parameter estimates using a forward-feeding oxygen mass balance equation call FWOXY [ForWard OXYgen, @Arriola22].
-For FWOXY, a dissolved oxygen time series is simulated by providing known parameter values in an equation similar to @eq-ebase.
+The default priors were based on those that returned reasonable parameter estimates using a forward-feeding oxygen mass balance equation called Fwoxy [ForWard OXYgen, @Arriola22].
+For Fwoxy, a dissolved oxygen time series is simulated by providing known parameter values in an equation similar to @eq-ebase.
 All of the physical parameters are held constant, e.g., wind speed, temperature, and salinity.
@@
 An initial oxygen concentration is provided and the simulated time series is created by forward addition of the inputs.
-As such, the dissolved oxygen output simulated by FWOXY as change per unit time were modelled by EBASE using the constant physical parameters as input.
-Reasonable prior distributions by EBASE were identified (Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]) based on the ability to return the known parameters from FWOXY:
+As such, the dissolved oxygen output simulated by Fwoxy as change per unit time were modelled by EBASE using the constant physical parameters as input.
+Reasonable prior distributions by EBASE were identified (Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]) based on the ability to return the known parameters from Fwoxy:
 
@@
 
-A preliminary assessment of the ability of EBASE to produce reasonable parameter estimates for $a$, $r$, and $b$ (@eq-ebase) was conducted using a simulated time series created with FWOXY.
+A preliminary assessment of the ability of EBASE to produce reasonable parameter estimates for $a$, $r$, and $b$ (@eq-ebase) was conducted using a simulated time series created with Fwoxy.
 These assessments evaluated if 1) EBASE estimates similar values for the known parameters, and 2) how the estimated values change with different model configurations.
@@
 As such, the results suggest that the default prior distributions used in EBASE may be appropriate in some cases, but increasing the prior variance for $a$ and $r$ and decreasing the variance for $b$ will likely produce more optimal solutions.
-The actual model results for the best performing combination of prior distributions for the one-day and 7-day optimization is shown in @fig-optex (all results are averaged at the time-step of the optimization period for comparability).
+The actual model results for the best performing combination of prior distributions for the one-day and 7-day optimization are shown in @fig-optex (all results are averaged at the time-step of the optimization period for comparability).
+The best performing models were those with the highest mean $R^2$ values across all parameter estimates in @fig-apasumdat.
+For both models, the metabolic parameters were estimated with high precision, and in some cases nearly perfect precision (i.e., dissolved oxygen and gas exchange, D).
+Estimates for the fixed $b$ parameter were most variable, but not biased as the mean for both models was close to 0.252 $\buni$.
 
@@
 
-![Comparison of EBASE results with known metabolic parameters from a simulated time series.
+![Comparison of EBASE results with known metabolic results from a simulated time series created with Fwoxy.
 Results are compared for unique combination of prior distributions, where the variances are evaluated from low (L), medium (M), to high (H) for a one-day and seven-day model optimization period.
@@
 The $b$ parameter is not evaluated because it is a fixed constant for the simulated data.](figs/apasumdat.png){#fig-apasumdat}
+
+\newpage
+
+![Comparison of EBASE results for the best performing model with known metabolic results from a simulated time series created with Fwoxy.
+The best performing model for the one-day (left, ndays = 1) and seven-day (right, ndays = 7) optimization period is based on the EBASE results with prior distributions that produced the highest mean $R^2$ values across all parameter estimates (@fig-apasumdat).
+The standard deviations of the prior distributions of each parameter for the best performing model are shown at the top.
+The results are averaged at the time step of the model optimization period.
+The $b$ parameter is fixed at 0.251 $\buni$.](figs/optex.png){#fig-optex fig-align="center"}
 

### 2023-01-11 manu-draft.qmd
methods and results text first draft maybe done

@@
 The best performing models were those with the highest mean $R^2$ values across all parameter estimates in @fig-apasumdat.
-For both models, the metabolic parameters were estimated with high precision, and in some cases nearly perfect precision (i.e., dissolved oxygen and gas exchange, D).
+For both models, the metabolic parameters were estimated with high precision, and in some cases with nearly perfect precision (i.e., dissolved oxygen and gas exchange, D).
 Estimates for the fixed $b$ parameter were most variable, but not biased as the mean for both models was close to 0.252 $\buni$.
@@
 Metabolic estimates from EBASE were further compared to those from existing methods to benchmark and evaluate potential differences from alternative model formulations used by each approach.
-The same one year continuous time series described above for water quality and weather data from Apalachicola Bay was used to compare metabolic estimates from EBASE, BASEmetab, and the Odum open-water method [@Odum56].
-The Odum method used an approach described in @Murrell18 created for estuarine application that used a gas-exchange approach described in @Thebault08.
+The observed one year continuous time series described above for water quality and weather data from Apalachicola Bay was used to compare metabolic estimates from EBASE, BASEmetab, and the Odum open-water method [@Odum56].
+The Odum method used an approach described in @Murrell18 created for estuarine application that uses a gas-exchange approach described in @Thebault08.
 For better comparison with the other methods, the gas exchange parmaterization in the Odum method was modified to use the @Wanninkhof14 approach [as available in the WtRegDO R package, @Beck21].
@@
 All other prior distributions for BASEmetab and EBASE were set as the software default.
-A one-day model optimization period was used for EBASE.
+A one-day model optimization period was used for EBASE for comparability with BASEmetab.
 
-The common set of metabolic parameters returned by the Odum, BASEmetab, and EBASE methods are net ecosystem metabolism, gross production, respiration, and gas exchange.
+The common set of metabolic parameters returned by the EBASE, BASEmetab, and Odum EBASE methods are net ecosystem metabolism, gross production, respiration, and gas exchange.
 Pairwise comparisons of the daily estimates returned by each method were evaluated using simple summary statistics including Pearson correlation coefficients and linear regression fits to evaluate potential biases in each method.
 For the latter analysis, differences in intercept estimates less than or greater than zero suggested a bias in magnitude, whereas differences in slope estimates less than or greater than one suggested a bias that depended on relative value of each observation (e.g., bias may differ between lower or higher values).
-Comparisons of the slope and estimates values to zero and one, respectively, were based on confidence intervals from the standard errors of each estimate.
+Comparisons of the intercept and slope estimates to zero and one, respectively, were based on confidence intervals from the standard errors of each.
 Results were assessed separately for estimates based on the observed or detided dissolved oxygen time series.
@@
 @fig-apacmpfig and `r run_reference(id = "apacmptab")` provide a comparison of the EBASE metabolic estimates with those from the Odum and BASEmetab methods.
+Correlations for all estimates, regardless of method or the input time series, were positive and significantly correlated.
+The only exception was the comparison of respiration for the Odum and EBASE methods ($\rho$ = 0.11), where the former produced a much larger range of estimates.
+The intercept values were all significant, suggesting some level of bias when comparing estimates between the methods.
+Slope estimates were all significantly different from one for the comparisons using the observed dissolved oxygen time series as input, whereas only three of the comparisons were significantly different using the detided time series.
+This latter comparison suggests that detiding the time series produces more comparable estimates between methods.
+Similarly, the standard deviation ($\rho$) of the residuals of each regression comparison were generally lower in the detided results for all comparisons.
+Variation in the gas exchange estimates were also more similar comparing the Odum and EBASE methods that used the same formulation [@Wanninkhof14] regardless of input time series.
 
-### Application to representative NERR sites
+As noted in the previous paragraph, metabolic estimates between the methods were more similar for results using the detided dissolved oxygen time series ([@fig-apacmpfig]b) as compared to using the observed dissolved oxygen as input ([@fig-apacmpfig]b).
+This suggests that tidal effects on continuous monitoring data can severely violate assumptions required for methods of estimating ecosystem metabolism [as described in detail in @Beck15].
+These effects are easily identified with estimates from the Odum method as negative values for production (P) and positive values for respiration (R), i.e., "anomalous" values.
+The effects of tidal advection on estimates from EBASE and BASEmetab are more subtle as anomalous values cannot be returned based on constraints in the mass-balance equations (e.g., it is impossible for respiration to be positive).
+However, the effects of tidal advection produce a wider range of estimates for NEM in EBASE and BASEmetab for the observed dissolved oxygen time series ([@fig-apacmpfig]a), caused by greater variation in the production and respiration estimates.
+As such, detiding the dissolved oxygen time series prior to estimating metabolic rates produces more stable (lower standard deviation) and smaller ranges (`r run_reference(id = "apacmptab")`), regardless of the method used.
 
@@
 
-<!-- \newpage -->
+\newpage
 
 ## Figures {.unnumbered}
-
-Figure 1: Fwoxy comparisons
-
-Figure 2: Apalachicola comparisons
-
-Figure 3: Sensitivity analyses
-
-Figure 4: NERRS application
-
-\newpage
 
@@
 ![Comparison of EBASE results for the best performing model with known metabolic results from a simulated time series created with Fwoxy.
-The best performing model for the one-day (left, ndays = 1) and seven-day (right, ndays = 7) optimization period is based on the EBASE results with prior distributions that produced the highest mean $R^2$ values across all parameter estimates (@fig-apasumdat).
+The best performing model for the one-day (a, ndays = 1) and seven-day (b, ndays = 7) optimization period is based on the EBASE results with prior distributions that produced the highest mean $R^2$ values across all parameter estimates (@fig-apasumdat).
 The standard deviations of the prior distributions of each parameter for the best performing model are shown at the top.

### 2023-01-11 R/figs.R

@@ -95,15 +95,15 @@ fwdatcmp <- fwdat %>%
   ) %>% 
   select(Date, DateTimeStamp, DO_obs, a, b, Pg_vol, Rt_vol, D)
 
-subttl <- expression(paste('ndays = 1, ', italic(a), ' (sd) = 1, ', italic(r), ' (sd) = 50, ', italic(b), ' (sd) = 0.001'))
+subttl <- expression(paste('(a) ndays = 1, ', italic(a), ' (sd) = 1, ', italic(r), ' (sd) = 50, ', italic(b), ' (sd) = 0.001'))
 p1 <- optex(apagrd, fwdatcmp, asdin = 1, rsdin = 50, bsdin = 0.001, ndaysin = 1, subttl = subttl)
 
-subttl <- expression(paste('ndays = 7, ', italic(a), ' (sd) = 0.01, ', italic(r), ' (sd) = 50, ', italic(b), ' (sd) = 0.001'))
+subttl <- expression(paste('(b) ndays = 7, ', italic(a), ' (sd) = 0.01, ', italic(r), ' (sd) = 50, ', italic(b), ' (sd) = 0.001'))
 p2 <- optex(apagrd, fwdatcmp, asdin = 0.01, rsdin = 50, bsdin = 0.001, ndaysin = 7, subttl = subttl, ylbs = F)
 
 p <- ((p1 + plot_layout(ncol = 1)) | (p2 + plot_layout(ncol = 1)))  + plot_layout(ncol = 2, guides = 'collect') & theme(legend.position = 'top')
 
-png(here('figs/optex.png'), height = 9, width = 8, family = 'serif', units = 'in', res = 500)
+png(here('figs/optex.png'), height = 8.75, width = 8, family = 'serif', units = 'in', res = 500)
 print(p)
 dev.off()
 

### 2023-01-11 R/tabs.R

@@ -3,6 +3,10 @@ library(tidyverse)
 library(lubridate)
 library(here)
 
+source(file = here('R/funcs.R'))
+
+# comparison of EBASE to other methods with actual apa data -----------------------------------
+
 load(file = url('https://github.com/fawda123/BASEmetab_script/raw/master/data/apacmp.RData'))
 
 tosum <- apacmp %>% 

### 2023-01-12 manu-draft.qmd
model io table done

@@
 
+load(file = here('tabs/ebaseiotab.RData'))
 load(file = here('tabs/apacmptab.RData'))
@@
 The estimated dissolved oxygen concentration is also returned, which can be compared to observed as one measure of model performance.
+All parameter inputs and outputs for EBASE are shown in `r run_reference(id = "ebaseiotab")`.
 
@@
 
-Table 1: EBASE parameters and outputs
-
-Table 2: Summary stats of Fwoxy comparison
-
-Table 3: Summary stats of Apalachicola comparison
-
-Table 4: Key parameters form NERRS application
+```{r}
+#| label: ebaseiotab
+#| tbl-cap: "EBASE model parameters showing the required input, parameters derived from the model input, and model output.
+The model notation shows the components decribed in @eq-ebase.
+The model notation for production, respiration, and gas exchange shows both the simple parameter name and the components of @eq-ebase used to estimate the parameter."
+ebaseiotab
+```
 

### 2023-01-12 R/tabs.R

@@ -5,6 +5,65 @@ library(here)
 
 source(file = here('R/funcs.R'))
 
+# EBASE i/o -----------------------------------------------------------------------------------
+
+totab <- tibble(
+  Type = c('Input', 'Input', 'Input', 'Input', 'Input', 'Input',
+           'Input-derived', 'Input-derived', 'Input-derived',
+           'Output', 'Output', 'Output', 'Output', 'Output', 'Output'),
+  Description = c('Dissolved oxygen', 'Water temperature', 'Salinity', 'Total photosynthetically active radiation', 'Wind speed', 'Water column depth',
+                  'Wind speed at 10 meter height, squared', 'Schmidt number (from water temperature and salinity)', 'Dissolved oxygen at saturation (from water temperature and salinity)',
+                  'Dissolved oxygen (modelled)', 'Production', 'Respiration', 'Gas exchange', 'Light efficiency', 'b'),
+  `Model notation` =  c('C_d', '-', '-', 'PAR', '-', 'H', 
+                        'U102', 'Schmidt number', 'C_sat', 
+                        'C_mod', 'Pg_vol', 'Rt_vol',  'D', 'a', 'b'),
+  Units = c('mg/L', 'C', 'psu', 'W/m2/s', 'm/s', 'm', 
+            'm2', 'unitless', 'mmol/m3',
+            'mmol/m3', 'mmol O2 m3/d', 'mmol O2 m3/d', 'mmol O2 m3/d', '(mmol/m3/d)/(W/m2)', '(cm/hr)/(m2/s2)')
+) %>% 
+  flextable::as_grouped_data(groups = 'Type')
+
+eqsz <- fp_text_default(font.size = 9)
+# col2wid <- 1
+# col3wid <- 2
+# col14wid <- 6.5 - (col2wid + col3wid)
+ebaseiotab <- flextable(totab) %>% 
+  flextable::compose(i = 2, j = 3, value = as_paragraph(as_equation('C_d', props = eqsz))) %>% 
+  flextable::compose(i = 5, j = 3, value = as_paragraph(as_equation('PAR', props = eqsz))) %>% 
+  flextable::compose(i = 7, j = 3, value = as_paragraph(as_equation('H', props = eqsz))) %>% 
+  flextable::compose(i = 9, j = 3, value = as_paragraph(as_equation('U_{10}^2', props = eqsz))) %>% 
[…]

### 2023-01-15 manu-draft.qmd
proofreading results and methods

@@
 * Overview of metabolism - importance
-* Existing methods: Odum/WtRegDO, BASEmetab, references in @Grace15, applications to lakes and streams
+* Existing methods: Odum/WtRegDO, BASE, references in @Grace15, applications to lakes and streams
 * Applications to estuaries - challenges
@@
 The Estuarine BAyesian Single station Estimation (EBASE) method provides a simple approach to quantify metabolic parameters from continuous dissolved oxygen time series.
-The EBASE method is a modified version of the BASEmetab mass balance equation described in @Grace15.
+The EBASE method is a modified version of the BASE mass balance equation described in @Grace15.
 The original equation is:
@@
 
-where the terms in the square brackets are production (P), respiration (R), and gas exchange (D), respectively, modeled as a function of the change in dissolved oxygen ($\delta{C_d}$) per unit time of the input data ($\delta{t}$).
-Five unknown parameters are estimated by the model: $a$, $p$, $r$, $\Theta$, and $K$.
-The required inputs are the time series of surface irradiance ($PAR$, $\mu$mol/m$^2$/s), water temperature ($T$, C), atmospheric pressure (atm), diel oxygen ($C_d$, mmol/m$^2$), and oxygen saturation ($C_{sat}$, mmol/m$^3$).
+where the terms in the square brackets are production (P), respiration (R), and gas exchange (D), respectively, modeled as a function of the change in dissolved oxygen ($\delta{C_d}$) per unit time ($\delta{t}$) of the input data.
+Five unknown parameters are estimated from the data: $a$, $p$, $r$, $\Theta$, and $K$.
+The required inputs are the time series of surface irradiance ($PAR$, $\mu$mol/m$^2$/s), water temperature ($T$, C), barometric pressure (atm), diel oxygen ($C_d$, mmol/m$^2$), and oxygen saturation ($C_{sat}$, mmol/m$^3$).
 Dissolved oxygen at saturation $C_{sat}$ is calculated from the input data as a function of water temperature, salinity, and barometric pressure.
@@
 
-The proportionality parameter, $k_w$, usually called the gas transfer velocity because it has the dimensions of length over time, increases as the water column in the vicinity of the air–water interface gets more turbulent.
+The proportionality parameter, $k_w$, usually called the gas transfer velocity because it has the dimensions of length over time, increases as the water column in the vicinity of the air–water interface becomes more turbulent.
 As such, $k_w$  is usually modeled as a function of measurable variables that can potentially predict turbulent mixing of the surface water, such as wind speed, current velocity, and water depth [@Ho16].
@@
 
-Instead of one of the traditional wind-based parameterization used in oceanography [e.g., @Wanninkhof14], @Grace15 uses the following for the gas transfer in BASEmetab:
+Instead of one of the traditional wind-based parameterization used in oceanography [e.g., @Wanninkhof14], @Grace15 used the following for the gas transfer in BASE:
 
@@
 The second modification to @eq-base for EBASE is a simplification to remove the $p$ and $\Theta$ parameters from the production (P) and respiration (R) components, respectively, of the mass balance equation.
-The BASEmetab method provides options to estimate both $\Theta$ and $p$ from the data or to set as fixed values defined in the model file ($\Theta = 1.07177$, $p=1$).
+The BASE method provides options to estimate both $\Theta$ and $p$ from the data or to set as fixed values defined in the model file ($\Theta = 1.07177$, $p=1$).
 For EBASE, the temperature dependency of respiration is removed by setting $\Theta^{(T-\bar{T})}$ = 1.
 The $p$ parameter, which is dimensionless, is also set to 1 as in @Holtgrieve10.
-Removing these components mproved model stability by creating more reasonable estimates for the other parameters.
+Removing these components improved model stability by creating more reasonable estimates for the other parameters.
 This further simplifies @eq-basewan to the final form for EBASE:
@@
 
-As for BASEmetab, the metabolic estimates in EBASE are defined by the change in dissolved oxygen over the time step, $\delta{C_d}/\delta{t}$, where gross production is provided by $aPAR$, respiration is provided by $r$, and gas exchange is provided by the remainder [as in @Wanninkhof14].
+As for BASE, the metabolic estimates in EBASE are defined by the change in dissolved oxygen over the time step, $\delta{C_d}/\delta{t}$, where gross production is provided by $aPAR$, respiration is provided by $r$, and gas exchange is provided by the remainder [as in @Wanninkhof14].
 Units for each of the metabolic estimates are $\runi$.
@@
 The remaining three parameters $a$ (light efficiency, $\auni$), $r$ (respiration, $\runi$), and $b$ ($\buni$) are estimated by likelihood given the observed data and prior distributions for the parameters.
-The JAGS software [@Plummer03] is used with the EBASE R package [described below, @Beck22;@RCT22] to estimate the unknown parameters using Markov Chain Monte Carlo simulations in a Bayesian framework.
+The JAGS software [as for BASE, @Plummer03] is used with the EBASE R package [described below, @Beck22;@RCT22] to estimate the unknown parameters using Markov Chain Monte Carlo simulations in a Bayesian framework.
 At each time step, the change in oxygen concentration between time steps is calculated from the equation using model inputs and parameter guesses, and then a finite difference approximation is used to estimate modeled oxygen concentration.
 Estimations proceed using a forward in time integration where the dissolved oxygen at each time step is informed by parameter estimates from the preceding time steps.
-The estimated dissolved oxygen concentration is also returned, which can be compared to observed as one measure of model performance.
+The estimated dissolved oxygen concentration is also returned, which can be compared to the observed values as one measure of model performance.
 All parameter inputs and outputs for EBASE are shown in `r run_reference(id = "ebaseiotab")`.
@@
 The metabolism estimates and their parameters returned by the Bayesian routine implemented in JAGS are affected by the prior distributions assigned to each.
-As for the BASEmetab model, relatively uninformed prior distributions following a normal Gaussian distribution are used by default, although the priors can of course be changed based on previous knowledge of parameters specific to an ecosystem or as informed by other metabolic modeling approaches.
+As for the BASE model, relatively uninformed prior distributions following a normal Gaussian distribution are used by default, although the priors can be changed based on previous knowledge of parameters specific to an ecosystem or as informed by other metabolic modeling approaches.
 The default priors were based on those that returned reasonable parameter estimates using a forward-feeding oxygen mass balance equation called Fwoxy [ForWard OXYgen, @Arriola22].
@@
 
+Each follows a normal distribution with means and standard deviation defined in the parentheses (@fig-priorplot).
 The $T$ notation is JAGS model syntax indicating the distributions are truncated to a minimum (left) and maximum (right) value defined in the parentheses.
@@
 Initial development of EBASE also showed that using an undefined upper limit of $b$ led to unstable and unreasonable parameter estimates.
-The density distributions for the default priors in EBASE are shown in @fig-priorplot.
 
@@
 An additional option also allows the user to specify the model optimization period, where the time period can vary from a minimum of one day to the maximum of the entire length of the time series.
-This differs from the BASEmetab approach where the optimization period is set at one day.
+This differs from the BASE approach where the optimization period is set at one day.
 Sensitivity of the EBASE method to different optimization periods is described in @sec-cmpmod.
@@
 Known values for $a$ and $r$ parameters were returned for each time step of the simulated data, as well as all metabolic estimates for production, respiration, and gas exchange ([@fig-simapa]b and c).
-The only parameter that could not be simulated by Fwoxy was $b$, which was set to 0.251 $\buni$.
+The only parameter that could not be simulated by Fwoxy was $b$, which was set to 0.251 $\buni$ [@Wanninkhof14].
 
@@
 First, results were compared to the known values after changing characteristics of their prior distributions.
-Second, results were compared using a model period of optimization for one day and one week of observations.
+Second, results were compared using a model period of optimization for one day and seven days.
 Details of these analyses are as follows.
@@
 The optimization period used for a time series determines how much data are used to estimate the three EBASE parameters ($a$, $r$, and $b$).
-Unlike BASEmetab, the EBASE software allows flexibility in the period of time used for optimization, with a minimum period of time being one day and a maximum period of being the length of the time series.
+Unlike BASE, the EBASE software allows flexibility in the period of time used for optimization, with a minimum period of time being one day and a maximum period being the length of the time series.
 For a chosen optimization period, a single estimate for $r$ and $b$ is returned by the model, whereas $n$ number of observations equal to the number of time steps in each optimization period are returned for the $a$ parameter (e.g., 96 values for a 24 hour time period if the sampling interval is 15 minutes).
-Short periods of observation may return unstable parameter estimates if forcing factors (e.g., wind, temperature) that affect estuarine biogeochemical rates occur at time scales longer the chosen period of optimization for EBASE.
+A short period of optimization (e.g., one day) may return unstable parameter estimates if forcing factors (e.g., wind, temperature) that affect estuarine biogeochemical rates occur at time scales longer the chosen period of optimization for EBASE.
 As such, potentially longer optimization periods may return more stable and robust estimates given that more data are used for parameter estimation.
@@
 Changing the variance of the $a$ parameter had little effect on any of the comparisons.
-Smaller variance of the prior distribution for the $r$ parameter generally contributed to lower $R^2$ for all parameters, with the precision being worse for the 7-day optimization period (e.g., the $a$ parameter).
-Larger variance of the prior distribution for the $b$ parameter generally contributed to lower $R^2$ only for the gas exchange estimate (D), although this effect was minimized with the 7-day optimization period.
-As such, the results suggest that the default prior distributions used in EBASE may be appropriate in some cases, but increasing the prior variance for $a$ and $r$ and decreasing the variance for $b$ will likely produce more optimal solutions.
-The actual model results for the best performing combination of prior distributions for the one-day and 7-day optimization are shown in @fig-optex (all results are averaged at the time-step of the optimization period for comparability).
+Smaller variance of the prior distribution for the $r$ parameter generally contributed to lower $R^2$ for all parameters, although the precision was worse for the seven day optimization period (e.g., the $a$ parameter).
+Conversely, larger variance of the prior distribution for the $b$ parameter generally contributed to lower $R^2$ only for the gas exchange estimate (D), although this effect was minimized with the seven day optimization period.
+
+The results suggest that the default prior distributions used in EBASE may be appropriate in some cases, but increasing the prior variance for $a$ and $r$ and decreasing the variance for $b$ will likely produce more optimal solutions.
+The actual model results for the best performing combination of prior distributions for the one day and seven day optimization are shown in @fig-optex (all results are averaged at the time-step of the optimization period for comparability).
 The best performing models were those with the highest mean $R^2$ values across all parameter estimates in @fig-apasumdat.
 For both models, the metabolic parameters were estimated with high precision, and in some cases with nearly perfect precision (i.e., dissolved oxygen and gas exchange, D).
-Estimates for the fixed $b$ parameter were most variable, but not biased as the mean for both models was close to 0.252 $\buni$.
+Estimates for the fixed $b$ parameter were most variable, but not biased as the mean for both models was close to 0.251 $\buni$.
 
@@
 Metabolic estimates from EBASE were further compared to those from existing methods to benchmark and evaluate potential differences from alternative model formulations used by each approach.
-The observed one year continuous time series described above for water quality and weather data from Apalachicola Bay was used to compare metabolic estimates from EBASE, BASEmetab, and the Odum open-water method [@Odum56].
+The observed one year continuous time series described above for water quality and weather data from Apalachicola Bay was used to compare metabolic estimates from EBASE, BASE, and the Odum open-water method [@Odum56].
 The Odum method used an approach described in @Murrell18 created for estuarine application that uses a gas-exchange approach described in @Thebault08.
-For better comparison with the other methods, the gas exchange parmaterization in the Odum method was modified to use the @Wanninkhof14 approach [as available in the WtRegDO R package, @Beck21].
-The input data were provided to each method after converting the units as required (e.g., BASEmetab requires PAR as $\mu$mol m$^{-2}$ s$^{-1}$).
-Water column depth at Cat Point is approximately 1.8 m and all daily metabolic estimates returned by each method were converted to volumetric rates by dividing by the depth ($\mu$mol O$_2$ m$^{-3}$ d$^{-1}$).
-Each comparison also evaluated metabolic estimates using the observed dissolved oxygen time series and a detided time series using weighted regression [@Beck15].
+For better comparison with the EBASE methods, the gas exchange parmaterization in the Odum method was modified to use the @Wanninkhof14 approach [as available in the WtRegDO R package, @Beck21].
+The input data were provided to each method after converting the units as required (e.g., BASE requires PAR as $\mu$mol/m$^{2}$/s).
+Water column depth at Cat Point is approximately 1.8 m and all daily metabolic estimates returned by each method were converted to volumetric rates by dividing by the depth (i.e., $\runi$).
+Each comparison also evaluated metabolic estimates using the observed dissolved oxygen time series and a detided time series using weighted regression [@Beck15; available from @Beck21].
 The latter comparison provided an assessment of metabolic rates that are expected to be minimally influenced by tidal advection.
-All metabolic estimates were obtained using the default settings for each of the three methods, with an exception that the prior distribution for the reaeration coefficient ($K$) in BASEmetab was informed by results from the Odum output.
-Initial assessments indicated that unstable and inaccurate results are obtained from BASEmetab using an uninformed prior distribution for $K$.
-All other prior distributions for BASEmetab and EBASE were set as the software default.
-A one-day model optimization period was used for EBASE for comparability with BASEmetab.
+All metabolic estimates were obtained using the default settings for each of the three methods, with an exception that the prior distribution for the reaeration coefficient ($K$) in BASE was informed by results from the Odum output.
+Initial assessments indicated that unstable and inaccurate results for the time series are obtained from BASE using an uninformed prior distribution for $K$.
+All other prior distributions for BASE and EBASE were set as the software default.
+A one day model optimization period was used for EBASE for comparability with BASE.
 
-The common set of metabolic parameters returned by the EBASE, BASEmetab, and Odum EBASE methods are net ecosystem metabolism, gross production, respiration, and gas exchange.
-Pairwise comparisons of the daily estimates returned by each method were evaluated using simple summary statistics including Pearson correlation coefficients and linear regression fits to evaluate potential biases in each method.
+The common set of metabolic parameters returned by the Odum, BASE, and EBASE methods are net ecosystem metabolism, gross production, respiration, and gas exchange.
+Pairwise comparisons of the daily estimates returned by each method were evaluated using simple summary statistics, including Pearson correlation coefficients and linear regression fits to evaluate potential biases in each method.
 For the latter analysis, differences in intercept estimates less than or greater than zero suggested a bias in magnitude, whereas differences in slope estimates less than or greater than one suggested a bias that depended on relative value of each observation (e.g., bias may differ between lower or higher values).
@@
 
-@fig-apacmpfig and `r run_reference(id = "apacmptab")` provide a comparison of the EBASE metabolic estimates with those from the Odum and BASEmetab methods.
-Correlations for all estimates, regardless of method or the input time series, were positive and significantly correlated.
+@fig-apacmpfig and `r run_reference(id = "apacmptab")` provide a comparison of the EBASE metabolic estimates with those from the Odum and BASE methods.
+Correlations for all estimates, regardless of method or the input time series, were positive and significantly correlated, as expected.
 The only exception was the comparison of respiration for the Odum and EBASE methods ($\rho$ = 0.11), where the former produced a much larger range of estimates.
 The intercept values were all significant, suggesting some level of bias when comparing estimates between the methods.
-Slope estimates were all significantly different from one for the comparisons using the observed dissolved oxygen time series as input, whereas only three of the comparisons were significantly different using the detided time series.
+Slope estimates were all significantly different from one for the comparisons using the observed dissolved oxygen time series as input, whereas only three of the eight comparisons were significantly different using the detided time series.
 This latter comparison suggests that detiding the time series produces more comparable estimates between methods.
@@
 
-As noted in the previous paragraph, metabolic estimates between the methods were more similar for results using the detided dissolved oxygen time series ([@fig-apacmpfig]b) as compared to using the observed dissolved oxygen as input ([@fig-apacmpfig]b).
+As noted in the previous paragraph, metabolic estimates between the methods were more similar for results using the detided dissolved oxygen time series ([@fig-apacmpfig]b) as compared to using the observed dissolved oxygen as input ([@fig-apacmpfig]a).
 This suggests that tidal effects on continuous monitoring data can severely violate assumptions required for methods of estimating ecosystem metabolism [as described in detail in @Beck15].
 These effects are easily identified with estimates from the Odum method as negative values for production (P) and positive values for respiration (R), i.e., "anomalous" values.
-The effects of tidal advection on estimates from EBASE and BASEmetab are more subtle as anomalous values cannot be returned based on constraints in the mass-balance equations (e.g., it is impossible for respiration to be positive).
-However, the effects of tidal advection produce a wider range of estimates for NEM in EBASE and BASEmetab for the observed dissolved oxygen time series ([@fig-apacmpfig]a), caused by greater variation in the production and respiration estimates.
-As such, detiding the dissolved oxygen time series prior to estimating metabolic rates produces more stable (lower standard deviation) and smaller ranges (`r run_reference(id = "apacmptab")`), regardless of the method used.
+The effects of tidal advection on estimates from EBASE and BASE are more subtle as anomalous values cannot be returned based on constraints in the mass-balance equations (e.g., it is impossible for respiration to be positive).
+However, the effects of tidal advection produced a wider range of estimates for NEM in EBASE and BASE compared to the Odum method ([@fig-apacmpfig]a), caused by greater variation in the production and respiration estimates.
+Overall, detiding the dissolved oxygen time series prior to estimating metabolic rates produces more stable (lower standard deviation) and smaller ranges (`r run_reference(id = "apacmptab")`), regardless of the method used.
 
@@
 
-![Simulated time series using the FWOXY model for one year of continuous data at Apalachicola Bay.
-Simulated input to EBASE was the (a) dissolved oxygen time series, with results compared to the (b) simulated $a$ parameter and (c) simulatd metabolic estimates for production (P), respiration (R), and gas exchange (D).](figs/simapa.png){#fig-simapa}
+![Simulated time series using the Fwoxy model for one year of continuous data at Apalachicola Bay.
+Simulated input to EBASE was the (a) dissolved oxygen time series, with results compared to the (b) simulated $a$ parameter and (c) simulated metabolic estimates for production (P), respiration (R), and gas exchange (D).](figs/simapa.png){#fig-simapa}
 
@@
 
-![Comparison of EBASE results with known metabolic results from a simulated time series created with Fwoxy.
-Results are compared for unique combination of prior distributions, where the variances are evaluated from low (L), medium (M), to high (H) for a one-day and seven-day model optimization period.
+![Comparison of EBASE results to a simulated time series created with known metabolic parameters.
+Results are compared for unique combination of prior distributions, where the standard deviations are evaluated from low (L), medium (M), to high (H) for a one day and seven day model optimization period.
 The metabolic results for dissolved oxygen (DO), production (P), respiration (R), gas exchange (D), and the $a$ parameter are evaluated based on the $R^2$ value of the linear fit between the EBASE results and simulated values.
-The $b$ parameter is not evaluated because it is a fixed constant for the simulated data.](figs/apasumdat.png){#fig-apasumdat}
+The $b$ parameter is a fixed constant and was not evaluated.](figs/apasumdat.png){#fig-apasumdat}
 
@@
 
-![Comparison of EBASE results for the best performing model with known metabolic results from a simulated time series created with Fwoxy.
-The best performing model for the one-day (a, ndays = 1) and seven-day (b, ndays = 7) optimization period is based on the EBASE results with prior distributions that produced the highest mean $R^2$ values across all parameter estimates (@fig-apasumdat).
+![Comparison of EBASE results for the best performing model to a simulated time series created with known metabolic parameters.
+The best performing model for the one day (a, ndays = 1) and seven day (b, ndays = 7) optimization period is based on the EBASE results with prior distributions that produced the highest mean $R^2$ values across all parameter estimates (@fig-apasumdat).
 The standard deviations of the prior distributions of each parameter for the best performing model are shown at the top.
@@
 
-![Comparisons of daily metabolic estimates from the Odum, BASEmetab, and EBASE methods for one year of continuous monitoring data at Apalachicola Bay.
+![Comparisons of daily metabolic estimates from the Odum, BASE, and EBASE methods for one year of continuous monitoring data at Apalachicola Bay.
 Results for net ecosystem metabolism (NEM), gross production (P), respiration (R), and gas exchange (D) are compared using the (a) observed and (b) detided dissolved oxygen time series as input to each model.
 All values are volumetric as mmol O$_2$ m$^3$ d$^{-1}$ and are grouped by the dry or wet season.
-Black lines show the 1:1 comparison, the blue line shows the linea regression fit, and the Pearson correlation values are shown in parentheses for each facet title.](figs/apacmpfig.png){#fig-apacmpfig fig-align="center"}
+Black lines show the 1:1 comparison and the red lines show the linear regression fit.](figs/apacmpfig.png){#fig-apacmpfig fig-align="center"}
 
@@
 #| tbl-cap: "EBASE model parameters showing the required input, parameters derived from the model input, and model output.
-The model notation shows the components decribed in @eq-ebase.
-The model notation for production, respiration, and gas exchange shows both the simple parameter name and the components of @eq-ebase used to estimate the parameter."
+The model notation for production, respiration, and gas exchange shows both the simple parameter name and the full model component used to estimate the parameter."
 ebaseiotab
@@
 #| label: apacmptab
-#| tbl-cap: "Summary statisics of daily metabolic estimates comparing Odum and BASEmetab to EBASE for net ecosystem metabolism (NEM), gross production (P), respiration (R), and gas exchange (D).
+#| tbl-cap: "Summary statisics of daily metabolic estimates comparing Odum and BASE to EBASE for net ecosystem metabolism (NEM), gross production (P), respiration (R), and gas exchange (D).
 Separate comparisons are made for estimates based on observed and detided dissolved oxygen.

### 2023-01-15 R/figs.R

@@ -4,6 +4,7 @@ library(here)
 library(patchwork)
 library(EBASE)
 library(readr)
+library(scales)
 
 source(here('R/funcs.R'))
 
@@ -53,6 +54,10 @@ p2 <- ggplot(fwdatcmp, aes(x = DateTimeStamp, y = a)) +
   )
 
 p3 <- ebase_plot(fwdatcmp, instantaneous = F) +
+  ggplot2::scale_color_discrete(
+    breaks = c('Pg_vol', 'Rt_vol', 'D'),
+    labels = c('P', 'R', 'D')
+  ) +
   scale_x_date(breaks = "1 month", labels = date_format("%b")) +
   labs(title = '(c) Simulated metabolic estimates')
 
@@ -113,6 +118,7 @@ load(file = url('https://github.com/fawda123/BASEmetab_script/raw/master/data/ap
 
 toplo <- apacmp %>% 
   mutate(
+    typ = ifelse(typ == 'BASEmetab', 'BASE', typ),
     val = case_when(
       var == 'Rt_vol' ~ -1 * val, 
       T ~ val
@@ -129,9 +135,9 @@ toplo <- apacmp %>%
   pivot_wider(names_from = 'typ', values_from = 'val')
 
 p1a <- apacmp_plo(toplo, xlb = NULL, ylb = 'Odum', dotyp = 'observed', addtitle = T)
-p1b <- apacmp_plo(toplo, xlb = NULL, ylb = 'BASEmetab', dotyp = 'observed', addtitle = F)
+p1b <- apacmp_plo(toplo, xlb = NULL, ylb = 'BASE', dotyp = 'observed', addtitle = F)
 p2a <- apacmp_plo(toplo, xlb = NULL, ylb = 'Odum', dotyp = 'detided', addtitle = T)
-p2b <- apacmp_plo(toplo, xlb = 'EBASE', ylb = 'BASEmetab', dotyp = 'detided', addtitle = F)
+p2b <- apacmp_plo(toplo, xlb = 'EBASE', ylb = 'BASE', dotyp = 'detided', addtitle = F)
 
 p <- p1a[[1]] + p1a[[2]] + p1a[[3]] + p1a[[4]] + p1b[[1]] + p1b[[2]] + p1b[[3]] + p1b[[4]] + 
   p2a[[1]] + p2a[[2]] + p2a[[3]] + p2a[[4]] + p2b[[1]] + p2b[[2]] + p2b[[3]] + p2b[[4]] + 

### 2023-01-15 R/funcs.R

@@ -155,7 +155,7 @@ priorcomp <- function(dat, ind){
     scale_y_reverse(expand = c(0, 0), breaks = toplo1$ind, labels = toplo1$def) + 
     labs(
       y = NULL, 
-      x = 'Variance of prior',
+      x = 'St. dev. of prior',
       caption = '* EBASE default'
     )
   

### 2023-01-15 R/tabs.R

@@ -35,8 +35,8 @@ ebaseiotab <- flextable(totab) %>%
   flextable::compose(i = 10, j = 3, value = as_paragraph(as_equation('S_c', props = eqsz))) %>% 
   flextable::compose(i = 11, j = 3, value = as_paragraph(as_equation('C_{sat}', props = eqsz))) %>% 
   flextable::compose(i = 13, j = 3, value = as_paragraph(as_equation('C_{mod}', props = eqsz))) %>% 
-  flextable::compose(i = 14, j = 3, value = as_paragraph(as_equation('Pg_{vol}, aPAR', props = eqsz))) %>% 
-  flextable::compose(i = 15, j = 3, value = as_paragraph(as_equation('Rt_{vol}, r', props = eqsz))) %>% 
+  flextable::compose(i = 14, j = 3, value = as_paragraph(as_equation('P, aPAR', props = eqsz))) %>% 
+  flextable::compose(i = 15, j = 3, value = as_paragraph(as_equation('R, r', props = eqsz))) %>% 
   flextable::compose(i = 16, j = 3, value = as_paragraph(as_equation('D, \\frac{1}{H}\\left[-bU_{10}^2\\left(\\frac{s_c}{600} \\right)^{-0.5} \\left(C_{Sat} - C_d\\right )\\right]', props = eqsz))) %>% 
   flextable::compose(i = 17, j = 3, value = as_paragraph(as_equation('a', props = eqsz))) %>% 
   flextable::compose(i = 18, j = 3, value = as_paragraph(as_equation('b', props = eqsz))) %>% 
@@ -70,6 +70,7 @@ load(file = url('https://github.com/fawda123/BASEmetab_script/raw/master/data/ap
 
 tosum <- apacmp %>% 
   mutate(
+    typ = ifelse(typ == 'BASEmetab', 'BASE', typ),
     val = case_when(
       var == 'Rt_vol' ~ -1 * val, 
       T ~ val
@@ -84,7 +85,7 @@ tosum <- apacmp %>%
 grd <- crossing(
   dotyp = unique(tosum$dotyp), 
   var = levels(tosum$var), 
-  comp = c('Odum v EBASE', 'BASEmetab v EBASE')
+  comp = c('Odum v EBASE', 'BASE v EBASE')
   ) %>% 
   mutate(
     corv = NA, 
@@ -144,7 +145,7 @@ totab <- grd %>%
[…]

### 2023-01-17 manu-draft.qmd
change delta to partial

@@
 $$
-\frac{\delta{C_d}}{\delta{t}} = [\,aPAR^p]\, - [\,r\Theta^{T - \bar{T}}]\, - \left[-K\cdot1.0241^{T-\bar{T}}\left(C_{sat} - C_d \right )\right]
+\frac{\partial{C_d}}{\partial{t}} = [\,aPAR^p]\, - [\,r\Theta^{T - \bar{T}}]\, - \left[-K\cdot1.0241^{T-\bar{T}}\left(C_{sat} - C_d \right )\right]
 $$ {#eq-base}
 
-where the terms in the square brackets are production (P), respiration (R), and gas exchange (D), respectively, modeled as a function of the change in dissolved oxygen ($\delta{C_d}$) per unit time ($\delta{t}$) of the input data.
+where the terms in the square brackets are production (P), respiration (R), and gas exchange (D), respectively, modeled as a function of the change in dissolved oxygen ($\partial{C_d}$) per unit time ($\partial{t}$) of the input data.
 Five unknown parameters are estimated from the data: $a$, $p$, $r$, $\Theta$, and $K$.
@@
 $$
-\frac{\delta{C_d}}{\delta{t}} = [\,aPAR^p]\, - [\,r\Theta^{T - \bar{T}}]\, - \frac{1}{H}\left[-bU_{10}^2\left(\frac{s_c}{600} \right)^{-0.5} \left(C_{Sat} - C_d \right )\right]
+\frac{\partial{C_d}}{\partial{t}} = [\,aPAR^p]\, - [\,r\Theta^{T - \bar{T}}]\, - \frac{1}{H}\left[-bU_{10}^2\left(\frac{s_c}{600} \right)^{-0.5} \left(C_{Sat} - C_d \right )\right]
 $$ {#eq-basewan}
@@
 $$
-\frac{\delta{C_d}}{\delta{t}} = [\,aPAR]\, - [\,r]\, - \frac{1}{H}\left[-bU_{10}^2\left(\frac{s_c}{600} \right)^{-0.5} \left(C_{Sat} - C_d \right )\right]
+\frac{\partial{C_d}}{\partial{t}} = [\,aPAR]\, - [\,r]\, - \frac{1}{H}\left[-bU_{10}^2\left(\frac{s_c}{600} \right)^{-0.5} \left(C_{Sat} - C_d \right )\right]
 $$ {#eq-ebase}
 
-As for BASE, the metabolic estimates in EBASE are defined by the change in dissolved oxygen over the time step, $\delta{C_d}/\delta{t}$, where gross production is provided by $aPAR$, respiration is provided by $r$, and gas exchange is provided by the remainder [as in @Wanninkhof14].
+As for BASE, the metabolic estimates in EBASE are defined by the change in dissolved oxygen over the time step, $\partial{C_d}/\partial{t}$, where gross production is provided by $aPAR$, respiration is provided by $r$, and gas exchange is provided by the remainder [as in @Wanninkhof14].
 Units for each of the metabolic estimates are $\runi$.

### 2023-01-22 manu-draft.qmd
table 1 and figs 1 and 2 redone with areal units

@@
 Continuous meteorological observations from the nearby East Bay station (29.7909$^{\circ}$ N, -84.8834$^{\circ}$ W, ~10km north of Cat Point) were combined with the water quality data at the same time step and included air temperature (C), wind speed (m/s), and PAR (mmol/m$^2$ as total per 15 minute observation).
-An oxygen time series was generated from Fwoxy using the inputs from Apalachicola Bay ([@fig-simapa]a), such that the simulated dissolved oxygen was created as close as possible to the observed given the inputs.
-Known values for $a$ and $r$ parameters were returned for each time step of the simulated data, as well as all metabolic estimates for production, respiration, and gas exchange ([@fig-simapa]b and c).
+An oxygen time series was generated from Fwoxy using the inputs from Apalachicola Bay ([@fig-synapa]a), such that the simulated dissolved oxygen was created as close as possible to the observed given the inputs.
+Known values for $a$ and $r$ parameters were returned for each time step of the simulated data, as well as all metabolic estimates for production, respiration, and gas exchange ([@fig-synapa]b and c).
 The only parameter that could not be simulated by Fwoxy was $b$, which was set to 0.251 $\buni$ [@Wanninkhof14].
@@
 
-![Simulated time series using the Fwoxy model for one year of continuous data at Apalachicola Bay.
-Simulated input to EBASE was the (a) dissolved oxygen time series, with results compared to the (b) simulated $a$ parameter and (c) simulated metabolic estimates for production (P), respiration (R), and gas exchange (D).](figs/simapa.png){#fig-simapa}
+![Synthetic time series using the Fwoxy model for one year of continuous data at Apalachicola Bay.
+Synthetic input to EBASE was the (a) dissolved oxygen time series, with results compared to the (b) simulated $a$ parameter and (c) simulated metabolic estimates for production (P), respiration (R), and gas exchange (D).
+The $a$ parameter is shown as an areal value for comparison with the metabolic results.](figs/synapa.png){#fig-synapa}
 

### 2023-01-22 R/figs.R

@@ -25,13 +25,13 @@ fwdatcmp <- fwdat %>%
     DateTimeStamp = dmy_hms(datet, tz = 'America/Jamaica'),
     Date = as.Date(DateTimeStamp, tz = 'America/Jamaica'),
     DO_obs = `oxy,mmol/m3`, 
-    a = `aparam,(mmolO2/m2/d)/(W/m2)` / `ht,m`,
-    Rt_vol = `er,mmol/m2/d` / `ht,m`,
-    Pg_vol = `gpp,mmol/m2/d` / `ht,m`,
-    D = -1 * `gasex,mmol/m2/d` / `ht,m`,
+    a = `aparam,(mmolO2/m2/d)/(W/m2)`,
+    R = `er,mmol/m2/d`,
+    P = `gpp,mmol/m2/d`,
+    D = -1 * `gasex,mmol/m2/d`,
     b = 100 * 3600 * `kw,m/s` / `wspd2,m2/s2` / (`sc,dimensionless` / 660) ^ -0.5 # (m/s)/(m2/s2) to (cm/hr) / (m2/s2)
   ) %>% 
-  select(Date, DateTimeStamp, DO_obs, a, b, Pg_vol, Rt_vol, D)
+  select(Date, DateTimeStamp, DO_obs, a, b, P, R, D)
 
 p1 <- ggplot(fwdatcmp, aes(x = DateTimeStamp, y = DO_obs)) + 
   geom_line(linewidth = 0.2) + 
@@ -40,7 +40,7 @@ p1 <- ggplot(fwdatcmp, aes(x = DateTimeStamp, y = DO_obs)) +
   labs(
     x = NULL, 
     y = expression(paste(O [2], ' (mmol ', m^-3, ')')),
-    title = '(a) Simulated dissolved oxygen'
+    title = '(a) Synthetic dissolved oxygen'
   )
 
 p2 <- ggplot(fwdatcmp, aes(x = DateTimeStamp, y = a)) + 
@@ -49,21 +49,20 @@ p2 <- ggplot(fwdatcmp, aes(x = DateTimeStamp, y = a)) +
   theme_minimal() + 
   labs(
     x = NULL, 
-    y = expression(paste(italic(a), ' (mmol ', m^-3, d^-1, ') / (W ', m^{-2}, ')')), 
-    title = expression(paste('(b) Simulated ', italic(a), ' parameter'))
+    y = expression(paste(italic(a), ' (mmol ', m^-2, d^-1, ') / (W ', m^{-2}, ')')), 
[…]

### 2023-01-22 R/tabs.R

@@ -16,7 +16,7 @@ totab <- tibble(
                   'Dissolved oxygen (modelled)', 'Production', 'Respiration', 'Gas exchange', 'Light efficiency', 'b'),
   `Model notation` =  c('C_d', '-', '-', 'PAR', '-', 'H', 
                         'U102', 'Schmidt number', 'C_sat', 
-                        'C_mod', 'Pg_vol', 'Rt_vol',  'D', 'a', 'b'),
+                        'C_mod', 'P', 'R',  'D', 'a', 'b'),
   Units = c('mg/L', 'C', 'psu', 'W/m2/s', 'm/s', 'm', 
             'm2', 'unitless', 'mmol/m3',
             'mmol/m3', 'mmol O2 m3/d', 'mmol O2 m3/d', 'mmol O2 m3/d', '(mmol/m3/d)/(W/m2)', '(cm/hr)/(m2/s2)')
@@ -35,25 +35,25 @@ ebaseiotab <- flextable(totab) %>%
   flextable::compose(i = 10, j = 3, value = as_paragraph(as_equation('S_c', props = eqsz))) %>% 
   flextable::compose(i = 11, j = 3, value = as_paragraph(as_equation('C_{sat}', props = eqsz))) %>% 
   flextable::compose(i = 13, j = 3, value = as_paragraph(as_equation('C_{mod}', props = eqsz))) %>% 
-  flextable::compose(i = 14, j = 3, value = as_paragraph(as_equation('P, aPAR', props = eqsz))) %>% 
-  flextable::compose(i = 15, j = 3, value = as_paragraph(as_equation('R, r', props = eqsz))) %>% 
-  flextable::compose(i = 16, j = 3, value = as_paragraph(as_equation('D, \\frac{1}{H}\\left[-bU_{10}^2\\left(\\frac{s_c}{600} \\right)^{-0.5} \\left(C_{Sat} - C_d\\right )\\right]', props = eqsz))) %>% 
+  flextable::compose(i = 14, j = 3, value = as_paragraph(as_equation('P,\\,aPAR', props = eqsz))) %>% 
+  flextable::compose(i = 15, j = 3, value = as_paragraph(as_equation('R,\\,r', props = eqsz))) %>% 
[…]

### 2023-01-22 manu-draft.qmd
remove mention of Fwoxy in text except for package citation, corrected text about one a value per time step

@@
 As for the BASE model, relatively uninformed prior distributions following a normal Gaussian distribution are used by default, although the priors can be changed based on previous knowledge of parameters specific to an ecosystem or as informed by other metabolic modeling approaches.
-The default priors were based on those that returned reasonable parameter estimates using a forward-feeding oxygen mass balance equation called Fwoxy [ForWard OXYgen, @Arriola22].
-For Fwoxy, a dissolved oxygen time series is simulated by providing known parameter values in an equation similar to @eq-ebase.
-All of the physical parameters are held constant, e.g., wind speed, temperature, and salinity.
-The other required inputs are the $a$, $r$, and $b$ parameters, which are set by the user.
-An initial oxygen concentration is provided and the simulated time series is created by forward addition of the inputs.
-As such, the dissolved oxygen output simulated by Fwoxy as change per unit time were modelled by EBASE using the constant physical parameters as input.
-Reasonable prior distributions by EBASE were identified (Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]) based on the ability to return the known parameters from Fwoxy:
+The default priors were based on those that returned reasonable parameter estimates using a known time series created with forward-feeding oxygen mass balance inputs to @eq-ebase.
+[Fwoxy R package, @Arriola22].
+All of the physical parameters were held constant, e.g., wind speed, temperature, and salinity.
+The other required inputs were the $a$, $r$, and $b$ parameters, chosen as 0.2 $\auni$, 20 $\runi$, and 0.251 $\buni$, respectively.
+An initial oxygen concentration was provided as 250 mmol/m$^3$ and the time series was created by forward addition of the inputs.
+As such, the dissolved oxygen output as change per unit time was modelled by EBASE using the constant the same physical parameters as input.
+Reasonable prior distributions by EBASE were identified (Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]) based on the ability to return the known parameters:
 
@@
 
-A preliminary assessment of the ability of EBASE to produce reasonable parameter estimates for $a$, $r$, and $b$ (@eq-ebase) was conducted using a simulated time series created with Fwoxy.
+A preliminary assessment of the ability of EBASE to produce reasonable parameter estimates for $a$, $r$, and $b$ (@eq-ebase) was conducted using a synthetic time series with actual physical data as inputs.
 These assessments evaluated if 1) EBASE estimates similar values for the known parameters, and 2) how the estimated values change with different model configurations.
-The simulated time series used inputs from an approximate one year of continuous data from Apalachicola Bay, Florida.
+The synthetic time series used inputs from an approximate one year of continuous data from Apalachicola Bay, Florida.
 The input data were 2012 observations from February 22nd to December 19th at the Cat Point water quality monitoring station (29.7021$^{\circ}$ N, -84.8802$^{\circ}$ W) and included 15 minute observations of dissolved oxygen (mg/L), water temperature (C), and salinity (psu).
 Continuous meteorological observations from the nearby East Bay station (29.7909$^{\circ}$ N, -84.8834$^{\circ}$ W, ~10km north of Cat Point) were combined with the water quality data at the same time step and included air temperature (C), wind speed (m/s), and PAR (mmol/m$^2$ as total per 15 minute observation).
-An oxygen time series was generated from Fwoxy using the inputs from Apalachicola Bay ([@fig-synapa]a), such that the simulated dissolved oxygen was created as close as possible to the observed given the inputs.
-Known values for $a$ and $r$ parameters were returned for each time step of the simulated data, as well as all metabolic estimates for production, respiration, and gas exchange ([@fig-synapa]b and c).
-The only parameter that could not be simulated by Fwoxy was $b$, which was set to 0.251 $\buni$ [@Wanninkhof14].
+An oxygen time series was generated using the inputs from Apalachicola Bay ([@fig-synapa]a), such that the synthetic dissolved oxygen was created as close as possible to the observed given the inputs.
+Known values for $a$ and $r$ parameters were returned for each time step of the synthetic data, as well as all metabolic estimates for production, respiration, and gas exchange ([@fig-synapa]b and c).
+The only parameter that could not be created as a continuous time series was $b$, which was set to 0.251 $\buni$ [@Wanninkhof14].
 
 Several configurations of the EBASE model can influence how the parameters and metabolic rates are estimated.
-Two types of sensitivity analyses were conducted to evaluate changes in the results returned by EBASE, where in all cases, the results were compared to the known values in the simulated time series.
+Two types of sensitivity analyses were conducted to evaluate changes in the results returned by EBASE, where in all cases, the results were compared to the known values in the synthetic time series.
 First, results were compared to the known values after changing characteristics of their prior distributions.
@@
 For each parameter, low, medium, and high values for the standard deviation were assessed as 10%, 100%, and 1000% of the default values.
-Each unique combination of low, medium, and high values for the standard deviation of each parameter were evaluated, creating 3$^3$ = 27 different combinations of prior distributions that were used with the simulated time series in EBASE.
+Each unique combination of low, medium, and high values for the standard deviation of each parameter were evaluated, creating 3$^3$ = 27 different combinations of prior distributions that were used with the synthetic time series in EBASE.
 
@@
 Unlike BASE, the EBASE software allows flexibility in the period of time used for optimization, with a minimum period of time being one day and a maximum period being the length of the time series.
-For a chosen optimization period, a single estimate for $r$ and $b$ is returned by the model, whereas $n$ number of observations equal to the number of time steps in each optimization period are returned for the $a$ parameter (e.g., 96 values for a 24 hour time period if the sampling interval is 15 minutes).
+For a chosen optimization period, single estimates for $a$, $r$ and $b$ are returned by the model.
 A short period of optimization (e.g., one day) may return unstable parameter estimates if forcing factors (e.g., wind, temperature) that affect estuarine biogeochemical rates occur at time scales longer the chosen period of optimization for EBASE.
 As such, potentially longer optimization periods may return more stable and robust estimates given that more data are used for parameter estimation.
-An optimization period of one day was compared to results using an optimization period of seven days for the simulated one-year time series.
+An optimization period of one day was compared to results using an optimization period of seven days for the synthetic one-year time series.
 Each optimization period was evaluated for every unique combination of prior distributions, totaling 54 (27 times the two optimization periods) unique analyses for the assessment of changing standard deviations of all parameters.
 
-Changing the variance of each of the prior distributions affected the ability of EBASE to return the known parameters from the simulated time series.
-@fig-apasumdat shows a summary of the comparisons of the simulated data with EBASE for each unique combination of prior distributions and two optimization periods of one and seven days.
-The results are compared using R-squared ($R^2$) values from the linear fit between the simulated data and results from EBASE.
+Changing the variance of each of the prior distributions affected the ability of EBASE to return the known parameters from the synthetic time series.
+@fig-apasumdat shows a summary of the comparisons of the synthetic data with EBASE for each unique combination of prior distributions and two optimization periods of one and seven days.
+The results are compared using R-squared ($R^2$) values from the linear fit between the synthetic data and results from EBASE.
 Darker colors represent lower $R^2$ values and can be used to assess how changing the variance of the priors affects the ability of EBASE to estimate the true parameter.
@@
 
+* Describe how EBASE deals with gaps
+
 ## Comments and Recommendations
@@
 
-![Synthetic time series using the Fwoxy model for one year of continuous data at Apalachicola Bay.
-Synthetic input to EBASE was the (a) dissolved oxygen time series, with results compared to the (b) simulated $a$ parameter and (c) simulated metabolic estimates for production (P), respiration (R), and gas exchange (D).
+![Synthetic time series for one year of continuous data at Apalachicola Bay.
+Synthetic input to EBASE was the (a) dissolved oxygen time series, with results compared to the (b) synthetic $a$ parameter and (c) synthetic metabolic estimates for production (P), respiration (R), and gas exchange (D).
 The $a$ parameter is shown as an areal value for comparison with the metabolic results.](figs/synapa.png){#fig-synapa}
@@
 
-![Comparison of EBASE results to a simulated time series created with known metabolic parameters.
+![Comparison of EBASE results to a synthetic time series created with known metabolic parameters.
 Results are compared for unique combination of prior distributions, where the standard deviations are evaluated from low (L), medium (M), to high (H) for a one day and seven day model optimization period.
-The metabolic results for dissolved oxygen (DO), production (P), respiration (R), gas exchange (D), and the $a$ parameter are evaluated based on the $R^2$ value of the linear fit between the EBASE results and simulated values.
+The metabolic results for dissolved oxygen (DO), production (P), respiration (R), gas exchange (D), and the $a$ parameter are evaluated based on the $R^2$ value of the linear fit between the EBASE results and synthetic values.
 The $b$ parameter is a fixed constant and was not evaluated.](figs/apasumdat.png){#fig-apasumdat}
@@
 
-![Comparison of EBASE results for the best performing model to a simulated time series created with known metabolic parameters.
+![Comparison of EBASE results for the best performing model to a synthetic time series created with known metabolic parameters.
 The best performing model for the one day (a, ndays = 1) and seven day (b, ndays = 7) optimization period is based on the EBASE results with prior distributions that produced the highest mean $R^2$ values across all parameter estimates (@fig-apasumdat).

### 2023-01-23 R/funcs.R
redo apacmpfig without dry/wet grouping

@@ -19,7 +19,7 @@ simpleCap <- function(x) {
 }
 
 # create subplots for apalachicola obs v dtd metab comparisons
-apacmp_plo <- function(dat, xlb = 'EBASE', ylb = 'Odum', dotyp = 'observed', addtitle = T, alph = 0.8, cols = c('darkgrey', 'black')){
+apacmp_plo <- function(dat, xlb = 'EBASE', ylb = 'Odum', dotyp = 'observed', addtitle = T, alph = 0.5, col = 'black'){
 
   thm <- theme_minimal() + 
     theme(
@@ -60,9 +60,9 @@ apacmp_plo <- function(dat, xlb = 'EBASE', ylb = 'Odum', dotyp = 'observed', add
     lims <- range(toplotmp[, c('yval', 'EBASE')], na.rm = T)
     
     ptmp <- ggplot(toplotmp, aes(x = EBASE, y = yval)) + 
-      geom_point(aes(colour = seas), alpha = alph) + 
+      geom_point(color = col, alpha = alph, stroke = 0, size = 2) + 
       geom_abline(intercept = 0, slope = 1) +
-      geom_smooth(method = 'lm', se = F, colour = 'tomato1', size = 0.7) +
+      geom_smooth(formula = y ~ x ,method = 'lm', se = F, colour = 'tomato1', size = 0.7) +
       facet_wrap(~var, ncol = 1) +
       scale_colour_manual(values = cols) + 
       scale_x_continuous(limits = lims) + 

### 2023-01-23 manu-draft.qmd
redo apacmpfig without dry/wet grouping
rework methods to put ebase equation first and shortened

@@
 The Estuarine BAyesian Single station Estimation (EBASE) method provides a simple approach to quantify metabolic parameters from continuous dissolved oxygen time series.
-The EBASE method is a modified version of the BASE mass balance equation described in @Grace15.
-The original equation is:
+The EBASE mass balance equation is:
+
+$$
+\frac{\partial{C_d}}{\partial{t}} = [\,aPAR]\, - [\,r]\, - \frac{1}{H}\left[-bU_{10}^2\left(\frac{s_c}{600} \right)^{-0.5} \left(C_{Sat} - C_d \right )\right]
+$$ {#eq-ebase}
+
+where the terms in the square brackets are production (P), respiration (R), and gas exchange (D), respectively, modeled as a function of the change in dissolved oxygen ($\partial{C_d}$) per unit time ($\partial{t}$) of the input data.
+Production is modeled linearly as a function of surface irradiance ($PAR$, Watts m$^{-2}$) and $a$ (light efficiency, $\auni$).
+Respiration is modeled with the single parameter $r$ ($\runi$).
+Gas exchange is modeled using the formulation in @Wanninkhof14 as a wind-based parameterization for the gas transfer velocity [@Ho16] multiplied by the oxygen concentration gradient as the difference between the concentration measured in the surface water layer, $C_d$ (mmol m$^{-3}$), and the saturation (or equilibrium) concentration, $C_{Sat}$ (mmol m$^{-3}$), calculated from salinity (psu) and temperature ($^\circ$C) [@Millero81; @Garcia92].
+The gas transfer velocity is modeled using wind speed squared at 10 meters above the water ($U_{10}^2$, m$^2$ s$^{-2}$), water column depth ($H$ in meters, as a fixed parameter or time series from a pressure sensor), and the Schmidt number ($S_c$) defined as the ratio of the kinematic viscosity of water to the molecular diffusivity of the gas for oxygen [from water temperature and salinity using the polynomial fit in @Wanninkhof14].
+The $b$ parameter is estimated from the data and has units of $\buni$.
+As such, the three parameters estimated by EBASE are $a$, $r$, and $b$ using the required input data that includes time series of dissolved oxygen, water temperature, salinity, PAR, wind speed, and water column depth (`r run_reference(id = "ebaseiotab")`).
+
+The EBASE approach was based on several modifications to BAyesian Single-station Estimation (BASE) the mass balance equation described in @Grace15:
 
@@
 
-where the terms in the square brackets are production (P), respiration (R), and gas exchange (D), respectively, modeled as a function of the change in dissolved oxygen ($\partial{C_d}$) per unit time ($\partial{t}$) of the input data.
-Five unknown parameters are estimated from the data: $a$, $p$, $r$, $\Theta$, and $K$.
-The required inputs are the time series of surface irradiance ($PAR$, $\mu$mol/m$^2$/s), water temperature ($T$, C), barometric pressure (atm), diel oxygen ($C_d$, mmol/m$^2$), and oxygen saturation ($C_{sat}$, mmol/m$^3$).
-Dissolved oxygen at saturation $C_{sat}$ is calculated from the input data as a function of water temperature, salinity, and barometric pressure.
-Several changes, outlined and justified below, were made to @eq-base for EBASE.
+where all parameters are defined as in @eq-ebase, with the additional parameters $p$, $\Theta$, and $K$ (the rearation coefficient).
+The BASE approach provides options to estimate $p$ and $\Theta$ from the data or to set as fixed values.
+A temperature dependency, $T$, is also included where respiration and gas exchange vary with temperature.
+The changes to @eq-base for EBASE were necessary for estuarine application and included 1) a linear relationship of production to PAR ($aPAR$) by setting $p = 1$; 2) constant respiration throughout the time series by removing the temperature dependency; and 3) use of a gas exchange parameterization commonly used in coastal or oceanic settings [i.e., @Grace15 used a littoral parameterization].
+The justification for these changes is as follows.
 
-### Modification of the gas exchange parameterization
+### Changes to BASE
 
 The underlying idea behind most gas exchange parameterizations for metabolism estimates is that the transfer of gases of relatively low solubility, like oxygen, is proportional to the concentration gradient in a thin film of water at the air–water interface.
-It is further assumed that the water at the top of the film is saturated with respect to the atmosphere, so that the concentration gradient across the thin film can be approximated as the difference between oxygen concentration measured in the surface water layer, $C_d$, and the saturation (or equilibrium) concentration $C_{Sat}$, which is usually calculated as a function of water temperature and salinity (the negative sign is needed because positive flux is set to be from water to air):
+It is also assumed that the water at the top of the film is saturated with respect to the atmosphere, so that the concentration gradient across the thin film can be approximated as the difference between oxygen concentration measured in the surface water layer, $C_d$, and the saturation (or equilibrium) concentration $C_{Sat}$, which is usually calculated as a function of water temperature and salinity (the negative sign is needed because positive flux is set to be from water to air):
 
@@
 
-The proportionality parameter, $k_w$, usually called the gas transfer velocity because it has the dimensions of length over time, increases as the water column in the vicinity of the air–water interface becomes more turbulent.
-As such, $k_w$  is usually modeled as a function of measurable variables that can potentially predict turbulent mixing of the surface water, such as wind speed, current velocity, and water depth [@Ho16].
-$k_w$ also increases with temperature because as the random motion of gas and water molecules in the thin film speeds up, the film thins out and the transfer of molecules across the film becomes easier.
-This dependence of $k_w$ on temperature is usually modeled with the Schmidt number, which combines information on the viscosity of the fluid and the diffusivity of the gas.
-The Schmidt number, in turn, is often parameterized as a gas-specific polynomial function of water temperature and salinity.
+The gas transfer velocity, $k_w$, has dimensions of length over time and increases as the water column in the vicinity of the air–water interface becomes more turbulent.
+As such, $k_w$  is usually modeled as a function of measurable variables that can potentially predict turbulent mixing of the surface water, e.g., wind speed, current velocity, and water depth [@Ho16].
+$k_w$ also increases with temperature because the film of water at the air-sea interface becomes more thin as the random motion of gas and water molecules increases at higher temperatures and transfer of molecules across the film increases.
+This dependence of $k_w$ on temperature is usually modeled with the Schmidt number as in @eq-ebase, which combines information on the viscosity of the fluid and the diffusivity of the gas.
 
-Instead of one of the traditional wind-based parameterization used in oceanography [e.g., @Wanninkhof14], @Grace15 used the following for the gas transfer in BASE:
+@Grace15 used the following for the gas transfer in BASE:
 
@@
 
-where $\bar{T}$ is the average water temperature over each 24-hour period in the diel oxygen time series and $K$ is the optimization parameter, often called the reaeration coefficient, which has dimensions of inverse time.
+where $\bar{T}$ is the average water temperature over each 24-hour period in the diel oxygen time series and $K$ is the reaeration coefficient estimated from the data.
 @eq-gasexbase indicates that $k_w$ increases with temperature and depth.
 The temperature dependence is consistent with the known decrease of $k_w$ with the Schmidt number, but the depth dependence is difficult to justify.
-One expects that, for a given current speed, $k_w$ would increase as the water depth decreases because turbulence generated at the sediment–water interface by currents would more easily reach the surface [@Ho16].
+An expectation is that $k_w$ would increase as water depth decreases for a given current speed because turbulence generated at the sediment–water interface by currents would more easily reach the surface [@Ho16].
+Based on these differences, the wind-based parameterization for gas exchange was used for EBASE  [@Wanninkhof14].
+Additionally, $C_{sat}$ was calculated using only temperature and salinity, removing the need for barometric pressure used in BASE.
 
-Based on the above differences, the first modification to @eq-base for EBASE is use of a wind-based parameterization for gas exchange that is more appropriate for estuaries [@Wanninkhof14]:
+The additional modifications to @eq-base for EBASE were removing the $p$ and $\Theta$ parameters from the production (P) and respiration (R) components, respectively.
+The BASE method provides options to estimate both $\Theta$ and $p$ from the data or to set as fixed values defined in the model file ($\Theta = 1.07177$, $p=1$).
+For EBASE, the temperature dependency of respiration is removed by setting $\Theta^{(T-\bar{T})}$ = 1 and it is assumed that respiration is constant as in other methods [@Odum56;@Murrel18].
+The $p$ parameter, which is dimensionless, is also set to 1 as in @Holtgrieve10, such that a linear relationship of production to $PAR$ was used.
+This removes the exponential dependence of production with increasing $PAR$.
+Removing these components also improved model stability by creating more reasonable estimates for the other parameters.
 
-$$
-\frac{\partial{C_d}}{\partial{t}} = [\,aPAR^p]\, - [\,r\Theta^{T - \bar{T}}]\, - \frac{1}{H}\left[-bU_{10}^2\left(\frac{s_c}{600} \right)^{-0.5} \left(C_{Sat} - C_d \right )\right]
-$$ {#eq-basewan}
-
-where $U_{10}^2$ is the wind speed at 10-m height, $H$ is water column depth (m, as a fixed parameter or time series from a pressure sensor) and $S_c$ is the Schmidt number defined as the ratio of the kinematic viscosity of water to the molecular diffusivity of the gas for oxygen [from water temperature and salinity using the polynomial fit of @Wanninkhof14].
-The parameter $b$ ($\buni$) must be estimated from the data.
-
-The remaining variables in @eq-basewan are defined as in @eq-base, with the exception that $C_{sat}$ is calculated using only temperature (C) and salinity (psu), removing the need for barometric pressure.
-Methods from @Garcia92 are used as follows:
-
-$$
-C_{sat} = \exp(C_O) \cdot \rho_{T, sal} \cdot 1e-3
-$$ {#eq-csat}
-
-where $C_O$ is thr solubility of dissolved oxygen per volume of seawater at the given temperature and $\rho$ (kg/m$^3$) is seawater density at zero elevation given temperature and salinity [@Millero81].
-$C_O$ is calculated using a polynomial equation:
-
-$$
-C_O = A_0 + A_1 \cdot T_s + A_2 \cdot T_s^2 + A_3 \cdot T_s^3 + A_4 \cdot T_s^4 + A_5 \cdot T_s^5 + sal \cdot (B_0 + B_1 \cdot T_s + B_2 \cdot T_s^2 + B_3 \cdot T_s^3) + C_0 \cdot sal^2
-$$ {#eq-co}
-
-where the coefficients are $A_0$ = 5.80818, $A_1$ = 3.20684, $A_2$ = 4.11890, $A_3$ = 4.93845, $A_4$ = 1.01567, $A_5$ = 1.41575, $B_0$ = -7.01211e-3, $B_1$ = -7.25958e-3, $B_2$ = -7.93334e-3, $B_3$ = -5.54491e-3, and $C_0$ = -1.32412e-7.
-$T_s$ is temperature scaled
-as input to @eq-co defined as:
-
-$$
-T_s = \log((298.15 - T)/(273.15 + T)
-$$ {#eq-ts}
-
-### Modification of the production and respiration parameterization
-
-The second modification to @eq-base for EBASE is a simplification to remove the $p$ and $\Theta$ parameters from the production (P) and respiration (R) components, respectively, of the mass balance equation.
-The BASE method provides options to estimate both $\Theta$ and $p$ from the data or to set as fixed values defined in the model file ($\Theta = 1.07177$, $p=1$).
-For EBASE, the temperature dependency of respiration is removed by setting $\Theta^{(T-\bar{T})}$ = 1.
-The $p$ parameter, which is dimensionless, is also set to 1 as in @Holtgrieve10.
-Removing these components improved model stability by creating more reasonable estimates for the other parameters.
-This further simplifies @eq-basewan to the final form for EBASE:
-
-$$
-\frac{\partial{C_d}}{\partial{t}} = [\,aPAR]\, - [\,r]\, - \frac{1}{H}\left[-bU_{10}^2\left(\frac{s_c}{600} \right)^{-0.5} \left(C_{Sat} - C_d \right )\right]
-$$ {#eq-ebase}
-
-As for BASE, the metabolic estimates in EBASE are defined by the change in dissolved oxygen over the time step, $\partial{C_d}/\partial{t}$, where gross production is provided by $aPAR$, respiration is provided by $r$, and gas exchange is provided by the remainder [as in @Wanninkhof14].
-Units for each of the metabolic estimates are $\runi$.
+### EBASE model estimation
 
@@
 The JAGS software [as for BASE, @Plummer03] is used with the EBASE R package [described below, @Beck22;@RCT22] to estimate the unknown parameters using Markov Chain Monte Carlo simulations in a Bayesian framework.
-At each time step, the change in oxygen concentration between time steps is calculated from the equation using model inputs and parameter guesses, and then a finite difference approximation is used to estimate modeled oxygen concentration.
+At each time step, the change in oxygen concentration between time steps is calculated from the equation using model inputs and parameter guesses based on the prior distributions, and then a finite difference approximation is used to estimate modeled oxygen concentration.
 Estimations proceed using a forward in time integration where the dissolved oxygen at each time step is informed by parameter estimates from the preceding time steps.
@@
 All parameter inputs and outputs for EBASE are shown in `r run_reference(id = "ebaseiotab")`.
-
-### Bayesian prior
+Areal rates for $P$, $R$, and $D$ (mmol m$^-2$ d$^{-1}$) are returned for comparability of estimates across locations where water column depth may vary.
 
@@
 
-### EBASE R package
-
 For ease of application, an R package was developed to implement the EBASE methods described above [@Beck22].
@@
 Users can also specify alternative prior distributions from those in Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior] for each parameter.
-Functions are available for viewing model results, including a plot of the modelled dissolved oxygen with the observed, a time series plot of the primary metabolic estimates (P, R, and D), and a time series plot of the credible intervals for the $a$, $r$, and $b$ parameters.
+Functions are available for viewing model results, including a plot of the modeled dissolved oxygen with the observed, a time series plot of the primary metabolic estimates (P, R, and D, as areal rates with units $\runi$), and a time series plot of the credible intervals for the $a$, $r$, and $b$ parameters.
 An additional option also allows the user to specify the model optimization period, where the time period can vary from a minimum of one day to the maximum of the entire length of the time series.
@@
 Results for net ecosystem metabolism (NEM), gross production (P), respiration (R), and gas exchange (D) are compared using the (a) observed and (b) detided dissolved oxygen time series as input to each model.
-All values are volumetric as mmol O$_2$ m$^3$ d$^{-1}$ and are grouped by the dry or wet season.
+All values are volumetric as mmol O$_2$ m$^3$ d$^{-1}$.
 Black lines show the 1:1 comparison and the red lines show the linear regression fit.](figs/apacmpfig.png){#fig-apacmpfig fig-align="center"}

### 2023-01-23 manu-draft.qmd
minor text edits

@@
 
-The underlying idea behind most gas exchange parameterizations for metabolism estimates is that the transfer of gases of relatively low solubility, like oxygen, is proportional to the concentration gradient in a thin film of water at the air–water interface.
+A common theory behind most gas exchange parameterizations for metabolism estimates is that the transfer of gases of relatively low solubility, like oxygen, is proportional to the concentration gradient in a thin film of water at the air–water interface.
 It is also assumed that the water at the top of the film is saturated with respect to the atmosphere, so that the concentration gradient across the thin film can be approximated as the difference between oxygen concentration measured in the surface water layer, $C_d$, and the saturation (or equilibrium) concentration $C_{Sat}$, which is usually calculated as a function of water temperature and salinity (the negative sign is needed because positive flux is set to be from water to air):
@@
 [Fwoxy R package, @Arriola22].
-All of the physical parameters were held constant, e.g., wind speed, temperature, and salinity.
+All of the physical parameters were held constant, i.e., wind speed, temperature, and salinity.
 The other required inputs were the $a$, $r$, and $b$ parameters, chosen as 0.2 $\auni$, 20 $\runi$, and 0.251 $\buni$, respectively.

### 2023-02-26 manu-draft.qmd
update apasumdat figure and text for mean and sd prior comps

@@
 
+```{=tex}
 \newcommand{\auni}{\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)}
@@
 \newcommand{\buni}{\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)}
-
+```
 ## Abstract {.unnumbered}
@@
 
-* Overview of metabolism - importance
-* Existing methods: Odum/WtRegDO, BASE, references in @Grace15, applications to lakes and streams
-* Applications to estuaries - challenges
-* Goals and objectives
+-   Overview of metabolism - importance
+-   Existing methods: Odum/WtRegDO, BASE, references in @Grace15, applications to lakes and streams
+-   Applications to estuaries - challenges
+-   Goals and objectives
 
@@
 where the terms in the square brackets are production (P), respiration (R), and gas exchange (D), respectively, modeled as a function of the change in dissolved oxygen ($\partial{C_d}$) per unit time ($\partial{t}$) of the input data.
-Production is modeled linearly as a function of surface irradiance ($PAR$, Watts m$^{-2}$) and $a$ (light efficiency, $\auni$).
-Respiration is modeled with the single parameter $r$ ($\runi$).
+Production is modeled linearly as a function of surface irradiance ($PAR$, Watts m$^{-2}$) and $a$ (light efficiency, $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$).
+Respiration is modeled with the single parameter $r$ ($\text{mmol}/\text{m}^3/\text{d}$).
 Gas exchange is modeled using the formulation in @Wanninkhof14 as a wind-based parameterization for the gas transfer velocity [@Ho16] multiplied by the oxygen concentration gradient as the difference between the concentration measured in the surface water layer, $C_d$ (mmol m$^{-3}$), and the saturation (or equilibrium) concentration, $C_{Sat}$ (mmol m$^{-3}$), calculated from salinity (psu) and temperature ($^\circ$C) [@Millero81; @Garcia92].
 The gas transfer velocity is modeled using wind speed squared at 10 meters above the water ($U_{10}^2$, m$^2$ s$^{-2}$), water column depth ($H$ in meters, as a fixed parameter or time series from a pressure sensor), and the Schmidt number ($S_c$) defined as the ratio of the kinematic viscosity of water to the molecular diffusivity of the gas for oxygen [from water temperature and salinity using the polynomial fit in @Wanninkhof14].
-The $b$ parameter is estimated from the data and has units of $\buni$.
+The $b$ parameter is estimated from the data and has units of $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$.
 As such, the three parameters estimated by EBASE are $a$, $r$, and $b$ using the required input data that includes time series of dissolved oxygen, water temperature, salinity, PAR, wind speed, and water column depth (`r run_reference(id = "ebaseiotab")`).
@@
 
-A common theory behind most gas exchange parameterizations for metabolism estimates is that the transfer of gases of relatively low solubility, like oxygen, is proportional to the concentration gradient in a thin film of water at the air–water interface.
+A common theory behind most gas exchange parameterizations for metabolism estimates is that the transfer of gases of relatively low solubility, like oxygen, is proportional to the concentration gradient in a thin film of water at the air--water interface.
 It is also assumed that the water at the top of the film is saturated with respect to the atmosphere, so that the concentration gradient across the thin film can be approximated as the difference between oxygen concentration measured in the surface water layer, $C_d$, and the saturation (or equilibrium) concentration $C_{Sat}$, which is usually calculated as a function of water temperature and salinity (the negative sign is needed because positive flux is set to be from water to air):
@@
 
-The gas transfer velocity, $k_w$, has dimensions of length over time and increases as the water column in the vicinity of the air–water interface becomes more turbulent.
+The gas transfer velocity, $k_w$, has dimensions of length over time and increases as the water column in the vicinity of the air--water interface becomes more turbulent.
 As such, $k_w$ is usually modeled as a function of measurable variables that can potentially predict turbulent mixing of the surface water, e.g., wind speed, current velocity, and water depth [@Ho16].
@@
 The temperature dependence is consistent with the known decrease of $k_w$ with the Schmidt number, but the depth dependence is difficult to justify.
-An expectation is that $k_w$ would increase as water depth decreases for a given current speed because turbulence generated at the sediment–water interface by currents would more easily reach the surface [@Ho16].
+An expectation is that $k_w$ would increase as water depth decreases for a given current speed because turbulence generated at the sediment--water interface by currents would more easily reach the surface [@Ho16].
 Based on these differences, the wind-based parameterization for gas exchange was used for EBASE [@Wanninkhof14].
@@
 The BASE method provides options to estimate both $\Theta$ and $p$ from the data or to set as fixed values defined in the model file ($\Theta = 1.07177$, $p=1$).
-For EBASE, the temperature dependency of respiration is removed by setting $\Theta^{(T-\bar{T})}$ = 1 and it is assumed that respiration is constant as in other methods [@Odum56;@Murrel18].
+For EBASE, the temperature dependency of respiration is removed by setting $\Theta^{(T-\bar{T})}$ = 1 and it is assumed that respiration is constant as in other methods [@Odum56; @Murrel18].
 The $p$ parameter, which is dimensionless, is also set to 1 as in @Holtgrieve10, such that a linear relationship of production to $PAR$ was used.
@@
 
-### EBASE model estimation
+### EBASE model estimation {#sec-ebasemod}
 
-The remaining three parameters $a$ (light efficiency, $\auni$), $r$ (respiration, $\runi$), and $b$ ($\buni$) are estimated by likelihood given the observed data and prior distributions for the parameters.
-The JAGS software [as for BASE, @Plummer03] is used with the EBASE R package [described below, @Beck22;@RCT22] to estimate the unknown parameters using Markov Chain Monte Carlo simulations in a Bayesian framework.
+The remaining three parameters $a$ (light efficiency, $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$), $r$ (respiration, $\text{mmol}/\text{m}^3/\text{d}$), and $b$ ($\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$) are estimated by likelihood given the observed data and prior distributions for the parameters.
+The JAGS software [as for BASE, @Plummer03] is used with the EBASE R package [described below, @Beck22; @RCT22] to estimate the unknown parameters using Markov Chain Monte Carlo simulations in a Bayesian framework.
 At each time step, the change in oxygen concentration between time steps is calculated from the equation using model inputs and parameter guesses based on the prior distributions, and then a finite difference approximation is used to estimate modeled oxygen concentration.
@@
 All of the physical parameters were held constant, i.e., wind speed, temperature, and salinity.
-The other required inputs were the $a$, $r$, and $b$ parameters, chosen as 0.2 $\auni$, 20 $\runi$, and 0.251 $\buni$, respectively.
+The other required inputs were the $a$, $r$, and $b$ parameters, chosen as 0.2 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$, 20 $\text{mmol}/\text{m}^3/\text{d}$, and 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$, respectively.
 An initial oxygen concentration was provided as 250 mmol/m$^3$ and the time series was created by forward addition of the inputs.
@@
 All prior distributions were constrained to positive values based on known physical constraints and requirements of the model formula in @eq-ebase.
-Additionally, the prior distribution for $b$ was constrained to an upper limit of 0.504 (~twice the default mean).
+Additionally, the prior distribution for $b$ was constrained to an upper limit of 0.504 (\~twice the default mean).
 @Wanninkhof14 states that the $b$ parameter typically does not vary by more than 20% and the standard deviation of the default prior is sufficiently high to expand beyond this range.
@@
 Users can also specify alternative prior distributions from those in Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior] for each parameter.
-Functions are available for viewing model results, including a plot of the modeled dissolved oxygen with the observed, a time series plot of the primary metabolic estimates (P, R, and D, as areal rates with units $\runi$), and a time series plot of the credible intervals for the $a$, $r$, and $b$ parameters.
+Functions are available for viewing model results, including a plot of the modeled dissolved oxygen with the observed, a time series plot of the primary metabolic estimates (P, R, and D, as areal rates with units $\text{mmol}/\text{m}^3/\text{d}$), and a time series plot of the credible intervals for the $a$, $r$, and $b$ parameters.
 An additional option also allows the user to specify the model optimization period, where the time period can vary from a minimum of one day to the maximum of the entire length of the time series.
@@
 
-A preliminary assessment of the ability of EBASE to produce reasonable parameter estimates for $a$, $r$, and $b$ (@eq-ebase) was conducted using a synthetic time series with actual physical data as inputs.
+A preliminary assessment of the ability of EBASE to produce reasonable parameter estimates for $a$, $r$, and $b$ (@eq-ebase) was conducted using a synthetic time series with actual data as inputs.
 These assessments evaluated if 1) EBASE estimates similar values for the known parameters, and 2) how the estimated values change with different model configurations.
@@
 The input data were 2012 observations from February 22nd to December 19th at the Cat Point water quality monitoring station (29.7021$^{\circ}$ N, -84.8802$^{\circ}$ W) and included 15 minute observations of dissolved oxygen (mg/L), water temperature (C), and salinity (psu).
-Continuous meteorological observations from the nearby East Bay station (29.7909$^{\circ}$ N, -84.8834$^{\circ}$ W, ~10km north of Cat Point) were combined with the water quality data at the same time step and included air temperature (C), wind speed (m/s), and PAR (mmol/m$^2$ as total per 15 minute observation).
+Continuous meteorological observations from the nearby East Bay station (29.7909$^{\circ}$ N, -84.8834$^{\circ}$ W, \~10km north of Cat Point) were combined with the water quality data at the same time step and included air temperature (C), wind speed (m/s), and PAR (mmol/m$^2$ as total per 15 minute observation).
 An oxygen time series was generated using the inputs from Apalachicola Bay ([@fig-synapa]a), such that the synthetic dissolved oxygen was created as close as possible to the observed given the inputs.
 Known values for $a$ and $r$ parameters were returned for each time step of the synthetic data, as well as all metabolic estimates for production, respiration, and gas exchange ([@fig-synapa]b and c).
-The only parameter that could not be created as a continuous time series was $b$, which was set to 0.251 $\buni$ [@Wanninkhof14].
+The only parameter that could not be created as a continuous time series was $b$, which was set to 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ [@Wanninkhof14].
 
@@
 The standard prior distributions for the EBASE parameters follow normal distributions with defined means and standard deviations (Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]).
-The sensitivity analysis evaluated changes in the distributions by varying the standard deviations from small to large to test the effect of more or less constrained ranges, respectively, on the priors.
-For each parameter, low, medium, and high values for the standard deviation were assessed as 10%, 100%, and 1000% of the default values.
-Each unique combination of low, medium, and high values for the standard deviation of each parameter were evaluated, creating 3$^3$ = 27 different combinations of prior distributions that were used with the synthetic time series in EBASE.
+The sensitivity analysis evaluated changes in the distributions by varying the means and standard deviations from small to large to test the effect of changes in the central tendencies and more or less constrained ranges, respectively, on the priors.
+For each parameter, low and high values for the means and standard deviation were assessed as changes approximately equal to two orders of magnitude.
+For example the default value for the mean of the $a$ prior distribution is 0.2 and low and high values were assessed as 0 and 2, respectively.
+The only exception was the high value that was tested for the mean of the $b$, which was set at 0.504 or twice the default value (justification noted in @sec-ebasemod).
+Each unique combination of low and high values for the means and standard deviations of each parameter were evaluated, creating 2$^6$ = 64 different combinations of prior distributions that were used with the synthetic time series in EBASE.
 
@@
 An optimization period of one day was compared to results using an optimization period of seven days for the synthetic one-year time series.
-Each optimization period was evaluated for every unique combination of prior distributions, totaling 54 (27 times the two optimization periods) unique analyses for the assessment of changing standard deviations of all parameters.
+Each optimization period was evaluated for every unique combination of prior distributions, totaling 128 (64 times the two optimization periods) unique analyses for the assessment of changing priors of all parameters.
 
-Changing the variance of each of the prior distributions affected the ability of EBASE to return the known parameters from the synthetic time series.
+Changing the prior distributions affected the ability of EBASE to return the known parameters from the synthetic time series, with results varying by optimization period.
 @fig-apasumdat shows a summary of the comparisons of the synthetic data with EBASE for each unique combination of prior distributions and two optimization periods of one and seven days.
@@
 For both models, the metabolic parameters were estimated with high precision, and in some cases with nearly perfect precision (i.e., dissolved oxygen and gas exchange, D).
-Estimates for the fixed $b$ parameter were most variable, but not biased as the mean for both models was close to 0.251 $\buni$.
+Estimates for the fixed $b$ parameter were most variable, but not biased as the mean for both models was close to 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$.
 
@@
 The input data were provided to each method after converting the units as required (e.g., BASE requires PAR as $\mu$mol/m$^{2}$/s).
-Water column depth at Cat Point is approximately 1.8 m and all daily metabolic estimates returned by each method were converted to volumetric rates by dividing by the depth (i.e., $\runi$).
+Water column depth at Cat Point is approximately 1.8 m and all daily metabolic estimates returned by each method were converted to volumetric rates by dividing by the depth (i.e., $\text{mmol}/\text{m}^3/\text{d}$).
 Each comparison also evaluated metabolic estimates using the observed dissolved oxygen time series and a detided time series using weighted regression [@Beck15; available from @Beck21].
@@
 
-* Describe how EBASE deals with gaps
+-   Describe how EBASE deals with gaps
 
@@
 ![Comparison of EBASE results to a synthetic time series created with known metabolic parameters.
-Results are compared for unique combination of prior distributions, where the standard deviations are evaluated from low (L), medium (M), to high (H) for a one day and seven day model optimization period.
+Results are compared for unique combination of prior distributions, where the means and standard deviations are evaluated from low (L) to high (H) for a one day and seven day model optimization period.
 The metabolic results for dissolved oxygen (DO), production (P), respiration (R), gas exchange (D), and the $a$ parameter are evaluated based on the $R^2$ value of the linear fit between the EBASE results and synthetic values.
@@
 The results are averaged at the time step of the model optimization period.
-The $b$ parameter is fixed at 0.251 $\buni$.](figs/optex.png){#fig-optex fig-align="center"}
+The $b$ parameter is fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$.](figs/optex.png){#fig-optex fig-align="center"}
 

### 2023-02-26 R/funcs.R

@@ -87,7 +87,7 @@ apacmp_plo <- function(dat, xlb = 'EBASE', ylb = 'Odum', dotyp = 'observed', add
 
 # Fwoxy apa comparison to EBASE for different priors sd only
 priorcomp <- function(dat, ind){
-  
+
   met <- tibble(
     lbs = c('r2', 'rmse', 'aved'),
     lbspr = c('italic(R)^2', 'RMSE', 'Ave.\nDiff.'), 
@@ -99,13 +99,12 @@ priorcomp <- function(dat, ind){
   direc <- met$direc[ind]
   
   toplo <- dat %>% 
-    select(-amean, -rmean, -bmean) %>% 
     unnest('ests') %>% 
-    select(ndays, asd, rsd, bsd, var, matches(toshw)) %>%
+    select(ndays, amean, asd, rmean, rsd, bmean, bsd, var, matches(toshw)) %>%
     filter(!var %in% 'b') %>% 
     pivot_wider(names_from = 'var', values_from = !!toshw) %>% 
     mutate(
-      ind = sort(rep(1: (nrow(.) / 2), times = 2)), 
+      ind = rep(1: (nrow(.) / 2), times = 2), 
       ndays = case_when(
         ndays == 1 ~ paste(ndays, 'day'), 
         T ~ paste(ndays, 'days')
@@ -113,30 +112,23 @@ priorcomp <- function(dat, ind){
     )
   
   toplo1 <- toplo %>% 
-    select(ind, asd, rsd, bsd) %>% 
+    select(ind, amean, asd, rmean, rsd, bmean, bsd) %>% 
     unique() %>% 
-    mutate(
-      def = case_when(
-        asd == 0.1 & rsd == 5 & bsd == 0.01 ~ '*', 
-        T ~ ''
-      ),
-      asd = factor(asd, labels = c('L', 'M', 'H')), 
-      rsd = factor(rsd, labels = c('L', 'M', 'H')), 
-      bsd = factor(bsd, labels = c('L', 'M', 'H')), 
-    ) %>% 
-    pivot_longer(-c('ind', 'def'), names_to = 'var', values_to = 'val') %>% 
+    mutate_at(-ind, factor, labels = c('L', 'H')) %>% 
+    pivot_longer(-c('ind'), names_to = 'var', values_to = 'val') %>% 
     mutate(
       var = factor(var, 
-                   levels = c('asd', 'rsd', 'bsd'), 
-                   labels = c('italic(a)', 'italic(r)', 'italic(b)'))
[…]

### 2023-02-27 R/funcs.R
priorcomp function for apasumdat ordered by mean r2 rank

@@ -111,6 +111,30 @@ priorcomp <- function(dat, ind){
       )
     )
   
+  # ave r2 summary
+  optest <- dat %>% 
+    mutate(
+      ind = rep(1: (nrow(.) / 2), times = 2), 
+      ndays = case_when(
+        ndays == 1 ~ paste(ndays, 'day'), 
+        T ~ paste(ndays, 'days')
+      )
+    ) %>% 
+    unnest('ests') %>% 
+    summarise(
+      r2 = mean(r2), 
+      .by = c('ind', 'ndays')
+    ) %>% 
+    mutate(
+      rnkr2 = rank(-r2), 
+      .by = 'ndays'
+    )
+  
+  # get ndays = 1 only
+  optest <- optest %>% 
+    filter(ndays == '7 days') %>% 
+    select(-ndays)
+  
   toplo1 <- toplo %>% 
     select(ind, amean, asd, rmean, rsd, bmean, bsd) %>% 
     unique() %>% 
@@ -121,7 +145,10 @@ priorcomp <- function(dat, ind){
                    levels = c('amean', 'asd', 'rmean', 'rsd', 'bmean', 'bsd'), 
                    labels = c('italic(a)~mean', 'italic(a)~sd', 'italic(r)~mean', 'italic(r)~sd', 'italic(b)~mean', 'italic(b)~sd')), 
       fac = ''
-    ) 
+    ) %>% 
+    left_join(optest, by = 'ind', multiple= 'all') %>% 
+    arrange(rnkr2, var) %>% 
+    mutate(ind = factor(ind, levels = rev(unique(ind))))
   
   toplo2 <- toplo %>% 
     select(-amean, -asd, -rmean, -rsd, -bmean, -bsd) %>%
@@ -131,7 +158,10 @@ priorcomp <- function(dat, ind){
                    levels = c('DO_mod', 'P', 'R', 'D', 'a'), 
                    labels = c('DO', 'P', 'R', 'D', 'italic(a)')
       )
-    )
+    ) %>% 
+    left_join(optest, by = 'ind', multiple= 'all') %>% 
+    arrange(rnkr2, var) %>% 
+    mutate(ind = factor(ind, levels = rev(unique(ind))))
   
   p1 <- ggplot(toplo1, aes(y = ind, x = var, fill = val)) + 
     geom_tile(color = 'black') + 
@@ -147,7 +177,7 @@ priorcomp <- function(dat, ind){
     facet_wrap(~fac) + 
[…]

### 2023-02-28 manu-draft.qmd
update to apasumdat fig to show rankings #1

@@
 The metabolic results for dissolved oxygen (DO), production (P), respiration (R), gas exchange (D), and the $a$ parameter are evaluated based on the $R^2$ value of the linear fit between the EBASE results and synthetic values.
-The $b$ parameter is a fixed constant and was not evaluated.](figs/apasumdat.png){#fig-apasumdat}
+The $b$ parameter is a fixed constant and was not evaluated.
+Rankings of each combination as the average $R^2$ value across all parameters in a row are shown on the right, with the top three as black/bold and bottom three as black/italic.](figs/apasumdat.png){#fig-apasumdat}
 

### 2023-02-28 R/funcs.R

@@ -86,7 +86,7 @@ apacmp_plo <- function(dat, xlb = 'EBASE', ylb = 'Odum', dotyp = 'observed', add
 }
 
 # Fwoxy apa comparison to EBASE for different priors sd only
-priorcomp <- function(dat, ind){
+priorcomp <- function(dat, ind, topbot = 3){
 
   met <- tibble(
     lbs = c('r2', 'rmse', 'aved'),
@@ -111,6 +111,7 @@ priorcomp <- function(dat, ind){
       )
     )
   
+  
   # ave r2 summary
   optest <- dat %>% 
     mutate(
@@ -128,12 +129,17 @@ priorcomp <- function(dat, ind){
     mutate(
       rnkr2 = rank(-r2), 
       .by = 'ndays'
-    )
+    ) 
   
-  # get ndays = 1 only
+  topn <- sort(unique(optest$rnkr2))[1:topbot]
+  botn <- sort(unique(optest$rnkr2), decreasing = T)[1:topbot]
+  
   optest <- optest %>% 
-    filter(ndays == '7 days') %>% 
-    select(-ndays)
+    mutate(
+      fnts = ifelse(!rnkr2 %in% c(topn, botn), 'plain', ifelse(rnkr2 %in% topn, 'bold', 'italic')), 
+      cols = ifelse(rnkr2 %in% c(botn, topn), 'black', '#828282'), 
+      .by = 'ndays'
+    )
   
   toplo1 <- toplo %>% 
     select(ind, amean, asd, rmean, rsd, bmean, bsd) %>% 
@@ -145,10 +151,7 @@ priorcomp <- function(dat, ind){
                    levels = c('amean', 'asd', 'rmean', 'rsd', 'bmean', 'bsd'), 
                    labels = c('italic(a)~mean', 'italic(a)~sd', 'italic(r)~mean', 'italic(r)~sd', 'italic(b)~mean', 'italic(b)~sd')), 
       fac = ''
-    ) %>% 
-    left_join(optest, by = 'ind', multiple= 'all') %>% 
-    arrange(rnkr2, var) %>% 
-    mutate(ind = factor(ind, levels = rev(unique(ind))))
+    )
   
   toplo2 <- toplo %>% 
     select(-amean, -asd, -rmean, -rsd, -bmean, -bsd) %>%
@@ -159,9 +162,7 @@ priorcomp <- function(dat, ind){
[…]

### 2023-03-10 R/figs.R
update synapa and apasumdat fig

@@ -18,7 +18,7 @@ dev.off()
 
 # simulated apa data --------------------------------------------------------------------------
 
-fwdat <- read_csv(file = url('https://raw.githubusercontent.com/fawda123/BASEmetab_script/master/data/apafwoxy.csv'))
+fwdat <- read_csv(file = url('https://raw.githubusercontent.com/fawda123/BASEmetab_script/master/data/apafwoxy2.csv'))
 
 fwdatcmp <- fwdat %>% 
   mutate(

### 2023-03-10 R/funcs.R

@@ -149,7 +149,7 @@ priorcomp <- function(dat, ind, topbot = 3){
     mutate(
       var = factor(var, 
                    levels = c('amean', 'asd', 'rmean', 'rsd', 'bmean', 'bsd'), 
-                   labels = c('italic(a)~mean', 'italic(a)~sd', 'italic(r)~mean', 'italic(r)~sd', 'italic(b)~mean', 'italic(b)~sd')), 
+                   labels = c('italic(a)~mu', 'italic(a)~sigma', 'italic(r)~mu', 'italic(r)~sigma', 'italic(b)~mu', 'italic(b)~sigma')), 
       fac = ''
     )
   
@@ -163,7 +163,7 @@ priorcomp <- function(dat, ind, topbot = 3){
       )
     ) %>% 
     left_join(optest, by = c('ind', 'ndays'), multiple= 'all')
-  
+
   p1 <- ggplot(toplo1, aes(y = ind, x = var, fill = val)) + 
     geom_tile(color = 'black') + 
     theme(

### 2023-03-10 manu-draft.qmd
fixing apasumdat and optex plots for rmse and a m2

@@
 
-![Comparison of EBASE results for the best performing model to a synthetic time series created with known metabolic parameters.
-The best performing model for the one day (a, ndays = 1) and seven day (b, ndays = 7) optimization period is based on the EBASE results with prior distributions that produced the highest mean $R^2$ values across all parameter estimates (@fig-apasumdat).
-The standard deviations of the prior distributions of each parameter for the best performing model are shown at the top.
+![Comparison of EBASE results for the best and worst performing models to a synthetic time series created with known metabolic parameters.
+The best and worst performing models for the one day (a, b) and seven day (c, d) optimization periods are based on the EBASE results with prior distributions that produced the highest mean $R^2$ values across all parameter estimates (@fig-apasumdat).
 The results are averaged at the time step of the model optimization period.

### 2023-03-10 R/figs.R

@@ -72,40 +72,61 @@ fl <- paste0(tempdir(), '/apasumdat.RData')
 download.file('https://github.com/fawda123/BASEmetab_script/raw/master/data/apasumdat.RData', destfile = fl)
 load(file = fl)
 
-p1 <- priorcomp(apasumdat, 1)
+p1 <- priorcomp(apasumdat, met = 'rmse')
 
 png(here('figs/apasumdat.png'), height = 6, width = 6, family = 'serif', units = 'in', res = 500)
 print(p1)
 dev.off()
 
-# best fwoxy apa comparison -------------------------------------------------------------------
+# best/worst fwoxy apa comp -------------------------------------------------------------------
 
-fl <- paste0(tempdir(), '/apagrd.RData')
-download.file('https://github.com/fawda123/BASEmetab_script/raw/master/data/apagrd.RData', destfile = fl)
+# grid ests
+fl1a <- paste0(tempdir(), '/apagrd1a.RData')
+download.file('https://github.com/fawda123/BASEmetab_script/raw/master/data/apagrd1a.RData', destfile = fl1a)
+load(file = fl1a)
+fl1b <- paste0(tempdir(), '/apagrd1b.RData')
+download.file('https://github.com/fawda123/BASEmetab_script/raw/master/data/apagrd1b.RData', destfile = fl1b)
+load(file = fl1b)
+fl7a <- paste0(tempdir(), '/apagrd7a.RData')
+download.file('https://github.com/fawda123/BASEmetab_script/raw/master/data/apagrd7a.RData', destfile = fl7a)
+load(file = fl7a)
+fl7b <- paste0(tempdir(), '/apagrd7b.RData')
+download.file('https://github.com/fawda123/BASEmetab_script/raw/master/data/apagrd7b.RData', destfile = fl7b)
+load(file = fl7b)
+
+apagrd <- bind_rows(apagrd1a, apagrd1b, apagrd7a, apagrd7b)
+
+# summary ests
+fl <- paste0(tempdir(), '/apasumdat.RData')
+download.file('https://github.com/fawda123/BASEmetab_script/raw/master/data/apasumdat.RData', destfile = fl)
[…]

### 2023-03-10 R/funcs.R

@@ -86,17 +86,22 @@ apacmp_plo <- function(dat, xlb = 'EBASE', ylb = 'Odum', dotyp = 'observed', add
 }
 
 # Fwoxy apa comparison to EBASE for different priors sd only
-priorcomp <- function(dat, ind, topbot = 3){
+priorcomp <- function(dat, met, topbot = 3){
 
-  met <- tibble(
-    lbs = c('r2', 'rmse', 'aved'),
-    lbspr = c('italic(R)^2', 'RMSE', 'Ave.\nDiff.'), 
-    direc = c(-1, 1, 1)
-  )
-  
-  toshw <- met$lbs[ind]
-  leglb <- met$lbspr[ind]
-  direc <- met$direc[ind]
+  metsel <- tibble(
+      met = c('r2', 'rmse', 'mape', 'mae'),
+      lbspr = c('italic(R)^2', 'RMSE', 'MAPE', 'mae'), 
+      direc = c(-1, 1, 1 ,1),
+      limts = list(c(0, 100), NULL, NULL, NULL),
+      trans = c('identity', 'log10', 'log10', 'log10')
+    ) %>% 
+    filter(met == !!met)
+ 
+  toshw <- metsel$met
+  leglb <- metsel$lbspr
+  direc <- metsel$direc
+  limts <- metsel$limts[[1]]
+  trans <- metsel$trans
   
   toplo <- dat %>% 
     unnest('ests') %>% 
@@ -111,40 +116,29 @@ priorcomp <- function(dat, ind, topbot = 3){
       )
     )
   
-  
   # ave r2 summary
-  optest <- dat %>% 
+  optest <- meansum_fun(dat, met) %>% 
     mutate(
-      ind = rep(1: (nrow(.) / 2), times = 2), 
       ndays = case_when(
         ndays == 1 ~ paste(ndays, 'day'), 
         T ~ paste(ndays, 'days')
       )
-    ) %>% 
-    unnest('ests') %>% 
-    summarise(
-      r2 = mean(r2), 
-      .by = c('ind', 'ndays')
-    ) %>% 
-    mutate(
-      rnkr2 = rank(-r2), 
-      .by = 'ndays'
-    ) 
+    )
   
-  topn <- sort(unique(optest$rnkr2))[1:topbot]
-  botn <- sort(unique(optest$rnkr2), decreasing = T)[1:topbot]
+  topn <- sort(unique(optest$rnkmetsum))[1:topbot]
[…]

### 2023-03-15 manu-draft.qmd
maybe better prior comp figure #1

@@
 $$
-a \sim N(0.2, 0.1)T(0,)
+a \sim N(0, 1)T(0,)
 $$ {#eq-aprior}
@@
 $$
-r \sim N(20, 5)T(0,)
+r \sim N(0, 50)T(0,)
 $$ {#eq-rprior}
@@
 $$
-b \sim N(0.251, 0.01)T(0,0.504)
+b \sim N(0, 0.1)T(0,0.502)
 $$ {#eq-bprior}
@@
 A preliminary assessment of the ability of EBASE to produce reasonable parameter estimates for $a$, $r$, and $b$ (@eq-ebase) was conducted using a synthetic time series with actual data as inputs.
-These assessments evaluated if 1) EBASE estimates similar values for the known parameters, and 2) how the estimated values change with different model configurations.
+These assessments evaluated if 1) EBASE estimates similar values for the known parameters in the synthetic time series, and 2) how the estimated values change with different model configurations.
 The synthetic time series used inputs from an approximate one year of continuous data from Apalachicola Bay, Florida.
-The input data were 2012 observations from February 22nd to December 19th at the Cat Point water quality monitoring station (29.7021$^{\circ}$ N, -84.8802$^{\circ}$ W) and included 15 minute observations of dissolved oxygen (mg/L), water temperature (C), and salinity (psu).
+The input data were 2012 observations at the Cat Point water quality monitoring station (29.7021$^{\circ}$ N, -84.8802$^{\circ}$ W) and included 15 minute observations of water temperature (C) and salinity (psu).
 Continuous meteorological observations from the nearby East Bay station (29.7909$^{\circ}$ N, -84.8834$^{\circ}$ W, \~10km north of Cat Point) were combined with the water quality data at the same time step and included air temperature (C), wind speed (m/s), and PAR (mmol/m$^2$ as total per 15 minute observation).
-An oxygen time series was generated using the inputs from Apalachicola Bay ([@fig-synapa]a), such that the synthetic dissolved oxygen was created as close as possible to the observed given the inputs.
-Known values for $a$ and $r$ parameters were returned for each time step of the synthetic data, as well as all metabolic estimates for production, respiration, and gas exchange ([@fig-synapa]b and c).
+Missing observations in the water quality and meteorological data were filled using autoregressive modelling of the actual parameters versus time to create a complete dataset from January 1st to December 31st.
+An oxygen time series was then generated using the inputs from Apalachicola Bay ([@fig-synapa]a), such that the synthetic dissolved oxygen was created using @eq-ebase as a forward calculating model and user-defined values for $a$, $r$, and $b$.
+All metabolic rates for production, respiration, and gas exchange ([@fig-synapa]b and c) were also estimated with the synthetic time series.
 The only parameter that could not be created as a continuous time series was $b$, which was set to 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ [@Wanninkhof14].
+Accordingly, the synthetic dissolved oxygen time series and the observed water quality and meteorological data were used with EBASE to determine if reasonable estimates for the known parameters and metabolic rates were recovered.
 
@@
 First, results were compared to the known values after changing characteristics of their prior distributions.
-Second, results were compared using a model period of optimization for one day and seven days.
+Second, results were compared using a model period of optimization for one day and seven days to assess how results varied based on the amount of data used for the Bayesian estimation.
 Details of these analyses are as follows.
 
-The standard prior distributions for the EBASE parameters follow normal distributions with defined means and standard deviations (Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]).
+The prior distributions for the EBASE parameters follow truncated normal distributions with defined means and standard deviations (Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]).
 The sensitivity analysis evaluated changes in the distributions by varying the means and standard deviations from small to large to test the effect of changes in the central tendencies and more or less constrained ranges, respectively, on the priors.
-For each parameter, low and high values for the means and standard deviation were assessed as changes approximately equal to two orders of magnitude.
-For example the default value for the mean of the $a$ prior distribution is 0.2 and low and high values were assessed as 0 and 2, respectively.
-The only exception was the high value that was tested for the mean of the $b$, which was set at 0.504 or twice the default value (justification noted in @sec-ebasemod).
+For each parameter, low and high values for the means and standard deviations were assessed as changes approximately equal to two orders of magnitude of the means and standard deviations of the known parameters of the synthetic time series.
+The only exception was the $b$ parameter where the low values created priors with a distribution close to zero and the high values were equivalent to a uniform prior distribution in the range of 0 to 0.504 (justification noted in @sec-ebasemod).
 Each unique combination of low and high values for the means and standard deviations of each parameter were evaluated, creating 2$^6$ = 64 different combinations of prior distributions that were used with the synthetic time series in EBASE.
@@
 The optimization period used for a time series determines how much data are used to estimate the three EBASE parameters ($a$, $r$, and $b$).
-Unlike BASE, the EBASE software allows flexibility in the period of time used for optimization, with a minimum period of time being one day and a maximum period being the length of the time series.
+Unlike BASE, the EBASE software allows flexibility in the period of time used for optimization, with a minimum being one day and a maximum being the length of the time series.
 For a chosen optimization period, single estimates for $a$, $r$ and $b$ are returned by the model.
@@
 
-Changing the prior distributions affected the ability of EBASE to return the known parameters from the synthetic time series, with results varying by optimization period.
-@fig-apasumdat shows a summary of the comparisons of the synthetic data with EBASE for each unique combination of prior distributions and two optimization periods of one and seven days.
-The results are compared using R-squared ($R^2$) values from the linear fit between the synthetic data and results from EBASE.
-Darker colors represent lower $R^2$ values and can be used to assess how changing the variance of the priors affects the ability of EBASE to estimate the true parameter.
+Changing the prior distributions affected the ability of EBASE to return the known parameters from the synthetic time series, with results varying somewhat by optimization period.
+@fig-priorcomp shows a summary of the comparisons of the synthetic data with EBASE for each unique combination of prior distributions and two optimization periods of one and seven days.
+The results are shown using Root Mean Squared Errors (RMSE) values comparing the synthetic estimates with those from EBASE.
+Darker colors represent higher RMSE values and can be used to assess how changing the priors affects the ability of EBASE to estimate the true parameter.
 
-Several conclusions can be made from the results in @fig-apasumdat based on the patterns of the $R^2$ values.
+Several conclusions can be made from the results in @fig-priorcomp based on the patterns of the $R^2$ values.
 Overall, the model was able to estimate the $a$ parameter with reasonable precision regardless of any of the prior distributions, whereas most of the variation was observed in the $r$ (respiration, R) and $b$ parameters.
@@
 The actual model results for the best performing combination of prior distributions for the one day and seven day optimization are shown in @fig-optex (all results are averaged at the time-step of the optimization period for comparability).
-The best performing models were those with the highest mean $R^2$ values across all parameter estimates in @fig-apasumdat.
+The best performing models were those with the highest mean $R^2$ values across all parameter estimates in @fig-priorcomp.
 For both models, the metabolic parameters were estimated with high precision, and in some cases with nearly perfect precision (i.e., dissolved oxygen and gas exchange, D).
@@
 
-\newpage
+{{< pagebreak >}}
 
@@
 The $b$ parameter is a fixed constant and was not evaluated.
-Rankings of each combination as the average $R^2$ value across all parameters in a row are shown on the right, with the top three as black/bold and bottom three as black/italic.](figs/apasumdat.png){#fig-apasumdat}
+Rankings of each combination as the average $R^2$ value across all parameters in a row are shown on the right, with the top three as black/bold and bottom three as black/italic.](figs/priorcomp.png){#fig-priorcomp}
 
-\newpage
+{{< pagebreak >}}
+
+![Add caption.](figs/priorsumcomp.png){#fig-priorsumcomp}
+
+{{< pagebreak >}}
 
 ![Comparison of EBASE results for the best and worst performing models to a synthetic time series created with known metabolic parameters.
-The best and worst performing models for the one day (a, b) and seven day (c, d) optimization periods are based on the EBASE results with prior distributions that produced the highest mean $R^2$ values across all parameter estimates (@fig-apasumdat).
+The best and worst performing models for the one day (a, b) and seven day (c, d) optimization periods are based on the EBASE results with prior distributions that produced the highest mean $R^2$ values across all parameter estimates (@fig-priorcomp).
 The results are averaged at the time step of the model optimization period.
@@
 
-\newpage
+{{< pagebreak >}}
 
-<!-- \newpage -->
+<!-- {{< pagebreak >}} -->
 
@@
 
-<!-- \newpage -->
+<!-- {{< pagebreak >}} -->
 

### 2023-03-15 R/figs.R

@@ -74,7 +74,19 @@ load(file = fl)
 
 p1 <- priorcomp(apasumdat, met = 'rmse')
 
-png(here('figs/apasumdat.png'), height = 6, width = 6, family = 'serif', units = 'in', res = 500)
+png(here('figs/priorcomp.png'), height = 6, width = 6, family = 'serif', units = 'in', res = 500)
+print(p1)
+dev.off()
+
+# Fwoxy apa comparison RMSE sum -------------------------------------------
+
+fl <- paste0(tempdir(), '/apasumdat.RData')
+download.file('https://github.com/fawda123/BASEmetab_script/raw/master/data/apasumdat.RData', destfile = fl)
+load(file = fl)
+
+p1 <- priorsumcomp(apasumdat)
+
+png(here('figs/priorsumcomp.png'), height = 6, width = 5, family = 'serif', units = 'in', res = 500)
 print(p1)
 dev.off()
 

### 2023-03-15 R/funcs.R

@@ -210,6 +210,57 @@ priorcomp <- function(dat, met, topbot = 3){
   
 }
 
+# Fwoxy apa comparison to EBASE for different priors, rmse summary
+priorsumcomp <- function(dat){
+  
+  toplo <- dat %>% 
+    unnest(ests) %>% 
+    select(-ind, -r2, -mape, -mae) %>% 
+    mutate_at(vars(matches('mean$|sd$')), factor, labels = c('L', 'H')) %>% 
+    mutate(
+      ndays = case_when(
+        ndays == 1 ~ paste(ndays, 'day'), 
+        T ~ paste(ndays, 'days')
+      )
+    ) %>% 
+    pivot_longer(cols = matches('mean$|sd$'), names_to = 'prior', values_to = 'val') %>% 
+    summarise(
+      medianrmse = median(rmse), 
+      iqr = IQR(rmse),
+      .by = c('ndays', 'prior', 'val')
+    ) %>% 
+    mutate(
+      param = gsub('mean$|sd$', '', prior),
+      param = factor(param, levels = c('a', 'r', 'b')),
+      prior = gsub('^a|^r|^b', '', prior), 
+      prior = factor(prior, levels = c('mean', 'sd'), labels = c('mu', 'sigma')), 
+      ndays = factor(ndays, levels = c('1 day', '7 days'), labels = c('1~day', '7~days'))
+    )
+  
+  p <- ggplot(toplo, aes(x = param, group = val)) + 
+    geom_point(aes(y = medianrmse, fill = val, size = iqr), pch = 21) + #, position = position_dodge(width = wd)) +
+    facet_grid(prior~ndays, labeller = label_parsed) + 
+    scale_fill_brewer(palette = 'Blues') + 
+    scale_size(range = c(1, 8), breaks = c(20, 60, 100, 140), labels = c('20', '60', '100', '140')) + 
+    guides(fill = guide_legend(override.aes = list(size = 5))) +
+    theme_bw() + 
+    theme(
+      axis.text.x = element_text(size = 14, face = 'italic'), 
+      strip.text.y = element_text(size = 14, face = 'italic'), 
+      strip.text.x = element_text(size = 12),
+      strip.background = element_blank()
+    ) +
+    labs(
+      x = NULL, 
+      y = 'Median RMSE', 
+      fill = NULL, 
+      size = 'IQR'
+    )
+  
+  return(p)
+  
+}
+
[…]

### 2023-03-19 R/figs.R
comparind cfd and r2 figs

@@ -72,7 +72,7 @@ fl <- paste0(tempdir(), '/apasumdat.RData')
 download.file('https://github.com/fawda123/BASEmetab_script/raw/master/data/apasumdat.RData', destfile = fl)
 load(file = fl)
 
-p1 <- priorcomp(apasumdat, met = 'rmse')
+p1 <- priorcomp(apasumdat, met = 'r2')
 
 png(here('figs/priorcomp.png'), height = 6, width = 6, family = 'serif', units = 'in', res = 500)
 print(p1)
@@ -84,7 +84,7 @@ fl <- paste0(tempdir(), '/apasumdat.RData')
 download.file('https://github.com/fawda123/BASEmetab_script/raw/master/data/apasumdat.RData', destfile = fl)
 load(file = fl)
 
-p1 <- priorsumcomp(apasumdat)
+p1 <- priorsumcomp(apasumdat, met = 'r2')
 
 png(here('figs/priorsumcomp.png'), height = 6, width = 5, family = 'serif', units = 'in', res = 500)
 print(p1)

### 2023-03-19 R/funcs.R

@@ -89,11 +89,11 @@ apacmp_plo <- function(dat, xlb = 'EBASE', ylb = 'Odum', dotyp = 'observed', add
 priorcomp <- function(dat, met, topbot = 3){
 
   metsel <- tibble(
-      met = c('r2', 'rmse', 'mape', 'mae'),
-      lbspr = c('italic(R)^2', 'RMSE', 'MAPE', 'mae'), 
-      direc = c(-1, 1, 1 ,1),
-      limts = list(c(0, 100), NULL, NULL, NULL),
-      trans = c('identity', 'log10', 'log10', 'log10')
+      met = c('r2', 'rmse', 'mape', 'mae', 'cfd'),
+      lbspr = c('italic(R)^2', 'RMSE', 'MAPE', 'mae', 'Coef.~of~Det.'), 
+      direc = c(-1, 1, 1, 1, -1),
+      limts = list(c(0, 100), NULL, NULL, NULL, c(0, 1)),
+      trans = c('identity', 'log10', 'log10', 'log10', 'identity')
     ) %>% 
     filter(met == !!met)
  
@@ -211,11 +211,12 @@ priorcomp <- function(dat, met, topbot = 3){
 }
 
 # Fwoxy apa comparison to EBASE for different priors, rmse summary
-priorsumcomp <- function(dat){
-  
+priorsumcomp <- function(dat, met = 'r2'){
+
   toplo <- dat %>% 
     unnest(ests) %>% 
-    select(-ind, -r2, -mape, -mae) %>% 
+    select(amean, asd, rmean, rsd, bmean, bsd, ndays, ind, var, !!met) %>% 
+    rename(met = !!met) %>% 
     mutate_at(vars(matches('mean$|sd$')), factor, labels = c('L', 'H')) %>% 
     mutate(
       ndays = case_when(
@@ -225,8 +226,8 @@ priorsumcomp <- function(dat){
     ) %>% 
     pivot_longer(cols = matches('mean$|sd$'), names_to = 'prior', values_to = 'val') %>% 
     summarise(
-      medianrmse = median(rmse), 
-      iqr = IQR(rmse),
+      medv = median(met), 
+      iqr = IQR(met),
       .by = c('ndays', 'prior', 'val')
     ) %>% 
     mutate(
@@ -237,8 +238,15 @@ priorsumcomp <- function(dat){
       ndays = factor(ndays, levels = c('1 day', '7 days'), labels = c('1~day', '7~days'))
[…]

### 2023-03-21 R/figs.R
starting rgn revisions

@@ -72,19 +72,19 @@ fl <- paste0(tempdir(), '/apasumdat.RData')
 download.file('https://github.com/fawda123/BASEmetab_script/raw/master/data/apasumdat.RData', destfile = fl)
 load(file = fl)
 
-p1 <- priorcomp(apasumdat, met = 'r2')
+p1 <- priorcomp(apasumdat, met = 'nse')
 
 png(here('figs/priorcomp.png'), height = 6, width = 6, family = 'serif', units = 'in', res = 500)
 print(p1)
 dev.off()
 
-# Fwoxy apa comparison RMSE sum -------------------------------------------
+# Fwoxy apa comparison met sum --------------------------------------------
 
 fl <- paste0(tempdir(), '/apasumdat.RData')
 download.file('https://github.com/fawda123/BASEmetab_script/raw/master/data/apasumdat.RData', destfile = fl)
 load(file = fl)
 
-p1 <- priorsumcomp(apasumdat, met = 'r2')
+p1 <- priorsumcomp(apasumdat, met = 'nse')
 
 png(here('figs/priorsumcomp.png'), height = 6, width = 5, family = 'serif', units = 'in', res = 500)
 print(p1)
@@ -128,10 +128,10 @@ fwdatcmp <- fwdat %>%
   ) %>% 
   select(Date, DateTimeStamp, DO_obs, a, b, P, R, D)
 
-p1 <- optex(apagrd, fwdatcmp, apasumdat, rnkmetsum = 1, ndays = 1, met = 'rmse', subttl = '(a) Best priors, 1 day')
-p2 <- optex(apagrd, fwdatcmp, apasumdat, rnkmetsum = 64, ndays = 1, met = 'rmse', subttl = '(b) Worst priors, 1 day', ylbs = F)
-p3 <- optex(apagrd, fwdatcmp, apasumdat, rnkmetsum = 1, ndays = 7, met = 'rmse', subttl = '(c) Best priors, 7 day', ylbs = F)
-p4 <- optex(apagrd, fwdatcmp, apasumdat, rnkmetsum = 64, ndays = 7, met = 'rmse', subttl = '(d) Worst priors, 7 day', ylbs = F)
+p1 <- optex(apagrd, fwdatcmp, apasumdat, rnkmetsum = 1, ndays = 1, met = 'cfd', subttl = '(a) Best priors, 1 day')
[…]

### 2023-03-21 R/funcs.R

@@ -89,8 +89,8 @@ apacmp_plo <- function(dat, xlb = 'EBASE', ylb = 'Odum', dotyp = 'observed', add
 priorcomp <- function(dat, met, topbot = 3){
 
   metsel <- tibble(
-      met = c('r2', 'rmse', 'mape', 'mae', 'cfd'),
-      lbspr = c('italic(R)^2', 'RMSE', 'MAPE', 'mae', 'Coef.~of~Det.'), 
+      met = c('r2', 'rmse', 'mape', 'mae', 'nse'),
+      lbspr = c('italic(R)^2', 'RMSE', 'MAPE', 'mae', 'NSE'), 
       direc = c(-1, 1, 1, 1, -1),
       limts = list(c(0, 100), NULL, NULL, NULL, c(0, 1)),
       trans = c('identity', 'log10', 'log10', 'log10', 'identity')
@@ -117,7 +117,7 @@ priorcomp <- function(dat, met, topbot = 3){
     )
   
   # ave r2 summary
-  optest <- meansum_fun(dat, met) %>% 
+  optest <- metsum_fun(dat, met) %>% 
     mutate(
       ndays = case_when(
         ndays == 1 ~ paste(ndays, 'day'), 
@@ -158,6 +158,9 @@ priorcomp <- function(dat, met, topbot = 3){
     ) %>% 
     left_join(optest, by = c('ind', 'ndays'), multiple= 'all')
 
+  if(met == 'nse')
+    toplo2$val <- pmax(toplo2$val, 0)
+
   p1 <- ggplot(toplo1, aes(y = ind, x = var, fill = val)) + 
     geom_tile(color = 'black') + 
     theme(
@@ -193,7 +196,7 @@ priorcomp <- function(dat, met, topbot = 3){
       panel.background = element_blank()
     ) + 
     facet_wrap(~ndays, ncol = 2) + 
-    scale_fill_distiller(palette = 'YlOrRd', direction = direc, limits = NULL, trans = trans) + 
+    scale_fill_distiller(palette = 'YlOrRd', direction = direc, limits = limts, trans = trans) + 
     scale_x_discrete(position = 'top', expand = c(0, 0), labels = parse(text = levels(toplo2$var))) + 
     scale_y_reverse(expand = c(0, 0)) + 
     coord_cartesian(clip = 'off') + 
@@ -226,8 +229,8 @@ priorsumcomp <- function(dat, met = 'r2'){
[…]

### 2023-03-21 manu-draft.qmd
starting rgn revisions
more rgn edits

@@
   - TBEP: Tampa Bay Estuary Program, St. Petersburg, Florida 33701 USA
-  - PSU: Pennsylvania State University, University Park, Pennsylvania 16802 USA
+  - PSU: The Pennsylvania State University, University Park, Pennsylvania 16802 USA
 
@@
 $$
-\frac{\partial{C_d}}{\partial{t}} = [\,aPAR]\, - [\,r]\, - \frac{1}{H}\left[-bU_{10}^2\left(\frac{s_c}{600} \right)^{-0.5} \left(C_{Sat} - C_d \right )\right]
+H\frac{dC}{dt} = aPAR - r - \left[bU_{10}^2\left(\frac{Sc}{600} \right)^{-0.5} \left(C_{Sat} - C \right )\right]
 $$ {#eq-ebase}
 
-where the terms in the square brackets are production (P), respiration (R), and gas exchange (D), respectively, modeled as a function of the change in dissolved oxygen ($\partial{C_d}$) per unit time ($\partial{t}$) of the input data.
+where the terms in the square brackets are production (P), respiration (R), and gas exchange (D), respectively, modeled as a function of the change in dissolved oxygen ($dC$) per unit time ($dt$) of the input data.
 Production is modeled linearly as a function of surface irradiance ($PAR$, Watts m$^{-2}$) and $a$ (light efficiency, $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$).
 Respiration is modeled with the single parameter $r$ ($\text{mmol}/\text{m}^3/\text{d}$).
-Gas exchange is modeled using the formulation in @Wanninkhof14 as a wind-based parameterization for the gas transfer velocity [@Ho16] multiplied by the oxygen concentration gradient as the difference between the concentration measured in the surface water layer, $C_d$ (mmol m$^{-3}$), and the saturation (or equilibrium) concentration, $C_{Sat}$ (mmol m$^{-3}$), calculated from salinity (psu) and temperature ($^\circ$C) [@Millero81; @Garcia92].
-The gas transfer velocity is modeled using wind speed squared at 10 meters above the water ($U_{10}^2$, m$^2$ s$^{-2}$), water column depth ($H$ in meters, as a fixed parameter or time series from a pressure sensor), and the Schmidt number ($S_c$) defined as the ratio of the kinematic viscosity of water to the molecular diffusivity of the gas for oxygen [from water temperature and salinity using the polynomial fit in @Wanninkhof14].
+Gas exchange is modeled using the formulation in @Wanninkhof14 as a wind-based parameterization for the gas transfer velocity [@Ho16] multiplied by the oxygen concentration gradient as the difference between the concentration measured in the surface water layer, $C$ (mmol m$^{-3}$), and the saturation (or equilibrium) concentration, $C_{Sat}$ (mmol m$^{-3}$), calculated from salinity (psu) and temperature ($^\circ$C) [@Millero81; @Garcia92].
+The gas transfer velocity is modeled using wind speed squared at 10 meters above the water ($U_{10}^2$, m$^2$ s$^{-2}$), water column depth ($H$ in meters, as a fixed parameter or time series from a pressure sensor), and the Schmidt number ($Sc$) defined as the ratio of the kinematic viscosity of water to the molecular diffusivity of the gas for oxygen [from water temperature and salinity using the polynomial fit in @Wanninkhof14].
 The $b$ parameter is estimated from the data and has units of $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$.
@@
 $$
-\frac{\partial{C_d}}{\partial{t}} = [\,aPAR^p]\, - [\,r\Theta^{T - \bar{T}}]\, - \left[-K\cdot1.0241^{T-\bar{T}}\left(C_{sat} - C_d \right )\right]
+\frac{dC}{dt} = aPAR^p - r\Theta^{T - \bar{T}} + \left[K\cdot1.0241^{T-\bar{T}}\left(C_{sat} - C \right )\right]
 $$ {#eq-base}
@@
 A temperature dependency, $T$, is also included where respiration and gas exchange vary with temperature.
-The changes to @eq-base for EBASE were necessary for estuarine application and included 1) a linear relationship of production to PAR ($aPAR$) by setting $p = 1$; 2) constant respiration throughout the time series by removing the temperature dependency; and 3) use of a gas exchange parameterization commonly used in coastal or oceanic settings [i.e., @Grace15 used a littoral parameterization].
+The changes to @eq-base for EBASE were necessary for estuarine application and included 1) a linear relationship of production to PAR ($aPAR$) by setting $p = 1$; 2) constant respiration by removing the temperature dependency; and 3) use of a gas exchange parameterization commonly used in coastal or oceanic settings that depends on wind speed and temperature [i.e., @Grace15 used a littoral parameterization].
 The justification for these changes is as follows.
@@
 A common theory behind most gas exchange parameterizations for metabolism estimates is that the transfer of gases of relatively low solubility, like oxygen, is proportional to the concentration gradient in a thin film of water at the air--water interface.
-It is also assumed that the water at the top of the film is saturated with respect to the atmosphere, so that the concentration gradient across the thin film can be approximated as the difference between oxygen concentration measured in the surface water layer, $C_d$, and the saturation (or equilibrium) concentration $C_{Sat}$, which is usually calculated as a function of water temperature and salinity (the negative sign is needed because positive flux is set to be from water to air):
+It is also assumed that the water at the top of the film is saturated with respect to the atmosphere, so that the concentration difference across the thin film can be approximated as the difference between oxygen concentration measured in the surface water layer, $C$, and the saturation (or equilibrium) concentration $C_{Sat}$, which is usually calculated as a function of water temperature and salinity (the negative sign is needed because positive flux is set to be from water to air):
 
 $$
-D = -k_w \left(C_{sat}-C_d \right).
+D = -k_w \left(C_{sat}-C \right).
 $$ {#eq-gasex}
@@
 
-where $\bar{T}$ is the average water temperature over each 24-hour period in the diel oxygen time series and $K$ is the reaeration coefficient estimated from the data.
+where $\bar{T}$ is the average water temperature over each 24-hour period in the oxygen time series and $K$ is the reaeration coefficient estimated from the data.
 @eq-gasexbase indicates that $k_w$ increases with temperature and depth.
@@
 As for the BASE model, relatively uninformed prior distributions following a normal Gaussian distribution are used by default, although the priors can be changed based on previous knowledge of parameters specific to an ecosystem or as informed by other metabolic modeling approaches.
-The default priors were based on those that returned reasonable parameter estimates using a known time series created with forward-feeding oxygen mass balance inputs to @eq-ebase.
-[Fwoxy R package, @Arriola22].
-All of the physical parameters were held constant, i.e., wind speed, temperature, and salinity.
-The other required inputs were the $a$, $r$, and $b$ parameters, chosen as 0.2 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$, 20 $\text{mmol}/\text{m}^3/\text{d}$, and 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$, respectively.
-An initial oxygen concentration was provided as 250 mmol/m$^3$ and the time series was created by forward addition of the inputs.
-As such, the dissolved oxygen output as change per unit time was modelled by EBASE using the constant the same physical parameters as input.
-Reasonable prior distributions by EBASE were identified (Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]) based on the ability to return the known parameters:
+Reasonable uninformed prior distributions for EBASE were chosen based on normal distributions with mean values based on approximate estimates from the literature [@Wanninkhof14;@Grace15] and standard deviations that were sufficiently large to allow the Bayesian routine to search an unconstrained parameter space.
+The default priors for $a$, $r$, and $b$ are:
 
 $$
-a \sim N(0, 1)T(0,)
+a \sim N(0.2, 1)T(0,\infty) \;\; \left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)
 $$ {#eq-aprior}
@@
 $$
-r \sim N(0, 50)T(0,)
+r \sim N(20, 50)T(0,\infty) \;\; \text{mmol}/\text{m}^3/\text{d}
 $$ {#eq-rprior}
@@
 $$
-b \sim N(0, 0.1)T(0,0.502)
+b \sim N(0.251, 0.1)T(0,0.502) \;\; \left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)
 $$ {#eq-bprior}
 
-Each follows a normal distribution with means and standard deviation defined in the parentheses (@fig-priorplot).
-The $T$ notation is JAGS model syntax indicating the distributions are truncated to a minimum (left) and maximum (right) value defined in the parentheses.
+where $N(\alpha, \beta)$ indicates a normal distribution with mean $\alpha$ and standard deviation $\beta$ and $T(\gamma, \delta)$ is 1 between $\gamma$ and $\theta$ and 0 elsewhere; $T$ has the effect of truncating the normal distribution to minimum and maximum values.
 All prior distributions were constrained to positive values based on known physical constraints and requirements of the model formula in @eq-ebase.
-Additionally, the prior distribution for $b$ was constrained to an upper limit of 0.504 (\~twice the default mean).
-@Wanninkhof14 states that the $b$ parameter typically does not vary by more than 20% and the standard deviation of the default prior is sufficiently high to expand beyond this range.
-Initial development of EBASE also showed that using an undefined upper limit of $b$ led to unstable and unreasonable parameter estimates.
+Additionally, the prior distribution for $b$ was constrained to an upper limit of 0.502 (\~twice the default mean).
+Initial development of EBASE showed that using an undefined upper limit of $b$ led to unstable and unreasonable parameter estimates.
+Examples of the prior distributions in Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior] are shown in @fig-priorplot.
 

### 2023-03-22 manu-draft.qmd
more rgn edits, added nse methods

@@
 The BASE method provides options to estimate both $\Theta$ and $p$ from the data or to set as fixed values defined in the model file ($\Theta = 1.07177$, $p=1$).
-For EBASE, the temperature dependency of respiration is removed by setting $\Theta^{(T-\bar{T})}$ = 1 and it is assumed that respiration is constant as in other methods [@Odum56; @Murrel18].
+For EBASE, the temperature dependency of respiration is removed by setting $\Theta^{(T-\bar{T})}$ = 1 and it is assumed that respiration is constant as in other methods [@Odum56; @Murrell18].
 The $p$ parameter, which is dimensionless, is also set to 1 as in @Holtgrieve10, such that a linear relationship of production to $PAR$ was used.
@@
 As for the BASE model, relatively uninformed prior distributions following a normal Gaussian distribution are used by default, although the priors can be changed based on previous knowledge of parameters specific to an ecosystem or as informed by other metabolic modeling approaches.
-Reasonable uninformed prior distributions for EBASE were chosen based on normal distributions with mean values based on approximate estimates from the literature [@Wanninkhof14;@Grace15] and standard deviations that were sufficiently large to allow the Bayesian routine to search an unconstrained parameter space.
+Reasonable uninformed prior distributions for EBASE were chosen with mean values using approximate estimates from the literature [@Wanninkhof14;@Grace15] and standard deviations that were sufficiently large to allow the Bayesian routine to search an unconstrained parameter space.
 The default priors for $a$, $r$, and $b$ are:
@@
 Functions are available for viewing model results, including a plot of the modeled dissolved oxygen with the observed, a time series plot of the primary metabolic estimates (P, R, and D, as areal rates with units $\text{mmol}/\text{m}^3/\text{d}$), and a time series plot of the credible intervals for the $a$, $r$, and $b$ parameters.
-An additional option also allows the user to specify the model optimization period, where the time period can vary from a minimum of one day to the maximum of the entire length of the time series.
+An additional option allows the user to specify the model optimization period, where the time period can vary from a minimum of one day to the maximum of the entire length of the time series.
 This differs from the BASE approach where the optimization period is set at one day.
@@
 
-### EBASE comparison with modelled observations {#sec-cmpmod}
+### EBASE comparison with known results {#sec-cmpmod}
 
@@
 These assessments evaluated if 1) EBASE estimates similar values for the known parameters in the synthetic time series, and 2) how the estimated values change with different model configurations.
-The synthetic time series used inputs from an approximate one year of continuous data from Apalachicola Bay, Florida.
-The input data were 2012 observations at the Cat Point water quality monitoring station (29.7021$^{\circ}$ N, -84.8802$^{\circ}$ W) and included 15 minute observations of water temperature (C) and salinity (psu).
-Continuous meteorological observations from the nearby East Bay station (29.7909$^{\circ}$ N, -84.8834$^{\circ}$ W, \~10km north of Cat Point) were combined with the water quality data at the same time step and included air temperature (C), wind speed (m/s), and PAR (mmol/m$^2$ as total per 15 minute observation).
+The synthetic time series used inputs from one year of continuous data from Apalachicola Bay, Florida.
+The input data were 2012 observations at the Cat Point water quality monitoring station (29.7021$^{\circ}$ N, --84.8802$^{\circ}$ W) and included 15 minute observations of water temperature and salinity.
+Continuous meteorological observations from the nearby East Bay station (29.7909$^{\circ}$ N, --84.8834$^{\circ}$ W, \~10km north of Cat Point) were combined with the water quality data at the same time step and included air temperature, wind speed, and PAR.
 Missing observations in the water quality and meteorological data were filled using autoregressive modelling of the actual parameters versus time to create a complete dataset from January 1st to December 31st.
-An oxygen time series was then generated using the inputs from Apalachicola Bay ([@fig-synapa]a), such that the synthetic dissolved oxygen was created using @eq-ebase as a forward calculating model and user-defined values for $a$, $r$, and $b$.
+A synthetic oxygen time series was then generated using the inputs from Apalachicola Bay ([@fig-synapa]a) and @eq-ebase as a forward calculating model with user-defined values for $a$, $r$, and $b$.
 All metabolic rates for production, respiration, and gas exchange ([@fig-synapa]b and c) were also estimated with the synthetic time series.
@@
 First, results were compared to the known values after changing characteristics of their prior distributions.
-Second, results were compared using a model period of optimization for one day and seven days to assess how results varied based on the amount of data used for the Bayesian estimation.
+Second, results were compared using a model period of optimization for one, seven, and 30 days to assess how results varied based on the amount of data used for the Bayesian estimation.
 Details of these analyses are as follows.
@@
 A short period of optimization (e.g., one day) may return unstable parameter estimates if forcing factors (e.g., wind, temperature) that affect estuarine biogeochemical rates occur at time scales longer the chosen period of optimization for EBASE.
-As such, potentially longer optimization periods may return more stable and robust estimates given that more data are used for parameter estimation.
-An optimization period of one day was compared to results using an optimization period of seven days for the synthetic one-year time series.
-Each optimization period was evaluated for every unique combination of prior distributions, totaling 128 (64 times the two optimization periods) unique analyses for the assessment of changing priors of all parameters.
+As such, potentially longer optimization periods may return more stable and robust estimates given that more data are used for parameter estimation, although at the risk of overgeneralizing metabolic rates if variation occurs at time scales less than the optimization period.
+Results from optimization periods of one, seven, and thirty days were compared for the synthetic one-year time series.
+Each optimization period was also evaluated for every unique combination of prior distributions, totaling 192 (64 times the three optimization periods) unique analyses for the assessment of changing priors of all parameters.
 
-Changing the prior distributions affected the ability of EBASE to return the known parameters from the synthetic time series, with results varying somewhat by optimization period.
-@fig-priorcomp shows a summary of the comparisons of the synthetic data with EBASE for each unique combination of prior distributions and two optimization periods of one and seven days.
-The results are shown using Root Mean Squared Errors (RMSE) values comparing the synthetic estimates with those from EBASE.
-Darker colors represent higher RMSE values and can be used to assess how changing the priors affects the ability of EBASE to estimate the true parameter.
+Changing the prior distributions affected the ability of EBASE to return the known parameters from the synthetic time series, with results varying by optimization period.
+@fig-priorcomp shows a summary of the comparisons of the synthetic data with EBASE for each unique combination of prior distributions and the three optimization periods of one, seven, and thirty days.
+The results are shown using Nash-Sutcliffe Efficiency (NSE) values [@Nash70;@Moriasi07]:
+
+$$
+NSE = 1 - \frac{\sum\limits_{i=1}^n \left(Y_i^{Syn} - Y_i^{EBASE}\right)^2}{\sum\limits_{i=1}^n \left(Y_i^{Syn} - \bar{Y}^{Syn}\right)^2}
+$$ {#eq-nse}
+
+where $Y_i$ is the estimate from EBASE or the synthetic times series for the optimization period $i$, evaluated for each metabolic parameter, each unique combination of priors, and each optimization period.
+The value $n$ is the number of optimization periods in the one-year time series, i.e., $n=365$ for one day, $n \approx 52$ for seven days, and $n \approx 12$ for 30 days.
+All metabolic estimates were averaged within the number of days for each optimization period prior to calculating $NSE$ values, if appropriate.
+The $NSE$ value is conceptually similar to the coefficient of determination, but varies from $-\infty$ to 1.
+Values in the positive range from 0 to 1 are desirable with 1 being a perfect fit, whereas values in the negative range indicate the mean of the synthetic data is a better predictor than EBASE.
+
+<!-- edit -->
 
@@
 
+<!-- edit -->
 ### EBASE comparison with real observations
@@
 
--   Describe how EBASE deals with gaps
+-   Describe how EBASE deals with gaps and flexibility of the model to accommodate diff optimization periods
+-   b is sketchy, how to improve? Either fix or develop JAGS model further to include additional data, e.g. tidal height.
 

### 2023-03-22 R/funcs.R

@@ -204,7 +204,7 @@ priorcomp <- function(dat, met, topbot = 3){
     labs(
       y = NULL, 
       fill = parse(text = leglb),
-      x = 'Parameter from EBASE vs simulated,\nby optimization period'
+      x = 'Parameter from EBASE vs synthetic,\nby optimization period'
     )
   
   out <- p1 + p2 + plot_layout(ncol = 2, widths = c(0.45, 1))
@@ -304,7 +304,7 @@ optex <- function(apagrd, fwdatcmp, apasumdat, rnkmetsum, ndays, met, subttl, yl
     separate(var, c('var', 'mod'), sep = '\\.') %>%
     mutate(
       mod = case_when(
-        mod == 'x' ~ 'Fwoxy',
+        mod == 'x' ~ 'Synthetic',
         mod == 'y' ~ 'EBASE'
       )
     ) %>%

### 2023-03-23 R/funcs.R
working on nse heatmap plot #1

@@ -92,7 +92,7 @@ priorcomp <- function(dat, met, topbot = 3){
       met = c('r2', 'rmse', 'mape', 'mae', 'nse'),
       lbspr = c('italic(R)^2', 'RMSE', 'MAPE', 'mae', 'NSE'), 
       direc = c(-1, 1, 1, 1, -1),
-      limts = list(c(0, 100), NULL, NULL, NULL, c(0, 1)),
+      limts = list(c(0, 100), NULL, NULL, NULL, NULL),
       trans = c('identity', 'log10', 'log10', 'log10', 'identity')
     ) %>% 
     filter(met == !!met)
@@ -158,9 +158,6 @@ priorcomp <- function(dat, met, topbot = 3){
     ) %>% 
     left_join(optest, by = c('ind', 'ndays'), multiple= 'all')
 
-  if(met == 'nse')
-    toplo2$val <- pmax(toplo2$val, 0)
-
   p1 <- ggplot(toplo1, aes(y = ind, x = var, fill = val)) + 
     geom_tile(color = 'black') + 
     theme(
@@ -180,7 +177,7 @@ priorcomp <- function(dat, met, topbot = 3){
       y = NULL, 
       x = 'Prior parameters'
     )
-
+browser()
   p2 <- ggplot(toplo2, aes(y = ind, x = var, fill = val)) + 
     geom_tile(color = 'black') + 
     geom_text(aes(y = ind, x = 5.6, label = rnkmetsum), size = 2.35, hjust = 0, color = toplo2$cols, fontface = toplo2$fnts) + 
@@ -196,7 +193,7 @@ priorcomp <- function(dat, met, topbot = 3){
       panel.background = element_blank()
     ) + 
     facet_wrap(~ndays, ncol = 2) + 
-    scale_fill_distiller(palette = 'YlOrRd', direction = direc, limits = limts, trans = trans) + 
+    # scale_fill_distiller(palette = 'YlOrRd', direction = direc, limits = limts, trans = trans) + 
     scale_x_discrete(position = 'top', expand = c(0, 0), labels = parse(text = levels(toplo2$var))) + 
     scale_y_reverse(expand = c(0, 0)) + 
     coord_cartesian(clip = 'off') + 
@@ -207,6 +204,15 @@ priorcomp <- function(dat, met, topbot = 3){
       x = 'Parameter from EBASE vs synthetic,\nby optimization period'
     )
   
+  # special color palette for nse
[…]

### 2023-03-27 R/figs.R
fix priorcomp plot

@@ -105,8 +105,14 @@ load(file = fl7a)
 fl7b <- paste0(tempdir(), '/apagrd7b.RData')
 download.file('https://github.com/fawda123/BASEmetab_script/raw/master/data/apagrd7b.RData', destfile = fl7b)
 load(file = fl7b)
+fl30a <- paste0(tempdir(), '/apagrd30a.RData')
+download.file('https://github.com/fawda123/BASEmetab_script/raw/master/data/apagrd30a.RData', destfile = fl30a)
+load(file = fl30a)
+fl30b <- paste0(tempdir(), '/apagrd30b.RData')
+download.file('https://github.com/fawda123/BASEmetab_script/raw/master/data/apagrd30b.RData', destfile = fl30b)
+load(file = fl30b)
 
-apagrd <- bind_rows(apagrd1a, apagrd1b, apagrd7a, apagrd7b)
+apagrd <- bind_rows(apagrd1a, apagrd1b, apagrd7a, apagrd7b, apagrd30a, apagrd30b)
 
 # summary ests
 fl <- paste0(tempdir(), '/apasumdat.RData')
@@ -128,10 +134,12 @@ fwdatcmp <- fwdat %>%
   ) %>% 
   select(Date, DateTimeStamp, DO_obs, a, b, P, R, D)
 
-p1 <- optex(apagrd, fwdatcmp, apasumdat, rnkmetsum = 1, ndays = 1, met = 'cfd', subttl = '(a) Best priors, 1 day')
-p2 <- optex(apagrd, fwdatcmp, apasumdat, rnkmetsum = 64, ndays = 1, met = 'cfd', subttl = '(b) Worst priors, 1 day', ylbs = F)
-p3 <- optex(apagrd, fwdatcmp, apasumdat, rnkmetsum = 1, ndays = 7, met = 'cfd', subttl = '(c) Best priors, 7 day', ylbs = F)
-p4 <- optex(apagrd, fwdatcmp, apasumdat, rnkmetsum = 64, ndays = 7, met = 'cfd', subttl = '(d) Worst priors, 7 day', ylbs = F)
+p1 <- optex(apagrd, fwdatcmp, apasumdat, rnkmetsum = 1, ndays = 1, met = 'nse', subttl = '(a) Best priors, 1 day')
[…]

### 2023-03-27 R/funcs.R

@@ -109,20 +109,21 @@ priorcomp <- function(dat, met, topbot = 3){
     filter(!var %in% 'b') %>% 
     pivot_wider(names_from = 'var', values_from = !!toshw) %>% 
     mutate(
-      ind = rep(1: (nrow(.) / 2), times = 2), 
+      ind = rep(1: (nrow(.) / 3), times = 3), 
       ndays = case_when(
         ndays == 1 ~ paste(ndays, 'day'), 
         T ~ paste(ndays, 'days')
       )
     )
   
-  # ave r2 summary
+  # ave summary
   optest <- metsum_fun(dat, met) %>% 
     mutate(
       ndays = case_when(
         ndays == 1 ~ paste(ndays, 'day'), 
         T ~ paste(ndays, 'days')
-      )
+      ), 
+      ndays = factor(ndays, levels = unique(ndays))
     )
   
   topn <- sort(unique(optest$rnkmetsum))[1:topbot]
@@ -154,7 +155,8 @@ priorcomp <- function(dat, met, topbot = 3){
       var = factor(var, 
                    levels = c('DO_mod', 'P', 'R', 'D', 'a'), 
                    labels = c('DO', 'P', 'R', 'D', 'italic(a)')
-      )
+      ), 
+      ndays = factor(ndays, levels = unique(ndays))
     ) %>% 
     left_join(optest, by = c('ind', 'ndays'), multiple= 'all')
 
@@ -177,7 +179,13 @@ priorcomp <- function(dat, met, topbot = 3){
       y = NULL, 
       x = 'Prior parameters'
     )
-browser()
+
+  # if(met == 'nse')
+  #   toplo2 <- toplo2 %>% 
+  #     mutate(
+  #       val = pmax(val, -2)
+  #     )
+  
   p2 <- ggplot(toplo2, aes(y = ind, x = var, fill = val)) + 
     geom_tile(color = 'black') + 
     geom_text(aes(y = ind, x = 5.6, label = rnkmetsum), size = 2.35, hjust = 0, color = toplo2$cols, fontface = toplo2$fnts) + 
@@ -192,8 +200,7 @@ browser()
       panel.grid = element_blank(), 
       panel.background = element_blank()
     ) + 
-    facet_wrap(~ndays, ncol = 2) + 
[…]

### 2023-03-27 manu-draft.qmd
updating figures with nse results

@@
 The value $n$ is the number of optimization periods in the one-year time series, i.e., $n=365$ for one day, $n \approx 52$ for seven days, and $n \approx 12$ for 30 days.
-All metabolic estimates were averaged within the number of days for each optimization period prior to calculating $NSE$ values, if appropriate.
-The $NSE$ value is conceptually similar to the coefficient of determination, but varies from $-\infty$ to 1.
+All metabolic estimates were averaged within the number of days for each optimization period prior to calculating NSE values, if appropriate.
+The NSE value is conceptually similar to the coefficient of determination, but varies from $-\infty$ to 1.
 Values in the positive range from 0 to 1 are desirable with 1 being a perfect fit, whereas values in the negative range indicate the mean of the synthetic data is a better predictor than EBASE.
@@
 
-Several conclusions can be made from the results in @fig-priorcomp based on the patterns of the $R^2$ values.
+Several conclusions can be made from the results in @fig-priorcomp based on the patterns of the NSE values.
 Overall, the model was able to estimate the $a$ parameter with reasonable precision regardless of any of the prior distributions, whereas most of the variation was observed in the $r$ (respiration, R) and $b$ parameters.
@@
 ![Comparison of EBASE results to a synthetic time series created with known metabolic parameters.
-Results are compared for unique combination of prior distributions, where the means and standard deviations are evaluated from low (L) to high (H) for a one day and seven day model optimization period.
-The metabolic results for dissolved oxygen (DO), production (P), respiration (R), gas exchange (D), and the $a$ parameter are evaluated based on the $R^2$ value of the linear fit between the EBASE results and synthetic values.
+Results are compared for unique combination of prior distributions, where the means and standard deviations are evaluated from low (L) to high (H) for a one day, seven day, and thirty day model optimization period.
+The metabolic results for dissolved oxygen (DO), production (P), respiration (R), gas exchange (D), and the $a$ parameter are evaluated based on Nash-Sutcliffe Efficiency (NSE) values between the EBASE results and synthetic values.
+NSE values from zero to one (green) indicate EBASE has recovered the synthetic parameters with values closer to 1 being a better approximation, whereas negative values indicate the mean of the synthetic data is a better predictor than EBASE.
+The color scale for NSE is also exponentially transformed to minimize the influence of large negative values on the visual display.
 The $b$ parameter is a fixed constant and was not evaluated.
-Rankings of each combination as the average $R^2$ value across all parameters in a row are shown on the right, with the top three as black/bold and bottom three as black/italic.](figs/priorcomp.png){#fig-priorcomp}
+Rankings of each combination as the median NSE value across all parameters in a row are shown on the right, with the top three as black/bold and bottom three as black/italic.](figs/priorcomp.png){#fig-priorcomp}
 
@@
 
-![Add caption.](figs/priorsumcomp.png){#fig-priorsumcomp}
+![Summaries of median Nash-Sutcliffe Efficiency values across all parameters for the comparisons of EBASE and synthetic model outputs separated by different combinations of prior values.
+Each result on the x-axis represents the median values for which the combination prior applies, e.g., all results from @fig-priorcomp where $\mu$ is high or low for the $a$ parameter using a one day optimization period.
+Point size represents the inter-quartile range of median NSE values for each result.](figs/priorsumcomp.png){#fig-priorsumcomp}
 
@@
 ![Comparison of EBASE results for the best and worst performing models to a synthetic time series created with known metabolic parameters.
-The best and worst performing models for the one day (a, b) and seven day (c, d) optimization periods are based on the EBASE results with prior distributions that produced the highest mean $R^2$ values across all parameter estimates (@fig-priorcomp).
+The best and worst performing models for the one day (a, b), seven day (c, d), and thirty day (e, f) optimization periods are based on the EBASE results with prior distributions that produced the highest and lowest median Nash-Sutcliffe Efficenciy values across all parameter estimates (@fig-priorcomp).
 The results are averaged at the time step of the model optimization period.

### 2023-03-27 R/figs.R

@@ -86,7 +86,7 @@ load(file = fl)
 
 p1 <- priorsumcomp(apasumdat, met = 'nse')
 
-png(here('figs/priorsumcomp.png'), height = 6, width = 5, family = 'serif', units = 'in', res = 500)
+png(here('figs/priorsumcomp.png'), height = 5, width = 6, family = 'serif', units = 'in', res = 500)
 print(p1)
 dev.off()
 
@@ -134,21 +134,21 @@ fwdatcmp <- fwdat %>%
   ) %>% 
   select(Date, DateTimeStamp, DO_obs, a, b, P, R, D)
 
-p1 <- optex(apagrd, fwdatcmp, apasumdat, rnkmetsum = 1, ndays = 1, met = 'nse', subttl = '(a) Best priors, 1 day')
-p2 <- optex(apagrd, fwdatcmp, apasumdat, rnkmetsum = 64, ndays = 1, met = 'nse', subttl = '(b) Worst priors, 1 day', ylbs = F)
-p3 <- optex(apagrd, fwdatcmp, apasumdat, rnkmetsum = 1, ndays = 7, met = 'nse', subttl = '(c) Best priors, 7 day', ylbs = F)
-p4 <- optex(apagrd, fwdatcmp, apasumdat, rnkmetsum = 64, ndays = 7, met = 'nse', subttl = '(d) Worst priors, 7 day', ylbs = F)
-p5 <- optex(apagrd, fwdatcmp, apasumdat, rnkmetsum = 1, ndays = 30, met = 'nse', subttl = '(e) Best priors, 30 day', ylbs = F)
-p6 <- optex(apagrd, fwdatcmp, apasumdat, rnkmetsum = 64, ndays = 30, met = 'nse', subttl = '(f) Worst priors, 30 day', ylbs = F)
+p1 <- optex(apagrd, fwdatcmp, apasumdat, rnkmetsum = 1, ndays = 1, met = 'nse', subttl = '(a) Best, 1 day')
+p2 <- optex(apagrd, fwdatcmp, apasumdat, rnkmetsum = 64, ndays = 1, met = 'nse', subttl = '(b) Worst, 1 day', ylbs = F)
+p3 <- optex(apagrd, fwdatcmp, apasumdat, rnkmetsum = 1, ndays = 7, met = 'nse', subttl = '(c) Best, 7 day', ylbs = F)
[…]

### 2023-03-27 R/funcs.R

@@ -166,7 +166,6 @@ priorcomp <- function(dat, met, topbot = 3){
       axis.text.x = element_text(size = 9, angle = 45, hjust = 0), 
       axis.ticks = element_blank(), 
       legend.position = 'left', 
-      legend.title = element_blank(),
       axis.text.y = element_blank(), 
       strip.placement = 'outside', 
       strip.background = element_blank()
@@ -177,7 +176,8 @@ priorcomp <- function(dat, met, topbot = 3){
     scale_y_reverse(expand = c(0, 0)) + 
     labs(
       y = NULL, 
-      x = 'Prior parameters'
+      x = 'Prior parameters', 
+      fill = 'Prior\nvalue'
     )
 
   # if(met == 'nse')
@@ -190,7 +190,7 @@ priorcomp <- function(dat, met, topbot = 3){
     geom_tile(color = 'black') + 
     geom_text(aes(y = ind, x = 5.6, label = rnkmetsum), size = 2.35, hjust = 0, color = toplo2$cols, fontface = toplo2$fnts) + 
     theme(
-      axis.text.x = element_text(face = 'italic', size = 12), 
+      axis.text.x = element_text(face = 'italic', size = 8), 
       axis.text.y = element_blank(),
       axis.ticks = element_blank(), 
       legend.position = 'right', 
@@ -252,21 +252,21 @@ priorsumcomp <- function(dat, met = 'r2'){
       param = factor(param, levels = c('a', 'r', 'b')),
       prior = gsub('^a|^r|^b', '', prior), 
       prior = factor(prior, levels = c('mean', 'sd'), labels = c('mu', 'sigma')), 
-      ndays = factor(ndays, levels = c('1 day', '7 days'), labels = c('1~day', '7~days'))
+      ndays = factor(ndays, levels = c('1 day', '7 days', '30 days'), labels = c('1~day', '7~days', '30~days'))
     )
   
   metsel <- tibble(
     met = c('r2', 'rmse', 'mape', 'mae', 'nse'),
-    lbspr = c('italic(R)^2', 'RMSE', 'MAPE', 'mae', 'Coef.~of~Det.')
+    lbspr = c('italic(R)^2', 'RMSE', 'MAPE', 'mae', 'NSE')
   ) %>% 
     filter(met == !!met)
   ylab <- parse(text = paste0('Median~', metsel$lbspr))
-  
+
   p <- ggplot(toplo, aes(x = param, group = val)) + 
[…]

### 2023-03-27 manu-draft.qmd
fix ebaseiotab using equatabs

@@
 library(flextable)
+library(dplyr)
+
+source(here('R/funcs.R'))
 
@@
 load(file = here('tabs/apacmptab.RData'))
+load(file = here('data/apasumdat.RData'))
 ```
@@
 <!-- edit -->
+```{r}
+# percent negative nse not do
+nseall <- apasumdat %>% 
+  tidyr::unnest('ests') %>% 
+  filter(var != 'DO_mod') %>% 
+  na.omit %>% 
+  pull(nse)
+perneg <- (sum(nseall < 0) / length(nseall) * 100) %>% 
+  round() %>% 
+  as.character() %>% 
+  paste0('%')
 
-Several conclusions can be made from the results in @fig-priorcomp based on the patterns of the NSE values.
-Overall, the model was able to estimate the $a$ parameter with reasonable precision regardless of any of the prior distributions, whereas most of the variation was observed in the $r$ (respiration, R) and $b$ parameters.
-Changing the variance of the $a$ parameter had little effect on any of the comparisons.
-Smaller variance of the prior distribution for the $r$ parameter generally contributed to lower $R^2$ for all parameters, although the precision was worse for the seven day optimization period (e.g., the $a$ parameter).
-Conversely, larger variance of the prior distribution for the $b$ parameter generally contributed to lower $R^2$ only for the gas exchange estimate (D), although this effect was minimized with the seven day optimization period.
+# gasex nse
+prmall <- apasumdat %>% 
+  tidyr::unnest('ests') %>% 
+  filter(var != 'DO_mod') %>% 
+  na.omit %>% 
+  summarise(
+    nse = round(median(nse), 2),
+    .by = var
+  ) %>% 
+  tibble::deframe()
 
-The results suggest that the default prior distributions used in EBASE may be appropriate in some cases, but increasing the prior variance for $a$ and $r$ and decreasing the variance for $b$ will likely produce more optimal solutions.
-The actual model results for the best performing combination of prior distributions for the one day and seven day optimization are shown in @fig-optex (all results are averaged at the time-step of the optimization period for comparability).
-The best performing models were those with the highest mean $R^2$ values across all parameter estimates in @fig-priorcomp.
-For both models, the metabolic parameters were estimated with high precision, and in some cases with nearly perfect precision (i.e., dissolved oxygen and gas exchange, D).
-Estimates for the fixed $b$ parameter were most variable, but not biased as the mean for both models was close to 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$.
+# optimization nse
+optnse <- apasumdat %>% 
+  tidyr::unnest('ests') %>% 
+  na.omit() %>% 
+  filter(var != 'DO_mod') %>% 
+  summarise(
+    nse = round(median(nse), 2), 
+    .by = ndays
+  ) %>% 
+  tibble::deframe()
 
-<!-- edit -->
+# best mods by opt
+optmod <- metsum_fun(apasumdat, 'nse', parms = T) %>% 
+  filter(rnkmetsum %in% 1) %>% 
+  summarise(
+    metsum = round(median(metsum), 2),
+    .by  = 'ndays'
+  ) %>% 
+  tibble::deframe()
+```
+
+Several conclusions can be made from the results in @fig-priorcomp following the patterns of the NSE values.
+Overall, most models regardless of optimization period or priors were able to reproduce the dissolved oxygen time series.
+The only models where the NSE values were less than zero for dissolved oxygen were those where $\rho$ of the $r$ parameter was low.
+However, substantial variation in NSE values was observed among all other parameters.
+Not considering dissolved oxygen, gas exchange ($D$) was estimated with the highest NSE, with a median value across all optimization periods and priors slightly larger than zero (median NSE = `r prmall[['D']]`).
+Conversely, the $a$ parameter was reproduced the most poorly, with a median value across all optimization periods and priors less than zero (median NSE = `r prmall[['a']]`).
+By optimization period, median NSE values across all parameters decreased slightly as the number of days used for model optimization increased (1 day = `r optnse[['1']]`, 7 days = `r optnse[['7']]`, 30 days = `r optnse[['30']]`).
+The top three ranked models were identical for the seven and thirty day models and similar for models using the one day optimization period.
+Median NSE values across parameters for the top model in each optimization period were also similar (1 day = `r optmod[['1']]`, 7 days = `r optmod[['7']]`, 30 days = `r optmod[['30']]`).
+For the seven day and thirty day optimization periods, the NSE values for all parameters were greater than zero for the top models, where two of the parameters in the top model for the one day optimization period has NSE < 0.
+Most parameters regardless of optimization period or priors had NSE < 0 (`r perneg`).
+As such, the prior combinations that produced the highest NSE values across all parameters were those where the $\rho$ values were high for each prior distribution, regardless of the values for $\mu$, i.e., those with sufficient flexibility to search a larger parameter space regardless of $\mu$ produced metabolic parameter estimates most similar to the known values from the synthetic time series.
+
+@fig-priorsumcomp provides an alternative summary of how the prior distributions influenced the ability of EBASE to reproduce the synthetic parameters.
+Median NSE values across all models and each optimization period for the different subsets of the prior combinations are shown.
+For example, the bottom left panel with the x-axis value for the $a$ parameter shows the median NSE values for all models using a one-day optimization period where the $\rho$ value for the $a$ prior distribution was low or high.
+The size of the points indicate the inter-quartile range of NSE values for each subset.
+As such, each set of points denoted by the x-axis parameters in each panel include median NSE values for all models by optimization period, but subset by the low/high $\mu$ or $\rho$ values.
+Although most most of the median value are less than zero, indicating a poor ability of EBASE to reproduce the parameters, patterns emerge as to which characteristics of the prior lead to robust estimates.
+For example, large differences in the median NSE values are observed for the low and high values of $\rho$ for the $a$ and $r$ parameters, with positive median NSE values and smaller IQR values using a high value.
+A similar, but smaller effect is seen for the $\rho$ value of the prior for the $b$ parameter.
+A conclusion similar to that for @fig-priorcomp is that large standard deviations in the prior distributions lead to more accurate estimates of metabolic parameters.
+Conversely, model results are less senstive to low or high values for $\mu$ as the spread between median NSE values is less than those for $\rho$.
+This suggests that EBASE model sensitivity is most influenced by $\rho$ of the prior distributions, i.e., larger values produce more accurate and less variable estimates regardless of $\mu$.
+
+The actual model results for the best and worst performing combinations of prior distributions for the one day, seven day, and thirty day optimization periods are shown in @fig-optex (all results are averaged at the time-step of the optimization period for comparability).
+The best performing models were those with the highest median NSE values across all parameter estimates in @fig-priorcomp.
+The best performing models for each optimization period (@fig-optex a, c, e) showed similar results where the parameters estimates were similar to the synthetic time series, although more variation in the EBASE results were of course observed with the one day optimization period.
+However, the $b$ parameter was estimated with high variability for the one and seven day optimization periods (@fig-optex a, c, bottom row) and with a large positive bias for the 30 day optimization period (@fig-optex e, bottom row).
+These results were not unexpected because it was impossible to evaluate NSE estimates using a fixed value in the synthetic results and the accuracy was not considered in identifying the best performing model.
+Results using EBASE default priors (medium values for $\mu$, high values for $\rho$) showed that the $b$ parameter can be estimated with reasonable accuracy, although still with considerable variability.
+Finally, the worst performing model for each optimization period (@fig-optex b, d, f) demonstrates the negative NSE values that were obtained (i.e., EBASE values much lower or higher than the synthetic values).
+The worst performing models were caused by constraining EBASE to a small parameter space using low or high values for $\mu$ combined with low values for $\rho$.
+
 ### EBASE comparison with real observations
@@
 Synthetic input to EBASE was the (a) dissolved oxygen time series, with results compared to the (b) synthetic $a$ parameter and (c) synthetic metabolic estimates for production (P), respiration (R), and gas exchange (D).
-The $a$ parameter is shown as an areal value for comparison with the metabolic results.](figs/synapa.png){#fig-synapa}
+The $a$ parameter is shown as an areal value for comparison with the metabolic results.
+The $b$ parameter is also fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ and is not shown.](figs/synapa.png){#fig-synapa}
 
@@
 ![Comparison of EBASE results to a synthetic time series created with known metabolic parameters.
-Results are compared for unique combination of prior distributions, where the means and standard deviations are evaluated from low (L) to high (H) for a one day, seven day, and thirty day model optimization period.
+Results are compared for unique combinations of prior distributions and optimization period.
+The means and standard deviations are evaluated as low (L) or high (H) values (see text for details) and the optimization period is one day, seven days, or thirty days.
 The metabolic results for dissolved oxygen (DO), production (P), respiration (R), gas exchange (D), and the $a$ parameter are evaluated based on Nash-Sutcliffe Efficiency (NSE) values between the EBASE results and synthetic values.

### 2023-03-27 R/dat_proc.R

@@ -1,0 +1,7 @@
+library(tidyverse)
+library(here)
+
+# apasumdat ---------------------------------------------------------------
+
+fl <- here('data/apasumdat.RData')
+download.file('https://github.com/fawda123/BASEmetab_script/raw/master/data/apasumdat.RData', destfile = fl)

### 2023-03-27 R/figs.R

@@ -143,12 +143,11 @@ p6 <- optex(apagrd, fwdatcmp, apasumdat, rnkmetsum = 64, ndays = 30, met = 'nse'
 
 p <- ((p1 + plot_layout(ncol = 1)) | (p2 + plot_layout(ncol = 1)) | (p3 + plot_layout(ncol = 1)) | (p4 + plot_layout(ncol = 1))| (p5 + plot_layout(ncol = 1)) | (p6 + plot_layout(ncol = 1)))  + plot_layout(ncol = 6, guides = 'collect') & 
   theme(
-    legend.position = 'top', 
-    axis.text.x = element_text(size = 8)
+    legend.position = 'top'
   ) & 
   scale_x_date(date_labels = '%b')
 
-png(here('figs/optex.png'), height = 7.5, width = 9, family = 'serif', units = 'in', res = 500)
+png(here('figs/optex.png'), height = 8.7, width = 9, family = 'serif', units = 'in', res = 500)
 print(p)
 dev.off()
 

### 2023-03-27 R/funcs.R

@@ -396,6 +396,7 @@ optex <- function(apagrd, fwdatcmp, apasumdat, rnkmetsum, ndays, met, subttl, yl
       strip.background = element_blank(), 
       legend.position = 'top', 
       legend.title = element_blank(), 
+      axis.text.x = element_text(size = 8),
       strip.text = element_text(size = rel(0.7))
     ) + 
     labs(
@@ -429,7 +430,7 @@ metsum_fun <- function(apasumdat, met, parms = F){
     mutate(
       ind = rep(1: (nrow(.) / 3), times = 3)
     ) %>% 
-    unnest('ests') %>% 
+    tidyr::unnest('ests') %>% 
     rename(met = !!met) %>% 
     summarise(
       metsum = median(met, na.rm = T), 

### 2023-03-29 manu-draft.qmd
almost done with rgn edits

@@
 The $b$ parameter is estimated from the data and has units of $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$.
-As such, the three parameters estimated by EBASE are $a$, $r$, and $b$ using the required input data that includes time series of dissolved oxygen, water temperature, salinity, PAR, wind speed, and water column depth (`r run_reference(id = "ebaseiotab")`).
+As such, the three parameters estimated by EBASE are $a$, $r$, and $b$ using the required input data that includes time series of dissolved oxygen, water temperature, salinity, PAR, wind speed, and water column depth (`r run_reference(id = "tbl-ebaseiotab")`).
 
@@
 The estimated dissolved oxygen concentration is also returned, which can be compared to the observed values as one measure of model performance.
-All parameter inputs and outputs for EBASE are shown in `r run_reference(id = "ebaseiotab")`.
+All parameter inputs and outputs for EBASE are shown in `r run_reference(id = "tbl-ebaseiotab")`.
 Areal rates for $P$, $R$, and $D$ (mmol m$^-2$ d$^{-1}$) are returned for comparability of estimates across locations where water column depth may vary.
@@
 
-<!-- edit -->
 ```{r}
@@
 These results were not unexpected because it was impossible to evaluate NSE estimates using a fixed value in the synthetic results and the accuracy was not considered in identifying the best performing model.
-Results using EBASE default priors (medium values for $\mu$, high values for $\rho$) showed that the $b$ parameter can be estimated with reasonable accuracy, although still with considerable variability.
 Finally, the worst performing model for each optimization period (@fig-optex b, d, f) demonstrates the negative NSE values that were obtained (i.e., EBASE values much lower or higher than the synthetic values).
@@
 
-### EBASE comparison with real observations
+### EBASE and Odum comparison with real observations
 
 Metabolic estimates from EBASE were further compared to those from existing methods to benchmark and evaluate potential differences from alternative model formulations used by each approach.
-The observed one year continuous time series described above for water quality and weather data from Apalachicola Bay was used to compare metabolic estimates from EBASE, BASE, and the Odum open-water method [@Odum56].
+The observed one-year continuous time series described above for water quality and weather data from Apalachicola Bay was used to compare metabolic estimates from EBASE and the Odum open-water method [@Odum56].
 The Odum method used an approach described in @Murrell18 created for estuarine application that uses a gas-exchange approach described in @Thebault08.
-For better comparison with the EBASE methods, the gas exchange parmaterization in the Odum method was modified to use the @Wanninkhof14 approach [as available in the WtRegDO R package, @Beck21].
-The input data were provided to each method after converting the units as required (e.g., BASE requires PAR as $\mu$mol/m$^{2}$/s).
-Water column depth at Cat Point is approximately 1.8 m and all daily metabolic estimates returned by each method were converted to volumetric rates by dividing by the depth (i.e., $\text{mmol}/\text{m}^3/\text{d}$).
-Each comparison also evaluated metabolic estimates using the observed dissolved oxygen time series and a detided time series using weighted regression [@Beck15; available from @Beck21].
-The latter comparison provided an assessment of metabolic rates that are expected to be minimally influenced by tidal advection.
-All metabolic estimates were obtained using the default settings for each of the three methods, with an exception that the prior distribution for the reaeration coefficient ($K$) in BASE was informed by results from the Odum output.
-Initial assessments indicated that unstable and inaccurate results for the time series are obtained from BASE using an uninformed prior distribution for $K$.
-All other prior distributions for BASE and EBASE were set as the software default.
-A one day model optimization period was used for EBASE for comparability with BASE.
+For better comparison with the EBASE methods, the gas exchange parameterization in the Odum method was modified to use the @Wanninkhof14 approach, the the $b$ parameter fixed at 0.251 $\buni$ [as available in the WtRegDO R package, @Beck21].
+An additional and fundamental difference between EBASE and the Odum method is the former is likelihood based with metabolic rates estimated by fitting the model to the data, whereas the latter is strictly arithmetic based primarily on integration of the diel dissolved oxygen curve.
+Thus, potentially different results are expected between the approaches based on differences in formulation of the mass balance equations for metabolism as well as alternative statistical approaches to estimating the parameters.
+Each comparison also evaluated differneces in the metabolic estimates using the observed dissolved oxygen time series and a detided time series using weighted regression [@Beck15; available from @Beck21].
+The latter comparison provided an assessment of metabolic rates that are expected to be minimally influenced by tidal advection, as a fundamental assumption of metabolic models using *in situ* dissolved oxygen time series from fixed monitoring stations.
+The prior distributions for EBASE were uninformed using the same large $\rho$ values from @sec-cmpmod and moderate $\mu$ values ($\mu$ = 0.2 $\auni$ for $a$, $\mu$ = 20 $\runi$ for $r$, $\mu$ = 0.251 $\buni$ for $b$).
+A one day model optimization period was used for EBASE for comparability with the Odum method.
 
-The common set of metabolic parameters returned by the Odum, BASE, and EBASE methods are net ecosystem metabolism, gross production, respiration, and gas exchange.
+The common set of metabolic parameters returned by the EBASE and Odum methods are net ecosystem metabolism, gross production, respiration, and gas exchange.
 Pairwise comparisons of the daily estimates returned by each method were evaluated using simple summary statistics, including Pearson correlation coefficients and linear regression fits to evaluate potential biases in each method.
-For the latter analysis, differences in intercept estimates less than or greater than zero suggested a bias in magnitude, whereas differences in slope estimates less than or greater than one suggested a bias that depended on relative value of each observation (e.g., bias may differ between lower or higher values).
+For the latter analysis, differences in intercept estimates less than or greater than zero suggested a bias in magnitude, whereas differences in slope estimates less than or greater than one suggested a bias that depended on relative value of each observation (e.g., bias may differ between lower or higher metabolic estimates).
 Comparisons of the intercept and slope estimates to zero and one, respectively, were based on confidence intervals from the standard errors of each.
+Root mean square errors (RMSE) of each linear model were also evaluated.
 Results were assessed separately for estimates based on the observed or detided dissolved oxygen time series.
 
-@fig-apacmpfig and `r run_reference(id = "apacmptab")` provide a comparison of the EBASE metabolic estimates with those from the Odum and BASE methods.
+<!-- edit -->
+
+@fig-apacmpfig and `r run_reference(id = "tbl-apacmptab")` provide a comparison of the metabolic estimates from the EBASE and Odum methods.
 Correlations for all estimates, regardless of method or the input time series, were positive and significantly correlated, as expected.
-The only exception was the comparison of respiration for the Odum and EBASE methods ($\rho$ = 0.11), where the former produced a much larger range of estimates.
-The intercept values were all significant, suggesting some level of bias when comparing estimates between the methods.
-Slope estimates were all significantly different from one for the comparisons using the observed dissolved oxygen time series as input, whereas only three of the eight comparisons were significantly different using the detided time series.
+The intercept values were all significant, excluding production and respiration with the observed dissolved oxygen, suggesting some level of bias when comparing estimates between the methods.
+Half of the slope estimates were significantly different for estimates from both the observed and detided dissolved oxygen, with production and gas exchange differing for the observed input and net metabolism and gas exchange differing for the detided input.
+The RMSE of each regression comparison were generally lower in the detided results for all comparisons.
 This latter comparison suggests that detiding the time series produces more comparable estimates between methods.
-Similarly, the standard deviation ($\rho$) of the residuals of each regression comparison were generally lower in the detided results for all comparisons.
-Variation in the gas exchange estimates were also more similar comparing the Odum and EBASE methods that used the same formulation [@Wanninkhof14] regardless of input time series.
 
@@
 These effects are easily identified with estimates from the Odum method as negative values for production (P) and positive values for respiration (R), i.e., "anomalous" values.
-The effects of tidal advection on estimates from EBASE and BASE are more subtle as anomalous values cannot be returned based on constraints in the mass-balance equations (e.g., it is impossible for respiration to be positive).
-However, the effects of tidal advection produced a wider range of estimates for NEM in EBASE and BASE compared to the Odum method ([@fig-apacmpfig]a), caused by greater variation in the production and respiration estimates.
-Overall, detiding the dissolved oxygen time series prior to estimating metabolic rates produces more stable (lower standard deviation) and smaller ranges (`r run_reference(id = "apacmptab")`), regardless of the method used.
+The effects of tidal advection on estimates from EBASE were more subtle as anomalous values cannot be returned based on constraints of the priors (i.e., $a$ and $r$ cannot be negative, Equations [-@eq-aprior] and [-@eq-rprior]).
+Additionally, the effects of tidal advection produced a wider range of estimates for NEM in both methods, whereas the detided results were mostly negative indicating that conclusions about net autotrophy or heterotrophy may be inaccurate using observed dissolved oxygen as input.
+Overall, detiding the dissolved oxygen time series prior to estimating metabolic rates produces more stable (lower standard deviation) and smaller ranges (`r run_reference(id = "tbl-apacmptab")`), regardless of the method used.
+
+<!-- edit -->
 
@@
 
-![The default prior distributions for the $a$, $r$, and $b$ parameters used in EBASE.
+![The default prior distributions for $a$, $r$, and $b$ used in EBASE.
 All priors are normal distributions and truncated to positive values.
-The $b$ parameter is also truncated to a maximum of 0.504.](figs/priorplot.png){#fig-priorplot fig-align="center"}
+$b$ is also truncated to a maximum of 0.504.](figs/priorplot.png){#fig-priorplot fig-align="center"}
 
@@
 
-![Comparisons of daily metabolic estimates from the Odum, BASE, and EBASE methods for one year of continuous monitoring data at Apalachicola Bay.
+![Comparisons of daily metabolic estimates from the EBASE and Odum methods for one year of continuous monitoring data at Apalachicola Bay.
 Results for net ecosystem metabolism (NEM), gross production (P), respiration (R), and gas exchange (D) are compared using the (a) observed and (b) detided dissolved oxygen time series as input to each model.
-All values are volumetric as mmol O$_2$ m$^3$ d$^{-1}$.
+All values are areal as mmol O$_2$ m$^2$ d$^{-1}$.
 Black lines show the 1:1 comparison and the red lines show the linear regression fit.](figs/apacmpfig.png){#fig-apacmpfig fig-align="center"}
@@
 ```{r}
-#| label: ebaseiotab
+#| label: tbl-ebaseiotab
 #| tbl-cap: "EBASE model parameters showing the required input, parameters derived from the model input, and model output.
@@
 ```{r}
-#| label: apacmptab
-#| tbl-cap: "Summary statisics of daily metabolic estimates comparing Odum and BASE to EBASE for net ecosystem metabolism (NEM), gross production (P), respiration (R), and gas exchange (D).
+#| label: tbl-apacmptab
+#| tbl-cap: "Summary statistics of daily metabolic estimates comparing the EBASE and Odum methods for net ecosystem metabolism (NEM), gross primary production (P), respiration (R), and gas exchange (D).
 Separate comparisons are made for estimates based on observed and detided dissolved oxygen.
-Summary statistics include Pearson correlation coefficients and intercept, slope, and residual standard deviation of a linear fit between the same metabolic estimates obtained from each method.
+Summary statistics include Pearson correlation coefficients and intercept, slope, and root mean square error (RMSE) of a linear fit between the same metabolic estimates obtained from each method.
 Significance of the p-values for the intercept and slope indicates if the estimate is different from zero or one, respectively."
@@
 
-<!-- {{< pagebreak >}} -->
+{{< pagebreak >}}
 

### 2023-03-29 R/figs.R

@@ -8,16 +8,6 @@ library(scales)
 
 source(here('R/funcs.R'))
 
-# prior distribution plot ---------------------------------------------------------------------
-
-p <- prior_plot(n = 1e5)
-
-png(here('figs/priorplot.png'), height = 3, width = 8, family = 'serif', units = 'in', res = 500)
-print(p)
-dev.off()
-
-# simulated apa data --------------------------------------------------------------------------
-
 fwdat <- read_csv(file = url('https://raw.githubusercontent.com/fawda123/BASEmetab_script/master/data/apafwoxy2.csv'))
 
 fwdatcmp <- fwdat %>% 
@@ -32,6 +22,16 @@ fwdatcmp <- fwdat %>%
     b = 100 * 3600 * `kw,m/s` / `wspd2,m2/s2` / (`sc,dimensionless` / 660) ^ -0.5 # (m/s)/(m2/s2) to (cm/hr) / (m2/s2)
   ) %>% 
   select(Date, DateTimeStamp, DO_obs, a, b, P, R, D)
+
+# prior distribution plot ---------------------------------------------------------------------
+
+p <- prior_plot(n = 2e6)
+
+png(here('figs/priorplot.png'), height = 3, width = 8, family = 'serif', units = 'in', res = 500)
+print(p)
+dev.off()
+
+# simulated apa data --------------------------------------------------------------------------
 
 p1 <- ggplot(fwdatcmp, aes(x = DateTimeStamp, y = DO_obs)) + 
   geom_line(linewidth = 0.2) + 
@@ -119,21 +119,6 @@ fl <- paste0(tempdir(), '/apasumdat.RData')
 download.file('https://github.com/fawda123/BASEmetab_script/raw/master/data/apasumdat.RData', destfile = fl)
 load(file = fl)
 
-fwdat <- read_csv(file = url('https://raw.githubusercontent.com/fawda123/BASEmetab_script/master/data/apafwoxy2.csv'))
-
-fwdatcmp <- fwdat %>% 
-  mutate(
-    DateTimeStamp = dmy_hms(datet, tz = 'America/Jamaica'),
-    Date = as.Date(DateTimeStamp, tz = 'America/Jamaica'),
[…]

### 2023-03-29 R/funcs.R

@@ -286,24 +286,8 @@ priorsumcomp <- function(dat, met = 'r2'){
   
 }
 
-# comparison of fwoxy and ebase results for ranked model by prior at opt of ndays
-optex <- function(apagrd, fwdatcmp, apasumdat, rnkmetsum, ndays, met, subttl, ylbs = TRUE){
-
-  # find the prior comparison
-  rnkfnd <- metsum_fun(apasumdat, met, parms = T) %>% 
-    filter(rnkmetsum == !!rnkmetsum) %>% 
-    filter(ndays == !!ndays)
-
-  res <- apagrd %>% 
-    filter(
-      amean == rnkfnd$amean & asd == rnkfnd$asd & 
-      rmean == rnkfnd$rmean & rsd == rnkfnd$rsd & 
-      bmean == rnkfnd$bmean & bsd == rnkfnd$bsd & 
-      ndays == rnkfnd$ndays
-    ) %>% 
-    pull(out) %>% 
-    .[[1]] %>% 
-    .[[1]]
+# plot for comparing Fwoxy and EBASE, used in optext and defex
+cmpplo <- function(res, fwdatcmp, subttl, ylbs = TRUE){
   
   cmp <- inner_join(fwdatcmp, res, by = c('Date', 'DateTimeStamp')) %>%
     mutate(
@@ -412,8 +396,50 @@ optex <- function(apagrd, fwdatcmp, apasumdat, rnkmetsum, ndays, met, subttl, yl
       theme(strip.text = element_blank())
     
   }
-    
+  
   p1 + p2
+  
+}
+
+# comparison of fwoxy and ebase results for ranked model by prior at opt of ndays
+optex <- function(apagrd, fwdatcmp, apasumdat, rnkmetsum, ndays, met, subttl, ylbs = TRUE){
+
+  # find the prior comparison
+  rnkfnd <- metsum_fun(apasumdat, met, parms = T) %>% 
+    filter(rnkmetsum == !!rnkmetsum) %>% 
+    filter(ndays == !!ndays)
+
+  res <- apagrd %>% 
+    filter(
+      amean == rnkfnd$amean & asd == rnkfnd$asd & 
+      rmean == rnkfnd$rmean & rsd == rnkfnd$rsd & 
+      bmean == rnkfnd$bmean & bsd == rnkfnd$bsd & 
+      ndays == rnkfnd$ndays
+    ) %>% 
+    pull(out) %>% 
[…]

### 2023-03-29 R/tabs.R

@@ -71,13 +71,7 @@ load(file = url('https://github.com/fawda123/BASEmetab_script/raw/master/data/ap
 tosum <- apacmp %>% 
   mutate(
     typ = ifelse(typ == 'BASEmetab', 'BASE', typ),
-    val = case_when(
-      var == 'Rt_vol' ~ -1 * val, 
-      T ~ val
-    ), 
-    var = factor(var, 
-                 levels = c('NEM', 'Pg_vol', 'Rt_vol', 'D'), 
-                 labels = c('NEM', 'P', 'R', 'D')
+    var = factor(var, levels = c('NEM', 'P', 'R', 'D'), 
     )
   ) %>% 
   pivot_wider(names_from = 'typ', values_from = 'val')
@@ -93,7 +87,8 @@ grd <- crossing(
     slo = NA, 
     slolo = NA,
     slohi = NA,
-    rse = NA
+    rse = NA,
+    rmse = NA
   )
 
 for(i in 1:nrow(grd)){
@@ -120,7 +115,7 @@ for(i in 1:nrow(grd)){
   # get summaries
   corv <- cor.test(tocmp$xval, tocmp$yval)
   corv <- paste0(round(corv$estimate, 2), p_ast(corv$p.value))
-
+ 
   lmmod <- lm(yval ~ xval, data = tocmp)
   lmmodsum <- summary(lmmod)
   tval <- (abs(lmmodsum$coefficients[2, 1] - 1)) / (lmmodsum$coefficients[2, 2]) 
@@ -130,6 +125,7 @@ for(i in 1:nrow(grd)){
   slolo <- confint(lmmod)[2, 1]
   slohi <- confint(lmmod)[2, 2]
   rse <- round(sigma(lmmod), 2)
+  rmse <- round(modelr::rmse(lmmod, tocmp), 2)
   
   # append to output
   grd[i, 'corv'] <- corv
@@ -138,17 +134,19 @@ for(i in 1:nrow(grd)){
   grd[i, 'slolo'] <- slolo
   grd[i, 'slohi'] <- slohi
   grd[i, 'rse'] <- rse
+  grd[i, 'rmse'] <- rmse
   
 }
 
 totab <- grd %>% 
-  select(dotyp, comp, var, corv, int, slo, rse) %>% 
+  select(dotyp, comp, var, corv, int, slo, rmse) %>% 
   mutate(
     dotyp = factor(dotyp, levels = c('observed', 'detided'), labels = c('Observed', 'Detided')), 
[…]

### 2023-04-18 manu-draft.qmd
minor text edit

@@
 The size of the points indicate the inter-quartile range of NSE values for each subset.
-As such, each set of points denoted by the x-axis parameters in each panel include median NSE values for all models by optimization period, but subset by the low/high $\mu$ or $\rho$ values.
-Although most most of the median value are less than zero, indicating a poor ability of EBASE to reproduce the parameters, patterns emerge as to which characteristics of the prior lead to robust estimates.
-For example, large differences in the median NSE values are observed for the low and high values of $\rho$ for the $a$ and $r$ parameters, with positive median NSE values and smaller IQR values using a high value.
+As such, each set of points denoted by the x-axis parameters in each panel include median NSE values for all models by optimization period, but subset by the low/high $\mu$ or $\rho$ values for each parameter.
+Although most of the median value are less than zero, indicating a poor ability of EBASE to reproduce the parameters, patterns emerge as to which characteristics of the prior lead to robust estimates.
+For example, large differences in the median NSE values are observed for the low and high values of $\rho$ for the $a$ and $r$ parameters, with positive median NSE values and smaller IQR values using a high value using the 7 and 30 day optimization periods.
 A similar, but smaller effect is seen for the $\rho$ value of the prior for the $b$ parameter.
 A conclusion similar to that for @fig-priorcomp is that large standard deviations in the prior distributions lead to more accurate estimates of metabolic parameters.
-Conversely, model results are less senstive to low or high values for $\mu$ as the spread between median NSE values is less than those for $\rho$.
+Conversely, model results are less sensitive to low or high values for $\mu$ as the spread between median NSE values is less than those for $\rho$.
 This suggests that EBASE model sensitivity is most influenced by $\rho$ of the prior distributions, i.e., larger values produce more accurate and less variable estimates regardless of $\mu$.
@@
 The observed one-year continuous time series described above for water quality and weather data from Apalachicola Bay was used to compare metabolic estimates from EBASE and the Odum open-water method [@Odum56].
-The Odum method used an approach described in @Murrell18 created for estuarine application that uses a gas-exchange approach described in @Thebault08.
-For better comparison with the EBASE methods, the gas exchange parameterization in the Odum method was modified to use the @Wanninkhof14 approach, the the $b$ parameter fixed at 0.251 $\buni$ [as available in the WtRegDO R package, @Beck21].
+The Odum method used an approach described in @Murrell18 created for estuarine application that uses a gas-exchange approach from @Thebault08.
+For better comparison with the EBASE methods, the gas exchange parameterization in the Odum method was modified to use the @Wanninkhof14 approach, with the $b$ parameter fixed at 0.251 $\buni$ [as available in the WtRegDO R package, @Beck21].
 An additional and fundamental difference between EBASE and the Odum method is the former is likelihood based with metabolic rates estimated by fitting the model to the data, whereas the latter is strictly arithmetic based primarily on integration of the diel dissolved oxygen curve.
 Thus, potentially different results are expected between the approaches based on differences in formulation of the mass balance equations for metabolism as well as alternative statistical approaches to estimating the parameters.
-Each comparison also evaluated differneces in the metabolic estimates using the observed dissolved oxygen time series and a detided time series using weighted regression [@Beck15; available from @Beck21].
+Each comparison also evaluated differences in the metabolic estimates using the observed dissolved oxygen time series and a detided time series using weighted regression [@Beck15; available from @Beck21].
 The latter comparison provided an assessment of metabolic rates that are expected to be minimally influenced by tidal advection, as a fundamental assumption of metabolic models using *in situ* dissolved oxygen time series from fixed monitoring stations.

### 2023-04-18 manu-draft.qmd
fix eqn commands and replace others in text

@@
 
-```{=tex}
 \newcommand{\auni}{\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)}
@@
 \newcommand{\buni}{\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)}
-```
+\newcommand{\area}{\text{mmol}/\text{m}^2/\text{d}}
+
 ## Abstract {.unnumbered}
@@
 where the terms in the square brackets are production (P), respiration (R), and gas exchange (D), respectively, modeled as a function of the change in dissolved oxygen ($dC$) per unit time ($dt$) of the input data.
-Production is modeled linearly as a function of surface irradiance ($PAR$, Watts m$^{-2}$) and $a$ (light efficiency, $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$).
-Respiration is modeled with the single parameter $r$ ($\text{mmol}/\text{m}^3/\text{d}$).
+Production is modeled linearly as a function of surface irradiance ($PAR$, Watts m$^{-2}$) and $a$ (light efficiency, $\auni$).
+Respiration is modeled with the single parameter $r$ ($\runi$).
 Gas exchange is modeled using the formulation in @Wanninkhof14 as a wind-based parameterization for the gas transfer velocity [@Ho16] multiplied by the oxygen concentration gradient as the difference between the concentration measured in the surface water layer, $C$ (mmol m$^{-3}$), and the saturation (or equilibrium) concentration, $C_{Sat}$ (mmol m$^{-3}$), calculated from salinity (psu) and temperature ($^\circ$C) [@Millero81; @Garcia92].
 The gas transfer velocity is modeled using wind speed squared at 10 meters above the water ($U_{10}^2$, m$^2$ s$^{-2}$), water column depth ($H$ in meters, as a fixed parameter or time series from a pressure sensor), and the Schmidt number ($Sc$) defined as the ratio of the kinematic viscosity of water to the molecular diffusivity of the gas for oxygen [from water temperature and salinity using the polynomial fit in @Wanninkhof14].
-The $b$ parameter is estimated from the data and has units of $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$.
+The $b$ parameter is estimated from the data and has units of $\buni$.
 As such, the three parameters estimated by EBASE are $a$, $r$, and $b$ using the required input data that includes time series of dissolved oxygen, water temperature, salinity, PAR, wind speed, and water column depth (`r run_reference(id = "tbl-ebaseiotab")`).
@@
 
-The remaining three parameters $a$ (light efficiency, $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$), $r$ (respiration, $\text{mmol}/\text{m}^3/\text{d}$), and $b$ ($\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$) are estimated by likelihood given the observed data and prior distributions for the parameters.
+The remaining three parameters $a$ (light efficiency, $\auni$), $r$ (respiration, $\runi$), and $b$ ($\buni$) are estimated by likelihood given the observed data and prior distributions for the parameters.
 The JAGS software [as for BASE, @Plummer03] is used with the EBASE R package [described below, @Beck22; @RCT22] to estimate the unknown parameters using Markov Chain Monte Carlo simulations in a Bayesian framework.
@@
 All parameter inputs and outputs for EBASE are shown in `r run_reference(id = "tbl-ebaseiotab")`.
-Areal rates for $P$, $R$, and $D$ (mmol m$^-2$ d$^{-1}$) are returned for comparability of estimates across locations where water column depth may vary.
+Areal rates for $P$, $R$, and $D$ ($\area$) are returned for comparability of estimates across locations where water column depth may vary.
 
@@
 Users can also specify alternative prior distributions from those in Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior] for each parameter.
-Functions are available for viewing model results, including a plot of the modeled dissolved oxygen with the observed, a time series plot of the primary metabolic estimates (P, R, and D, as areal rates with units $\text{mmol}/\text{m}^3/\text{d}$), and a time series plot of the credible intervals for the $a$, $r$, and $b$ parameters.
+Functions are available for viewing model results, including a plot of the modeled dissolved oxygen with the observed, a time series plot of the primary metabolic estimates (P, R, and D, as areal rates with units $\area$), and a time series plot of the credible intervals for the $a$, $r$, and $b$ parameters.
 An additional option allows the user to specify the model optimization period, where the time period can vary from a minimum of one day to the maximum of the entire length of the time series.
@@
 All metabolic rates for production, respiration, and gas exchange ([@fig-synapa]b and c) were also estimated with the synthetic time series.
-The only parameter that could not be created as a continuous time series was $b$, which was set to 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ [@Wanninkhof14].
+The only parameter that could not be created as a continuous time series was $b$, which was set to 0.251 $\buni$ [@Wanninkhof14].
 Accordingly, the synthetic dissolved oxygen time series and the observed water quality and meteorological data were used with EBASE to determine if reasonable estimates for the known parameters and metabolic rates were recovered.

### 2023-04-18 manu-draft.qmd
edit to final ebase v odum paragraphs

@@
 Thus, potentially different results are expected between the approaches based on differences in formulation of the mass balance equations for metabolism as well as alternative statistical approaches to estimating the parameters.
-Each comparison also evaluated differences in the metabolic estimates using the observed dissolved oxygen time series and a detided time series using weighted regression [@Beck15; available from @Beck21].
+Each comparison also evaluated differences in the metabolic estimates using the observed dissolved oxygen time series and a detided time series using weighted regression [@Beck15; @Beck21].
 The latter comparison provided an assessment of metabolic rates that are expected to be minimally influenced by tidal advection, as a fundamental assumption of metabolic models using *in situ* dissolved oxygen time series from fixed monitoring stations.
@@
 Root mean square errors (RMSE) of each linear model were also evaluated.
-Results were assessed separately for estimates based on the observed or detided dissolved oxygen time series.
-
-<!-- edit -->
+Results were assessed separately for estimates using the observed or detided dissolved oxygen time series.
 
@@
 Correlations for all estimates, regardless of method or the input time series, were positive and significantly correlated, as expected.
-The intercept values were all significant, excluding production and respiration with the observed dissolved oxygen, suggesting some level of bias when comparing estimates between the methods.
-Half of the slope estimates were significantly different for estimates from both the observed and detided dissolved oxygen, with production and gas exchange differing for the observed input and net metabolism and gas exchange differing for the detided input.
-The RMSE of each regression comparison were generally lower in the detided results for all comparisons.
+The intercept values were all significant, excluding production with the observed dissolved oxygen, suggesting some level of bias when comparing estimates between the methods.
+The slope estimates for respiration and gas exchange were significantly different for estimates from the observed dissolved oxygen, where all but respiration were signicantly different using the detided dissolved oxygen.
+The RMSE values of each regression comparison were generally lower in the detided results for all comparisons, excluding gas exchange which was higher with the detided dissolved oxygen compared to the observed.
 This latter comparison suggests that detiding the time series produces more comparable estimates between methods.
 
-As noted in the previous paragraph, metabolic estimates between the methods were more similar for results using the detided dissolved oxygen time series ([@fig-apacmpfig]b) as compared to using the observed dissolved oxygen as input ([@fig-apacmpfig]a).
+As noted in the previous paragraph, metabolic estimates between the methods were generally more similar based on RMSE results using the detided dissolved oxygen time series ([@fig-apacmpfig]b) as compared to using the observed dissolved oxygen as input ([@fig-apacmpfig]a).
 This suggests that tidal effects on continuous monitoring data can severely violate assumptions required for methods of estimating ecosystem metabolism [as described in detail in @Beck15].
@@
 The effects of tidal advection on estimates from EBASE were more subtle as anomalous values cannot be returned based on constraints of the priors (i.e., $a$ and $r$ cannot be negative, Equations [-@eq-aprior] and [-@eq-rprior]).
-Additionally, the effects of tidal advection produced a wider range of estimates for NEM in both methods, whereas the detided results were mostly negative indicating that conclusions about net autotrophy or heterotrophy may be inaccurate using observed dissolved oxygen as input.
-Overall, detiding the dissolved oxygen time series prior to estimating metabolic rates produces more stable (lower standard deviation) and smaller ranges (`r run_reference(id = "tbl-apacmptab")`), regardless of the method used.
-
-<!-- edit -->
+Overall, the results suggest that detiding the dissolved oxygen time series prior to estimating metabolic rates produces more stable estimates with smaller ranges (`r run_reference(id = "tbl-apacmptab")`), regardless of the method used.
 

### 2023-04-18 manu-draft.qmd
added JMA comments

@@
 
-where the terms in the square brackets are production (P), respiration (R), and gas exchange (D), respectively, modeled as a function of the change in dissolved oxygen ($dC$) per unit time ($dt$) of the input data.
+where the terms on the right side of the equation are gross primary production (P or $aPAR$), ecosystem respiration (R), and gas exchange (D, remainder), respectively, modeled as a function of the change in dissolved oxygen ($dC$) per unit time ($dt$) of the input data.
 Production is modeled linearly as a function of surface irradiance ($PAR$, Watts m$^{-2}$) and $a$ (light efficiency, $\auni$).
@@
 $$ {#eq-base}
-
-where all parameters are defined as in @eq-ebase, with the additional parameters $p$, $\Theta$, and $K$ (the rearation coefficient).
-The BASE approach provides options to estimate $p$ and $\Theta$ from the data or to set as fixed values.
-A temperature dependency, $T$, is also included where respiration and gas exchange vary with temperature.
-The changes to @eq-base for EBASE were necessary for estuarine application and included 1) a linear relationship of production to PAR ($aPAR$) by setting $p = 1$; 2) constant respiration by removing the temperature dependency; and 3) use of a gas exchange parameterization commonly used in coastal or oceanic settings that depends on wind speed and temperature [i.e., @Grace15 used a littoral parameterization].
-The justification for these changes is as follows.
 
@@
 The synthetic time series used inputs from one year of continuous data from Apalachicola Bay, Florida.
-The input data were 2012 observations at the Cat Point water quality monitoring station (29.7021$^{\circ}$ N, --84.8802$^{\circ}$ W) and included 15 minute observations of water temperature and salinity.
+The input data were observations from the year 2012 at the Cat Point water quality monitoring station (29.7021$^{\circ}$ N, --84.8802$^{\circ}$ W) and included 15-minute observations of water temperature and salinity.
 Continuous meteorological observations from the nearby East Bay station (29.7909$^{\circ}$ N, --84.8834$^{\circ}$ W, \~10km north of Cat Point) were combined with the water quality data at the same time step and included air temperature, wind speed, and PAR.
@@
 
-The prior distributions for the EBASE parameters follow truncated normal distributions with defined means and standard deviations (Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]).
+The prior distributions for the EBASE parameters follow truncated normal distributions with defined means ($\mu$) and standard deviations ($\sigma$)(Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]).
 The sensitivity analysis evaluated changes in the distributions by varying the means and standard deviations from small to large to test the effect of changes in the central tendencies and more or less constrained ranges, respectively, on the priors.
 For each parameter, low and high values for the means and standard deviations were assessed as changes approximately equal to two orders of magnitude of the means and standard deviations of the known parameters of the synthetic time series.
-The only exception was the $b$ parameter where the low values created priors with a distribution close to zero and the high values were equivalent to a uniform prior distribution in the range of 0 to 0.504 (justification noted in @sec-ebasemod).
+The only exception was the $b$ parameter where the low values created priors with a distribution close to zero and the high values were equivalent to a uniform prior distribution in the range of 0 to 0.502 (justification noted in @sec-ebasemod).
 Each unique combination of low and high values for the means and standard deviations of each parameter were evaluated, creating 2$^6$ = 64 different combinations of prior distributions that were used with the synthetic time series in EBASE.
@@
 Overall, most models regardless of optimization period or priors were able to reproduce the dissolved oxygen time series.
-The only models where the NSE values were less than zero for dissolved oxygen were those where $\rho$ of the $r$ parameter was low.
+The only models where the NSE values were less than zero for dissolved oxygen were those where $\sigma$ of the $r$ parameter was low.
 However, substantial variation in NSE values was observed among all other parameters.
@@
 Most parameters regardless of optimization period or priors had NSE < 0 (`r perneg`).
-As such, the prior combinations that produced the highest NSE values across all parameters were those where the $\rho$ values were high for each prior distribution, regardless of the values for $\mu$, i.e., those with sufficient flexibility to search a larger parameter space regardless of $\mu$ produced metabolic parameter estimates most similar to the known values from the synthetic time series.
+As such, the prior combinations that produced the highest NSE values across all parameters were those where the $\sigma$ values were high for each prior distribution, regardless of the values for $\mu$, i.e., those with sufficient flexibility to search a larger parameter space regardless of $\mu$ produced metabolic parameter estimates most similar to the known values from the synthetic time series.
 
@@
 Median NSE values across all models and each optimization period for the different subsets of the prior combinations are shown.
-For example, the bottom left panel with the x-axis value for the $a$ parameter shows the median NSE values for all models using a one-day optimization period where the $\rho$ value for the $a$ prior distribution was low or high.
+For example, the bottom left panel with the x-axis value for the $a$ parameter shows the median NSE values for all models using a one-day optimization period where the $\sigma$ value for the $a$ prior distribution was low or high.
 The size of the points indicate the inter-quartile range of NSE values for each subset.
-As such, each set of points denoted by the x-axis parameters in each panel include median NSE values for all models by optimization period, but subset by the low/high $\mu$ or $\rho$ values for each parameter.
-Although most of the median value are less than zero, indicating a poor ability of EBASE to reproduce the parameters, patterns emerge as to which characteristics of the prior lead to robust estimates.
-For example, large differences in the median NSE values are observed for the low and high values of $\rho$ for the $a$ and $r$ parameters, with positive median NSE values and smaller IQR values using a high value using the 7 and 30 day optimization periods.
-A similar, but smaller effect is seen for the $\rho$ value of the prior for the $b$ parameter.
+As such, each set of points denoted by the x-axis parameters in each panel include median NSE values for all models by optimization period, but subset by the low/high $\mu$ or $\sigma$ values for each parameter.
+Although most of the median NSE values are less than zero, indicating a poor ability of EBASE to reproduce the parameters, patterns emerge as to which characteristics of the prior lead to robust estimates.
+For example, large differences in the median NSE values are observed for low and high $\sigma$ for the $a$ and $r$ parameters, with positive median NSE values and smaller IQR values when using a high $\sigma$ value using the 7 and 30 day optimization periods.
+A similar, but smaller effect is seen for the $\sigma$ value of the prior for the $b$ parameter.
 A conclusion similar to that for @fig-priorcomp is that large standard deviations in the prior distributions lead to more accurate estimates of metabolic parameters.
-Conversely, model results are less sensitive to low or high values for $\mu$ as the spread between median NSE values is less than those for $\rho$.
-This suggests that EBASE model sensitivity is most influenced by $\rho$ of the prior distributions, i.e., larger values produce more accurate and less variable estimates regardless of $\mu$.
+Conversely, model results are less sensitive to low or high values for $\mu$ as the spread between median NSE values is less than those for $\sigma$.
+This suggests that EBASE model sensitivity is most influenced by $\sigma$ of the prior distributions, i.e., larger values produce more accurate and less variable estimates regardless of $\mu$.
 
@@
 The best performing models were those with the highest median NSE values across all parameter estimates in @fig-priorcomp.
-The best performing models for each optimization period (@fig-optex a, c, e) showed similar results where the parameters estimates were similar to the synthetic time series, although more variation in the EBASE results were of course observed with the one day optimization period.
+The best performing models for each optimization period (@fig-optex a, c, e) showed similar results where the parameter estimates were similar to the synthetic time series, although more variation in the EBASE results were of course observed with the one day optimization period.
 However, the $b$ parameter was estimated with high variability for the one and seven day optimization periods (@fig-optex a, c, bottom row) and with a large positive bias for the 30 day optimization period (@fig-optex e, bottom row).
@@
 Finally, the worst performing model for each optimization period (@fig-optex b, d, f) demonstrates the negative NSE values that were obtained (i.e., EBASE values much lower or higher than the synthetic values).
-The worst performing models were caused by constraining EBASE to a small parameter space using low or high values for $\mu$ combined with low values for $\rho$.
+The worst performing models were caused by constraining EBASE to a small parameter space using low or high values for $\mu$ combined with low values for $\sigma$.
 
@@
 The Odum method used an approach described in @Murrell18 created for estuarine application that uses a gas-exchange approach from @Thebault08.
-For better comparison with the EBASE methods, the gas exchange parameterization in the Odum method was modified to use the @Wanninkhof14 approach, with the $b$ parameter fixed at 0.251 $\buni$ [as available in the WtRegDO R package, @Beck21].
-An additional and fundamental difference between EBASE and the Odum method is the former is likelihood based with metabolic rates estimated by fitting the model to the data, whereas the latter is strictly arithmetic based primarily on integration of the diel dissolved oxygen curve.
-Thus, potentially different results are expected between the approaches based on differences in formulation of the mass balance equations for metabolism as well as alternative statistical approaches to estimating the parameters.
+For a better comparison with the EBASE methods, the gas exchange parameterization in the Odum method was modified to use the @Wanninkhof14 approach and the $b$ parameter was fixed at 0.251 $\buni$ [as available in the WtRegDO R package, @Beck21].
+An additional and fundamental difference between EBASE and the Odum method is that the former is likelihood based with metabolic rates estimated by fitting the model to the data, whereas the latter is strictly arithmetic based primarily on integration of the diel dissolved oxygen curve.
+Thus, potentially different results are expected between the approaches based on differences in formulation of the mass balance equations for metabolism, as well as alternative statistical approaches to estimating the parameters.
 Each comparison also evaluated differences in the metabolic estimates using the observed dissolved oxygen time series and a detided time series using weighted regression [@Beck15; @Beck21].
 The latter comparison provided an assessment of metabolic rates that are expected to be minimally influenced by tidal advection, as a fundamental assumption of metabolic models using *in situ* dissolved oxygen time series from fixed monitoring stations.
-The prior distributions for EBASE were uninformed using the same large $\rho$ values from @sec-cmpmod and moderate $\mu$ values ($\mu$ = 0.2 $\auni$ for $a$, $\mu$ = 20 $\runi$ for $r$, $\mu$ = 0.251 $\buni$ for $b$).
+The prior distributions for EBASE were uninformed using the same large $\sigma$ values from @sec-cmpmod and moderate $\mu$ values ($\mu$ = 0.2 $\auni$ for $a$, $\mu$ = 20 $\runi$ for $r$, $\mu$ = 0.251 $\buni$ for $b$).
 A one day model optimization period was used for EBASE for comparability with the Odum method.
 
-The common set of metabolic parameters returned by the EBASE and Odum methods are net ecosystem metabolism, gross production, respiration, and gas exchange.
-Pairwise comparisons of the daily estimates returned by each method were evaluated using simple summary statistics, including Pearson correlation coefficients and linear regression fits to evaluate potential biases in each method.
+The common set of metabolic parameters returned by the EBASE and Odum methods are net ecosystem metabolism (NEM), gross primary production (P), ecosystem respiration (R), and gas exchange (D).
+Pairwise comparisons of the daily estimates returned by each method were evaluated using simple summary statistics, including Pearson correlation coefficients ($\rho$) and linear regression fits to evaluate potential biases in each method.
 For the latter analysis, differences in intercept estimates less than or greater than zero suggested a bias in magnitude, whereas differences in slope estimates less than or greater than one suggested a bias that depended on relative value of each observation (e.g., bias may differ between lower or higher metabolic estimates).
@@
 @fig-apacmpfig and `r run_reference(id = "tbl-apacmptab")` provide a comparison of the metabolic estimates from the EBASE and Odum methods.
-Correlations for all estimates, regardless of method or the input time series, were positive and significantly correlated, as expected.
-The intercept values were all significant, excluding production with the observed dissolved oxygen, suggesting some level of bias when comparing estimates between the methods.
-The slope estimates for respiration and gas exchange were significantly different for estimates from the observed dissolved oxygen, where all but respiration were signicantly different using the detided dissolved oxygen.
-The RMSE values of each regression comparison were generally lower in the detided results for all comparisons, excluding gas exchange which was higher with the detided dissolved oxygen compared to the observed.
-This latter comparison suggests that detiding the time series produces more comparable estimates between methods.
+Correlations for all metabolic estimates, regardless of method or the input time series, were positive and significantly correlated, as expected.
+The intercept values were all significant, excluding P with the observed dissolved oxygen, suggesting some level of bias when comparing metabolic estimates between the methods.
+The slope estimates for R and D were significantly different for metabolic estimates from the observed dissolved oxygen, where all but R were significantly different using the detided dissolved oxygen.
+The RMSE values of each regression comparing the metabolic estimates between methods were generally lower in the detided results, excluding D which was higher with the detided dissolved oxygen relative to the observed.
+This result suggests that detiding the time series produces more comparable metabolic estimates between methods.
 
-As noted in the previous paragraph, metabolic estimates between the methods were generally more similar based on RMSE results using the detided dissolved oxygen time series ([@fig-apacmpfig]b) as compared to using the observed dissolved oxygen as input ([@fig-apacmpfig]a).
+As noted in the previous paragraph, metabolic estimates between the methods were generally more similar based on RMSE results using the detided dissolved oxygen time series ([@fig-apacmpfig]b) than using the observed dissolved oxygen as input ([@fig-apacmpfig]a).
 This suggests that tidal effects on continuous monitoring data can severely violate assumptions required for methods of estimating ecosystem metabolism [as described in detail in @Beck15].
-These effects are easily identified with estimates from the Odum method as negative values for production (P) and positive values for respiration (R), i.e., "anomalous" values.
-The effects of tidal advection on estimates from EBASE were more subtle as anomalous values cannot be returned based on constraints of the priors (i.e., $a$ and $r$ cannot be negative, Equations [-@eq-aprior] and [-@eq-rprior]).
+These effects are easily identified with estimates from the Odum method as negative values for P and positive values for R, i.e., "anomalous" values.
+The effects of tidal advection on estimates from EBASE were more subtle as anomalous values cannot be returned based on constraints of the priors (i.e., none can be negative).
 Overall, the results suggest that detiding the dissolved oxygen time series prior to estimating metabolic rates produces more stable estimates with smaller ranges (`r run_reference(id = "tbl-apacmptab")`), regardless of the method used.
@@
 All priors are normal distributions and truncated to positive values.
-$b$ is also truncated to a maximum of 0.504.](figs/priorplot.png){#fig-priorplot fig-align="center"}
+$b$ is also truncated to a maximum of 0.502.](figs/priorplot.png){#fig-priorplot fig-align="center"}
 
@@
 ![Synthetic time series for one year of continuous data at Apalachicola Bay.
-Synthetic input to EBASE was the (a) dissolved oxygen time series, with results compared to the (b) synthetic $a$ parameter and (c) synthetic metabolic estimates for production (P), respiration (R), and gas exchange (D).
+Synthetic input to EBASE was the (a) dissolved oxygen time series, with results compared to the (b) synthetic $a$ parameter and (c) synthetic metabolic estimates for gross primary production (P), ecosystem respiration (R), and gas exchange (D).
 The $a$ parameter is shown as an areal value for comparison with the metabolic results.
@@
 Results are compared for unique combinations of prior distributions and optimization period.
-The means and standard deviations are evaluated as low (L) or high (H) values (see text for details) and the optimization period is one day, seven days, or thirty days.
-The metabolic results for dissolved oxygen (DO), production (P), respiration (R), gas exchange (D), and the $a$ parameter are evaluated based on Nash-Sutcliffe Efficiency (NSE) values between the EBASE results and synthetic values.
+The means ($\mu$) and standard deviations ($\sigma$) are evaluated as low (L) or high (H) values (see text for details) and the optimization period is one day, seven days, or thirty days.
+The metabolic results for dissolved oxygen (DO), gross primary production (P), ecosystem respiration (R), gas exchange (D), and the $a$ parameter are evaluated based on Nash-Sutcliffe Efficiency (NSE) values between the EBASE results and synthetic values.
 NSE values from zero to one (green) indicate EBASE has recovered the synthetic parameters with values closer to 1 being a better approximation, whereas negative values indicate the mean of the synthetic data is a better predictor than EBASE.
@@
 
-![Summaries of median Nash-Sutcliffe Efficiency values across all parameters for the comparisons of EBASE and synthetic model outputs separated by different combinations of prior values.
+![Summaries of median Nash-Sutcliffe Efficiency (NSE) values across all parameters for the comparisons of EBASE and synthetic model outputs separated by different combinations of prior values.
 Each result on the x-axis represents the median values for which the combination prior applies, e.g., all results from @fig-priorcomp where $\mu$ is high or low for the $a$ parameter using a one day optimization period.
@@
 ![Comparisons of daily metabolic estimates from the EBASE and Odum methods for one year of continuous monitoring data at Apalachicola Bay.
-Results for net ecosystem metabolism (NEM), gross production (P), respiration (R), and gas exchange (D) are compared using the (a) observed and (b) detided dissolved oxygen time series as input to each model.
+Results for net ecosystem metabolism (NEM), gross primary production (P), ecosystem respiration (R), and gas exchange (D) are compared using the (a) observed and (b) detided dissolved oxygen time series as input to each model.
 All values are areal as mmol O$_2$ m$^2$ d$^{-1}$.

### 2023-04-18 manu-draft.qmd
added some initial text to discussion

@@
 These effects are easily identified with estimates from the Odum method as negative values for P and positive values for R, i.e., "anomalous" values.
-The effects of tidal advection on estimates from EBASE were more subtle as anomalous values cannot be returned based on constraints of the priors (i.e., none can be negative).
+The effects of tidal advection on estimates from EBASE were more subtle as anomalous values cannot be returned based on constraints of the priors (i.e., none can be negative), although the range of values for the metabolic estimates was much larger with observed dissolved oxygen.
 Overall, the results suggest that detiding the dissolved oxygen time series prior to estimating metabolic rates produces more stable estimates with smaller ranges (`r run_reference(id = "tbl-apacmptab")`), regardless of the method used.
@@
 
+EBASE provides several computational advantages over BASE implemented in the R package that was created for the purposes herein.
+Most importantly, EBASE was developed to allow different model optimization lengths, whereas the BASE method was developed to estimate results for only a single day.
+Allowing EBASE to incorporate more than a day of observations in the optimization can likely produce more accurate and precise estimates for the metabolic parameters.
+Physical and biological characteristics of an ecosystem that affect the metabolic rates are likely acting at time scales longer than a day, e.g., wind patterns that can affect gas exchange as manifested in the $b$ parameter may persist for several days.
+Exposing the model to additional observations may produce more stable results as the parameters for the best fit are estimated by the Bayesian routine as a function of the data that are evaluated each optimization period.
+However, we cannot provide a precise recommendation on the number of days to use for the optimization period as the exact length of time that ecosystem characteristics can affect metabolic rates may not be known and likely varies by location.
+Our example using the synthetic time series from Apalachicola Bay suggests that results were similar for the 7 and 30 day optimization period.
+As such, using one week of data for each optimization period is likely appropriate for this location.
+Using one day is not recommended given the issues stated above (i.e., unstable results), whereas an optimization period longer than a month may miss important events in the time series that can affect the metabolic estimates.
+
+The EBASE R package can also accommodate missing observations in the input data.
+This allows the estimation of metabolic rates for time series that cover relatively long periods of time (e.g., more than year), when data gaps are likely to occur for several reasons (e.g., equipment malfunction, routine maintenance, etc.).
+The R package accommodates data gaps using linear interpolation for any required parameter used by EBASE.
+This allows the Bayesian routine to provide continuous estimates across the time series.
+The results are then post-processed such that any metabolic estimates that exceed a threshold for interpolated observations are removed from the output.
+The analyst can determine the appropriate length of time to use for excluding results based on interpolated values.
+The default setting for EBASE is 12 hours, where results for an entire optimization period are removed if any interpolated period exceeded 12 hours.
+Although this does not apply to the synthetic time series, some of the estimates using the 2012 observed data for Apalachicola Bay were excluded from the analysis.
+This feature allows EBASE to be used more readily because an analyst does not need to manually screen data that include missing values.
+
+-   Primary modifications (Production is linearly related to PAR (aPAR), not saturation point Respiration is constant (not temp dependent), Uses a gas exchange commonly used in oceanic systems (not littoral - streams/lakes as in Grace))
 -   Describe how EBASE deals with gaps and flexibility of the model to accommodate diff optimization periods
 -   b is sketchy, how to improve? Either fix or develop JAGS model further to include additional data, e.g. tidal height.
+-   detiding
+-   other main conclusions (from RGN comments on draft?)
 

### 2023-04-21 manu-draft.qmd
first cut of abstract

@@
     number-sections: true
-title: "EBASE manuscript"
+title: "Fitting metabolic models to dissolved oxygen data: the Estuarine BAyesian Single-station Estimation method"
 bibliography: refs.bib
@@
 ## Abstract {.unnumbered}
+
+Continuous monitoring data for dissolved oxygen have been used to estimate the contribution of gross primary production and ecosystem respiration to net ecosystem metabolism.
+Most models for estimating metabolic processes make several assumptions on how biological and physical factors affect the expression of metabolism in dissolved oxygen concentrations, which can lead to poorly constrained errors that influence the interpretation of results.
+In estuaries, these methods can inform our understanding of spatial and temporal patterns in trophic state, yet very few models have been developed that address challenges of estimating metabolism in these dynamic systems.
+This paper describes the development, assessment, and interpretation of results from a new approach for estimating metabolic rates in estuaries, or the Estuarine BAyesian Single-station Estimation (EBASE) method.
+EBASE provides several advantages over existing models by 1) using a linear relationship between irradiance and production, 2) removing the temperature-dependency of respiration, and 3) applying a wind-based gas exchange formulation method applicable to estuaries.
+Moreover, the approach uses a Bayesian framework where critical model parameters are fit as a function of the data and informed using a set of prior distributions.
+We demonstrate that EBASE can reproduce known metabolic parameters from a synthetic time series created with continuous monitoring data from the Apalachicola estuarine reserve and provide a quantitative assessment of how the prior distributions affect the model results.
+The approach is also bench-marked with more conventional open-water methods for estimating metabolism, with critical differences attributed to formulation of the metabolic mass-balance equations and the underlying statistical framework (i.e., Bayesian vs simple integration).
+The EBASE approach is available using open source software and can be readily applied to multiple years of long-term monitoring data that are available in many estuaries.
+Overall, EBASE provides an accessible method to parameterize a simple metabolic model appropriate for estuarine systems and will provide additional understanding of processes that influence ecosystem status and condition.
 

### 2023-04-21 manu-draft.qmd
minor abstract edits

@@
 Continuous monitoring data for dissolved oxygen have been used to estimate the contribution of gross primary production and ecosystem respiration to net ecosystem metabolism.
-Most models for estimating metabolic processes make several assumptions on how biological and physical factors affect the expression of metabolism in dissolved oxygen concentrations, which can lead to poorly constrained errors that influence the interpretation of results.
+Most models for estimating metabolic processes make several assumptions on how biological and physical factors affect the expression of metabolism in dissolved oxygen data, which can lead to poorly constrained errors that influence the interpretation of results.
 In estuaries, these methods can inform our understanding of spatial and temporal patterns in trophic state, yet very few models have been developed that address challenges of estimating metabolism in these dynamic systems.
-This paper describes the development, assessment, and interpretation of results from a new approach for estimating metabolic rates in estuaries, or the Estuarine BAyesian Single-station Estimation (EBASE) method.
-EBASE provides several advantages over existing models by 1) using a linear relationship between irradiance and production, 2) removing the temperature-dependency of respiration, and 3) applying a wind-based gas exchange formulation method applicable to estuaries.
+This paper describes the development, assessment, and interpretation of results from a new approach for estimating metabolic rates in estuaries, the Estuarine BAyesian Single-station Estimation (EBASE) method.
+EBASE provides several advantages over existing models by 1) using a linear relationship between irradiance and production, 2) removing the temperature-dependency of respiration, and 3) applying a wind-based gas exchange formula for open water systems.
 Moreover, the approach uses a Bayesian framework where critical model parameters are fit as a function of the data and informed using a set of prior distributions.
 We demonstrate that EBASE can reproduce known metabolic parameters from a synthetic time series created with continuous monitoring data from the Apalachicola estuarine reserve and provide a quantitative assessment of how the prior distributions affect the model results.
-The approach is also bench-marked with more conventional open-water methods for estimating metabolism, with critical differences attributed to formulation of the metabolic mass-balance equations and the underlying statistical framework (i.e., Bayesian vs simple integration).
+The approach is also bench-marked with more conventional methods for estimating metabolism, with critical differences attributed to formulation of the metabolic mass-balance equations and the underlying statistical framework (i.e., Bayesian vs simple integration).
 The EBASE approach is available using open source software and can be readily applied to multiple years of long-term monitoring data that are available in many estuaries.

### 2023-05-01 manu-draft.qmd
added synthetic and synthetic noise comparison plot

@@
 
+As a final evaluation of EBASE to return known metabolic parameters, a new time series was created that added estimated effects of instrument and tidal noise to the synthetic time series.
+This assessment provided information on the influence of common noise signals in dissolved oxygen time series on the estimation process for EBASE.
+An assumption of the analysis is that EBASE will return similar parameters for a noisy time series as for those in the time series used above that did not include noise.
+The instrument and tidal noise time series were first estimated by using a weighted regression analysis [@Beck15] on an actual observed dissolved oxygen time series for Apalachicola Bay using data from 2021.
+
 ### EBASE and Odum comparison with real observations
@@
 
+![A subset of the synthetic time series and the synthetic time series with noise for evaluating the ability of EBASE to recover known metabolic parameters.
+The original synthetic time series and the same time series with instrument and tidal noise is shown in (a) and the estimated instrument and tidal noise is shown in (b).](figs/synapanos.png){#fig-synapanos fig-align="center"}
+
+{{< pagebreak >}}
+
+![Comparisons of modelled dissolved oxygen and metabolic parameters from EBASE for a synthetic time series and a synthetic time series with instrument and tidal noise.
+Black lines show the 1:1 comparison and the red lines show the linear regression fit.
+See @fig-synapanos for an example of the dissolved oxygen for both input time series.](figs/synapanoscmp.png){#fig-synapanoscmp fig-align="center"}
+
+{{< pagebreak >}}
+
 ![Comparisons of daily metabolic estimates from the EBASE and Odum methods for one year of continuous monitoring data at Apalachicola Bay.

### 2023-05-01 R/figs.R

@@ -23,6 +23,25 @@ fwdatcmp <- fwdat %>%
   ) %>% 
   select(Date, DateTimeStamp, DO_obs, a, b, P, R, D)
 
+# fwoxy for input to ebase
+fwdatinp <- fwdat %>% 
+  mutate(
+    datet = dmy_hms(datet, tz = 'America/Jamaica')
+  ) %>% 
+  select(
+    DateTimeStamp = datet,
+    DO_obs = `oxy,mmol/m3`, 
+    Temp = `temp,degC`, 
+    Sal = `salt,ppt`, 
+    PAR = `par,W/m2`, 
+    WSpd = `wspd2,m2/s2`, 
+    H = `ht,m`
+  ) %>% 
+  mutate(
+    DO_obs = DO_obs / 1000 * 32, # to mg/L
+    WSpd = sqrt(WSpd)
+  )
+
 # prior distribution plot ---------------------------------------------------------------------
 
 p <- prior_plot(n = 2e6)
@@ -142,16 +161,96 @@ dev.off()
 # fl <- paste0(tempdir(), '/ebasedefault.RData')
 # download.file('https://github.com/fawda123/BASEmetab_script/raw/master/data/ebasedefault.RData', destfile = fl)
 # load(file = fl)
-# 
+
 # p1 <- defex(ebasedefault, fwdatcmp, ndays = 1, subttl = '(a) 1 day')
 # p2 <- defex(ebasedefault, fwdatcmp, ndays = 7, subttl = '(b) 7 day', ylbs = F)
 # p3 <- defex(ebasedefault, fwdatcmp, ndays = 30, subttl = '(c) 30 day', ylbs = F)
 # 
-# p <- ((p1 + plot_layout(ncol = 1)) | (p2 + plot_layout(ncol = 1)) | (p3 + plot_layout(ncol = 1)))  + plot_layout(ncol = 3, guides = 'collect') & 
+# p <- ((p1 + plot_layout(ncol = 1)) | (p2 + plot_layout(ncol = 1)) | (p3 + plot_layout(ncol = 1)))  + plot_layout(ncol = 3, guides = 'collect') &
 #   theme(
 #     legend.position = 'top'
-#   ) & 
+#   ) &
 #   scale_x_date(date_labels = '%b')
+
+# synthetic with noise ------------------------------------------------------------------------
+
+# load apadb data run through wtreg
+fl <- paste0(tempdir(), '/apadbdtd.RData')
[…]

### 2023-05-01 R/funcs.R

@@ -347,7 +347,7 @@ cmpplo <- function(res, fwdatcmp, subttl, ylbs = TRUE){
       subtitle = subttl
     )
   
-  labs <- c('DO~(mmol~m^{3}~d^{-1})',
+  labs <- c('O[2]~(mmol~m^{-3})',
             'italic(a)~(mmol~m^{-3}~d^{-1})/(W~m^{-2})', 
             'italic(b)~(cm~hr^{-1})/(m^{2}~s^{-2})'
   )
@@ -474,5 +474,80 @@ metsum_fun <- function(apasumdat, met, parms = F){
       select(-matches('mean$|sd$'))
   
   return(out)
+  
+}
+
+# compare synthetic and synthetic + noise
+syncomp_plo <- function(resobs, resnos){
+  
+  res <- list(
+    'Synthetic' = resobs,
+    'Synthetic + noise' = resnos
+  ) %>% 
+    enframe(name = 'dotyp') %>% 
+    unnest('value') %>% 
+    select(-matches('hi$|lo$|converge|rsq|^Date$|dDO|^H$')) %>% 
+    pivot_longer(-c('dotyp', 'DateTimeStamp', 'grp')) %>% 
+    summarise(
+      value = mean(value, na.rm = T),
+      DateTimeStamp = min(DateTimeStamp),
+      .by = c('dotyp', 'grp', 'name')
+    ) %>% 
+    filter(!name == 'DO_obs') %>% 
+    mutate(
+      name = factor(name, 
+                    levels = c('DO_mod', 'a', 'b', 'P', 'R', 'D'), 
+                    labels = c('O[2]~(mmol~m^{-3})', 'italic(a)~(mmol~m^{-3}~d^{-1})/(W~m^{-2})', 'italic(b)~(cm~hr^{-1})/(m^{2}~s^{-2})', 'P~(mmol~m^{2}~d^{-1})', 'R~(mmol~m^{2}~d^{-1})', 'D~(mmol~m^{2}~d^{-1})')
+      )
+    ) %>% 
+    pivot_wider(names_from = dotyp, values_from = value)
+  
+  vrs <- levels(res$name)
+  
+  for(i in seq_along(vrs)){
+    
+    vr <- vrs[i]
+
+    ylab <- NULL
+    xlab <- NULL
+    if(i %in% c(1, 4))
+      ylab <- 'Synthetic + noise'
+    if(i %in% 4:6)
+      xlab <- 'Synthetic'
+    
+    toplo <- res %>% filter(name == vr)
+    
[…]

### 2023-05-02 manu-draft.qmd
rework text about apa actual comparisons, no more slope/intercept analyses

@@
 The synthetic time series used inputs from one year of continuous data from Apalachicola Bay, Florida.
-The input data were observations from the year 2012 at the Cat Point water quality monitoring station (29.7021$^{\circ}$ N, --84.8802$^{\circ}$ W) and included 15-minute observations of water temperature and salinity.
+The input data were observations from the year 2021 at the Cat Point water quality monitoring station (29.7021$^{\circ}$ N, --84.8802$^{\circ}$ W) and included 15-minute observations of water temperature and salinity.
 Continuous meteorological observations from the nearby East Bay station (29.7909$^{\circ}$ N, --84.8834$^{\circ}$ W, \~10km north of Cat Point) were combined with the water quality data at the same time step and included air temperature, wind speed, and PAR.
@@
 The common set of metabolic parameters returned by the EBASE and Odum methods are net ecosystem metabolism (NEM), gross primary production (P), ecosystem respiration (R), and gas exchange (D).
-Pairwise comparisons of the daily estimates returned by each method were evaluated using simple summary statistics, including Pearson correlation coefficients ($\rho$) and linear regression fits to evaluate potential biases in each method.
-For the latter analysis, differences in intercept estimates less than or greater than zero suggested a bias in magnitude, whereas differences in slope estimates less than or greater than one suggested a bias that depended on relative value of each observation (e.g., bias may differ between lower or higher metabolic estimates).
-Comparisons of the intercept and slope estimates to zero and one, respectively, were based on confidence intervals from the standard errors of each.
-Root mean square errors (RMSE) of each linear model were also evaluated.
+Pairwise comparisons of the daily estimates returned by each method were evaluated using simple summary statistics, including Pearson correlation coefficients ($\rho$) and root mean square errors (RMSE) of the linear regression fit comparing the same metabolic parameter between methods.
 Results were assessed separately for estimates using the observed or detided dissolved oxygen time series.
@@
 @fig-apacmpfig and `r run_reference(id = "tbl-apacmptab")` provide a comparison of the metabolic estimates from the EBASE and Odum methods.
-Correlations for all metabolic estimates, regardless of method or the input time series, were positive and significantly correlated, as expected.
-The intercept values were all significant, excluding P with the observed dissolved oxygen, suggesting some level of bias when comparing metabolic estimates between the methods.
-The slope estimates for R and D were significantly different for metabolic estimates from the observed dissolved oxygen, where all but R were significantly different using the detided dissolved oxygen.
+Correlations for all metabolic estimates, regardless of the input time series, were positive and significantly correlated, as expected.
+However, correlations were generally stronger for the metabolic results based on the detided dissolved oxygen time series.
+The only exception for the correlation estimates was gas exchange (D), which had a slightly weaker correlation using the detided results.
+@fit-apacmpfig shows that the Odum results for gas exchange were more variable for the same values returned as no or small gas exchange for the EBASE method based on the detided dissolved oxygen.
 The RMSE values of each regression comparing the metabolic estimates between methods were generally lower in the detided results, excluding D which was higher with the detided dissolved oxygen relative to the observed.
-This result suggests that detiding the time series produces more comparable metabolic estimates between methods.
 
-As noted in the previous paragraph, metabolic estimates between the methods were generally more similar based on RMSE results using the detided dissolved oxygen time series ([@fig-apacmpfig]b) than using the observed dissolved oxygen as input ([@fig-apacmpfig]a).
-This suggests that tidal effects on continuous monitoring data can severely violate assumptions required for methods of estimating ecosystem metabolism [as described in detail in @Beck15].
-These effects are easily identified with estimates from the Odum method as negative values for P and positive values for R, i.e., "anomalous" values.
-The effects of tidal advection on estimates from EBASE were more subtle as anomalous values cannot be returned based on constraints of the priors (i.e., none can be negative), although the range of values for the metabolic estimates was much larger with observed dissolved oxygen.
-Overall, the results suggest that detiding the dissolved oxygen time series prior to estimating metabolic rates produces more stable estimates with smaller ranges (`r run_reference(id = "tbl-apacmptab")`), regardless of the method used.
+The comparisons in @fig-apacmpfig and `r run_reference(id = "tbl-apacmptab")` suggest that tidal effects on continuous monitoring data can severely violate assumptions required for methods of estimating ecosystem metabolism [as described in detail in @Beck15].
+These effects are easily identified by stronger correlations and lower RMSE using the detided dissolved oxygen.
+Further, effects of tidal advection can be seen with estimates from the Odum method as negative values for P and R, i.e., "anomalous" values.
+The effects of tidal advection on estimates from EBASE were more subtle as anomalous values cannot be returned based on constraints of the priors (i.e., none can be negative), although the range of values for the metabolic estimates were much larger with observed dissolved oxygen.
+Overall, the results suggest that detiding the dissolved oxygen time series prior to estimating metabolic rates reduces tidal noise and produces more stable estimates with smaller ranges, regardless of the method used.
 
@@
 The default setting for EBASE is 12 hours, where results for an entire optimization period are removed if any interpolated period exceeded 12 hours.
-Although this does not apply to the synthetic time series, some of the estimates using the 2012 observed data for Apalachicola Bay were excluded from the analysis.
+Although this does not apply to the synthetic time series, some of the estimates using the 2021 observed data for Apalachicola Bay were excluded from the analysis.
 This feature allows EBASE to be used more readily because an analyst does not need to manually screen data that include missing values.
@@
 Separate comparisons are made for estimates based on observed and detided dissolved oxygen.
-Summary statistics include Pearson correlation coefficients and intercept, slope, and root mean square error (RMSE) of a linear fit between the same metabolic estimates obtained from each method.
-Significance of the p-values for the intercept and slope indicates if the estimate is different from zero or one, respectively."
+Summary statistics include Pearson correlation coefficients and root mean square error (RMSE) of a linear fit between the same metabolic estimates obtained from each method."
 apacmptab

### 2023-05-02 R/tabs.R

@@ -70,7 +70,6 @@ load(file = url('https://github.com/fawda123/BASEmetab_script/raw/master/data/ap
 
 tosum <- apacmp %>% 
   mutate(
-    typ = ifelse(typ == 'BASEmetab', 'BASE', typ),
     var = factor(var, levels = c('NEM', 'P', 'R', 'D'), 
     )
   ) %>% 
@@ -79,7 +78,7 @@ tosum <- apacmp %>%
 grd <- crossing(
   dotyp = unique(tosum$dotyp), 
   var = levels(tosum$var), 
-  comp = c('Odum v EBASE', 'BASE v EBASE')
+  comp = c('Odum v EBASE')
   ) %>% 
   mutate(
     corv = NA, 
@@ -142,11 +141,10 @@ totab <- grd %>%
   select(dotyp, comp, var, corv, int, slo, rmse) %>% 
   mutate(
     dotyp = factor(dotyp, levels = c('observed', 'detided'), labels = c('Observed', 'Detided')), 
-    comp = factor(comp, levels = c('Odum v EBASE', 'BASE v EBASE')), 
+    comp = factor(comp, levels = c('Odum v EBASE')), 
     var = factor(var, levels = c('NEM', 'P', 'R', 'D'))
   ) %>% 
   arrange(dotyp, comp, var) %>%
-  filter(comp != 'BASE v EBASE') %>%  # remove comparison to BASE
   group_by(dotyp) %>% 
   mutate(
     comp = ifelse(duplicated(comp), '', as.character(comp))
@@ -160,7 +158,7 @@ totab <- grd %>%
     Slope = slo,
     RMSE = rmse
   ) %>% 
-  select(-Comparison) %>% 
+  select(-Comparison, -Intercept, -Slope) %>% 
   flextable::as_grouped_data(groups = c('Dissolved Oxygen'))
 
 apacmptab <- totab %>%
@@ -170,9 +168,9 @@ apacmptab <- totab %>%
   padding(padding = 1, part = 'all') %>% 
   width(width = 6.5 / ncol_keys(.)) %>% 
   flextable::compose(part = 'header', j = 'RMSE', value = as_paragraph(as_equation("\\text{RMSE}~(\\text{mmol}~\\text{O}_2/\\text{m}^2/\\text{d})"))) %>% 
[…]

### 2023-05-02 manu-draft.qmd
added text about synthetic plus noise methods and results

@@
 As a final evaluation of EBASE to return known metabolic parameters, a new time series was created that added estimated effects of instrument and tidal noise to the synthetic time series.
-This assessment provided information on the influence of common noise signals in dissolved oxygen time series on the estimation process for EBASE.
-An assumption of the analysis is that EBASE will return similar parameters for a noisy time series as for those in the time series used above that did not include noise.
-The instrument and tidal noise time series were first estimated by using a weighted regression analysis [@Beck15] on an actual observed dissolved oxygen time series for Apalachicola Bay using data from 2021.
+The objective of the analysis is to develop an expectation of how metabolic estimates may change using in situ data with noise that may substantially mask a biological signal in dissolved oxygen measurements.
+The instrument and tidal noise were estimated using a weighted regression analysis [@Beck15; @Beck21] on actual observed dissolved oxygen time series for Apalachicola Bay using data from 2021.
+This approach models dissolved oxygen as a function of time, salinity, and season using a moving window approach with weights of the independent variables appropriate for each time period within the windows.
+The results of this analysis return a predicted and detided dissolved oxygen time series.
+The instrument noise was estimated as the difference between the observed and predicted time series and the tidal noise was estimated as the difference between the predicted and detided time series.
+Both the instrument and tidal noise were added to the same synthetic time series above (@fig-synapanos).
+EBASE was then used with unconstrained priors (large $\sigma$) on the synthetic time series with noise, then compared with results from the original synthetic time series.
+
+@fig-synapanoscmp shows the comparison of the EBASE results for the synthetic time series with instrument and tidal noise to those from the original synthetic time series.
+Results for both time series produced similar dissolved oxygen estimates (top-left panel, @fig-synapanoscmp), although notable differences were observed in the remaining parameters.
+In particular, gross primary production (P) and ecosystem respiration (R) were generally higher and more variable for results from the noisy time series, with differences in the former attributed to higher estimates in the $a$ parameter.
+As for the previous results, the $b$ parameter was estimated with low precision for both input time series, although results were centered around the true value of 0.251 $\buni$.
+Gas exchange (D) was similar between the input time series, suggesting lack of precision in the $b$ parameter regardless of noise had a minimal influence.
 
@@
 The only exception for the correlation estimates was gas exchange (D), which had a slightly weaker correlation using the detided results.
-@fit-apacmpfig shows that the Odum results for gas exchange were more variable for the same values returned as no or small gas exchange for the EBASE method based on the detided dissolved oxygen.
+@fig-apacmpfig shows that the Odum results for gas exchange were more variable for the same values returned as no or small gas exchange for the EBASE method based on the detided dissolved oxygen.
 The RMSE values of each regression comparing the metabolic estimates between methods were generally lower in the detided results, excluding D which was higher with the detided dissolved oxygen relative to the observed.
@@
 The effects of tidal advection on estimates from EBASE were more subtle as anomalous values cannot be returned based on constraints of the priors (i.e., none can be negative), although the range of values for the metabolic estimates were much larger with observed dissolved oxygen.
+A similar result (larger estimates for P and R) was observed using the synthetic time series with instrument and tidal noise (@fig-synapanoscmp).
 Overall, the results suggest that detiding the dissolved oxygen time series prior to estimating metabolic rates reduces tidal noise and produces more stable estimates with smaller ranges, regardless of the method used.
@@
 -   b is sketchy, how to improve? Either fix or develop JAGS model further to include additional data, e.g. tidal height.
--   detiding
+-   detiding - noise may produce higher and more variable metabolic rates
 -   other main conclusions (from RGN comments on draft?)
@@
 Results are compared for unique combinations of prior distributions and optimization period.
-The means ($\mu$) and standard deviations ($\sigma$) are evaluated as low (L) or high (H) values (see text for details) and the optimization period is one day, seven days, or thirty days.
+The means ($\mu$) and standard deviations ($\sigma$) are evaluated as low (L) or high (H) values (see text for details) and the optimization period for EBASE is one day, seven days, or thirty days.
 The metabolic results for dissolved oxygen (DO), gross primary production (P), ecosystem respiration (R), gas exchange (D), and the $a$ parameter are evaluated based on Nash-Sutcliffe Efficiency (NSE) values between the EBASE results and synthetic values.
-NSE values from zero to one (green) indicate EBASE has recovered the synthetic parameters with values closer to 1 being a better approximation, whereas negative values indicate the mean of the synthetic data is a better predictor than EBASE.
+NSE values from zero to one (green) indicate EBASE has recovered the synthetic parameters with values closer to 1 being a better approximation, whereas negative values indicate poor recovery.
 The color scale for NSE is also exponentially transformed to minimize the influence of large negative values on the visual display.

### 2023-05-03 R/figs.R
rework optex fig to keep y-axis scales consistent across rows

@@ -138,39 +138,11 @@ fl <- paste0(tempdir(), '/apasumdat.RData')
 download.file('https://github.com/fawda123/BASEmetab_script/raw/master/data/apasumdat.RData', destfile = fl)
 load(file = fl)
 
-p1 <- optex(apagrd, fwdatcmp, apasumdat, rnkmetsum = 1, ndays = 1, met = 'nse', subttl = '(a) Best, 1 day')
-p2 <- optex(apagrd, fwdatcmp, apasumdat, rnkmetsum = 64, ndays = 1, met = 'nse', subttl = '(b) Worst, 1 day', ylbs = F)
-p3 <- optex(apagrd, fwdatcmp, apasumdat, rnkmetsum = 1, ndays = 7, met = 'nse', subttl = '(c) Best, 7 day', ylbs = F)
-p4 <- optex(apagrd, fwdatcmp, apasumdat, rnkmetsum = 64, ndays = 7, met = 'nse', subttl = '(d) Worst, 7 day', ylbs = F)
-p5 <- optex(apagrd, fwdatcmp, apasumdat, rnkmetsum = 1, ndays = 30, met = 'nse', subttl = '(e) Best, 30 day', ylbs = F)
-p6 <- optex(apagrd, fwdatcmp, apasumdat, rnkmetsum = 64, ndays = 30, met = 'nse', subttl = '(f) Worst, 30 day', ylbs = F)
-
-p <- ((p1 + plot_layout(ncol = 1)) | (p2 + plot_layout(ncol = 1)) | (p3 + plot_layout(ncol = 1)) | (p4 + plot_layout(ncol = 1))| (p5 + plot_layout(ncol = 1)) | (p6 + plot_layout(ncol = 1)))  + plot_layout(ncol = 6, guides = 'collect') & 
-  theme(
-    legend.position = 'top'
-  ) & 
-  scale_x_date(date_labels = '%b')
+p <- optex(apagrd, fwdatcmp, apasumdat, rnkmetsum = c(1, 64), met = 'nse')
 
 png(here('figs/optex.png'), height = 8.7, width = 9, family = 'serif', units = 'in', res = 500)
 print(p)
 dev.off()
-
-# # Fwoxy apa EBASE defaults ------------------------------------------------
-# 
-# # grid ests
-# fl <- paste0(tempdir(), '/ebasedefault.RData')
[…]

### 2023-05-03 R/funcs.R

@@ -402,44 +402,93 @@ cmpplo <- function(res, fwdatcmp, subttl, ylbs = TRUE){
 }
 
 # comparison of fwoxy and ebase results for ranked model by prior at opt of ndays
-optex <- function(apagrd, fwdatcmp, apasumdat, rnkmetsum, ndays, met, subttl, ylbs = TRUE){
+optex <- function(apagrd, fwdatcmp, apasumdat, rnkmetsum, met){
 
   # find the prior comparison
-  rnkfnd <- metsum_fun(apasumdat, met, parms = T) %>% 
-    filter(rnkmetsum == !!rnkmetsum) %>% 
-    filter(ndays == !!ndays)
-
+  rnkfnd <- metsum_fun(apasumdat, met = met, parms = T) %>% 
+    filter(rnkmetsum %in% !!rnkmetsum) %>% 
+    select(-ind, -metsum)
+  
   res <- apagrd %>% 
-    filter(
-      amean == rnkfnd$amean & asd == rnkfnd$asd & 
-      rmean == rnkfnd$rmean & rsd == rnkfnd$rsd & 
-      bmean == rnkfnd$bmean & bsd == rnkfnd$bsd & 
-      ndays == rnkfnd$ndays
+    inner_join(rnkfnd, by = c("amean", "asd", "rmean", "rsd", "bmean", "bsd", "ndays")) %>% 
+    arrange(ndays, rnkmetsum) %>% 
+    mutate(
+      subfig = paste0('(', letters[1:nrow(.)], ')'), 
+      subfig = case_when(
+        rnkmetsum == 1 ~ paste(subfig, 'Best', ndays, 'day', sep = '~'), 
+        rnkmetsum == 64 ~ paste(subfig, 'Worst', ndays, 'day', sep = '~') 
+      ), 
+      subfig = factor(subfig)
     ) %>% 
-    pull(out) %>% 
-    .[[1]] %>% 
-    .[[1]]
+    select(subfig, out) %>% 
+    unnest(out) %>% 
+    unnest(out)
   
-  out <- cmpplo(res, fwdatcmp, subttl, ylbs)
-  
-  return(out)
-  
-}
-
-# comparison of fwoxy and ebase results using default EBASE priors
-defex <- function(ebasedefault, fwdatcmp, ndays, subttl, ylbs = TRUE){
-  
-  res <- ebasedefault %>% 
-    filter(
-      ndays == !!ndays
[…]

### 2023-05-03 manu-draft.qmd
update fig and text for priorsumcomp

@@
 @fig-priorsumcomp provides an alternative summary of how the prior distributions influenced the ability of EBASE to reproduce the synthetic parameters.
-Median NSE values across all models and each optimization period for the different subsets of the prior combinations are shown.
-For example, the bottom left panel with the x-axis value for the $a$ parameter shows the median NSE values for all models using a one-day optimization period where the $\sigma$ value for the $a$ prior distribution was low or high.
+Median NSE values across all models, parameters, and each optimization period for the different subsets of the prior combinations are shown relative to the grand median NSE value across all comparisons.
+For example, the bottom left panel with the x-axis value for the $a$ parameter shows the difference in the median NSE values for all models using a one-day optimization period where the $\sigma$ value for the $a$ prior distribution was low or high.
 The size of the points indicate the inter-quartile range of NSE values for each subset.
 As such, each set of points denoted by the x-axis parameters in each panel include median NSE values for all models by optimization period, but subset by the low/high $\mu$ or $\sigma$ values for each parameter.
-Although most of the median NSE values are less than zero, indicating a poor ability of EBASE to reproduce the parameters, patterns emerge as to which characteristics of the prior lead to robust estimates.
-For example, large differences in the median NSE values are observed for low and high $\sigma$ for the $a$ and $r$ parameters, with positive median NSE values and smaller IQR values when using a high $\sigma$ value using the 7 and 30 day optimization periods.
-A similar, but smaller effect is seen for the $\sigma$ value of the prior for the $b$ parameter.
-A conclusion similar to that for @fig-priorcomp is that large standard deviations in the prior distributions lead to more accurate estimates of metabolic parameters.
+Positive values indicate relative performance greater than the overall median NSE values and negative values indicate relative performance worse than the overall median.
+Dissolved oxygen was generally modeled with high accuracy for all priors and it was excluded to not skew the assessment.
+The patterns suggest that EBASE results are more sensitive to changes in the $\sigma$ parameter as shown by the larger spread in the median results compared to those for $\mu$.
+This result is especially true for the priors of the $a$ and $r$ parameters.
 Conversely, model results are less sensitive to low or high values for $\mu$ as the spread between median NSE values is less than those for $\sigma$.
 This suggests that EBASE model sensitivity is most influenced by $\sigma$ of the prior distributions, i.e., larger values produce more accurate and less variable estimates regardless of $\mu$.
+Patterns by optimization period were not apparent.
 
@@
 ![Summaries of median Nash-Sutcliffe Efficiency (NSE) values across all parameters for the comparisons of EBASE and synthetic model outputs separated by different combinations of prior values.
+Summaries are shown relative to the difference from the grand median across all comparisons.
 Each result on the x-axis represents the median values for which the combination prior applies, e.g., all results from @fig-priorcomp where $\mu$ is high or low for the $a$ parameter using a one day optimization period.

### 2023-05-03 R/funcs.R

@@ -241,6 +241,7 @@ priorsumcomp <- function(dat, met = 'r2'){
         T ~ paste(ndays, 'days')
       )
     ) %>% 
+    filter(!var %in% 'DO_mod') %>% 
     pivot_longer(cols = matches('mean$|sd$'), names_to = 'prior', values_to = 'val') %>% 
     summarise(
       medv = median(met, na.rm = T), 
@@ -248,11 +249,13 @@ priorsumcomp <- function(dat, met = 'r2'){
       .by = c('ndays', 'prior', 'val')
     ) %>% 
     mutate(
-      param = gsub('mean$|sd$', '', prior),
-      param = factor(param, levels = c('a', 'r', 'b')),
-      prior = gsub('^a|^r|^b', '', prior), 
-      prior = factor(prior, levels = c('mean', 'sd'), labels = c('mu', 'sigma')), 
-      ndays = factor(ndays, levels = c('1 day', '7 days', '30 days'), labels = c('1~day', '7~days', '30~days'))
+           grandmed = median(medv),
+           medv = medv - grandmed,
+           param = gsub('mean$|sd$', '', prior),
+           param = factor(param, levels = c('a', 'r', 'b')),
+           prior = gsub('^a|^r|^b', '', prior), 
+           prior = factor(prior, levels = c('mean', 'sd'), labels = c('mu', 'sigma')), 
+           ndays = factor(ndays, levels = c('1 day', '7 days', '30 days'), labels = c('1~day', '7~days', '30~days'))
     )
   
   metsel <- tibble(
@@ -260,20 +263,26 @@ priorsumcomp <- function(dat, met = 'r2'){
     lbspr = c('italic(R)^2', 'RMSE', 'MAPE', 'mae', 'NSE')
   ) %>% 
     filter(met == !!met)
-  ylab <- parse(text = paste0('Median~', metsel$lbspr))
-
+  ylab <- parse(text = paste0('Median~', metsel$lbspr,'~-Grand~Median'))
+  
+  wd <- 0.3
+  
   p <- ggplot(toplo, aes(x = param, group = val)) + 
-    geom_point(aes(y = medv, fill = val, size = iqr), pch = 21) + #, position = position_dodge(width = wd)) +
+    geom_linerange(aes(ymin = 0, ymax = medv, x = param), position = position_dodge(width = wd), linetype = 'dashed') +
+    geom_hline(yintercept = 0) + 
[…]

### 2023-05-04 manu-draft.qmd
scratch text to discussion

@@
 
+EBASE demonstrated on synthetic and obseved data that reasonable estimates ..
+EBASe was developed to address several critical needs in existing methods for estimating metabolism by 1) using a Bayesian framework, and 2) modifying several components of the underlying model with parameterizations more appropriate for estuarine settings.
+First, the Bayesian framework follows a an approach where the model is fit to the available data and allow for the incorporation of prior knowledge in teh model fitting process.
+...
+The advantages of such an approach has led to widespread adoption of Bayesian approaches over frequentist approaches, especially in the coastal and marine community,  yet they have not been explored in do data for estuaries.
+We have demonstrated that even using uninformed priors, EBASE can reproduce .
+However, an exception is b which was unstable .
+This result may be explained simply by lack of gas exchange signal in teh DO time series that could be dominated by production and respiration.
+Caffrey et al.
+suggest...
+EBASE could be applied in other settings where gas exchange may be expected to have a strong influence (windy).
+Or, more constrained priors could be used.
+
+The second improvement of EBASE over alternative methods was the several modifications to
+
+Our results also showed that tidal advection... led to positive bias, more variation, previously shown by .... solution is to detide or further develop model to accomodate tidal heigh contribution
+
 EBASE provides several computational advantages over BASE implemented in the R package that was created for the purposes herein.
@@
 
--   Primary modifications (Production is linearly related to PAR (aPAR), not saturation point Respiration is constant (not temp dependent), Uses a gas exchange commonly used in oceanic systems (not littoral - streams/lakes as in Grace))
--   Describe how EBASE deals with gaps and flexibility of the model to accommodate diff optimization periods
--   b is sketchy, how to improve? Either fix or develop JAGS model further to include additional data, e.g. tidal height.
--   detiding - noise may produce higher and more variable metabolic rates
 -   other main conclusions (from RGN comments on draft?)

### 2023-05-04 manu-draft.qmd
additional scratch text to discussion

@@
 
-The second improvement of EBASE over alternative methods was the several modifications to
+The second improvement of EBASE over alternative methods was the several modifications to ....
+Primary modifications (Production is linearly related to PAR (aPAR), not saturation point Respiration is constant (not temp dependent), Uses a gas exchange commonly used in oceanic systems (not littoral - streams/lakes as in Grace)).
+this also reduces the data requirements by not needing...
 
@@
 This feature allows EBASE to be used more readily because an analyst does not need to manually screen data that include missing values.
+
+Some limitations of EBASE include still relatively long time to fit the model compared to other methods, e.g., Odum, but this is not unique to EBASE and other MCMC Bayesian approaches have similar constraints.
+Some options can be changed to decrease time for model fitting, or multiple cores can be used.
+Another limitation is the need for the user to choose priors and optimization periods, although we provide some suggestions above.
+Our extensive analysis evaluating effects on priors and optimization period provide some guidance.
+As a comparison, the detiding method requires a user to choose windows and the guidance is xyz, although the sensitivity analysis was not informed by known results.
+Our use of the EBASE model formula to generate a synthetic time series with known parameters is a significant improvement over the former analysis and our suggestions can be take with more confidence.
+Although this is limited to Apalachicola data, the broad suggestion to user unconstrained priors will likely expand to other system with different characteristics.
 

### 2023-05-05 manu-draft.qmd
ebase advantages paragraphs to discussion

@@
 \newcommand{\area}{\text{mmol}/\text{m}^2/\text{d}}
-
+```
 ## Abstract {.unnumbered}
@@
 where the terms on the right side of the equation are gross primary production (P or $aPAR$), ecosystem respiration (R), and gas exchange (D, remainder), respectively, modeled as a function of the change in dissolved oxygen ($dC$) per unit time ($dt$) of the input data.
-Production is modeled linearly as a function of surface irradiance ($PAR$, Watts m$^{-2}$) and $a$ (light efficiency, $\auni$).
-Respiration is modeled with the single parameter $r$ ($\runi$).
+Production is modeled linearly as a function of surface irradiance ($PAR$, Watts m$^{-2}$) and $a$ (light efficiency, $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$).
+Respiration is modeled with the single parameter $r$ ($\text{mmol}/\text{m}^3/\text{d}$).
 Gas exchange is modeled using the formulation in @Wanninkhof14 as a wind-based parameterization for the gas transfer velocity [@Ho16] multiplied by the oxygen concentration gradient as the difference between the concentration measured in the surface water layer, $C$ (mmol m$^{-3}$), and the saturation (or equilibrium) concentration, $C_{Sat}$ (mmol m$^{-3}$), calculated from salinity (psu) and temperature ($^\circ$C) [@Millero81; @Garcia92].
 The gas transfer velocity is modeled using wind speed squared at 10 meters above the water ($U_{10}^2$, m$^2$ s$^{-2}$), water column depth ($H$ in meters, as a fixed parameter or time series from a pressure sensor), and the Schmidt number ($Sc$) defined as the ratio of the kinematic viscosity of water to the molecular diffusivity of the gas for oxygen [from water temperature and salinity using the polynomial fit in @Wanninkhof14].
-The $b$ parameter is estimated from the data and has units of $\buni$.
+The $b$ parameter is estimated from the data and has units of $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$.
 As such, the three parameters estimated by EBASE are $a$, $r$, and $b$ using the required input data that includes time series of dissolved oxygen, water temperature, salinity, PAR, wind speed, and water column depth (`r run_reference(id = "tbl-ebaseiotab")`).
@@
 
-### Changes to BASE
+### Changes to BASE {#sec-basemodif}
 
@@
 
-The remaining three parameters $a$ (light efficiency, $\auni$), $r$ (respiration, $\runi$), and $b$ ($\buni$) are estimated by likelihood given the observed data and prior distributions for the parameters.
+The remaining three parameters $a$ (light efficiency, $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$), $r$ (respiration, $\text{mmol}/\text{m}^3/\text{d}$), and $b$ ($\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$) are estimated by likelihood given the observed data and prior distributions for the parameters.
 The JAGS software [as for BASE, @Plummer03] is used with the EBASE R package [described below, @Beck22; @RCT22] to estimate the unknown parameters using Markov Chain Monte Carlo simulations in a Bayesian framework.
@@
 All parameter inputs and outputs for EBASE are shown in `r run_reference(id = "tbl-ebaseiotab")`.
-Areal rates for $P$, $R$, and $D$ ($\area$) are returned for comparability of estimates across locations where water column depth may vary.
+Areal rates for $P$, $R$, and $D$ ($\text{mmol}/\text{m}^2/\text{d}$) are returned for comparability of estimates across locations where water column depth may vary.
 
@@
 As for the BASE model, relatively uninformed prior distributions following a normal Gaussian distribution are used by default, although the priors can be changed based on previous knowledge of parameters specific to an ecosystem or as informed by other metabolic modeling approaches.
-Reasonable uninformed prior distributions for EBASE were chosen with mean values using approximate estimates from the literature [@Wanninkhof14;@Grace15] and standard deviations that were sufficiently large to allow the Bayesian routine to search an unconstrained parameter space.
+Reasonable uninformed prior distributions for EBASE were chosen with mean values using approximate estimates from the literature [@Wanninkhof14; @Grace15] and standard deviations that were sufficiently large to allow the Bayesian routine to search an unconstrained parameter space.
 The default priors for $a$, $r$, and $b$ are:
@@
 Users can also specify alternative prior distributions from those in Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior] for each parameter.
-Functions are available for viewing model results, including a plot of the modeled dissolved oxygen with the observed, a time series plot of the primary metabolic estimates (P, R, and D, as areal rates with units $\area$), and a time series plot of the credible intervals for the $a$, $r$, and $b$ parameters.
+Functions are available for viewing model results, including a plot of the modeled dissolved oxygen with the observed, a time series plot of the primary metabolic estimates (P, R, and D, as areal rates with units $\text{mmol}/\text{m}^2/\text{d}$), and a time series plot of the credible intervals for the $a$, $r$, and $b$ parameters.
 An additional option allows the user to specify the model optimization period, where the time period can vary from a minimum of one day to the maximum of the entire length of the time series.
@@
 All metabolic rates for production, respiration, and gas exchange ([@fig-synapa]b and c) were also estimated with the synthetic time series.
-The only parameter that could not be created as a continuous time series was $b$, which was set to 0.251 $\buni$ [@Wanninkhof14].
+The only parameter that could not be created as a continuous time series was $b$, which was set to 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ [@Wanninkhof14].
 Accordingly, the synthetic dissolved oxygen time series and the observed water quality and meteorological data were used with EBASE to determine if reasonable estimates for the known parameters and metabolic rates were recovered.
@@
 @fig-priorcomp shows a summary of the comparisons of the synthetic data with EBASE for each unique combination of prior distributions and the three optimization periods of one, seven, and thirty days.
-The results are shown using Nash-Sutcliffe Efficiency (NSE) values [@Nash70;@Moriasi07]:
+The results are shown using Nash-Sutcliffe Efficiency (NSE) values [@Nash70; @Moriasi07]:
 
@@
 Median NSE values across parameters for the top model in each optimization period were also similar (1 day = `r optmod[['1']]`, 7 days = `r optmod[['7']]`, 30 days = `r optmod[['30']]`).
-For the seven day and thirty day optimization periods, the NSE values for all parameters were greater than zero for the top models, where two of the parameters in the top model for the one day optimization period has NSE < 0.
-Most parameters regardless of optimization period or priors had NSE < 0 (`r perneg`).
+For the seven day and thirty day optimization periods, the NSE values for all parameters were greater than zero for the top models, where two of the parameters in the top model for the one day optimization period has NSE \< 0.
+Most parameters regardless of optimization period or priors had NSE \< 0 (`r perneg`).
 As such, the prior combinations that produced the highest NSE values across all parameters were those where the $\sigma$ values were high for each prior distribution, regardless of the values for $\mu$, i.e., those with sufficient flexibility to search a larger parameter space regardless of $\mu$ produced metabolic parameter estimates most similar to the known values from the synthetic time series.
@@
 In particular, gross primary production (P) and ecosystem respiration (R) were generally higher and more variable for results from the noisy time series, with differences in the former attributed to higher estimates in the $a$ parameter.
-As for the previous results, the $b$ parameter was estimated with low precision for both input time series, although results were centered around the true value of 0.251 $\buni$.
+As for the previous results, the $b$ parameter was estimated with low precision for both input time series, although results were centered around the true value of 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$.
 Gas exchange (D) was similar between the input time series, suggesting lack of precision in the $b$ parameter regardless of noise had a minimal influence.
@@
 The Odum method used an approach described in @Murrell18 created for estuarine application that uses a gas-exchange approach from @Thebault08.
-For a better comparison with the EBASE methods, the gas exchange parameterization in the Odum method was modified to use the @Wanninkhof14 approach and the $b$ parameter was fixed at 0.251 $\buni$ [as available in the WtRegDO R package, @Beck21].
+For a better comparison with the EBASE methods, the gas exchange parameterization in the Odum method was modified to use the @Wanninkhof14 approach and the $b$ parameter was fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ [as available in the WtRegDO R package, @Beck21].
 An additional and fundamental difference between EBASE and the Odum method is that the former is likelihood based with metabolic rates estimated by fitting the model to the data, whereas the latter is strictly arithmetic based primarily on integration of the diel dissolved oxygen curve.
@@
 The latter comparison provided an assessment of metabolic rates that are expected to be minimally influenced by tidal advection, as a fundamental assumption of metabolic models using *in situ* dissolved oxygen time series from fixed monitoring stations.
-The prior distributions for EBASE were uninformed using the same large $\sigma$ values from @sec-cmpmod and moderate $\mu$ values ($\mu$ = 0.2 $\auni$ for $a$, $\mu$ = 20 $\runi$ for $r$, $\mu$ = 0.251 $\buni$ for $b$).
+The prior distributions for EBASE were uninformed using the same large $\sigma$ values from @sec-cmpmod and moderate $\mu$ values ($\mu$ = 0.2 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$ for $a$, $\mu$ = 20 $\text{mmol}/\text{m}^3/\text{d}$ for $r$, $\mu$ = 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ for $b$).
 A one day model optimization period was used for EBASE for comparability with the Odum method.
@@
 
-EBASE demonstrated on synthetic and obseved data that reasonable estimates ..
-EBASe was developed to address several critical needs in existing methods for estimating metabolism by 1) using a Bayesian framework, and 2) modifying several components of the underlying model with parameterizations more appropriate for estuarine settings.
-First, the Bayesian framework follows a an approach where the model is fit to the available data and allow for the incorporation of prior knowledge in teh model fitting process.
-...
-The advantages of such an approach has led to widespread adoption of Bayesian approaches over frequentist approaches, especially in the coastal and marine community,  yet they have not been explored in do data for estuaries.
-We have demonstrated that even using uninformed priors, EBASE can reproduce .
-However, an exception is b which was unstable .
-This result may be explained simply by lack of gas exchange signal in teh DO time series that could be dominated by production and respiration.
-Caffrey et al.
-suggest...
-EBASE could be applied in other settings where gas exchange may be expected to have a strong influence (windy).
-Or, more constrained priors could be used.
+The above analyses demonstrated that EBASE can successfully recover known metabolic parameters from a synthetic time series and provides robust estimates on real data covering a year of observations.
+The success of these results was predicated on the requirements of EBASE to address several critical needs that existing methods for estimating metabolism do not address for coastal applications.
+These needs were addressed by 1) using a Bayesian framework that estimates the best fit of the model to the data with posterior probabilities describing the likelihood of the fit, and 2) modifying the underlying mass-balance model for dissolved with several components more appropriate for estuarine settings.
+First, the Bayesian framework enables hypothesis testing of how a model can be fit to data by allowing the use of prior knowledge in the model fitting process, which results in posterior probabilities describing the certainty of the parameter estimates [@Hilborn13].
+The advantages of the Bayesian approach have led to widespread adoption over frequentist approaches in many scientific applications, especially in the coastal and marine community [e.g., @myers2001; @borsuk2001; @brown2017], yet these methods have not been rigorously explored for estimating metabolic rates in estuaries.
+Our results demonstrated that even using uninformed priors, EBASE can reproduce known metabolic parameters and rates with reasonable certainty.
+However, an exception was the ability of the model to estimate stable values for the $b$ parameter, although the resulting estimate of gas exchange (D) was similar to the known values.
+This result could be explained the dominance of the relative processes that manifest signals in the dissolved oxygen time series.
+That is, estuarine dissolved oxygen at Apalachicola Bay may be dominated by production and respiration, and less so from gas exchange, such that the available data provided insufficient information for EBASE to estimate $b$.
+This result aligns with others that have suggested gas exchange may be a relatively small component of the metabolic signal in shallow estuaries [@Murrell18].
+In such cases, more constrained prior distributions for $b$ could be appropriate.
 
-The second improvement of EBASE over alternative methods was the several modifications to ....
+The second advantage of EBASE that makes the model more appropriate for estuarine settings included several modification to the underlying mass-balance equation for dissolved oxygen (@eq-ebase).
+As described in @sec-basemodif, EBASE is an estuarine adaptation of BASE (@eq-base) that was developed for stream applications and many assumptions of the underlying model of the latter were inappropriate for the former.
+Specifically, EBASE was modified to address three primary concerns of using BASE in an estuarine setting.
+First, the estimation of gross primary production as a function of PAR assumed a linear relationship such that the increase of production per unit increase of PAR was consistent for any level of the latter.
+The BASE model includes an exponential parameter, $p$, that defines a non-linear relationship between P and PAR (although it should be noted that BASE provides an option to set $p = 1$).
+@Holtgrieve10 describes a simple linear photosynthesis-irradiance curve as appropriate for systems that are never light-saturated and this assumption likely holds true for estuaries that are commonly more turbid than open-ocean or freshwater locations.
+Second, EBASE removed the temperature-dependency of respiration and assumed that respiration did not change with water temperature, consistent with applications of the Odum method to other shallow estuaries [@Beck15; @Murrell18].
+This modification addresses recent work that suggests respiration in streams may be reduced during the night with lower temperatures as the respiration of labile autocthonous organic matter may contribute to changes in the dissolved oxygen time series [@sadro2014; @schindler2017].
+There is currently no justification for this assumption in estuaries.
+Finally, the gas exchange in EBASE used the parameterization in @Wanninkhof14 that is primarily wind-based and developed for the open-ocean, which is clearly more appropriate for estuarine settings.
+This modification also removes a dependency of gas exchange on temperature and depth, further simplifying the approach.
+Overall, these changes also reduce the required inputs for its use and many monitoring locations nationwide include the requisite data for applying EBASE (i.e., dissolved oxygen, water temperature, salinity, PAR, and wind speed).
+
 Primary modifications (Production is linearly related to PAR (aPAR), not saturation point Respiration is constant (not temp dependent), Uses a gas exchange commonly used in oceanic systems (not littoral - streams/lakes as in Grace)).

### 2023-05-05 manu-draft.qmd
more text to discussion

@@
 The remaining three parameters $a$ (light efficiency, $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$), $r$ (respiration, $\text{mmol}/\text{m}^3/\text{d}$), and $b$ ($\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$) are estimated by likelihood given the observed data and prior distributions for the parameters.
-The JAGS software [as for BASE, @Plummer03] is used with the EBASE R package [described below, @Beck22; @RCT22] to estimate the unknown parameters using Markov Chain Monte Carlo simulations in a Bayesian framework.
+The JAGS software [as for BASE, @Plummer03] is used with the EBASE R package [described below, @Beck22; @RCT22] to estimate the unknown parameters using Markov Chain Monte Carlo (MCMC) simulations in a Bayesian framework.
 At each time step, the change in oxygen concentration between time steps is calculated from the equation using model inputs and parameter guesses based on the prior distributions, and then a finite difference approximation is used to estimate modeled oxygen concentration.
@@
 Second, EBASE removed the temperature-dependency of respiration and assumed that respiration did not change with water temperature, consistent with applications of the Odum method to other shallow estuaries [@Beck15; @Murrell18].
-This modification addresses recent work that suggests respiration in streams may be reduced during the night with lower temperatures as the respiration of labile autocthonous organic matter may contribute to changes in the dissolved oxygen time series [@sadro2014; @schindler2017].
+This modification addresses recent work that suggests respiration in streams may be reduced during the night with lower temperatures as the respiration of labile autochthonous organic matter may contribute to changes in the dissolved oxygen time series [@sadro2014; @schindler2017].
 There is currently no justification for this assumption in estuaries.
-Finally, the gas exchange in EBASE used the parameterization in @Wanninkhof14 that is primarily wind-based and developed for the open-ocean, which is clearly more appropriate for estuarine settings.
-This modification also removes a dependency of gas exchange on temperature and depth, further simplifying the approach.
+Finally, the gas exchange in EBASE used the parameterization in @Wanninkhof14 that is primarily wind-based and developed for the open-ocean, which is clearly more appropriate than the prior formulation in BASE that also includes a temperature and depth dependency.
 Overall, these changes also reduce the required inputs for its use and many monitoring locations nationwide include the requisite data for applying EBASE (i.e., dissolved oxygen, water temperature, salinity, PAR, and wind speed).
 
-Primary modifications (Production is linearly related to PAR (aPAR), not saturation point Respiration is constant (not temp dependent), Uses a gas exchange commonly used in oceanic systems (not littoral - streams/lakes as in Grace)).
-this also reduces the data requirements by not needing...
-
-Our results also showed that tidal advection... led to positive bias, more variation, previously shown by .... solution is to detide or further develop model to accomodate tidal heigh contribution
+Our results also confirmed that the effects of tidal advection on metabolic estimates can lead to biases in understanding gross primary production and ecosystem respiration.
+This result has previously been shown using the Odum open water method [@kemp1980; @caffrey2003; @Beck15] and confirmed herein, whereby results using observed time series can produce over-estimates of metabolic rates with more variation than those without tidal influence.
+Our assessment of the synthetic time series with noise and use of the detided dissolved oyxgen time series from the observed data at Apalachicola Bay to estimate metabolism both demonstrated this bias (Figures [-@fig-synapanoscmp] and [-@fig-apacmpfig]).
+As a result, detiding dissolved oxygen data is recommended prior to estimating metabolism, regardless of the method used.
+In addition to inaccurate and imprecise metabolic estimates, tidal influences are also most likely observed as negative (or anomalous) values in the production and respiration rates using the Odum method [@caffrey2003].
+Tidal effects may be less obvious using EBASE since negative values cannot be obtained based on constraints of the mass-balance model.
+A potential future approach for developing EBASE could be an explicit formulation of tidal advection in the model.
+A synoptic tidal height time series could be used in the Bayesian framework as a separate estimate of the signal of advection on the observed dissolved oxygen.
+This has the advantage of both detiding and estimating metabolic rates using the same model, whereas using weighted regression prior to EBASE as used herein, requires additional analysis.
+However, field measurements [e.g., control volume approaches, @falter2008; @nidzieko2014] may be required to provide reasonable constraints on the prior distributions for any parameters that quantify advective influences on dissolved oxygen in a Bayesian framework.
 
@@
 
-Some limitations of EBASE include still relatively long time to fit the model compared to other methods, e.g., Odum, but this is not unique to EBASE and other MCMC Bayesian approaches have similar constraints.
-Some options can be changed to decrease time for model fitting, or multiple cores can be used.
-Another limitation is the need for the user to choose priors and optimization periods, although we provide some suggestions above.
-Our extensive analysis evaluating effects on priors and optimization period provide some guidance.
-As a comparison, the detiding method requires a user to choose windows and the guidance is xyz, although the sensitivity analysis was not informed by known results.
-Our use of the EBASE model formula to generate a synthetic time series with known parameters is a significant improvement over the former analysis and our suggestions can be take with more confidence.
-Although this is limited to Apalachicola data, the broad suggestion to user unconstrained priors will likely expand to other system with different characteristics.
+While the EBASE theory and software provide advantages for metabolic estimates, the approach is not without limitations.
+Like most models, explicit choices are needed to apply EBASE to time series data.
+First, the use of unconstrained prior distributions was shown to reproduce known metabolic parameters with reasonable certainty, although the $b$ parameter was still relatively unstable and unconstrained priors may not provide useful estimates in estuarine settings outside of the location evaluated herein.
+This may require further constraining the priors, although doing so may produce biased results in the absence of additional data to justify a constraint.
+Alternative metabolic models could be used for informing priors, such as using the Odum method to inform the prior for the reareation coefficient, $K$, for the BASE method.
+Initial application of BASE to estuarine data demonstrated that unstable and unrealistic metabolic rates were returned with uninformed priors unless the $K$ parameter was constrained.
+A similar approach could be applied to create more accurate estimates from EBASE.
+An additional choice required to use EBASE is the model optimization period, although some guidance is provided above.
 
--   other main conclusions (from RGN comments on draft?)
+A final limitation of EBASE is the relatively long processing time when applying the method to time series longer than a few days.
+For example, the execution time applying EBASE to a year of data on a conventional laptop computer was approximately 40 minutes, depending on the model settings.
+This time is compared to the Odum method applied to the same dataset, where results are obtained in a matter of seconds.
+Given the Bayesian framework used by EBASE, not considering the additional statistical advantages, it is not reasonable to expect comparable processing times between the methods.
+Regardless, computation times could be reduced with additional hardware improvements (e.g., accessing multiple processing cores) or changing the model estimation methods (e.g., reducing the number of MCMC simulations used by JAGS), although the latter may negatively affect model accuracy.
 
 ## Comments and Recommendations
+
+-   other main conclusions (from RGN comments on draft)
 

### 2023-05-08 manu-draft.qmd
added comments and recommendations to discussion

@@
 The remaining three parameters $a$ (light efficiency, $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$), $r$ (respiration, $\text{mmol}/\text{m}^3/\text{d}$), and $b$ ($\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$) are estimated by likelihood given the observed data and prior distributions for the parameters.
-The JAGS software [as for BASE, @Plummer03] is used with the EBASE R package [described below, @Beck22; @RCT22] to estimate the unknown parameters using Markov Chain Monte Carlo (MCMC) simulations in a Bayesian framework.
+The JAGS software [as for BASE, @Plummer03] is used with the EBASE R package [described below, @Beck23; @RCT22] to estimate the unknown parameters using Markov Chain Monte Carlo (MCMC) simulations in a Bayesian framework.
 At each time step, the change in oxygen concentration between time steps is calculated from the equation using model inputs and parameter guesses based on the prior distributions, and then a finite difference approximation is used to estimate modeled oxygen concentration.
@@
 
-For ease of application, an R package was developed to implement the EBASE methods described above [@Beck22].
+For ease of application, an R package was developed to implement the EBASE methods described above [@Beck23].
 A detailed web page (<https://fawda123.github.io/EBASE/>) describes the required data inputs for estimating the metabolic parameters using the core R functions that pass inputs to the JAGS software.
@@
 The success of these results was predicated on the requirements of EBASE to address several critical needs that existing methods for estimating metabolism do not address for coastal applications.
-These needs were addressed by 1) using a Bayesian framework that estimates the best fit of the model to the data with posterior probabilities describing the likelihood of the fit, and 2) modifying the underlying mass-balance model for dissolved with several components more appropriate for estuarine settings.
-First, the Bayesian framework enables hypothesis testing of how a model can be fit to data by allowing the use of prior knowledge in the model fitting process, which results in posterior probabilities describing the certainty of the parameter estimates [@Hilborn13].
+These needs were addressed by 1) using a Bayesian framework that estimates the best fit of the model to the data with posterior probabilities describing the likelihood of the fit, and 2) modifying the underlying mass-balance model for dissolved oxygen with several components more appropriate for estuarine settings.
+First, the Bayesian framework enables hypothesis testing of how a model can be fit to data by using prior knowledge in the model fitting process, which results in posterior probabilities describing the certainty of the parameter estimates [@Hilborn13].
 The advantages of the Bayesian approach have led to widespread adoption over frequentist approaches in many scientific applications, especially in the coastal and marine community [e.g., @myers2001; @borsuk2001; @brown2017], yet these methods have not been rigorously explored for estimating metabolic rates in estuaries.
@@
 However, an exception was the ability of the model to estimate stable values for the $b$ parameter, although the resulting estimate of gas exchange (D) was similar to the known values.
-This result could be explained the dominance of the relative processes that manifest signals in the dissolved oxygen time series.
+This result could be explained by the dominance of the relative processes that manifest signals in the dissolved oxygen time series.
 That is, estuarine dissolved oxygen at Apalachicola Bay may be dominated by production and respiration, and less so from gas exchange, such that the available data provided insufficient information for EBASE to estimate $b$.
 This result aligns with others that have suggested gas exchange may be a relatively small component of the metabolic signal in shallow estuaries [@Murrell18].
-In such cases, more constrained prior distributions for $b$ could be appropriate.
+In such cases, more constrained prior distributions for $b$ could be appropriate or EBASE could be tested in settings with larger contributions of gas exchange expressed by dissolved oxygen.
 
@@
 There is currently no justification for this assumption in estuaries.
-Finally, the gas exchange in EBASE used the parameterization in @Wanninkhof14 that is primarily wind-based and developed for the open-ocean, which is clearly more appropriate than the prior formulation in BASE that also includes a temperature and depth dependency.
-Overall, these changes also reduce the required inputs for its use and many monitoring locations nationwide include the requisite data for applying EBASE (i.e., dissolved oxygen, water temperature, salinity, PAR, and wind speed).
+Finally, the gas exchange in EBASE used the parameterization in @Wanninkhof14 that is primarily wind-based and developed for the open-ocean, which is clearly more appropriate than the prior formulation in BASE that includes a temperature and depth dependency for streams.
+Overall, these changes also reduce the required inputs and many monitoring locations nationwide include the requisite data for applying EBASE (i.e., dissolved oxygen, water temperature, salinity, PAR, and wind speed).
 
@@
 Our assessment of the synthetic time series with noise and use of the detided dissolved oyxgen time series from the observed data at Apalachicola Bay to estimate metabolism both demonstrated this bias (Figures [-@fig-synapanoscmp] and [-@fig-apacmpfig]).
-As a result, detiding dissolved oxygen data is recommended prior to estimating metabolism, regardless of the method used.
+As a result, detiding dissolved oxygen data is recommended prior to estimating metabolism, regardless of whether EBASE or an alternative method is used.
 In addition to inaccurate and imprecise metabolic estimates, tidal influences are also most likely observed as negative (or anomalous) values in the production and respiration rates using the Odum method [@caffrey2003].
-Tidal effects may be less obvious using EBASE since negative values cannot be obtained based on constraints of the mass-balance model.
-A potential future approach for developing EBASE could be an explicit formulation of tidal advection in the model.
-A synoptic tidal height time series could be used in the Bayesian framework as a separate estimate of the signal of advection on the observed dissolved oxygen.
-This has the advantage of both detiding and estimating metabolic rates using the same model, whereas using weighted regression prior to EBASE as used herein, requires additional analysis.
-However, field measurements [e.g., control volume approaches, @falter2008; @nidzieko2014] may be required to provide reasonable constraints on the prior distributions for any parameters that quantify advective influences on dissolved oxygen in a Bayesian framework.
+Tidal effects may be less obvious using EBASE since negative values cannot be obtained based on constraints of the prior distributions, although as noted above, larger and more variable estimates are returned.
+A potential future approach for developing EBASE could be an explicit formulation of tidal advection in the model using a synoptic tidal height time series in the Bayesian framework.
+This has the advantage of both detiding and estimating metabolic rates using the same model, whereas using weighted regression prior to EBASE as used herein, requires additional analysis and assumptions.
+However, field measurements [e.g., control volume approaches, @falter2008; @nidzieko2014] may be required to provide reasonable constraints on the prior distributions for any parameters that quantify advective influences on dissolved oxygen.
 
-EBASE provides several computational advantages over BASE implemented in the R package that was created for the purposes herein.
-Most importantly, EBASE was developed to allow different model optimization lengths, whereas the BASE method was developed to estimate results for only a single day.
+EBASE provides several computational advantages over BASE implemented in the R package that was created for the purposes herein [@Beck23].
+Most importantly, EBASE was developed to allow different model optimization lengths, whereas the BASE method was developed to estimate results for a single day.
 Allowing EBASE to incorporate more than a day of observations in the optimization can likely produce more accurate and precise estimates for the metabolic parameters.
-Physical and biological characteristics of an ecosystem that affect the metabolic rates are likely acting at time scales longer than a day, e.g., wind patterns that can affect gas exchange as manifested in the $b$ parameter may persist for several days.
+Physical and biological characteristics of an ecosystem that affect the metabolic rates are likely acting at time scales longer than a day, e.g., wind patterns that can affect gas exchange as quantified by the $b$ parameter may persist for several days.
 Exposing the model to additional observations may produce more stable results as the parameters for the best fit are estimated by the Bayesian routine as a function of the data that are evaluated each optimization period.
@@
 The analyst can determine the appropriate length of time to use for excluding results based on interpolated values.
-The default setting for EBASE is 12 hours, where results for an entire optimization period are removed if any interpolated period exceeded 12 hours.
+The default setting for EBASE is 12 hours, where results for an entire optimization period are removed if any interpolated period exceeded 12 hours within that period.
 Although this does not apply to the synthetic time series, some of the estimates using the 2021 observed data for Apalachicola Bay were excluded from the analysis.
-This feature allows EBASE to be used more readily because an analyst does not need to manually screen data that include missing values.
+This feature allows EBASE to be applied more easily to other data setes because an analyst does not need to manually screen data that include missing values.
 
@@
 Initial application of BASE to estuarine data demonstrated that unstable and unrealistic metabolic rates were returned with uninformed priors unless the $K$ parameter was constrained.
-A similar approach could be applied to create more accurate estimates from EBASE.
-An additional choice required to use EBASE is the model optimization period, although some guidance is provided above.
+A similar approach could be applied to create more accurate estimates from EBASE depending on location.
+An additional choice for using EBASE is the model optimization period, although some guidance is provided above.
 
-A final limitation of EBASE is the relatively long processing time when applying the method to time series longer than a few days.
+A final limitation of EBASE is the relatively long processing time when applying the method to a time series longer than a few days.
 For example, the execution time applying EBASE to a year of data on a conventional laptop computer was approximately 40 minutes, depending on the model settings.
@@
 
--   other main conclusions (from RGN comments on draft)
+The development of EBASE represents a new approach for estimating metabolic rates in estuarine settings that leverages existing methods by adapting theoretical components of the underlying mass-balance model using a Bayesian framework.
+Our results have demonstrated that unknown metabolic parameters can be estimated with reasonable certainty using uninformed prior distributions and EBASE could be a valuable approach to improve the understanding of critical ecosystem processes as the method can be readily applied to other settings.
+A notable limitation of EBASE was an inability to preciseily recover the $b$ parameter from a synthetic time series, although this result may be explained by a minimal contribution of gas exchange as expressed in the dissolved oxygen time series.
+Effects of tidal advection on dissolved oxygen measurements were also observed, demonstrating a violation of the assumption that the measured data represent a water sample with the same metabolic history.
+Further development of EBASE to accommodate a tidal signal, or exploring alternative methods for detiding dissolved oxygen data, represent promising avenues of future research to more accurately quantify metabolic parameters in estuarines.
+Regardless, EBASE represents a potentially powerful tool that removes many of the assumptions of existing methods by allowing unknown metabolic parameters to be estimated by the data and can also incorporate existing knowledge for further exploration of estuarine ecosystem properties.
 

### 2023-05-08 manu-draft.qmd
adding draft content to intro

@@
 
--   Overview of metabolism - importance
--   Existing methods: Odum/WtRegDO, BASE, references in @Grace15, applications to lakes and streams
--   Applications to estuaries - challenges
--   Goals and objectives
+Estuaries play a key role in the global cycling of elements because of high rates of biogeochemical activity at the interface between land, ocean, and atmosphere.
+As materials are transported from land to ocean, estuaries profoundly transform or filter these materials through various processes, chief among which are primary production and respiration, collectively referred to as metabolism [@schubel1984].
+For example, estuarine respiration of organic carbon delivered from land produces carbon dioxide, which may be lost to the atmosphere via outgassing [e.g, @cai2011] .
+Similarly, estuarine primary production consumes dissolved inorganic carbon delivered from land, forming particulate organic carbon that may be buried in estuarine sediments [e.g., @hu2006] .
+Estuarine metabolic processes are fundamentally important, as illustrated by the dissolved inorganic carbon budget of eastern North American estuaries [@najjar2018]: net primary production (11.7 $\pm$ 3.0 Tg C yr$^{-1}$) and heterotrophic respiration (13.2 $\pm$ 3.1 Tg C yr$^{–1}$) far exceed inputs from upland sources (5.6 $\pm$ 1.2 Tg C yr$^{-1}$) and outputs to the atmosphere (4.2 $\pm$ 1.7 Tg C yr$^{–1}$) and ocean (2.9 $\pm$ 2.3 Tg C yr$^{–1}$).
+As such, small changes in either net primary production or heterotrophic respiration can have a large impact on the difference between the two, net ecosystem metabolism (NEM, also referred to as net ecosystem production and net community production), which ultimately governs the form and quantity of estuarine materials that are lost to the atmosphere, the sediments, and the ocean.
+
+Methods for estimating metabolic rates in aquatic environments are numerous .
+Kemp and Testa (2011) and Staehr et al.
+(2012) summarized the main approaches for quantifying ecosystem metabolism in aquatic systems (including incubations, the open-water method, ecosystem budgets, oxygen isotopes, and inert gases).
+
+Complementing the incubation-based methods are open-water methods (Odum, 1956), which exploit the diel cycle of dissolved oxygen and allow the simultaneous determination of whole-system gross primary production (GPP, equals net primary production + autotrophic respiration), ecosystem respiration (ER, equals heterotrophic + autotrophic respiration), and NEP (equals GPP -- ER).
+A global synthesis of warm-season, open-water metabolism studies (Hoellein et al., 2013) was based on a total of 47 estimates, almost all of which were derived from a single study of U.S.
+estuaries (Caffrey, 2004); a similar synthesis by Yvon-Durocher et al.
+(2012) was exclusively based on Caffrey (2004).
+As such, a unified model to explain the large variability of metabolic rates within and across estuaries is clearly lacking as a critical tool for understanding metabolic processes.
+Such a model should be informed by the available data at a given location and have the ability to incorporate prior information on key parameters if available.
+Bayesian techniques have not been fully leveraged in estuarine applications despite their potential to address the above needs.
+A potentially useful application developed for streams is the BAyesian Single-station Estimation method.
+
+Application of open-water techniques for estimating metabolism is especially problematic based on the assumptions and poor error constraints of existing methods that rely on continuous measurements of dissolved oxygen.
+Fundamental to these challenges is the high temporal and spatial variability noted in global syntheses of primary production and respiration.
+Cloern et al.
+(2014) found that annual phytoplankton primary production can vary 10-fold within an estuary and up to 5-fold from year-to-year.
+Hopkinson and Smith (2005) focused on cross-system variations and found benthic and pelagic respiration rates varying by more than factors of 30 and 40, respectively, ranges that are broadly consistent with wholesystem respiration rates based on the open-water method (Hoellein et al., 2013).
+
+The goal of this paper is to describe the development and application of a new data assimilation method for simultaneously determining gross primary production (P), ecosystem respiration (R), net ecosystem metabolism (NEM), and the gas exchange (D) in estuaries from high-resolution dissolved oxygen time series.
+We call this method EBASE (Estuarine BAyesian Single-station Estimation), which builds and improves on the BASE method [@Grace15] using an approach more appropriate for estuarine time series.
+The rationale and changes made to BASE are described in detail, followed by a demonstration of how EBASE can estimate known metabolic parameters with reasonable certainty from a synthetic time series.
+Lastly, EBASE is compared to metabolic estimates using the Odum open-water method to identify key differences related to both the theoretical and statistical differences associated with each method.
+These comparisons combine results using real observations and those based on detided inputs to demonstrate effects of tidal advection on metabolic estimates.
+EBASE is available as an open-source software package created with the R statistical programming language [@Beck23; @RCT23] and the results herein provide practical suggestions for use of the software on novel datasets.
 
@@
 The remaining three parameters $a$ (light efficiency, $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$), $r$ (respiration, $\text{mmol}/\text{m}^3/\text{d}$), and $b$ ($\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$) are estimated by likelihood given the observed data and prior distributions for the parameters.
-The JAGS software [as for BASE, @Plummer03] is used with the EBASE R package [described below, @Beck23; @RCT22] to estimate the unknown parameters using Markov Chain Monte Carlo (MCMC) simulations in a Bayesian framework.
+The JAGS software [as for BASE, @Plummer03] is used with the EBASE R package [described below, @Beck23; @RCT23] to estimate the unknown parameters using Markov Chain Monte Carlo (MCMC) simulations in a Bayesian framework.
 At each time step, the change in oxygen concentration between time steps is calculated from the equation using model inputs and parameter guesses based on the prior distributions, and then a finite difference approximation is used to estimate modeled oxygen concentration.

### 2023-05-09 manu-draft.qmd
adding citations to intro

@@
 
-Methods for estimating metabolic rates in aquatic environments are numerous .
-Kemp and Testa (2011) and Staehr et al.
-(2012) summarized the main approaches for quantifying ecosystem metabolism in aquatic systems (including incubations, the open-water method, ecosystem budgets, oxygen isotopes, and inert gases).
-
-Complementing the incubation-based methods are open-water methods (Odum, 1956), which exploit the diel cycle of dissolved oxygen and allow the simultaneous determination of whole-system gross primary production (GPP, equals net primary production + autotrophic respiration), ecosystem respiration (ER, equals heterotrophic + autotrophic respiration), and NEP (equals GPP -- ER).
+Methods for estimating metabolism in aquatic environments are numerous [@kemp2011; @staehr2011].
+Chief among these are open-water methods [@Odum56], which exploit the diel cycle of dissolved oxygen and allow the simultaneous determination of whole-system gross primary production (P, equals net primary production + autotrophic respiration), ecosystem respiration (R, equals heterotrophic + autotrophic respiration), and NE (equals P-R).
 A global synthesis of warm-season, open-water metabolism studies (Hoellein et al., 2013) was based on a total of 47 estimates, almost all of which were derived from a single study of U.S.
-estuaries (Caffrey, 2004); a similar synthesis by Yvon-Durocher et al.
-(2012) was exclusively based on Caffrey (2004).
+estuaries [@caffrey2004]; a similar synthesis by @yvon-durocher2012 was exclusively based on @caffrey2004.
 As such, a unified model to explain the large variability of metabolic rates within and across estuaries is clearly lacking as a critical tool for understanding metabolic processes.
@@
 Bayesian techniques have not been fully leveraged in estuarine applications despite their potential to address the above needs.
-A potentially useful application developed for streams is the BAyesian Single-station Estimation method.
+A potentially useful application developed for streams is the BAyesian Single-station Estimation method....
 
@@
 Hopkinson and Smith (2005) focused on cross-system variations and found benthic and pelagic respiration rates varying by more than factors of 30 and 40, respectively, ranges that are broadly consistent with wholesystem respiration rates based on the open-water method (Hoellein et al., 2013).
+Tidal variation...
 
-The goal of this paper is to describe the development and application of a new data assimilation method for simultaneously determining gross primary production (P), ecosystem respiration (R), net ecosystem metabolism (NEM), and the gas exchange (D) in estuaries from high-resolution dissolved oxygen time series.
+The goal of this paper is to describe the development and application of a new data assimilation method for simultaneously determining gross primary production, ecosystem respiration, net ecosystem metabolism (NEM), and the gas exchange (D) in estuaries from high-resolution dissolved oxygen time series.
 We call this method EBASE (Estuarine BAyesian Single-station Estimation), which builds and improves on the BASE method [@Grace15] using an approach more appropriate for estuarine time series.
@@
 Our results have demonstrated that unknown metabolic parameters can be estimated with reasonable certainty using uninformed prior distributions and EBASE could be a valuable approach to improve the understanding of critical ecosystem processes as the method can be readily applied to other settings.
-A notable limitation of EBASE was an inability to preciseily recover the $b$ parameter from a synthetic time series, although this result may be explained by a minimal contribution of gas exchange as expressed in the dissolved oxygen time series.
-Effects of tidal advection on dissolved oxygen measurements were also observed, demonstrating a violation of the assumption that the measured data represent a water sample with the same metabolic history.
-Further development of EBASE to accommodate a tidal signal, or exploring alternative methods for detiding dissolved oxygen data, represent promising avenues of future research to more accurately quantify metabolic parameters in estuarines.
-Regardless, EBASE represents a potentially powerful tool that removes many of the assumptions of existing methods by allowing unknown metabolic parameters to be estimated by the data and can also incorporate existing knowledge for further exploration of estuarine ecosystem properties.
+A notable limitation of EBASE was an inability to precisely recover the $b$ parameter from a synthetic time series, although this result may be explained by a minimal contribution of gas exchange as expressed in the dissolved oxygen time series.
+Effects of tidal advection on dissolved oxygen measurements were also observed, demonstrating a violation of the assumption that the measured data represent a water sample having a continuous metabolic history.
+Further development of EBASE to accommodate a tidal signal, or exploring alternative methods for detiding dissolved oxygen data, are promising avenues of future research to more accurately quantify metabolic parameters in estuaries.
+Overall, EBASE represents a potentially powerful tool that removes many of the assumptions of existing methods by allowing unknown metabolic parameters to be estimated by the data and can also incorporate existing knowledge for further exploration of estuarine ecosystem properties.
 

### 2023-05-15 manu-draft.qmd
better intro draft done, need to proof

@@
 Estuarine metabolic processes are fundamentally important, as illustrated by the dissolved inorganic carbon budget of eastern North American estuaries [@najjar2018]: net primary production (11.7 $\pm$ 3.0 Tg C yr$^{-1}$) and heterotrophic respiration (13.2 $\pm$ 3.1 Tg C yr$^{–1}$) far exceed inputs from upland sources (5.6 $\pm$ 1.2 Tg C yr$^{-1}$) and outputs to the atmosphere (4.2 $\pm$ 1.7 Tg C yr$^{–1}$) and ocean (2.9 $\pm$ 2.3 Tg C yr$^{–1}$).
-As such, small changes in either net primary production or heterotrophic respiration can have a large impact on the difference between the two, net ecosystem metabolism (NEM, also referred to as net ecosystem production and net community production), which ultimately governs the form and quantity of estuarine materials that are lost to the atmosphere, the sediments, and the ocean.
+As such, small changes in either net primary production or heterotrophic respiration can have a large impact on the difference between the two, net ecosystem metabolism (NEM), which ultimately governs the form and quantity of estuarine materials that are lost to the atmosphere, the sediments, and the ocean.
 
-Methods for estimating metabolism in aquatic environments are numerous [@kemp2011; @staehr2011].
-Chief among these are open-water methods [@Odum56], which exploit the diel cycle of dissolved oxygen and allow the simultaneous determination of whole-system gross primary production (P, equals net primary production + autotrophic respiration), ecosystem respiration (R, equals heterotrophic + autotrophic respiration), and NE (equals P-R).
-A global synthesis of warm-season, open-water metabolism studies (Hoellein et al., 2013) was based on a total of 47 estimates, almost all of which were derived from a single study of U.S.
-estuaries [@caffrey2004]; a similar synthesis by @yvon-durocher2012 was exclusively based on @caffrey2004.
-As such, a unified model to explain the large variability of metabolic rates within and across estuaries is clearly lacking as a critical tool for understanding metabolic processes.
-Such a model should be informed by the available data at a given location and have the ability to incorporate prior information on key parameters if available.
-Bayesian techniques have not been fully leveraged in estuarine applications despite their potential to address the above needs.
-A potentially useful application developed for streams is the BAyesian Single-station Estimation method....
+Methods for estimating metabolism in aquatic environments are numerous and include open-water techniques, bottle-based incubations, ecoystem budgets, oxygen isotopes, and use of inert gases [@kemp2011; @staehr2011].
+Based on the variety of assumptions and limitations of each method, intercomparisons of metabolic rates between ecosystem has been challenging.
+For example, an assessment in Randers Fjord (Denmark) showed agreement in terms of sign and magnitude among four different approaches [@gazeau2005].
+However, GPP and ER estimates from incubations were consistently lower than those from open-water methods, consistent with the estuarine respiration synthesis of @hopkinson2005.
+Even within one technique, there are numerous choices to be made that alter the outcome.
+For example, in the commonly used 14C technique for measuring primary production, @cloern2014 note that investigators have made divergent choices with regard to filtering mesozooplankton grazers, measuring 14C in dissolved organic carbon, accounting for changing spectral light quality with depth, and choosing the euphotic zone depth, the incubation time, and the incubation location (*in situ* vs.
+on-deck).
+These challenges are compounded in estuaries where high temporal and spatial variability have been noted in global syntheses of primary production and respiration.
+@cloern2014 found that annual phytoplankton primary production can vary 10-fold within an estuary and up to 5-fold from year-to-year.
+@hopkinson2005 focused on cross-system variations and found benthic and pelagic respiration rates varying by more than factors of 30 and 40, respectively, ranges that are broadly consistent with whole system respiration rates based on the open-water method [@hoellein2013].
 
-Application of open-water techniques for estimating metabolism is especially problematic based on the assumptions and poor error constraints of existing methods that rely on continuous measurements of dissolved oxygen.
-Fundamental to these challenges is the high temporal and spatial variability noted in global syntheses of primary production and respiration.
-Cloern et al.
-(2014) found that annual phytoplankton primary production can vary 10-fold within an estuary and up to 5-fold from year-to-year.
-Hopkinson and Smith (2005) focused on cross-system variations and found benthic and pelagic respiration rates varying by more than factors of 30 and 40, respectively, ranges that are broadly consistent with wholesystem respiration rates based on the open-water method (Hoellein et al., 2013).
-Tidal variation...
+Advances in sensor technology for continuous collection of dissolved oxygen time series suggest that the open-water method [@Odum56] could be leveraged for consolidation of approaches.
+This approach exploits the diel cycle of dissolved oxygen and allow the simultaneous determination of whole-system gross primary production (P, equals net primary production + autotrophic respiration), ecosystem respiration (R, equals heterotrophic + autotrophic respiration), and NEM (equals P-R).
+A global synthesis of warm-season, open-water metabolism studies [@hoellein2013] was based on a total of 47 estimates, almost all of which were derived from a single study of U.S.
+estuaries [@caffrey2004].
+A similar synthesis was done by @yvon-durocher2012.
+These studies evaluated the open-water method as applied to continuous monitoring data available for several decades across the National Estuarine Research Reserve System (NERRS).
+Thus, the approach can be broadly applied where data are available, although it is not without limitations.
+In particular, tidal advection can violate assumptions of the method that the measurements are from a parcel of water with a continuous metabolic history [@kemp1980].
+Striking evidence of this challenge is the large number of physically impossible negative GPP and ER estimates in the open-water analysis of 28 3-year time series by @caffrey2003.
+The computed rates were negative 23% of the time, on average, and as high as 69% for individual sites.
+Estimating gas exchange using the open-water method in estuaries has also been challenging.
+Most approaches estimate gas exchange using empirical formula dependent on wind speed and temperature or held at a constant value [e.g., @caffrey2004; @russell2007; @nagel2009; @Beck15; @Murrell18].
+Gas exchange is a first-order process in the estuarine dissolved oxygen budget and, therefore, inaccurate estimation can lead to poorly constrained errors that influence interpretation of results.
+
+A unified model to explain the large variability of metabolic rates within and across estuaries is clearly lacking as a critical tool for understanding metabolic processes.
+Bayesian techniques have not been fully explored in estuarine applications despite their potential to address the above [although see @ciavatta2008].
+Bayesian techniques allow for incorporation of prior knowledge about model parameters and robust estimation of parameter probability distributions [@Holtgrieve10] using highly efficient computational algorithms [@Grace15; @winslow2016].
+These techniques are very powerful because they not only estimate GPP and ER as standard open-water methods do, but they retrieve estimates and their uncertainties of gas exchange, GPP, and ER parameters (e.g., the initial slope of the photosynthesis--irradiance curve), which reveal quantitative information about the mechanisms of metabolism and gas exchange.
+A potentially useful application developed for streams is the BAyesian Single-station Estimation (BASE) method [@Grace15].
+BASE has been demonstrated to accurately determine metabolic parameters and their uncertainty using continuous monitoring data that are similar to those available at many estuarine locations.
+As such, modification of BASE to include model characterizations more appropriate for estuaries, while maintaining the fundamental Bayesian approach, could prove useful.
 
@@
 Our results have demonstrated that unknown metabolic parameters can be estimated with reasonable certainty using uninformed prior distributions and EBASE could be a valuable approach to improve the understanding of critical ecosystem processes as the method can be readily applied to other settings.
-A notable limitation of EBASE was an inability to precisely recover the $b$ parameter from a synthetic time series, although this result may be explained by a minimal contribution of gas exchange as expressed in the dissolved oxygen time series.
+A notable limitation of EBASE was an inability to precisely recover the $b$ parameter from a synthetic time series, although this result may be explained by a minimal contribution of gas exchange as expressed in the synthetic dissolved oxygen time series.
 Effects of tidal advection on dissolved oxygen measurements were also observed, demonstrating a violation of the assumption that the measured data represent a water sample having a continuous metabolic history.

### 2023-05-16 manu-draft.qmd
proofing intro and discussion, added acknowledgements and key words

@@
 
-*Key words*:
+*Key words*: Bayesian, dissolved oxygen, ecosystem metabolism, estuary, gas exchange, open source
 
@@
 Methods for estimating metabolism in aquatic environments are numerous and include open-water techniques, bottle-based incubations, ecoystem budgets, oxygen isotopes, and use of inert gases [@kemp2011; @staehr2011].
-Based on the variety of assumptions and limitations of each method, intercomparisons of metabolic rates between ecosystem has been challenging.
-For example, an assessment in Randers Fjord (Denmark) showed agreement in terms of sign and magnitude among four different approaches [@gazeau2005].
-However, GPP and ER estimates from incubations were consistently lower than those from open-water methods, consistent with the estuarine respiration synthesis of @hopkinson2005.
+Based on the variety of assumptions and limitations of each method, comparisons of metabolic rates within and between ecosystems hav been challenging.
+For example, an assessment in Randers Fjord (Denmark) showed agreement in both the sign and magnitude among four different approaches [@gazeau2005].
+However, gross primary production (P) and ecosystem respiration (R) estimates from incubations were consistently lower than those from open-water methods, consistent with the estuarine respiration synthesis of @hopkinson2005.
 Even within one technique, there are numerous choices to be made that alter the outcome.
-For example, in the commonly used 14C technique for measuring primary production, @cloern2014 note that investigators have made divergent choices with regard to filtering mesozooplankton grazers, measuring 14C in dissolved organic carbon, accounting for changing spectral light quality with depth, and choosing the euphotic zone depth, the incubation time, and the incubation location (*in situ* vs.
+For example, in the commonly used ${}^{14}$C technique for measuring primary production, @cloern2014 noted that investigators have made divergent choices with regard to filtering mesozooplankton grazers, measuring ${}^{14}$C in dissolved organic carbon, accounting for changing spectral light quality with depth, and choosing the euphotic zone depth, the incubation time, and the incubation location (*in situ* vs.
 on-deck).
-These challenges are compounded in estuaries where high temporal and spatial variability have been noted in global syntheses of primary production and respiration.
+These challenges are further compounded in estuaries where high temporal and spatial variability have been noted in global syntheses of P and R.
 @cloern2014 found that annual phytoplankton primary production can vary 10-fold within an estuary and up to 5-fold from year-to-year.
@@
 
-Advances in sensor technology for continuous collection of dissolved oxygen time series suggest that the open-water method [@Odum56] could be leveraged for consolidation of approaches.
-This approach exploits the diel cycle of dissolved oxygen and allow the simultaneous determination of whole-system gross primary production (P, equals net primary production + autotrophic respiration), ecosystem respiration (R, equals heterotrophic + autotrophic respiration), and NEM (equals P-R).
+Advances in sensor technology for continuous collection of dissolved oxygen time series suggest that the open-water method [@Odum56] could be leveraged for applications across multiple locations.
+This approach exploits the diel cycle of dissolved oxygen and allows the simultaneous determination of whole-system gross primary production (net primary production + autotrophic respiration), ecosystem respiration (equals heterotrophic + autotrophic respiration), and NEM (equals P-R).
 A global synthesis of warm-season, open-water metabolism studies [@hoellein2013] was based on a total of 47 estimates, almost all of which were derived from a single study of U.S.
@@
 Thus, the approach can be broadly applied where data are available, although it is not without limitations.
-In particular, tidal advection can violate assumptions of the method that the measurements are from a parcel of water with a continuous metabolic history [@kemp1980].
-Striking evidence of this challenge is the large number of physically impossible negative GPP and ER estimates in the open-water analysis of 28 3-year time series by @caffrey2003.
+In particular, tidal advection can violate assumptions of the method that the measurements are from the same parcel of water with a continuous metabolic history [@kemp1980].
+Striking evidence of this challenge is the large number of physically impossible negative P and R estimates in the open-water analysis of 28 3-year time series by @caffrey2003.
 The computed rates were negative 23% of the time, on average, and as high as 69% for individual sites.
@@
 Most approaches estimate gas exchange using empirical formula dependent on wind speed and temperature or held at a constant value [e.g., @caffrey2004; @russell2007; @nagel2009; @Beck15; @Murrell18].
-Gas exchange is a first-order process in the estuarine dissolved oxygen budget and, therefore, inaccurate estimation can lead to poorly constrained errors that influence interpretation of results.
+Gas exchange is a first-order process in the estuarine dissolved oxygen budget and, therefore, inaccurate estimations based on empirical formula or fixed values can lead to poorly constrained errors that influence interpretation of results.
 
@@
 Bayesian techniques allow for incorporation of prior knowledge about model parameters and robust estimation of parameter probability distributions [@Holtgrieve10] using highly efficient computational algorithms [@Grace15; @winslow2016].
-These techniques are very powerful because they not only estimate GPP and ER as standard open-water methods do, but they retrieve estimates and their uncertainties of gas exchange, GPP, and ER parameters (e.g., the initial slope of the photosynthesis--irradiance curve), which reveal quantitative information about the mechanisms of metabolism and gas exchange.
+These techniques are very powerful because they not only estimate P and R as standard open-water methods do, but they retrieve estimates and their uncertainties of gas exchange, P, and R parameters (e.g., the initial slope of the photosynthesis--irradiance curve), which reveals quantitative information about the mechanisms of metabolism and gas exchange.
 A potentially useful application developed for streams is the BAyesian Single-station Estimation (BASE) method [@Grace15].
@@
 
-The goal of this paper is to describe the development and application of a new data assimilation method for simultaneously determining gross primary production, ecosystem respiration, net ecosystem metabolism (NEM), and the gas exchange (D) in estuaries from high-resolution dissolved oxygen time series.
+The goal of this paper is to describe the development and application of a new Bayesian method for simultaneously determining gross primary production, ecosystem respiration, net ecosystem metabolism, and the gas exchange (D) in estuaries from high-resolution dissolved oxygen time series.
 We call this method EBASE (Estuarine BAyesian Single-station Estimation), which builds and improves on the BASE method [@Grace15] using an approach more appropriate for estuarine time series.
@@
 Our results demonstrated that even using uninformed priors, EBASE can reproduce known metabolic parameters and rates with reasonable certainty.
-However, an exception was the ability of the model to estimate stable values for the $b$ parameter, although the resulting estimate of gas exchange (D) was similar to the known values.
+However, an exception was the ability of the model to estimate stable values for the $b$ parameter, although the resulting estimate of gas exchange was similar to the known values.
 This result could be explained by the dominance of the relative processes that manifest signals in the dissolved oxygen time series.
 That is, estuarine dissolved oxygen at Apalachicola Bay may be dominated by production and respiration, and less so from gas exchange, such that the available data provided insufficient information for EBASE to estimate $b$.
-This result aligns with others that have suggested gas exchange may be a relatively small component of the metabolic signal in shallow estuaries [@Murrell18].
-In such cases, more constrained prior distributions for $b$ could be appropriate or EBASE could be tested in settings with larger contributions of gas exchange expressed by dissolved oxygen.
+This result aligns with others that have suggested gas exchange may be a relatively small component of the metabolic signal in shallow estuaries of the Gulf of Mexico [@Murrell18].
+In such cases, more constrained prior distributions for $b$ could be appropriate or EBASE could be tested in settings with expected larger contributions of gas exchange expressed by dissolved oxygen.
 
@@
 Finally, the gas exchange in EBASE used the parameterization in @Wanninkhof14 that is primarily wind-based and developed for the open-ocean, which is clearly more appropriate than the prior formulation in BASE that includes a temperature and depth dependency for streams.
-Overall, these changes also reduce the required inputs and many monitoring locations nationwide include the requisite data for applying EBASE (i.e., dissolved oxygen, water temperature, salinity, PAR, and wind speed).
+Overall, these changes also reduced the required inputs and many monitoring locations nationwide include the requisite data for applying EBASE (i.e., dissolved oxygen, water temperature, salinity, PAR, and wind speed, as collected by the NERRS).
 
@@
 A potential future approach for developing EBASE could be an explicit formulation of tidal advection in the model using a synoptic tidal height time series in the Bayesian framework.
-This has the advantage of both detiding and estimating metabolic rates using the same model, whereas using weighted regression prior to EBASE as used herein, requires additional analysis and assumptions.
+This has the advantage of simultaneously detiding and estimating metabolic rates using the same model, whereas using weighted regression prior to EBASE, as used herein, requires additional analysis and assumptions.
 However, field measurements [e.g., control volume approaches, @falter2008; @nidzieko2014] may be required to provide reasonable constraints on the prior distributions for any parameters that quantify advective influences on dissolved oxygen.
@@
 As such, using one week of data for each optimization period is likely appropriate for this location.
-Using one day is not recommended given the issues stated above (i.e., unstable results), whereas an optimization period longer than a month may miss important events in the time series that can affect the metabolic estimates.
+For other locations, using one day is not recommended given the issues stated above (i.e., unstable results), whereas an optimization period longer than a month may miss important events in the time series that can affect the metabolic estimates and will likely produce parameters that are overly generalized.
 
@@
 This allows the estimation of metabolic rates for time series that cover relatively long periods of time (e.g., more than year), when data gaps are likely to occur for several reasons (e.g., equipment malfunction, routine maintenance, etc.).
-The R package accommodates data gaps using linear interpolation for any required parameter used by EBASE.
+The R package accommodates data gaps using linear interpolation for any required input used by EBASE.
 This allows the Bayesian routine to provide continuous estimates across the time series.
-The results are then post-processed such that any metabolic estimates that exceed a threshold for interpolated observations are removed from the output.
-The analyst can determine the appropriate length of time to use for excluding results based on interpolated values.
-The default setting for EBASE is 12 hours, where results for an entire optimization period are removed if any interpolated period exceeded 12 hours within that period.
+The results are then automatically post-processed to remove any metabolic estimates that exceed a threshold for the number of interpolated observations.
+The analyst can determine the appropriate length of time to use for excluding results based, although the default setting for EBASE is 12 hours, where results for an entire optimization period are removed if any interpolated period exceeded 12 hours within that period.
 Although this does not apply to the synthetic time series, some of the estimates using the 2021 observed data for Apalachicola Bay were excluded from the analysis.
-This feature allows EBASE to be applied more easily to other data setes because an analyst does not need to manually screen data that include missing values.
+This feature allows EBASE to be applied more easily to other data sets because an analyst does not need to manually screen data that include missing values prior to analysis.
 
@@
 This may require further constraining the priors, although doing so may produce biased results in the absence of additional data to justify a constraint.
-Alternative metabolic models could be used for informing priors, such as using the Odum method to inform the prior for the reareation coefficient, $K$, for the BASE method.
-Initial application of BASE to estuarine data demonstrated that unstable and unrealistic metabolic rates were returned with uninformed priors unless the $K$ parameter was constrained.
+Field-based measurements or alternative metabolic models could be used for informing priors.
+As an example, initial application of BASE to estuarine data prior to developing EBASE demonstrated that unstable and unrealistic metabolic rates were returned with uninformed priors.
+The Odum method was used on the same data to develop an appropriate expectation of the range of values for the reaeration coefficient, $K$ , which was then used to inform the prior distribution using the BASE method.
+More stable and realistic metabolic estimates were then returned by BASE, demonstrating the value of the Bayesian approach that can incorporate prior knowledge.
 A similar approach could be applied to create more accurate estimates from EBASE depending on location.
@@
 Our results have demonstrated that unknown metabolic parameters can be estimated with reasonable certainty using uninformed prior distributions and EBASE could be a valuable approach to improve the understanding of critical ecosystem processes as the method can be readily applied to other settings.
-A notable limitation of EBASE was an inability to precisely recover the $b$ parameter from a synthetic time series, although this result may be explained by a minimal contribution of gas exchange as expressed in the synthetic dissolved oxygen time series.
+A notable limitation of EBASE was an inability to precisely recover the $b$ parameter from a synthetic time series, although this result may be explained by a minimal contribution of gas exchange as expressed in the dissolved oxygen time series.
 Effects of tidal advection on dissolved oxygen measurements were also observed, demonstrating a violation of the assumption that the measured data represent a water sample having a continuous metabolic history.
@@
 ## Acknowledgments {.unnumbered}
+
+We thank Jason Garwood and Ethan Borque of the Apalachicola National Estuarine Research Reserve for assisting with the deployment and processing of data used herein.
+We also thank Wade McGillis for assisting in this effort.
+We thank Darren Giling for initial conversations on applying BASE to estuarine data.
+This research was funded by award 1924559 from the National Science Foundation to co-PIs Ray Najjar, Maria Herrmann, and Kathleen Hill.
 

### 2023-05-17 manu-draft.qmd
added JMA edits

@@
 Methods for estimating metabolism in aquatic environments are numerous and include open-water techniques, bottle-based incubations, ecoystem budgets, oxygen isotopes, and use of inert gases [@kemp2011; @staehr2011].
-Based on the variety of assumptions and limitations of each method, comparisons of metabolic rates within and between ecosystems hav been challenging.
+Based on the variety of assumptions and limitations of each method, comparisons of metabolic rates within and between ecosystems have been challenging.
 For example, an assessment in Randers Fjord (Denmark) showed agreement in both the sign and magnitude among four different approaches [@gazeau2005].
@@
 
-As a final evaluation of EBASE to return known metabolic parameters, a new time series was created that added estimated effects of instrument and tidal noise to the synthetic time series.
+As a final evaluation of EBASE to return known metabolic parameters, a second synthetic time series was created that added estimated effects of instrument and tidal noise to the original synthetic time series.
 The objective of the analysis is to develop an expectation of how metabolic estimates may change using in situ data with noise that may substantially mask a biological signal in dissolved oxygen measurements.
@@
 The instrument noise was estimated as the difference between the observed and predicted time series and the tidal noise was estimated as the difference between the predicted and detided time series.
-Both the instrument and tidal noise were added to the same synthetic time series above (@fig-synapanos).
+Both the instrument and tidal noise were added to the original synthetic time series above (@fig-synapanos).
 EBASE was then used with unconstrained priors (large $\sigma$) on the synthetic time series with noise, then compared with results from the original synthetic time series.
@@
 As for the previous results, the $b$ parameter was estimated with low precision for both input time series, although results were centered around the true value of 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$.
-Gas exchange (D) was similar between the input time series, suggesting lack of precision in the $b$ parameter regardless of noise had a minimal influence.
+Gas exchange (D) was similar between the input time series, suggesting that lack of precision in the $b$ parameter had a minimal influence regardless of noise.
 
@@
 The only exception for the correlation estimates was gas exchange (D), which had a slightly weaker correlation using the detided results.
-@fig-apacmpfig shows that the Odum results for gas exchange were more variable for the same values returned as no or small gas exchange for the EBASE method based on the detided dissolved oxygen.
+@fig-apacmpfig shows that the Odum results for gas exchange were more variable than those for EBASE using the detided dissolved oxygen, where the latter often estimated gas exhange at much smaller values compared to the former.
 The RMSE values of each regression comparing the metabolic estimates between methods were generally lower in the detided results, excluding D which was higher with the detided dissolved oxygen relative to the observed.
@@
 
-The second advantage of EBASE that makes the model more appropriate for estuarine settings included several modification to the underlying mass-balance equation for dissolved oxygen (@eq-ebase).
+The second advantage of EBASE that makes the model more appropriate for estuarine settings included several modifications to the underlying mass-balance equation for dissolved oxygen (@eq-ebase).
 As described in @sec-basemodif, EBASE is an estuarine adaptation of BASE (@eq-base) that was developed for stream applications and many assumptions of the underlying model of the latter were inappropriate for the former.
@@
 As an example, initial application of BASE to estuarine data prior to developing EBASE demonstrated that unstable and unrealistic metabolic rates were returned with uninformed priors.
-The Odum method was used on the same data to develop an appropriate expectation of the range of values for the reaeration coefficient, $K$ , which was then used to inform the prior distribution using the BASE method.
+The Odum method was used on the same data to develop an appropriate expectation of the range of values for the reaeration coefficient, $K$, which was then used to inform the prior distribution using the BASE method.
 More stable and realistic metabolic estimates were then returned by BASE, demonstrating the value of the Bayesian approach that can incorporate prior knowledge.

### 2023-07-24 manu-draft.qmd
replace r with R for all text, figs, and tables

@@
     number-sections: true
-title: "Fitting metabolic models to dissolved oxygen data: the Estuarine BAyesian Single-station Estimation method"
+title: "Fitting metabolic models to dissolved oxygen data: The Estuarine BAyesian Single-station Estimation method"
 bibliography: refs.bib
@@
 This paper describes the development, assessment, and interpretation of results from a new approach for estimating metabolic rates in estuaries, the Estuarine BAyesian Single-station Estimation (EBASE) method.
-EBASE provides several advantages over existing models by 1) using a linear relationship between irradiance and production, 2) removing the temperature-dependency of respiration, and 3) applying a wind-based gas exchange formula for open water systems.
-Moreover, the approach uses a Bayesian framework where critical model parameters are fit as a function of the data and informed using a set of prior distributions.
+EBASE provides several advantages over existing models by using an integration formula appropriate for estuarine systems and using a Bayesian framework where critical model parameters are fit as a function of the data as informed by a set of prior distributions.
+This approach can also vary the time period over which critical model parameters are estimated.
 We demonstrate that EBASE can reproduce known metabolic parameters from a synthetic time series created with continuous monitoring data from the Apalachicola estuarine reserve and provide a quantitative assessment of how the prior distributions affect the model results.
@@
 As materials are transported from land to ocean, estuaries profoundly transform or filter these materials through various processes, chief among which are primary production and respiration, collectively referred to as metabolism [@schubel1984].
-For example, estuarine respiration of organic carbon delivered from land produces carbon dioxide, which may be lost to the atmosphere via outgassing [e.g, @cai2011] .
-Similarly, estuarine primary production consumes dissolved inorganic carbon delivered from land, forming particulate organic carbon that may be buried in estuarine sediments [e.g., @hu2006] .
-Estuarine metabolic processes are fundamentally important, as illustrated by the dissolved inorganic carbon budget of eastern North American estuaries [@najjar2018]: net primary production (11.7 $\pm$ 3.0 Tg C yr$^{-1}$) and heterotrophic respiration (13.2 $\pm$ 3.1 Tg C yr$^{–1}$) far exceed inputs from upland sources (5.6 $\pm$ 1.2 Tg C yr$^{-1}$) and outputs to the atmosphere (4.2 $\pm$ 1.7 Tg C yr$^{–1}$) and ocean (2.9 $\pm$ 2.3 Tg C yr$^{–1}$).
-As such, small changes in either net primary production or heterotrophic respiration can have a large impact on the difference between the two, net ecosystem metabolism (NEM), which ultimately governs the form and quantity of estuarine materials that are lost to the atmosphere, the sediments, and the ocean.
+For example, estuarine respiration of organic carbon delivered from land produces carbon dioxide, which may be lost to the atmosphere via outgassing [e.g, @cai2011].
+Similarly, estuarine primary production consumes dissolved inorganic carbon delivered from land, forming particulate organic carbon that may be buried in estuarine sediments [e.g., @hu2006].
+Estuarine metabolic processes are fundamentally important, as illustrated by the dissolved inorganic carbon budget of eastern North American estuaries [@najjar2018]: net primary production (11.7 $\pm$ 3.0 Tg C yr$^{-1}$, best estimate $\pm$ 2 standard errors) and heterotrophic respiration (13.2 $\pm$ 3.1 Tg C yr$^{–1}$) far exceed inputs from upland sources (5.6 $\pm$ 1.2 Tg C yr$^{-1}$) and outputs to the atmosphere (4.2 $\pm$ 1.7 Tg C yr$^{–1}$) and ocean (2.9 $\pm$ 2.3 Tg C yr$^{–1}$).
+As such, small changes in either net primary production or heterotrophic respiration can have a large impact on the difference between the two, net ecosystem metabolism (NEM, sometimes also referred to as net ecosystem production), which ultimately governs the form and quantity of estuarine materials that are lost to the atmosphere, the sediments, and the ocean.
 
-Methods for estimating metabolism in aquatic environments are numerous and include open-water techniques, bottle-based incubations, ecoystem budgets, oxygen isotopes, and use of inert gases [@kemp2011; @staehr2011].
+Methods for estimating metabolism in aquatic environments are numerous and include open-water techniques, bottle-based incubations, ecosystem budgets, oxygen isotopes, and use of inert gases [@kemp2011; @staehr2011].
 Based on the variety of assumptions and limitations of each method, comparisons of metabolic rates within and between ecosystems have been challenging.
@@
 Advances in sensor technology for continuous collection of dissolved oxygen time series suggest that the open-water method [@Odum56] could be leveraged for applications across multiple locations.
-This approach exploits the diel cycle of dissolved oxygen and allows the simultaneous determination of whole-system gross primary production (net primary production + autotrophic respiration), ecosystem respiration (equals heterotrophic + autotrophic respiration), and NEM (equals P-R).
+This approach exploits the diel cycle of dissolved oxygen and allows the simultaneous determination of whole-system gross primary production (net primary production $+$ autotrophic respiration), ecosystem respiration (heterotrophic $+$ autotrophic respiration), and NEM (P $-$ R).
 A global synthesis of warm-season, open-water metabolism studies [@hoellein2013] was based on a total of 47 estimates, almost all of which were derived from a single study of U.S.
@@
 Most approaches estimate gas exchange using empirical formula dependent on wind speed and temperature or held at a constant value [e.g., @caffrey2004; @russell2007; @nagel2009; @Beck15; @Murrell18].
-Gas exchange is a first-order process in the estuarine dissolved oxygen budget and, therefore, inaccurate estimations based on empirical formula or fixed values can lead to poorly constrained errors that influence interpretation of results.
+Gas exchange is a first-order process in the estuarine dissolved oxygen budget and, therefore, inaccurate estimations based on empirical formulae or fixed values can lead to poorly constrained errors that influence interpretation of results.
 
@@
 Bayesian techniques allow for incorporation of prior knowledge about model parameters and robust estimation of parameter probability distributions [@Holtgrieve10] using highly efficient computational algorithms [@Grace15; @winslow2016].
-These techniques are very powerful because they not only estimate P and R as standard open-water methods do, but they retrieve estimates and their uncertainties of gas exchange, P, and R parameters (e.g., the initial slope of the photosynthesis--irradiance curve), which reveals quantitative information about the mechanisms of metabolism and gas exchange.
+These techniques are very powerful because they not only estimate P and R as standard open-water methods do, but they retrieve estimates and their uncertainties of parameters related to gas exchange, P, and R (e.g., the initial slope of the photosynthesis--irradiance curve), which reveals quantitative information about the mechanisms of metabolism and gas exchange.
 A potentially useful application developed for streams is the BAyesian Single-station Estimation (BASE) method [@Grace15].
@@
 $$
-H\frac{dC}{dt} = aPAR - r - \left[bU_{10}^2\left(\frac{Sc}{600} \right)^{-0.5} \left(C_{Sat} - C \right )\right]
+H\frac{dC}{dt} = aPAR - R - \left[bU_{10}^2\left(\frac{Sc}{600} \right)^{-0.5} \left(C_{Sat} - C \right )\right]
 $$ {#eq-ebase}
@@
 Production is modeled linearly as a function of surface irradiance ($PAR$, Watts m$^{-2}$) and $a$ (light efficiency, $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$).
-Respiration is modeled with the single parameter $r$ ($\text{mmol}/\text{m}^3/\text{d}$).
+Respiration is modeled with the single parameter $R$ ($\text{mmol}/\text{m}^3/\text{d}$).
 Gas exchange is modeled using the formulation in @Wanninkhof14 as a wind-based parameterization for the gas transfer velocity [@Ho16] multiplied by the oxygen concentration gradient as the difference between the concentration measured in the surface water layer, $C$ (mmol m$^{-3}$), and the saturation (or equilibrium) concentration, $C_{Sat}$ (mmol m$^{-3}$), calculated from salinity (psu) and temperature ($^\circ$C) [@Millero81; @Garcia92].
@@
 The $b$ parameter is estimated from the data and has units of $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$.
-As such, the three parameters estimated by EBASE are $a$, $r$, and $b$ using the required input data that includes time series of dissolved oxygen, water temperature, salinity, PAR, wind speed, and water column depth (`r run_reference(id = "tbl-ebaseiotab")`).
+As such, the three parameters estimated by EBASE are $a$, $R$, and $b$ using the required input data that includes time series of dissolved oxygen, water temperature, salinity, PAR, wind speed, and water column depth (`r run_reference(id = "tbl-ebaseiotab")`).
 
@@
 
-The remaining three parameters $a$ (light efficiency, $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$), $r$ (respiration, $\text{mmol}/\text{m}^3/\text{d}$), and $b$ ($\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$) are estimated by likelihood given the observed data and prior distributions for the parameters.
+The remaining three parameters $a$ (light efficiency, $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$), $R$ (respiration, $\text{mmol}/\text{m}^3/\text{d}$), and $b$ ($\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$) are estimated by likelihood given the observed data and prior distributions for the parameters.
 The JAGS software [as for BASE, @Plummer03] is used with the EBASE R package [described below, @Beck23; @RCT23] to estimate the unknown parameters using Markov Chain Monte Carlo (MCMC) simulations in a Bayesian framework.
@@
 Reasonable uninformed prior distributions for EBASE were chosen with mean values using approximate estimates from the literature [@Wanninkhof14; @Grace15] and standard deviations that were sufficiently large to allow the Bayesian routine to search an unconstrained parameter space.
-The default priors for $a$, $r$, and $b$ are:
+The default priors for $a$, $R$, and $b$ are:
 
@@
 $$
-r \sim N(20, 50)T(0,\infty) \;\; \text{mmol}/\text{m}^3/\text{d}
+R \sim N(20, 50)T(0,\infty) \;\; \text{mmol}/\text{m}^3/\text{d}
 $$ {#eq-rprior}
@@
 Users can also specify alternative prior distributions from those in Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior] for each parameter.
-Functions are available for viewing model results, including a plot of the modeled dissolved oxygen with the observed, a time series plot of the primary metabolic estimates (P, R, and D, as areal rates with units $\text{mmol}/\text{m}^2/\text{d}$), and a time series plot of the credible intervals for the $a$, $r$, and $b$ parameters.
+Functions are available for viewing model results, including a plot of the modeled dissolved oxygen with the observed, a time series plot of the primary metabolic estimates (P, R, and D, as areal rates with units $\text{mmol}/\text{m}^2/\text{d}$), and a time series plot of the credible intervals for the $a$, $R$, and $b$ parameters.
 An additional option allows the user to specify the model optimization period, where the time period can vary from a minimum of one day to the maximum of the entire length of the time series.
@@
 
-A preliminary assessment of the ability of EBASE to produce reasonable parameter estimates for $a$, $r$, and $b$ (@eq-ebase) was conducted using a synthetic time series with actual data as inputs.
+A preliminary assessment of the ability of EBASE to produce reasonable parameter estimates for $a$, $R$, and $b$ (@eq-ebase) was conducted using a synthetic time series with actual data as inputs.
 These assessments evaluated if 1) EBASE estimates similar values for the known parameters in the synthetic time series, and 2) how the estimated values change with different model configurations.
@@
 The input data were observations from the year 2021 at the Cat Point water quality monitoring station (29.7021$^{\circ}$ N, --84.8802$^{\circ}$ W) and included 15-minute observations of water temperature and salinity.
-Continuous meteorological observations from the nearby East Bay station (29.7909$^{\circ}$ N, --84.8834$^{\circ}$ W, \~10km north of Cat Point) were combined with the water quality data at the same time step and included air temperature, wind speed, and PAR.
+Continuous meteorological observations from the nearby East Bay station (29.7694$^{\circ}$ N, --84.8815$^{\circ}$ W, \~10 km north of Cat Point) were combined with the water quality data at the same time step and included air temperature, wind speed, and PAR.
 Missing observations in the water quality and meteorological data were filled using autoregressive modelling of the actual parameters versus time to create a complete dataset from January 1st to December 31st.
-A synthetic oxygen time series was then generated using the inputs from Apalachicola Bay ([@fig-synapa]a) and @eq-ebase as a forward calculating model with user-defined values for $a$, $r$, and $b$.
+A synthetic oxygen time series was then generated using the inputs from Apalachicola Bay ([@fig-synapa]a) and @eq-ebase as a forward calculating model with user-defined values for $a$, $R$, and $b$.
 All metabolic rates for production, respiration, and gas exchange ([@fig-synapa]b and c) were also estimated with the synthetic time series.
@@
 
-The optimization period used for a time series determines how much data are used to estimate the three EBASE parameters ($a$, $r$, and $b$).
+The optimization period used for a time series determines how much data are used to estimate the three EBASE parameters ($a$, $R$, and $b$).
 Unlike BASE, the EBASE software allows flexibility in the period of time used for optimization, with a minimum being one day and a maximum being the length of the time series.
-For a chosen optimization period, single estimates for $a$, $r$ and $b$ are returned by the model.
+For a chosen optimization period, single estimates for $a$, $R$, and $b$ are returned by the model.
 A short period of optimization (e.g., one day) may return unstable parameter estimates if forcing factors (e.g., wind, temperature) that affect estuarine biogeochemical rates occur at time scales longer the chosen period of optimization for EBASE.
@@
 Overall, most models regardless of optimization period or priors were able to reproduce the dissolved oxygen time series.
-The only models where the NSE values were less than zero for dissolved oxygen were those where $\sigma$ of the $r$ parameter was low.
+The only models where the NSE values were less than zero for dissolved oxygen were those where $\sigma$ of the $R$ parameter was low.
 However, substantial variation in NSE values was observed among all other parameters.
@@
 The patterns suggest that EBASE results are more sensitive to changes in the $\sigma$ parameter as shown by the larger spread in the median results compared to those for $\mu$.
-This result is especially true for the priors of the $a$ and $r$ parameters.
+This result is especially true for the priors of the $a$ and $R$ parameters.
 Conversely, model results are less sensitive to low or high values for $\mu$ as the spread between median NSE values is less than those for $\sigma$.
@@
 The latter comparison provided an assessment of metabolic rates that are expected to be minimally influenced by tidal advection, as a fundamental assumption of metabolic models using *in situ* dissolved oxygen time series from fixed monitoring stations.
-The prior distributions for EBASE were uninformed using the same large $\sigma$ values from @sec-cmpmod and moderate $\mu$ values ($\mu$ = 0.2 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$ for $a$, $\mu$ = 20 $\text{mmol}/\text{m}^3/\text{d}$ for $r$, $\mu$ = 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ for $b$).
+The prior distributions for EBASE were uninformed using the same large $\sigma$ values from @sec-cmpmod and moderate $\mu$ values ($\mu$ = 0.2 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$ for $a$, $\mu$ = 20 $\text{mmol}/\text{m}^3/\text{d}$ for $R$, $\mu$ = 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ for $b$).
 A one day model optimization period was used for EBASE for comparability with the Odum method.
@@
 
-![The default prior distributions for $a$, $r$, and $b$ used in EBASE.
+![The default prior distributions for $a$, $R$, and $b$ used in EBASE.
 All priors are normal distributions and truncated to positive values.

### 2023-07-24 R/figs.R

@@ -44,7 +44,7 @@ fwdatinp <- fwdat %>%
 
 # prior distribution plot ---------------------------------------------------------------------
 
-p <- prior_plot(n = 2e6)
+p <- prior_plot(n = 3e6)
 
 png(here('figs/priorplot.png'), height = 3, width = 8, family = 'serif', units = 'in', res = 500)
 print(p)
@@ -146,12 +146,12 @@ dev.off()
 
 # synthetic with noise ------------------------------------------------------------------------
 
-# load apadb data run through wtreg
-fl <- paste0(tempdir(), '/apadbdtd.RData')
-download.file('https://github.com/fawda123/BASEmetab_script/raw/master/data/apadbdtd.RData', destfile = fl)
+# load apacp data run through wtreg
+fl <- paste0(tempdir(), '/apacpdtd.RData')
+download.file('https://github.com/fawda123/BASEmetab_script/raw/master/data/apacpdtd.RData', destfile = fl)
 load(file = fl)
 
-nosdat <- apadbdtd %>% 
+nosdat <- apacpdtd %>% 
   mutate(
     tidnoise = DO_prd - DO_nrm, 
     obsnoise = DO_obs - DO_prd

### 2023-07-24 R/funcs.R

@@ -144,7 +144,7 @@ priorcomp <- function(dat, met, topbot = 3){
     mutate(
       var = factor(var, 
                    levels = c('amean', 'asd', 'rmean', 'rsd', 'bmean', 'bsd'), 
-                   labels = c('italic(a)~mu', 'italic(a)~sigma', 'italic(r)~mu', 'italic(r)~sigma', 'italic(b)~mu', 'italic(b)~sigma')), 
+                   labels = c('italic(a)~mu', 'italic(a)~sigma', 'italic(R)~mu', 'italic(R)~sigma', 'italic(b)~mu', 'italic(b)~sigma')), 
       fac = ''
     )
   
@@ -252,7 +252,7 @@ priorsumcomp <- function(dat, met = 'r2'){
            grandmed = median(medv),
            medv = medv - grandmed,
            param = gsub('mean$|sd$', '', prior),
-           param = factor(param, levels = c('a', 'r', 'b')),
+           param = factor(param, levels = c('a', 'r', 'b'), labels = c('a', 'R', 'b')),
            prior = gsub('^a|^r|^b', '', prior), 
            prior = factor(prior, levels = c('mean', 'sd'), labels = c('mu', 'sigma')), 
            ndays = factor(ndays, levels = c('1 day', '7 days', '30 days'), labels = c('1~day', '7~days', '30~days'))

### 2023-07-24 R/tabs.R

@@ -36,7 +36,7 @@ ebaseiotab <- flextable(totab) %>%
   flextable::compose(i = 11, j = 3, value = as_paragraph(as_equation('C_{sat}', props = eqsz))) %>% 
   flextable::compose(i = 13, j = 3, value = as_paragraph(as_equation('C_{mod}', props = eqsz))) %>% 
   flextable::compose(i = 14, j = 3, value = as_paragraph(as_equation('P,\\,aPAR', props = eqsz))) %>% 
-  flextable::compose(i = 15, j = 3, value = as_paragraph(as_equation('R,\\,r', props = eqsz))) %>% 
+  flextable::compose(i = 15, j = 3, value = as_paragraph(as_equation('R', props = eqsz))) %>% 
   flextable::compose(i = 16, j = 3, value = as_paragraph(as_equation('D,\\,\\frac{1}{H}\\left[-bU_{10}^2\\left(\\frac{s_c}{600} \\right)^{-0.5} \\left(C_{Sat} - C_d\\right )\\right]', props = eqsz))) %>% 
   flextable::compose(i = 17, j = 3, value = as_paragraph(as_equation('a', props = eqsz))) %>% 
   flextable::compose(i = 18, j = 3, value = as_paragraph(as_equation('b', props = eqsz))) %>% 

### 2023-07-26 manu-draft.qmd
more revisions

@@
 $$
-H\frac{dC}{dt} = aPAR - R - \left[bU_{10}^2\left(\frac{Sc}{600} \right)^{-0.5} \left(C_{Sat} - C \right )\right]
+H\frac{dC}{dt} = aPAR - R - k_w\left(C_{Sat} - C \right )
 $$ {#eq-ebase}
 
-where the terms on the right side of the equation are gross primary production (P or $aPAR$), ecosystem respiration (R), and gas exchange (D, remainder), respectively, modeled as a function of the change in dissolved oxygen ($dC$) per unit time ($dt$) of the input data.
-Production is modeled linearly as a function of surface irradiance ($PAR$, Watts m$^{-2}$) and $a$ (light efficiency, $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$).
-Respiration is modeled with the single parameter $R$ ($\text{mmol}/\text{m}^3/\text{d}$).
-Gas exchange is modeled using the formulation in @Wanninkhof14 as a wind-based parameterization for the gas transfer velocity [@Ho16] multiplied by the oxygen concentration gradient as the difference between the concentration measured in the surface water layer, $C$ (mmol m$^{-3}$), and the saturation (or equilibrium) concentration, $C_{Sat}$ (mmol m$^{-3}$), calculated from salinity (psu) and temperature ($^\circ$C) [@Millero81; @Garcia92].
-The gas transfer velocity is modeled using wind speed squared at 10 meters above the water ($U_{10}^2$, m$^2$ s$^{-2}$), water column depth ($H$ in meters, as a fixed parameter or time series from a pressure sensor), and the Schmidt number ($Sc$) defined as the ratio of the kinematic viscosity of water to the molecular diffusivity of the gas for oxygen [from water temperature and salinity using the polynomial fit in @Wanninkhof14].
+The parameter $k_w$ is the gas transfer velocity defined as:
+
+$$
+k_w = bU_{10}^2\left(\frac{Sc}{600} \right)^{-0.5}
+$$ {#eq-kw}
+
+For @eq-ebase, the terms on the right side of the equation are gross primary production (P, $aPAR$), ecosystem respiration (R), and gas exchange (D, remainder), respectively, processes that result in the change in dissolved oxygen ($dC$) per unit time ($dt$); $H$ is the water depth.
+Production is modeled linearly as a function of surface irradiance ($PAR$, $\text{W}/\text{m}^2$) and $a$ (light efficiency, $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$).
+Respiration is modeled with the single parameter $R$ ($\text{mmol}/\text{m}^2/\text{d}$).
+Gas exchange is modeled using the formulation in @Wanninkhof14 as a wind-based parameterization for the gas transfer velocity [@Ho16] multiplied by the oxygen concentration gradient as the difference between the concentration measured in the surface water layer, $C$ ($\text{mmol}/\text{m}^3$) , and the saturation (or equilibrium) concentration, $C_{Sat}$ ($\text{mmol}/\text{m}^3$) calculated from salinity (psu) and temperature ($^\circ$C) [@Millero81; @Garcia92].
+The gas transfer velocity, $k_w$ in @eq-kw, is modeled using wind speed squared at 10 meters above the water ($U_{10}^2$, $\text{m}^2/\text{s}^2$) and the Schmidt number ($Sc$), which is the ratio of the kinematic viscosity of water to the molecular diffusivity of oxygen in water [$S_c$ is computed from water temperature and salinity using the polynomial fit in @Wanninkhof14].
 The $b$ parameter is estimated from the data and has units of $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$.

### 2023-07-27 manu-draft.qmd
more edits from rn

@@
 For @eq-ebase, the terms on the right side of the equation are gross primary production (P, $aPAR$), ecosystem respiration (R), and gas exchange (D, remainder), respectively, processes that result in the change in dissolved oxygen ($dC$) per unit time ($dt$); $H$ is the water depth.
-Production is modeled linearly as a function of surface irradiance ($PAR$, $\text{W}/\text{m}^2$) and $a$ (light efficiency, $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$).
+Production is modeled linearly as a function of surface irradiance (PAR, $\text{W}/\text{m}^2$) and $a$ (light efficiency, $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$).
 Respiration is modeled with the single parameter $R$ ($\text{mmol}/\text{m}^2/\text{d}$).
@@
 
-The EBASE approach was based on several modifications to BAyesian Single-station Estimation (BASE) the mass balance equation described in @Grace15:
-
-$$
-\frac{dC}{dt} = aPAR^p - r\Theta^{T - \bar{T}} + \left[K\cdot1.0241^{T-\bar{T}}\left(C_{sat} - C \right )\right]
-$$ {#eq-base}
-
 ### Changes to BASE {#sec-basemodif}
 
-A common theory behind most gas exchange parameterizations for metabolism estimates is that the transfer of gases of relatively low solubility, like oxygen, is proportional to the concentration gradient in a thin film of water at the air--water interface.
-It is also assumed that the water at the top of the film is saturated with respect to the atmosphere, so that the concentration difference across the thin film can be approximated as the difference between oxygen concentration measured in the surface water layer, $C$, and the saturation (or equilibrium) concentration $C_{Sat}$, which is usually calculated as a function of water temperature and salinity (the negative sign is needed because positive flux is set to be from water to air):
-
-$$
-D = -k_w \left(C_{sat}-C \right).
-$$ {#eq-gasex}
-
-The gas transfer velocity, $k_w$, has dimensions of length over time and increases as the water column in the vicinity of the air--water interface becomes more turbulent.
-As such, $k_w$ is usually modeled as a function of measurable variables that can potentially predict turbulent mixing of the surface water, e.g., wind speed, current velocity, and water depth [@Ho16].
-$k_w$ also increases with temperature because the film of water at the air-sea interface becomes more thin as the random motion of gas and water molecules increases at higher temperatures and transfer of molecules across the film increases.
+@eq-ebase is similar to the oxygen mass balance equation in BASE [@Grace15], with the main difference being the treatment of gas exchange.
+The gas transfer velocity, $k_w$, increases as the water column in the vicinity of the air--water interface becomes more turbulent.
+As such, $k_w$ is usually modeled as a function of measurable variables that can potentially predict turbulent mixing of the surface water, e.g., wind speed, current velocity, and water depth [@Ho16]; fetch and suspended solids have also been considered for estuarine systems [@Borges11].
+$k_w$ also increases with temperature because the film of water at the air-sea interface becomes thinner as the viscosity decreases and the random motion of gas molecules increases.
 This dependence of $k_w$ on temperature is usually modeled with the Schmidt number as in @eq-ebase, which combines information on the viscosity of the fluid and the diffusivity of the gas.
@@
 An expectation is that $k_w$ would increase as water depth decreases for a given current speed because turbulence generated at the sediment--water interface by currents would more easily reach the surface [@Ho16].
-Based on these differences, the wind-based parameterization for gas exchange was used for EBASE [@Wanninkhof14].
-Additionally, $C_{sat}$ was calculated using only temperature and salinity, removing the need for barometric pressure used in BASE.
+Based on these differences and the lack of availability of other possible drivers of the transfer velocity in coastal systems (e.g., currents, fetch, suspended solids, and wave height), the wind-based parameterization for gas exchange was used for EBASE [@Wanninkhof14].
+However, this parameterization is largely based on open-ocean data and so we allow $b$, which was esimated by @Wanninkhof14 to be 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$, to be a free parameter that may reflect spatial and temporal variability in factors other than wind and temperature that may influence transfer velocity.
 
-The additional modifications to @eq-base for EBASE were removing the $p$ and $\Theta$ parameters from the production (P) and respiration (R) components, respectively.
-The BASE method provides options to estimate both $\Theta$ and $p$ from the data or to set as fixed values defined in the model file ($\Theta = 1.07177$, $p=1$).
-For EBASE, the temperature dependency of respiration is removed by setting $\Theta^{(T-\bar{T})}$ = 1 and it is assumed that respiration is constant as in other methods [@Odum56; @Murrell18].
-The $p$ parameter, which is dimensionless, is also set to 1 as in @Holtgrieve10, such that a linear relationship of production to $PAR$ was used.
-This removes the exponential dependence of production with increasing $PAR$.
-Removing these components also improved model stability by creating more reasonable estimates for the other parameters.
+BASE was also modified by removing the temperature dependence of respiration and the potential for $P$ to depend nonlinearly on PAR.
+While there is evidence for both dependencies in aquatic systems, we found, in preliminary model runs, that model stability was enhanced when these dependencies were removed.
+It would be straightforward to include such dependencies in EBASE as our understanding of them improve and as data become available that will allow them to be robustly constrained.
 
@@
 The second advantage of EBASE that makes the model more appropriate for estuarine settings included several modifications to the underlying mass-balance equation for dissolved oxygen (@eq-ebase).
-As described in @sec-basemodif, EBASE is an estuarine adaptation of BASE (@eq-base) that was developed for stream applications and many assumptions of the underlying model of the latter were inappropriate for the former.
+As described in @sec-basemodif, EBASE is an estuarine adaptation of BASE [@Grace15] that was developed for stream applications and many assumptions of the underlying model of the latter were inappropriate for the former.
 Specifically, EBASE was modified to address three primary concerns of using BASE in an estuarine setting.

### 2023-07-28 manu-draft.qmd
text edits for default prior equations

@@
 At each time step, the change in oxygen concentration between time steps is calculated from the equation using model inputs and parameter guesses based on the prior distributions, and then a finite difference approximation is used to estimate modeled oxygen concentration.
-Estimations proceed using a forward in time integration where the dissolved oxygen at each time step is informed by parameter estimates from the preceding time steps.
+Estimations proceed using a forward-in-time integration where the dissolved oxygen at each time step is informed by parameter estimates from the preceding time steps.
 The estimated dissolved oxygen concentration is also returned, which can be compared to the observed values as one measure of model performance.
@@
 $$
-a \sim N(0.2, 1)T(0,\infty) \;\; \left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)
+a \sim N(4, 1)T(0,\infty) \;\; \left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)
 $$ {#eq-aprior}
@@
 $$
-R \sim N(20, 50)T(0,\infty) \;\; \text{mmol}/\text{m}^3/\text{d}
+R \sim N(300, 50)T(0,\infty) \;\; \text{mmol}/\text{m}^2/\text{d}
 $$ {#eq-rprior}

### 2023-08-01 manu-draft.qmd
added text about mean prior justification, added table with caffrey 2004 natl estimates

@@
 @eq-ebase is similar to the oxygen mass balance equation in BASE [@Grace15], with the main difference being the treatment of gas exchange.
-The gas transfer velocity, $k_w$, increases as the water column in the vicinity of the air--water interface becomes more turbulent.
+The gas transfer velocity, $k_w$ (@eq-kw), increases as the water column in the vicinity of the air--water interface becomes more turbulent.
 As such, $k_w$ is usually modeled as a function of measurable variables that can potentially predict turbulent mixing of the surface water, e.g., wind speed, current velocity, and water depth [@Ho16]; fetch and suspended solids have also been considered for estuarine systems [@Borges11].
@@
 
-The remaining three parameters $a$ (light efficiency, $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$), $R$ (respiration, $\text{mmol}/\text{m}^3/\text{d}$), and $b$ ($\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$) are estimated by likelihood given the observed data and prior distributions for the parameters.
+The remaining three parameters $a$ (light efficiency, $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$), $R$ (respiration, $\text{mmol}/\text{m}^2/\text{d}$), and $b$ ($\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$) are estimated by likelihood given the observed data and prior distributions for the parameters.
 The JAGS software [as for BASE, @Plummer03] is used with the EBASE R package [described below, @Beck23; @RCT23] to estimate the unknown parameters using Markov Chain Monte Carlo (MCMC) simulations in a Bayesian framework.
@@
 As for the BASE model, relatively uninformed prior distributions following a normal Gaussian distribution are used by default, although the priors can be changed based on previous knowledge of parameters specific to an ecosystem or as informed by other metabolic modeling approaches.
-Reasonable uninformed prior distributions for EBASE were chosen with mean values using approximate estimates from the literature [@Wanninkhof14; @Grace15] and standard deviations that were sufficiently large to allow the Bayesian routine to search an unconstrained parameter space.
-The default priors for $a$, $R$, and $b$ are:
+Reasonable uninformed prior distributions for EBASE were chosen with mean values using approximate estimates from the literature [@caffrey2004; @Wanninkhof14] and standard deviations that were sufficiently large to allow the Bayesian routine to search an unconstrained parameter space.
+Specifically, the prior mean for $R$ was chosen as 300 $\text{mmol}/\text{m}^2/\text{d}$ (\~10 $\text{g O}_2/\text{m}^2/\text{d}$) using a national synthesis of metabolic rates for US estuaries [Table 2 in @caffrey2004].
+Although @caffrey2004 focused on small, shallow estuaries that are part of the NERRS, this dataset is the largest, longest, and most consistently collected in the US and it is reasonable to expect others may apply EBASE to these data.
+The prior mean for $a$ was chosen assuming that $P$ and $R$ are comparable at long time scales, such that $aPAR$ = $R$, or $a$ = $300/PAR$.
+Using the global mean surface $PAR$ of 70 $W/m^2$, the prior mean for $a$ is reasonably set at 4 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$.
+Lastly, the prior mean for $b$ was chosen as 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ following @Wanninkhof14.
+For the prior standard deviations....
+Together, the default priors for $a$, $R$, and $b$ are:
 

### 2023-08-01 manu-draft.qmd
added info about missing data

@@
 load(file = here('data/apasumdat.RData'))
+load(file = here('data/misdat.RData'))
 ```
@@
 
-where $N(\alpha, \beta)$ indicates a normal distribution with mean $\alpha$ and standard deviation $\beta$ and $T(\gamma, \delta)$ is 1 between $\gamma$ and $\theta$ and 0 elsewhere; $T$ has the effect of truncating the normal distribution to minimum and maximum values.
+where $N(\alpha, \beta)$ indicates a normal distribution with mean $\alpha$ and standard deviation $\beta$ and $T(\gamma, \theta)$ is 1 between $\gamma$ and $\theta$ and 0 elsewhere; $T$ has the effect of truncating the normal distribution to minimum and maximum values.
 All prior distributions were constrained to positive values based on known physical constraints and requirements of the model formula in @eq-ebase.
-Additionally, the prior distribution for $b$ was constrained to an upper limit of 0.502 (\~twice the default mean).
+Additionally, the prior distribution for $b$ was constrained to an upper limit of 0.502 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ (\~twice the default mean).
 Initial development of EBASE showed that using an undefined upper limit of $b$ led to unstable and unreasonable parameter estimates.
@@
 
+```{r}
+wqmis <- misdat$Temp # same percentages across all wq parms
+mtmis <- misdat$totpar # same percentages across all met parms
+```
+
 A preliminary assessment of the ability of EBASE to produce reasonable parameter estimates for $a$, $R$, and $b$ (@eq-ebase) was conducted using a synthetic time series with actual data as inputs.
@@
 Continuous meteorological observations from the nearby East Bay station (29.7694$^{\circ}$ N, --84.8815$^{\circ}$ W, \~10 km north of Cat Point) were combined with the water quality data at the same time step and included air temperature, wind speed, and PAR.
-Missing observations in the water quality and meteorological data were filled using autoregressive modelling of the actual parameters versus time to create a complete dataset from January 1st to December 31st.
+Missing observations in the water quality and meteorological data (`r wqmis`% and `r mtmis`% of all observations in each dataset, respectively) were filled using autoregressive modelling of the actual parameters versus time to create a complete dataset from January 1st to December 31st.
 A synthetic oxygen time series was then generated using the inputs from Apalachicola Bay ([@fig-synapa]a) and @eq-ebase as a forward calculating model with user-defined values for $a$, $R$, and $b$.

### 2023-08-01 R/dat_proc.R

@@ -5,3 +5,8 @@ library(here)
 
 fl <- here('data/apasumdat.RData')
 download.file('https://github.com/fawda123/BASEmetab_script/raw/master/data/apasumdat.RData', destfile = fl)
+
+# misdat, missing data in apacp 2021 ----------------------------------------------------------
+
+fl <- here('data/misdat.RData')
+download.file('https://github.com/fawda123/BASEmetab_script/raw/master/data/misdat.RData', destfile = fl)

### 2023-08-04 R/tabs.R
fixing units in tabs and reproc misdat

@@ -17,9 +17,9 @@ totab <- tibble(
   `Model notation` =  c('C_d', '-', '-', 'PAR', '-', 'H', 
                         'U102', 'Schmidt number', 'C_sat', 
                         'C_mod', 'P', 'R',  'D', 'a', 'b'),
-  Units = c('mg/L', 'C', 'psu', 'W/m2/s', 'm/s', 'm', 
+  Units = c('mg/L', 'C', 'psu', 'W/m2', 'm/s', 'm', 
             'm2', 'unitless', 'mmol/m3',
-            'mmol/m3', 'mmol O2 m2/d', 'mmol O2 m2/d', 'mmol O2 m2/d', '(mmol/m3/d)/(W/m2)', '(cm/hr)/(m2/s2)')
+            'mmol/m3', 'mmol O2 m2/d', 'mmol O2 m2/d', 'mmol O2 m2/d', '(mmol/m2/d)/(W/m2)', '(cm/hr)/(m2/s2)')
 ) %>% 
   flextable::as_grouped_data(groups = 'Type')
 

### 2023-08-07 manu-draft.qmd
fixing units in tabs and reproc misdat
text jusification for priors updated following meeting

@@
 \newcommand{\auni}{\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)}
-\newcommand{\runi}{\text{mmol}/\text{m}^3/\text{d}}
+\newcommand{\runi}{\text{mmol}/\text{m}^2/\text{d}}
 \newcommand{\buni}{\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)}
-\newcommand{\area}{\text{mmol}/\text{m}^2/\text{d}}
 ```
@@
 Lastly, the prior mean for $b$ was chosen as 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ following @Wanninkhof14.
-For the prior standard deviations....
+The prior standard deviations were chosen similarly based on values in @caffrey2004 based on the standard deviation of the reported values.
+The standard deviation of $R$ was 171 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$ and the standard deviation for $a$ was set proportionally as 2.25 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$, again assuming $P$ = $aPAR$ = $R$.
+The standard deviation for $b$ was set as 50% of the mean as an uninformed and reasonable value given that there are no comparable global or national estimates for the range of values.
 Together, the default priors for $a$, $R$, and $b$ are:
@@
 wqmis <- misdat$Temp # same percentages across all wq parms
-mtmis <- misdat$totpar # same percentages across all met parms
+mtmis <- misdat$PAR # same percentages across all met parms
 ```
@@
 The sensitivity analysis evaluated changes in the distributions by varying the means and standard deviations from small to large to test the effect of changes in the central tendencies and more or less constrained ranges, respectively, on the priors.
-For each parameter, low and high values for the means and standard deviations were assessed as changes approximately equal to two orders of magnitude of the means and standard deviations of the known parameters of the synthetic time series.
-The only exception was the $b$ parameter where the low values created priors with a distribution close to zero and the high values were equivalent to a uniform prior distribution in the range of 0 to 0.502 (justification noted in @sec-ebasemod).
-Each unique combination of low and high values for the means and standard deviations of each parameter were evaluated, creating 2$^6$ = 64 different combinations of prior distributions that were used with the synthetic time series in EBASE.
+Low and high values for the means and standard deviations were assessed following reasonable ranges from @caffrey2004 using similar logic for the default priors.
+The minimum and maximum mean values for $R$ were evaluated as 138 and 1009 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$ and proportionally for $a$ as 2.0 and 14.4 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$.
+The minimum and maximum values for the standard deviations for $R$ and $a$ were chosen as 10% and 1000% of the default mean values to test a range from high precision to completely uninformed values.
+The mean and standard deviation for the $b$ parameter were not tested and set as the default values for all analyses given that there is minimal prior knowledge on expectations for this parameter.
+The default prior distribution is already uninformed and we have no appropriate expectations on methods for varying this parameter.
+Each unique combination of low and high values for the means and standard deviations of each parameter were evaluated, creating 2$^4$ = 16 different combinations of prior distributions that were used with the synthetic time series in EBASE.
 
@@
 Results from optimization periods of one, seven, and thirty days were compared for the synthetic one-year time series.
-Each optimization period was also evaluated for every unique combination of prior distributions, totaling 192 (64 times the three optimization periods) unique analyses for the assessment of changing priors of all parameters.
+Each optimization period was also evaluated for every unique combination of prior distributions, totaling 48 (16 times the three optimization periods) unique analyses for the assessment of changing priors of all parameters.
 

### 2023-08-08 manu-draft.qmd
minor methods text change

@@
 Lastly, the prior mean for $b$ was chosen as 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ following @Wanninkhof14.
-The prior standard deviations were chosen similarly based on values in @caffrey2004 based on the standard deviation of the reported values.
+The prior standard deviations were chosen similarly based on values in @caffrey2004 using the standard deviation of the reported values.
 The standard deviation of $R$ was 171 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$ and the standard deviation for $a$ was set proportionally as 2.25 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$, again assuming $P$ = $aPAR$ = $R$.
@@
 The sensitivity analysis evaluated changes in the distributions by varying the means and standard deviations from small to large to test the effect of changes in the central tendencies and more or less constrained ranges, respectively, on the priors.
-Low and high values for the means and standard deviations were assessed following reasonable ranges from @caffrey2004 using similar logic for the default priors.
+Low and high values for the means and standard deviations were assessed following reasonable ranges from @caffrey2004 using logic similar to that for choosing the default priors.
 The minimum and maximum mean values for $R$ were evaluated as 138 and 1009 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$ and proportionally for $a$ as 2.0 and 14.4 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$.

### 2023-08-14 R/figs.R
reproc relevant figs after new sensitivity analysis and obs/nose results #4

@@ -93,7 +93,7 @@ load(file = fl)
 
 p1 <- priorcomp(apasumdat, met = 'nse')
 
-png(here('figs/priorcomp.png'), height = 6, width = 6, family = 'serif', units = 'in', res = 500)
+png(here('figs/priorcomp.png'), height = 4, width = 6, family = 'serif', units = 'in', res = 500)
 print(p1)
 dev.off()
 
@@ -105,40 +105,31 @@ load(file = fl)
 
 p1 <- priorsumcomp(apasumdat, met = 'nse')
 
-png(here('figs/priorsumcomp.png'), height = 5, width = 6, family = 'serif', units = 'in', res = 500)
+png(here('figs/priorsumcomp.png'), height = 4, width = 5, family = 'serif', units = 'in', res = 500)
 print(p1)
 dev.off()
 
 # best/worst fwoxy apa comp -------------------------------------------------------------------
 
 # grid ests
-fl1a <- paste0(tempdir(), '/apagrd1a.RData')
-download.file('https://github.com/fawda123/BASEmetab_script/raw/master/data/apagrd1a.RData', destfile = fl1a)
-load(file = fl1a)
-fl1b <- paste0(tempdir(), '/apagrd1b.RData')
-download.file('https://github.com/fawda123/BASEmetab_script/raw/master/data/apagrd1b.RData', destfile = fl1b)
-load(file = fl1b)
-fl7a <- paste0(tempdir(), '/apagrd7a.RData')
-download.file('https://github.com/fawda123/BASEmetab_script/raw/master/data/apagrd7a.RData', destfile = fl7a)
-load(file = fl7a)
-fl7b <- paste0(tempdir(), '/apagrd7b.RData')
-download.file('https://github.com/fawda123/BASEmetab_script/raw/master/data/apagrd7b.RData', destfile = fl7b)
-load(file = fl7b)
-fl30a <- paste0(tempdir(), '/apagrd30a.RData')
-download.file('https://github.com/fawda123/BASEmetab_script/raw/master/data/apagrd30a.RData', destfile = fl30a)
-load(file = fl30a)
-fl30b <- paste0(tempdir(), '/apagrd30b.RData')
-download.file('https://github.com/fawda123/BASEmetab_script/raw/master/data/apagrd30b.RData', destfile = fl30b)
-load(file = fl30b)
+fl1 <- paste0(tempdir(), '/apagrd1.RData')
[…]

### 2023-08-14 R/funcs.R

@@ -105,7 +105,7 @@ priorcomp <- function(dat, met, topbot = 3){
   
   toplo <- dat %>% 
     unnest('ests') %>% 
-    select(ndays, amean, asd, rmean, rsd, bmean, bsd, var, matches(toshw)) %>%
+    select(ndays, amean, asd, rmean, rsd, var, matches(toshw)) %>%
     filter(!var %in% 'b') %>% 
     pivot_wider(names_from = 'var', values_from = !!toshw) %>% 
     mutate(
@@ -137,19 +137,19 @@ priorcomp <- function(dat, met, topbot = 3){
     )
   
   toplo1 <- toplo %>% 
-    select(ind, amean, asd, rmean, rsd, bmean, bsd) %>% 
+    select(ind, amean, asd, rmean, rsd) %>% 
     unique() %>% 
     mutate_at(vars(-matches('ind')), factor, labels = c('L', 'H')) %>% 
     pivot_longer(-c('ind'), names_to = 'var', values_to = 'val') %>% 
     mutate(
       var = factor(var, 
-                   levels = c('amean', 'asd', 'rmean', 'rsd', 'bmean', 'bsd'), 
-                   labels = c('italic(a)~mu', 'italic(a)~sigma', 'italic(R)~mu', 'italic(R)~sigma', 'italic(b)~mu', 'italic(b)~sigma')), 
+                   levels = c('amean', 'asd', 'rmean', 'rsd'), 
+                   labels = c('italic(a)~mu', 'italic(a)~sigma', 'italic(R)~mu', 'italic(R)~sigma')), 
       fac = ''
     )
   
   toplo2 <- toplo %>% 
-    select(-amean, -asd, -rmean, -rsd, -bmean, -bsd) %>%
+    select(-amean, -asd, -rmean, -rsd) %>%
     pivot_longer(-c(ind, ndays), names_to = 'var', values_to = 'val') %>% 
     mutate(
       var = factor(var, 
@@ -219,9 +219,9 @@ priorcomp <- function(dat, met, topbot = 3){
   if(met == 'nse')
     p2 <- p2 + 
       scale_fill_gradient2(low = 'red', mid = 'white', high = 'green', midpoint = 1, trans = 'exp' ,
-                           breaks = c(-448, -1, 0, 0.5, 0.9))
[…]

### 2023-08-15 R/figs.R
reproc priorplot with ebase update

@@ -1,3 +1,5 @@
+# setup ---------------------------------------------------------------------------------------
+
 library(tidyverse)
 library(lubridate)
 library(here)

### 2023-08-15 manu-draft.qmd
added info about synthetic time series

@@
 
-For @eq-ebase, the terms on the right side of the equation are gross primary production (P, $aPAR$), ecosystem respiration (R), and gas exchange (D, remainder), respectively, processes that result in the change in dissolved oxygen ($dC$) per unit time ($dt$); $H$ is the water depth.
+For @eq-ebase, the terms on the right side of the equation are gross primary production ($P$, $aPAR$), ecosystem respiration ($R$), and gas exchange ($D$, remainder), respectively, processes that result in the change in dissolved oxygen ($dC$) per unit time ($dt$); $H$ is the water depth.
 Production is modeled linearly as a function of surface irradiance (PAR, $\text{W}/\text{m}^2$) and $a$ (light efficiency, $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$).
@@
 The prior standard deviations were chosen similarly based on values in @caffrey2004 using the standard deviation of the reported values.
-The standard deviation of $R$ was 171 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$ and the standard deviation for $a$ was set proportionally as 2.25 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$, again assuming $P$ = $aPAR$ = $R$.
+The standard deviation of $R$ was 170 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$ and the standard deviation for $a$ was set proportionally as 2.25 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$, again assuming $aPAR$ = $R$.
 The standard deviation for $b$ was set as 50% of the mean as an uninformed and reasonable value given that there are no comparable global or national estimates for the range of values.
@@
 $$
-a \sim N(4, 1)T(0,\infty) \;\; \left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)
+a \sim N(4, 2.25)I(0,\infty) \;\; \left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)
 $$ {#eq-aprior}
@@
 $$
-R \sim N(300, 50)T(0,\infty) \;\; \text{mmol}/\text{m}^2/\text{d}
+R \sim N(300, 170)I(0,\infty) \;\; \text{mmol}/\text{m}^2/\text{d}
 $$ {#eq-rprior}
@@
 $$
-b \sim N(0.251, 0.1)T(0,0.502) \;\; \left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)
+b \sim N(0.251, 0.126)I(0,0.502) \;\; \left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)
 $$ {#eq-bprior}
 
-where $N(\alpha, \beta)$ indicates a normal distribution with mean $\alpha$ and standard deviation $\beta$ and $T(\gamma, \theta)$ is 1 between $\gamma$ and $\theta$ and 0 elsewhere; $T$ has the effect of truncating the normal distribution to minimum and maximum values.
+where $N(\alpha, \beta)$ indicates a normal distribution with mean $\alpha$ and standard deviation $\beta$ and $I(\gamma, \theta)$ is 1 between $\gamma$ and $\theta$ and 0 elsewhere; $I$ has the effect of truncating the normal distribution to minimum and maximum values.
 All prior distributions were constrained to positive values based on known physical constraints and requirements of the model formula in @eq-ebase.
 Additionally, the prior distribution for $b$ was constrained to an upper limit of 0.502 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ (\~twice the default mean).
-Initial development of EBASE showed that using an undefined upper limit of $b$ led to unstable and unreasonable parameter estimates.
+Initial development of EBASE showed that using an undefined upper limit of $b$ led to unstable and unreasonable estimates for all parameters.
 Examples of the prior distributions in Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior] are shown in @fig-priorplot.
@@
 Missing observations in the water quality and meteorological data (`r wqmis`% and `r mtmis`% of all observations in each dataset, respectively) were filled using autoregressive modelling of the actual parameters versus time to create a complete dataset from January 1st to December 31st.
-A synthetic oxygen time series was then generated using the inputs from Apalachicola Bay ([@fig-synapa]a) and @eq-ebase as a forward calculating model with user-defined values for $a$, $R$, and $b$.
+
+A synthetic oxygen time series was then generated using the inputs from Apalachicola Bay ([@fig-synapa]) and @eq-ebase as a forward calculating model with estimated values for $a$, $R$, and $b$.
 All metabolic rates for production, respiration, and gas exchange ([@fig-synapa]b and c) were also estimated with the synthetic time series.
+The time series was estimated using the observed data described above, the separate metabolic components and their parameters in @eq-ebase, and approximate metabolic rates from the Odum open-water method as implemented in the WtRegDO R package [@Beck21] as an approximate estimate for back-calculation of the EBASE parameters.
+In short, $P$, $R$, and $D$ as $\text{mmol}/\text{m}^2/\text{d}$ were estimated using the Odum method and the separate synthetic components for EBASE were created by estimating the corresponding parameters, such that $P /PAR = a$, $R$ as is, and $D = k_w\left(C_{sat} - C \right)$, where $k_w$ includes the $b$ parameter.
+Daily estimates for each were created as the daily average for each parameter fit with a polynomial spline to create a smoothed output.
+The synthetic dissolved oxygen time series was then created as the integrated sum of the separate components for $P$, $R$, and $D$ divided by water column depth.
 The only parameter that could not be created as a continuous time series was $b$, which was set to 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ [@Wanninkhof14].
@@
 
-The prior distributions for the EBASE parameters follow truncated normal distributions with defined means ($\mu$) and standard deviations ($\sigma$)(Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]).
-The sensitivity analysis evaluated changes in the distributions by varying the means and standard deviations from small to large to test the effect of changes in the central tendencies and more or less constrained ranges, respectively, on the priors.
+The prior distributions for the EBASE parameters follow truncated normal distributions with defined means ($\mu$) and standard deviations ($\sigma$) (Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]).
+The sensitivity analysis evaluated changes in the distributions by varying the means and standard deviations from small to large to test the effect of changes in the central tendencies and more or less constrained ranges, respectively, on the results.
 Low and high values for the means and standard deviations were assessed following reasonable ranges from @caffrey2004 using logic similar to that for choosing the default priors.
 The minimum and maximum mean values for $R$ were evaluated as 138 and 1009 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$ and proportionally for $a$ as 2.0 and 14.4 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$.
-The minimum and maximum values for the standard deviations for $R$ and $a$ were chosen as 10% and 1000% of the default mean values to test a range from high precision to completely uninformed values.
+The minimum and maximum values for the standard deviations for $R$ and $a$ were chosen as 10% and 1000% of the default standard deviations to test a range from high precision to completely uninformed values.
 The mean and standard deviation for the $b$ parameter were not tested and set as the default values for all analyses given that there is minimal prior knowledge on expectations for this parameter.
@@
 Results from optimization periods of one, seven, and thirty days were compared for the synthetic one-year time series.
-Each optimization period was also evaluated for every unique combination of prior distributions, totaling 48 (16 times the three optimization periods) unique analyses for the assessment of changing priors of all parameters.
+Each optimization period was also evaluated for every unique combination of prior distributions, totaling 48 (16 times three optimization periods) unique analyses for the assessment of changing priors of all parameters.
 
@@
 ![Synthetic time series for one year of continuous data at Apalachicola Bay.
-Synthetic input to EBASE was the (a) dissolved oxygen time series, with results compared to the (b) synthetic $a$ parameter and (c) synthetic metabolic estimates for gross primary production (P), ecosystem respiration (R), and gas exchange (D).
-The $a$ parameter is shown as an areal value for comparison with the metabolic results.
-The $b$ parameter is also fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ and is not shown.](figs/synapa.png){#fig-synapa}
+Synthetic input to EBASE was the (a) dissolved oxygen time series, with results compared to the (b) synthetic $a$ parameter and (c) synthetic metabolic estimates for gross primary production ($P$), the negative of ecosystem respiration ($-R$), and gas exchange ($D$).](figs/synapa.png){#fig-synapa}
 

### 2023-08-18 manu-draft.qmd
update prior and sensitivity rationale from RN suggestions

@@
 The metabolism estimates and their parameters returned by the Bayesian routine implemented in JAGS are affected by the prior distributions assigned to each.
-As for the BASE model, relatively uninformed prior distributions following a normal Gaussian distribution are used by default, although the priors can be changed based on previous knowledge of parameters specific to an ecosystem or as informed by other metabolic modeling approaches.
+As for BASE, relatively uninformed prior distributions following a normal Gaussian distribution are used by default, although the priors can be changed based on previous knowledge of parameters specific to an ecosystem or as informed by other metabolic modeling approaches.
 Reasonable uninformed prior distributions for EBASE were chosen with mean values using approximate estimates from the literature [@caffrey2004; @Wanninkhof14] and standard deviations that were sufficiently large to allow the Bayesian routine to search an unconstrained parameter space.
@@
 Although @caffrey2004 focused on small, shallow estuaries that are part of the NERRS, this dataset is the largest, longest, and most consistently collected in the US and it is reasonable to expect others may apply EBASE to these data.
-The prior mean for $a$ was chosen assuming that $P$ and $R$ are comparable at long time scales, such that $aPAR$ = $R$, or $a$ = $300/PAR$.
-Using the global mean surface $PAR$ of 70 $W/m^2$, the prior mean for $a$ is reasonably set at 4 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$.
+The prior mean for $a$ was chosen assuming that $P$ and $R$ are comparable at long time scales, such that $aPAR$ = $R$, or $a$ = $\left(300~\text{mmol}/\text{m}^2/\text{d}\right)/PAR$.
+Using the global mean surface $PAR$ of 80 $W/m^2$, which is based on a global mean surface total shortwave radiation of 200 $\text{W}/\text{m}^2$ [@kiehl1997] and a $PAR$ fraction of total shortwave radiation of 0.4 [@jacovides2004], the prior mean for $a$ is reasonably set at 4 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$.
 Lastly, the prior mean for $b$ was chosen as 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ following @Wanninkhof14.
-The prior standard deviations were chosen similarly based on values in @caffrey2004 using the standard deviation of the reported values.
-The standard deviation of $R$ was 170 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$ and the standard deviation for $a$ was set proportionally as 2.25 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$, again assuming $aPAR$ = $R$.
-The standard deviation for $b$ was set as 50% of the mean as an uninformed and reasonable value given that there are no comparable global or national estimates for the range of values.
+The prior standard deviation for $R$ was based on the standard deviation of the mean values of $R$ at the NERRS sites in @caffrey2004.
+The standard deviation of $R$ was 171 $\text{mmol}/\text{m}^2/\text{d}$, which was rounded to 50% of the mean, or 150 $\text{mmol}/\text{m}^2/\text{d}$.
+The standard deviation for $a$ was set proportionally to 2 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$.
+For the prior standad deviation of $b$, we considered using the 20% error estimated for $b$ by @Wanninkhof14.
+However, we chose a larger value, 50%, because the wind speed-based gas transfer velocity formulation of @Wanninkhof14 was mainly developed and calibrated for open-ocean applications, where the complications of surfactants [@frew1997] and tidal currents [@Ho16], as well as fetch and suspended matter [@Borges11], are expected to be minimal.
 Together, the default priors for $a$, $R$, and $b$ are:
@@
 $$
-a \sim N(4, 2.25)I(0,\infty) \;\; \left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)
+a \sim N(4, 2)I(0,\infty) \;\; \left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)
 $$ {#eq-aprior}
@@
 $$
-R \sim N(300, 170)I(0,\infty) \;\; \text{mmol}/\text{m}^2/\text{d}
+R \sim N(300, 150)I(0,\infty) \;\; \text{mmol}/\text{m}^2/\text{d}
 $$ {#eq-rprior}
@@
 $$
-b \sim N(0.251, 0.126)I(0,0.502) \;\; \left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)
+b \sim N(0.251, 0.125)I(0,0.502) \;\; \left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)
 $$ {#eq-bprior}
@@
 The sensitivity analysis evaluated changes in the distributions by varying the means and standard deviations from small to large to test the effect of changes in the central tendencies and more or less constrained ranges, respectively, on the results.
-Low and high values for the means and standard deviations were assessed following reasonable ranges from @caffrey2004 using logic similar to that for choosing the default priors.
-The minimum and maximum mean values for $R$ were evaluated as 138 and 1009 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$ and proportionally for $a$ as 2.0 and 14.4 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$.
-The minimum and maximum values for the standard deviations for $R$ and $a$ were chosen as 10% and 1000% of the default standard deviations to test a range from high precision to completely uninformed values.
+Low and high values for the means and standard deviations for $R$ were assessed following reasonable ranges from @caffrey2004 using logic similar to that for choosing the default priors.
+The minimum and maximum mean values for $R$ were evaluated as 138 and 1009 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$, which we rounded to $1/2$ and 3 times the mean (150 and 900 $\text{mmol}/\text{m}^2/\text{d}$) and proportionally for $a$ as 2.0 and 12 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$.
+The minimum and maximum values for the standard deviations for $R$ and $a$ were chosen as 10% and 1000% of the default mean values to test a range from high precision to completely uninformed values.
 The mean and standard deviation for the $b$ parameter were not tested and set as the default values for all analyses given that there is minimal prior knowledge on expectations for this parameter.

### 2023-08-25 manu-draft.qmd
reproc figs and tables for #4 but still jacked up

@@
 All priors are normal distributions and truncated to positive values.
-$b$ is also truncated to a maximum of 0.502.](figs/priorplot.png){#fig-priorplot fig-align="center"}
+$b$ is also truncated to a maximum of 0.502 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$.](figs/priorplot.png){#fig-priorplot fig-align="center"}
 
@@
 ![Synthetic time series for one year of continuous data at Apalachicola Bay.
-Synthetic input to EBASE was the (a) dissolved oxygen time series, with results compared to the (b) synthetic $a$ parameter and (c) synthetic metabolic estimates for gross primary production ($P$), the negative of ecosystem respiration ($-R$), and gas exchange ($D$).](figs/synapa.png){#fig-synapa}
+Synthetic input to EBASE was the (a) dissolved oxygen time series, with results compared to the (b) synthetic $a$ parameter and (c) synthetic metabolic estimates for gross primary production ($P$), the negative of ecosystem respiration ($-R$), and gas exchange ($-D$).](figs/synapa.png){#fig-synapa}
 
@@
 
-![Comparison of EBASE results to a synthetic time series created with known metabolic parameters.
+![Comparison of EBASE results to a synthetic time series created with known model parameters.
+($a$, $R$, and $b$).
 Results are compared for unique combinations of prior distributions and optimization period.
 The means ($\mu$) and standard deviations ($\sigma$) are evaluated as low (L) or high (H) values (see text for details) and the optimization period for EBASE is one day, seven days, or thirty days.
-The metabolic results for dissolved oxygen (DO), gross primary production (P), ecosystem respiration (R), gas exchange (D), and the $a$ parameter are evaluated based on Nash-Sutcliffe Efficiency (NSE) values between the EBASE results and synthetic values.
-NSE values from zero to one (green) indicate EBASE has recovered the synthetic parameters with values closer to 1 being a better approximation, whereas negative values indicate poor recovery.
+The model predictions of dissolved oxygen (DO), gross primary production (P), ecosystem respiration (R), gas exchange (D), and the $a$ parameter are evaluated based on Nash-Sutcliffe Efficiency (NSE) values between the EBASE results and synthetic values.
+NSE values from zero to one (green) indicate EBASE has some skill in recovering the synthetic values of DO, $D$, $P$, $R$ , and $a$, with values closer to 1 being a better approximation, whereas negative values indicate no skill.
 The color scale for NSE is also exponentially transformed to minimize the influence of large negative values on the visual display.
 The $b$ parameter is a fixed constant and was not evaluated.
-Rankings of each combination as the median NSE value across all parameters in a row are shown on the right, with the top three as black/bold and bottom three as black/italic.](figs/priorcomp.png){#fig-priorcomp}
+Rankings of each combination as the median NSE value across all five model outputs in a row are shown on the right, with the top three as black/bold and bottom three as black/italic.](figs/priorcomp.png){#fig-priorcomp}
 

### 2023-08-25 R/figs.R

@@ -47,8 +47,9 @@ fwdatinp <- fwdat %>%
 # prior distribution plot ---------------------------------------------------------------------
 
 p <- prior_plot(n = 3e6)
+p <- p + theme(axis.text = element_text(size = 10))
 
-png(here('figs/priorplot.png'), height = 3, width = 8, family = 'serif', units = 'in', res = 500)
+png(here('figs/priorplot.png'), height = 2.5, width = 6.5, family = 'serif', units = 'in', res = 500)
 print(p)
 dev.off()
 
@@ -75,15 +76,16 @@ p2 <- ggplot(fwdatcmp, aes(x = DateTimeStamp, y = a)) +
   )
 
 p3 <- ebase_plot(fwdatcmp, instantaneous = F) +
-  ggplot2::scale_color_discrete(
-    breaks = c('P', 'R', 'D')
-  ) +
   scale_x_date(breaks = "1 month", labels = date_format("%b")) +
   labs(title = '(c) Synthetic metabolic estimates')
 
-p <- p1 + p2 + p3 + plot_layout(ncol = 1) & theme(panel.grid.minor = element_blank())
+p <- p1 + p2 + p3 + plot_layout(ncol = 1) & 
+  theme(
+    panel.grid.minor = element_blank(),
+    axis.text = element_text(size = 10)
+    )
 
-png(here('figs/synapa.png'), height = 8, width = 8, family = 'serif', units = 'in', res = 500)
+png(here('figs/synapa.png'), height = 7, width = 7, family = 'serif', units = 'in', res = 500)
 print(p)
 dev.off()
 

### 2023-08-25 R/funcs.R

@@ -153,8 +153,8 @@ priorcomp <- function(dat, met, topbot = 3){
     pivot_longer(-c(ind, ndays), names_to = 'var', values_to = 'val') %>% 
     mutate(
       var = factor(var, 
-                   levels = c('DO_mod', 'P', 'R', 'D', 'a'), 
-                   labels = c('DO', 'P', 'R', 'D', 'italic(a)')
+                   levels = c('DO_mod', 'D', 'P', 'R', 'a'), 
+                   labels = c('DO', 'D', 'P', 'R', 'italic(a)')
       ), 
       ndays = factor(ndays, levels = unique(ndays))
     ) %>% 

### 2023-08-27 manu-draft.qmd
reworking some figures, optex now has options for ylim

@@
 
-![Summaries of median Nash-Sutcliffe Efficiency (NSE) values across all parameters for the comparisons of EBASE and synthetic model outputs separated by different combinations of prior values.
-Summaries are shown relative to the difference from the grand median across all comparisons.
-Each result on the x-axis represents the median values for which the combination prior applies, e.g., all results from @fig-priorcomp where $\mu$ is high or low for the $a$ parameter using a one day optimization period.
-Point size represents the inter-quartile range of median NSE values for each result.](figs/priorsumcomp.png){#fig-priorsumcomp}
+![Summaries of % change in Nash-Sutcliffe Efficiency (NSE) values across all the comparisons of EBASE and the five synthetic model outputs separated by different combinations of prior values.
+Summaries are shown relative to the change from the median for each comparison group (i.e., optimization period and prior parameter, $n$ = 32).
+Each result on the x-axis represents the median values for which the combination prior applies.
+For example, the points in the upper left panel that are fartheset to the left summarize the results from @fig-priorcomp where $\mu$ is high or low for the $a$ parameter using a one day optimization period.
+Point size represents the inter-quartile range of median NSE values for each result ($n$ = 8) .](figs/priorsumcomp.png){#fig-priorsumcomp}
 
@@
 ![Comparison of EBASE results for the best and worst performing models to a synthetic time series created with known metabolic parameters.
-The best and worst performing models for the one day (a, b), seven day (c, d), and thirty day (e, f) optimization periods are based on the EBASE results with prior distributions that produced the highest and lowest median Nash-Sutcliffe Efficenciy values across all parameter estimates (@fig-priorcomp).
+The best and worst performing models for the one day (a, b), seven day (c, d), and thirty day (e, f) optimization periods are based on the EBASE results with prior distributions that produced the highest and lowest median Nash-Sutcliffe Efficency values across all parameter estimates (@fig-priorcomp).
 The results are averaged at the time step of the model optimization period.

### 2023-08-27 R/figs.R

@@ -133,7 +133,14 @@ fl <- paste0(tempdir(), '/apasumdat.RData')
 download.file('https://github.com/fawda123/BASEmetab_script/raw/master/data/apasumdat.RData', destfile = fl)
 load(file = fl)
 
-p <- optex(apagrd, fwdatcmp, apasumdat, rnkmetsum = c(1, 16), met = 'nse')
+# row y axis limits
+lims <- list(
+  P = c(0, 600), 
+  R = c(0, 600), 
+  a = c(0, 11)
+)
+
+p <- optex(apagrd, fwdatcmp, apasumdat, rnkmetsum = c(1, 16), met = 'nse', lims = lims)
 
 png(here('figs/optex.png'), height = 8.7, width = 9, family = 'serif', units = 'in', res = 500)
 print(p)

### 2023-08-27 R/funcs.R

@@ -249,8 +249,11 @@ priorsumcomp <- function(dat, met = 'r2'){
       .by = c('ndays', 'prior', 'val')
     ) %>% 
     mutate(
-           grandmed = median(medv),
-           medv = medv - grandmed,
+      grandmed = median(medv), 
+      .by = c('ndays', 'val')
+    ) %>% 
+    mutate(
+           medv = 100 * (medv - grandmed) / ((grandmed + medv) / 2),
            param = gsub('mean$|sd$', '', prior),
            param = factor(param, levels = c('a', 'r'), labels = c('a', 'R')),
            prior = gsub('^a|^r|^b', '', prior), 
@@ -263,7 +266,7 @@ priorsumcomp <- function(dat, met = 'r2'){
     lbspr = c('italic(R)^2', 'RMSE', 'MAPE', 'mae', 'NSE')
   ) %>% 
     filter(met == !!met)
-  ylab <- parse(text = paste0('Median~', metsel$lbspr,'~-Grand~Median'))
+  ylab <- parse(text = paste0('`%`', '~change~', metsel$lbspr))
   
   wd <- 0.3
   
@@ -411,8 +414,13 @@ cmpplo <- function(res, fwdatcmp, subttl, ylbs = TRUE){
 }
 
 # comparison of fwoxy and ebase results for ranked model by prior at opt of ndays
-optex <- function(apagrd, fwdatcmp, apasumdat, rnkmetsum, met){
+optex <- function(apagrd, fwdatcmp, apasumdat, rnkmetsum, met, lims = NULL){
 
+  if(!is.null(lims))
+    lims <- lapply(lims, function(x) tibble(min = x[1], max = x[2])) %>% 
+      tibble::enframe(name = 'var') %>% 
+      unnest('value')
+  
   # find the prior comparison
   rnkfnd <- metsum_fun(apasumdat, met = met, parms = T) %>% 
     filter(rnkmetsum %in% !!rnkmetsum) %>% 
@@ -448,15 +456,16 @@ optex <- function(apagrd, fwdatcmp, apasumdat, rnkmetsum, met){
       )
     ) %>%
     pivot_wider(names_from = 'mod', values_from = 'val')
-  
+
   toplo <- cmp %>% 
+    left_join(lims, by = 'var') %>% 
     summarise(
       Synthetic = mean(Synthetic, na.rm = T),
       EBASE = mean(EBASE, na.rm = T),
       Date = min(Date),
[…]

### 2023-08-27 R/figs.R
redo synapanoscmp for ts comparison, need to edit and change caption

@@ -220,9 +220,9 @@ fl <- paste0(tempdir(), '/resnos.RData')
 download.file('https://github.com/fawda123/BASEmetab_script/raw/master/data/resnos.RData', destfile = fl)
 load(file = fl)
 
-p <- syncomp_plo(resobs, resnos) 
+p <- syncomp_plo(resobs, resnos, fwdatcmp) 
   
-png(here('figs/synapanoscmp.png'), height = 4.75, width = 6, family = 'serif', units = 'in', res = 500)
+png(here('figs/synapanoscmp.png'), height = 5, width = 6.5, family = 'serif', units = 'in', res = 500)
 print(p)
 dev.off()
 

### 2023-08-27 R/funcs.R

@@ -570,75 +570,124 @@ metsum_fun <- function(apasumdat, met, parms = F){
 }
 
 # compare synthetic and synthetic + noise
-syncomp_plo <- function(resobs, resnos){
+syncomp_plo <- function(resobs, resnos, fwdatcmp){
   
-  res <- list(
-    'Synthetic' = resobs,
-    'Synthetic + noise' = resnos
+  toplo1 <- list(
+    'Synthetic' = fwdatcmp,
+    'EBASE recovered' = resobs,
+    'EBASE recovered (+ noise)' = resnos
   ) %>% 
-    enframe(name = 'dotyp') %>% 
+    enframe() %>% 
     unnest('value') %>% 
-    select(-matches('hi$|lo$|converge|rsq|^Date$|dDO|^H$')) %>% 
-    pivot_longer(-c('dotyp', 'DateTimeStamp', 'grp')) %>% 
-    summarise(
-      value = mean(value, na.rm = T),
-      DateTimeStamp = min(DateTimeStamp),
-      .by = c('dotyp', 'grp', 'name')
-    ) %>% 
-    filter(!name == 'DO_obs') %>% 
+    select(matches('name|Date|DateTimeStamp|grp|^a$|^R$')) %>% 
+    pivot_longer(names_to = 'var', values_to = 'val', a:R) %>% 
+    group_nest(name, var) %>% 
     mutate(
-      name = factor(name, 
-                    levels = c('DO_mod', 'a', 'b', 'P', 'R', 'D'), 
-                    labels = c('O[2]~(mmol~m^{-3})', 'italic(a)~(mmol~m^{-2}~d^{-1})/(W~m^{-2})', 'italic(b)~(cm~hr^{-1})/(m^{2}~s^{-2})', 'P~(mmol~m^{2}~d^{-1})', 'R~(mmol~m^{2}~d^{-1})', 'D~(mmol~m^{2}~d^{-1})')
-      )
+      data = purrr::pmap(list(name, data), function(name, data){
+        
+        if(name == 'Synthetic')
+          return(data)
+        
+        out <- data %>% 
+          group_by(grp) %>% 
+          reframe(
+            val = unique(val, na.rm = T), 
+            DateTimeStamp = min(DateTimeStamp)
+          ) %>% 
+          na.omit()
+        
+        return(out)
+        
+      })
     ) %>% 
-    pivot_wider(names_from = dotyp, values_from = value)
+    unnest('data') %>% 
+    mutate(
+      var = factor(var, 
[…]

### 2023-08-28 manu-draft.qmd
fix figure caption for synapanoscmp

@@
 
-![Comparisons of modelled dissolved oxygen and metabolic parameters from EBASE for a synthetic time series and a synthetic time series with instrument and tidal noise.
-Black lines show the 1:1 comparison and the red lines show the linear regression fit.
+![Comparisons of EBASE to recover parameters $a$ (top) and $R$ (bottom) from a synthetic time series and a synthetic time series with residual and tidal noise included.
+The left plots show the time series and the right plots show a 1:1 comparison for each result.
 See @fig-synapanos for an example of the dissolved oxygen for both input time series.](figs/synapanoscmp.png){#fig-synapanoscmp fig-align="center"}

### 2023-08-28 R/figs.R

@@ -222,7 +222,7 @@ load(file = fl)
 
 p <- syncomp_plo(resobs, resnos, fwdatcmp) 
   
-png(here('figs/synapanoscmp.png'), height = 5, width = 6.5, family = 'serif', units = 'in', res = 500)
+png(here('figs/synapanoscmp.png'), height = 5.5, width = 7, family = 'serif', units = 'in', res = 500)
 print(p)
 dev.off()
 

### 2023-08-28 R/funcs.R

@@ -645,15 +645,15 @@ syncomp_plo <- function(resobs, resnos, fwdatcmp){
   
   toplo2a <- toplo2 %>% 
     filter(grepl('a', var))
-  maxva <- max(c(toplo2a$Synthetic, toplo2a$val)) 
+  rnga <- range(c(toplo2a$Synthetic, toplo2a$val)) 
   p2a <- ggplot(toplo2a, aes(x = Synthetic, y = val, color = EBASE)) + 
     geom_point(show.legend = F, size = 1) +
     geom_smooth(method = 'lm', se = F, formula = y~x, show.legend = F) +
     scale_color_manual(values = c("#00BA38", "#619CFF")) +
     geom_abline(intercept = 0, slope = 1) +
     coord_cartesian(
-      xlim = c(0, maxva), 
-      ylim = c(0, maxva)
+      xlim = c(rnga[1], rnga[2]), 
+      ylim = c(rnga[1], rnga[2])
     ) +
     theme(legend.position = 'none') +
     facet_wrap(~var, scales = 'free', labeller = label_parsed, strip.position = 'top') + 
@@ -664,15 +664,15 @@ syncomp_plo <- function(resobs, resnos, fwdatcmp){
   
   toplo2b <- toplo2 %>% 
     filter(grepl('R', var))
-  maxvb <- max(c(toplo2b$Synthetic, toplo2b$val)) 
+  rngb <- range(c(toplo2b$Synthetic, toplo2b$val)) 
   p2b <- ggplot(toplo2b, aes(x = Synthetic, y = val, color = EBASE)) + 
     geom_point(show.legend = F, size = 1) +
     geom_smooth(method = 'lm', se = F, formula = y~x, show.legend = F) +
     scale_color_manual(values = c("#00BA38", "#619CFF")) +
     geom_abline(intercept = 0, slope = 1) +
     coord_cartesian(
-      xlim = c(0, maxvb), 
-      ylim = c(0, maxvb)
+      xlim = c(rngb[1], rngb[2]), 
+      ylim = c(rngb[1], rngb[2])
     ) +
     theme(legend.position = 'none') +
     facet_wrap(~var, scales = 'free', labeller = label_parsed, strip.position = 'top') + 
@@ -687,7 +687,7 @@ syncomp_plo <- function(resobs, resnos, fwdatcmp){
                  strip.background = element_blank(), 
                  panel.grid.minor = element_blank(), 
                  strip.placement = 'outside', 
[…]

### 2023-08-28 manu-draft.qmd
fix apacmpfig for RN suggestions

@@
 ![Comparisons of daily metabolic estimates from the EBASE and Odum methods for one year of continuous monitoring data at Apalachicola Bay.
-Results for net ecosystem metabolism (NEM), gross primary production (P), ecosystem respiration (R), and gas exchange (D) are compared using the (a) observed and (b) detided dissolved oxygen time series as input to each model.
+Results for (a) gross primary production ($P$), (b) ecosystem respiration ($R$), and (c) net ecosystem metabolism ($NEM$) are compared using the observed and detided dissolved oxygen time series as input to each model.
 All values are areal as mmol O$_2$ m$^2$ d$^{-1}$.

### 2023-08-28 R/figs.R

@@ -230,31 +230,8 @@ dev.off()
 
 load(file = url('https://github.com/fawda123/BASEmetab_script/raw/master/data/apacmp.RData'))
 
-toplo <- apacmp %>% 
-  filter(typ != 'BASEmetab') %>% 
-  mutate(
-    val = case_when(
-      var == 'Rt_vol' ~ -1 * val, 
-      T ~ val
-    ), 
-    var = factor(var, 
-                 levels = c('NEM', 'P', 'R', 'D')
-    ), 
-    seas = case_when(
-      month(Date) %in% c(1:5, 10:12) ~ 'dry', 
-      T ~ 'wet'
-    )
-  ) %>% 
-  pivot_wider(names_from = 'typ', values_from = 'val')
+p <- apacmp_plo(apacmp)
 
-p1a <- apacmp_plo(toplo, xlb = NULL, ylb = 'Odum', dotyp = 'observed', addtitle = T)
-p2a <- apacmp_plo(toplo, xlb = 'EBASE', ylb = 'Odum', dotyp = 'detided', addtitle = T)
-
-p <- p1a[[1]] + p1a[[2]] + p1a[[3]] + p1a[[4]] + 
-  p2a[[1]] + p2a[[2]] + p2a[[3]] + p2a[[4]] + 
-  plot_layout(ncol = 4, guides = 'collect') & 
-  theme(legend.position = 'bottom')
-
-png(here('figs/apacmpfig.png'), height = 4.75, width = 8, family = 'serif', units = 'in', res = 500)
+png(here('figs/apacmpfig.png'), height = 7, width = 3.5, family = 'serif', units = 'in', res = 500)
 print(p)
 dev.off()

### 2023-08-28 R/funcs.R

@@ -11,80 +11,6 @@ p_ast <- function(x){
   
 }
 
-# capitalization function
-simpleCap <- function(x) {
-  s <- strsplit(x, " ")[[1]]
-  paste(toupper(substring(s, 1,1)), substring(s, 2),
-        sep="", collapse=" ")
-}
-
-# create subplots for apalachicola obs v dtd metab comparisons
-apacmp_plo <- function(dat, xlb = 'EBASE', ylb = 'Odum', dotyp = 'observed', addtitle = T, alph = 0.5, col = 'black'){
-
-  thm <- theme_minimal() + 
-    theme(
-      strip.background = element_blank(), 
-      legend.position = 'top', 
-      panel.grid.minor = element_blank(), 
-      axis.text = element_text(size = 6)
-    )
-  
-  names(dat)[names(dat) == ylb] <- 'yval'
-  
-  subfigs <- list(
-    observed = '(a)', 
-    detided = '(b)'
-  )
-  
-  dat <- dat %>% 
-    filter(dotyp == !!dotyp) %>% 
-    select(var, seas, yval, EBASE) %>% 
-    na.omit()
-  
-  vars <- levels(dat$var)
-  for(vr in vars){
-    
-    toplotmp <- dat %>% 
-      filter(var == vr)
-    ind <- which(vr == vars)
-  
-    if(ind != 1)
-      ylb <- NULL
-    
-    ttl <- NULL
-    if(addtitle & ind == 1){
-      subfig <- subfigs[[dotyp]]
-      ttl <- paste(subfig, simpleCap(dotyp), 'dissolved oxygen')
-    }
-    
-    lims <- range(toplotmp[, c('yval', 'EBASE')], na.rm = T)
-    
-    ptmp <- ggplot(toplotmp, aes(x = EBASE, y = yval)) + 
-      geom_point(color = col, alpha = alph, stroke = 0, size = 2) + 
-      geom_abline(intercept = 0, slope = 1) +
-      geom_smooth(formula = y ~ x ,method = 'lm', se = F, colour = 'tomato1', size = 0.7) +
-      facet_wrap(~var, ncol = 1) +
-      scale_colour_manual(values = cols) + 
-      scale_x_continuous(limits = lims) + 
-      scale_y_continuous(limits = lims) + 
-      thm + 
-      labs(
-        colour = 'Season', 
-        y = ylb, 
-        title = ttl, 
-        x = xlb
-      )
-    
[…]

### 2023-08-28 manu-draft.qmd
fixing tables from RN comments

@@
 #| label: tbl-apacmptab
-#| tbl-cap: "Summary statistics of daily metabolic estimates comparing the EBASE and Odum methods for net ecosystem metabolism (NEM), gross primary production (P), respiration (R), and gas exchange (D).
-Separate comparisons are made for estimates based on observed and detided dissolved oxygen.
-Summary statistics include Pearson correlation coefficients and root mean square error (RMSE) of a linear fit between the same metabolic estimates obtained from each method."
+#| tbl-cap: "Summary statistics of daily metabolic estimates comparing the EBASE and Odum methods for gross primary production ($P$), ecosystem respiration ($R$), and net ecosystem metabolism ($NEM$) using the observed (left values) and detided (right values) dissolved oxygen time series as input to each model.
+Summary statistics include Pearson correlation coefficients and root mean square difference (RMSD) between the metabolic estimates obtained from each method.
+P-values for correlation coefficients all less than 0.005."
 apacmptab

### 2023-08-28 R/tabs.R

@@ -13,7 +13,7 @@ totab <- tibble(
            'Output', 'Output', 'Output', 'Output', 'Output', 'Output'),
   Description = c('Dissolved oxygen', 'Water temperature', 'Salinity', 'Total photosynthetically active radiation', 'Wind speed', 'Water column depth',
                   'Wind speed at 10 meter height, squared', 'Schmidt number (from water temperature and salinity)', 'Dissolved oxygen at saturation (from water temperature and salinity)',
-                  'Dissolved oxygen (modelled)', 'Production', 'Respiration', 'Gas exchange', 'Light efficiency', 'b'),
+                  'Dissolved oxygen (modelled)', 'Production', 'Respiration', 'Gas exchange', 'Light efficiency', 'Sensitivity of gas transfer to wind speed'),
   `Model notation` =  c('C_d', '-', '-', 'PAR', '-', 'H', 
                         'U102', 'Schmidt number', 'C_sat', 
                         'C_mod', 'P', 'R',  'D', 'a', 'b'),
@@ -28,11 +28,11 @@ eqsz <- fp_text_default(font.size = 9)
 # col3wid <- 2
 # col14wid <- 6.5 - (col2wid + col3wid)
 ebaseiotab <- flextable(totab) %>% 
-  flextable::compose(i = 2, j = 3, value = as_paragraph(as_equation('C_d', props = eqsz))) %>% 
+  flextable::compose(i = 2, j = 3, value = as_paragraph(as_equation('C', props = eqsz))) %>% 
   flextable::compose(i = 5, j = 3, value = as_paragraph(as_equation('PAR', props = eqsz))) %>% 
   flextable::compose(i = 7, j = 3, value = as_paragraph(as_equation('H', props = eqsz))) %>% 
   flextable::compose(i = 9, j = 3, value = as_paragraph(as_equation('U_{10}^2', props = eqsz))) %>% 
-  flextable::compose(i = 10, j = 3, value = as_paragraph(as_equation('S_c', props = eqsz))) %>% 
+  flextable::compose(i = 10, j = 3, value = as_paragraph(as_equation('Sc', props = eqsz))) %>% 
[…]

### 2023-08-31 manu-draft.qmd
update sensitivity figs and text, fix #4

@@
 The minimum and maximum values for the standard deviations for $R$ and $a$ were chosen as 10% and 1000% of the default mean values to test a range from high precision to completely uninformed values.
-The mean and standard deviation for the $b$ parameter were not tested and set as the default values for all analyses given that there is minimal prior knowledge on expectations for this parameter.
-The default prior distribution is already uninformed and we have no appropriate expectations on methods for varying this parameter.
-Each unique combination of low and high values for the means and standard deviations of each parameter were evaluated, creating 2$^4$ = 16 different combinations of prior distributions that were used with the synthetic time series in EBASE.
+The mean and standard deviation for the $b$ parameter were not tested and the prior value was set as the fixed constant used in the synthetic time series (0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$).
+There is minimal prior knowledge on expectations for the $b$ parameter and initial assessments with the synthetic time series demonstrated that the parameter was poorly constrained, primarily because the fixed parameter could not be estimated using observed data (i.e., wind, water temperature) that affected the actual realization of $b$.
+As such, each unique combination of low and high values for the means and standard deviations for $a$ and $R$ were evaluated, creating 2$^4$ = 16 different combinations of prior distributions that were used with the synthetic time series in EBASE.
 
-The optimization period used for a time series determines how much data are used to estimate the three EBASE parameters ($a$, $R$, and $b$).
+The optimization period used for a time series determines how much data are used to estimate the $a$ and $R$ parameters.
 Unlike BASE, the EBASE software allows flexibility in the period of time used for optimization, with a minimum being one day and a maximum being the length of the time series.
-For a chosen optimization period, single estimates for $a$, $R$, and $b$ are returned by the model.
+For a chosen optimization period, single estimates for $a$ and $R$ are returned by the model ($b$ is also returned, but was fixed for the sensitivity analysis).
 A short period of optimization (e.g., one day) may return unstable parameter estimates if forcing factors (e.g., wind, temperature) that affect estuarine biogeochemical rates occur at time scales longer the chosen period of optimization for EBASE.
@@
 Overall, most models regardless of optimization period or priors were able to reproduce the dissolved oxygen time series.
-The only models where the NSE values were less than zero for dissolved oxygen were those where $\sigma$ of the $R$ parameter was low.
+The only models where the NSE values were less than zero for dissolved oxygen were those where $\sigma$ of the $a$ and $R$ parameters was low and $mu$ for one of the $a$ and $R$ parameters was high.
 However, substantial variation in NSE values was observed among all other parameters.
 Not considering dissolved oxygen, gas exchange ($D$) was estimated with the highest NSE, with a median value across all optimization periods and priors slightly larger than zero (median NSE = `r prmall[['D']]`).
-Conversely, the $a$ parameter was reproduced the most poorly, with a median value across all optimization periods and priors less than zero (median NSE = `r prmall[['a']]`).
-By optimization period, median NSE values across all parameters decreased slightly as the number of days used for model optimization increased (1 day = `r optnse[['1']]`, 7 days = `r optnse[['7']]`, 30 days = `r optnse[['30']]`).
-The top three ranked models were identical for the seven and thirty day models and similar for models using the one day optimization period.
-Median NSE values across parameters for the top model in each optimization period were also similar (1 day = `r optmod[['1']]`, 7 days = `r optmod[['7']]`, 30 days = `r optmod[['30']]`).
-For the seven day and thirty day optimization periods, the NSE values for all parameters were greater than zero for the top models, where two of the parameters in the top model for the one day optimization period has NSE \< 0.
-Most parameters regardless of optimization period or priors had NSE \< 0 (`r perneg`).
-As such, the prior combinations that produced the highest NSE values across all parameters were those where the $\sigma$ values were high for each prior distribution, regardless of the values for $\mu$, i.e., those with sufficient flexibility to search a larger parameter space regardless of $\mu$ produced metabolic parameter estimates most similar to the known values from the synthetic time series.
+Conversely, the $a$ parameter was reproduced the most poorly, with a median value less than zero across all optimization periods and priors (median NSE = `r prmall[['a']]`).
+By optimization period, median NSE values across all parameters increased as the number of days used for model optimization increased (1 day = `r optnse[['1']]`, 7 days = `r optnse[['7']]`, 30 days = `r optnse[['30']]`).
+The top three ranked models were identical across optimization periods and median NSE values across parameters for the top models increased with optimization period (1 day = `r optmod[['1']]`, 7 days = `r optmod[['7']]`, 30 days = `r optmod[['30']]`).
+The NSE values for all parameters were greater than zero for the top three models in the 30 day optimization period, whereas the $a$ parameter had NSE less than zero for the top model in the 7 day optimization period and $R$ and $a$ had negative NSE values for the top model in the one day optimization period.
+About half of the parameters regardless of optimization period or priors had NSE \< 0 (`r perneg`).
+As such, the prior combinations that produced the highest NSE values across all parameters were those where all prior distributions had low $\mu$ and $\sigma$ values for both $a$ and $R$, although the second and third best performing models had high $\sigma$ for $R$.
 
 @fig-priorsumcomp provides an alternative summary of how the prior distributions influenced the ability of EBASE to reproduce the synthetic parameters.
-Median NSE values across all models, parameters, and each optimization period for the different subsets of the prior combinations are shown relative to the grand median NSE value across all comparisons.
-For example, the bottom left panel with the x-axis value for the $a$ parameter shows the difference in the median NSE values for all models using a one-day optimization period where the $\sigma$ value for the $a$ prior distribution was low or high.
-The size of the points indicate the inter-quartile range of NSE values for each subset.
+Median NSE values across all models, parameters, and each optimization period for the different subsets of the prior combinations are shown.
+For example, the left-most bars in top left panel with the x-axis value for the $a$ parameter shows the median NSE values for all models using a one-day optimization period where the $\mu$ value for the $a$ prior distribution was low or high.
+The lines indicate the inter-quartile range of NSE values for each subset.
 As such, each set of points denoted by the x-axis parameters in each panel include median NSE values for all models by optimization period, but subset by the low/high $\mu$ or $\sigma$ values for each parameter.
-Positive values indicate relative performance greater than the overall median NSE values and negative values indicate relative performance worse than the overall median.
 Dissolved oxygen was generally modeled with high accuracy for all priors and it was excluded to not skew the assessment.
-The patterns suggest that EBASE results are more sensitive to changes in the $\sigma$ parameter as shown by the larger spread in the median results compared to those for $\mu$.
-This result is especially true for the priors of the $a$ and $R$ parameters.
-Conversely, model results are less sensitive to low or high values for $\mu$ as the spread between median NSE values is less than those for $\sigma$.
-This suggests that EBASE model sensitivity is most influenced by $\sigma$ of the prior distributions, i.e., larger values produce more accurate and less variable estimates regardless of $\mu$.
-Patterns by optimization period were not apparent.
+The patterns suggest that EBASE results are more sensitive to model optimization period as shown by higher median NSE values with longer optimization times (note the different y-axis scales, also confirmed in @fig-priorplot).
+Similarly, median values were much more variable for the one day optimization period, as shown by a mix of both positive and negative values in the interquartile ranges.
+Lower median NSE values were observed for higher values for $\mu$, where higher median values for were observed for higher values of $\sigma$, regardless of optimization period.
 
-The actual model results for the best and worst performing combinations of prior distributions for the one day, seven day, and thirty day optimization periods are shown in @fig-optex (all results are averaged at the time-step of the optimization period for comparability).
+The actual model results for the best and worst performing combinations of prior distributions for the one day, seven day, and thirty day optimization periods are shown in @fig-optex (all results are averaged at the time-step of the optimization period for visual comparability).
 The best performing models were those with the highest median NSE values across all parameter estimates in @fig-priorcomp.
 The best performing models for each optimization period (@fig-optex a, c, e) showed similar results where the parameter estimates were similar to the synthetic time series, although more variation in the EBASE results were of course observed with the one day optimization period.
-However, the $b$ parameter was estimated with high variability for the one and seven day optimization periods (@fig-optex a, c, bottom row) and with a large positive bias for the 30 day optimization period (@fig-optex e, bottom row).
-These results were not unexpected because it was impossible to evaluate NSE estimates using a fixed value in the synthetic results and the accuracy was not considered in identifying the best performing model.
-Finally, the worst performing model for each optimization period (@fig-optex b, d, f) demonstrates the negative NSE values that were obtained (i.e., EBASE values much lower or higher than the synthetic values).
-The worst performing models were caused by constraining EBASE to a small parameter space using low or high values for $\mu$ combined with low values for $\sigma$.
+The worst performing model for each optimization period (@fig-optex b, d, f) demonstrates the negative NSE values that were obtained (i.e., EBASE values much lower or higher than the synthetic values).
+The worst performing models were caused by constraining EBASE to a small parameter space, such as a high $\mu$ and low $\sigma$ for $a$ as in the bottom row for subplot b, d, and f.
 
-As a final evaluation of EBASE to return known metabolic parameters, a second synthetic time series was created that added estimated effects of instrument and tidal noise to the original synthetic time series.
+As a final evaluation of EBASE to return known metabolic parameters, a second synthetic time series was created that added estimated effects of noise to the original synthetic time series.
 The objective of the analysis is to develop an expectation of how metabolic estimates may change using in situ data with noise that may substantially mask a biological signal in dissolved oxygen measurements.
-The instrument and tidal noise were estimated using a weighted regression analysis [@Beck15; @Beck21] on actual observed dissolved oxygen time series for Apalachicola Bay using data from 2021.
+The noise estimates were created using a weighted regression analysis [@Beck15; @Beck21] on actual observed dissolved oxygen time series for Apalachicola Bay using data from 2021.
 This approach models dissolved oxygen as a function of time, salinity, and season using a moving window approach with weights of the independent variables appropriate for each time period within the windows.
@@
 The model predictions of dissolved oxygen (DO), gross primary production (P), ecosystem respiration (R), gas exchange (D), and the $a$ parameter are evaluated based on Nash-Sutcliffe Efficiency (NSE) values between the EBASE results and synthetic values.
-NSE values from zero to one (green) indicate EBASE has some skill in recovering the synthetic values of DO, $D$, $P$, $R$ , and $a$, with values closer to 1 being a better approximation, whereas negative values indicate no skill.
-The color scale for NSE is also exponentially transformed to minimize the influence of large negative values on the visual display.
 The $b$ parameter is a fixed constant and was not evaluated.
@@
 
-![Summaries of % change in Nash-Sutcliffe Efficiency (NSE) values across all the comparisons of EBASE and the five synthetic model outputs separated by different combinations of prior values.
-Summaries are shown relative to the change from the median for each comparison group (i.e., optimization period and prior parameter, $n$ = 32).
+![Summaries of Nash-Sutcliffe Efficiency (NSE) values across all the comparisons of EBASE and the five synthetic model outputs separated by different combinations of prior values (note the different y-axis scales).
 Each result on the x-axis represents the median values for which the combination prior applies.
-For example, the points in the upper left panel that are fartheset to the left summarize the results from @fig-priorcomp where $\mu$ is high or low for the $a$ parameter using a one day optimization period.
-Point size represents the inter-quartile range of median NSE values for each result ($n$ = 8) .](figs/priorsumcomp.png){#fig-priorsumcomp}
+For example, the bars in the upper left panel that are farthest to the left summarize the results from @fig-priorcomp where $\mu$ is high or low for the $a$ parameter using a one day optimization period.
+Bars are the medians and lines the interquartile ranges of NSE values for each result ($n$ = 8 for each bar) .](figs/priorsumcomp.png){#fig-priorsumcomp}
 
@@
 The best and worst performing models for the one day (a, b), seven day (c, d), and thirty day (e, f) optimization periods are based on the EBASE results with prior distributions that produced the highest and lowest median Nash-Sutcliffe Efficency values across all parameter estimates (@fig-priorcomp).
-The results are averaged at the time step of the model optimization period.
-The $b$ parameter is fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$.](figs/optex.png){#fig-optex fig-align="center"}
+The results are averaged at the time step of the model optimization period.](figs/optex.png){#fig-optex fig-align="center"}
 

### 2023-08-31 R/figs.R

@@ -109,7 +109,7 @@ load(file = fl)
 
 p1 <- priorsumcomp(apasumdat, met = 'nse')
 
-png(here('figs/priorsumcomp.png'), height = 4, width = 5, family = 'serif', units = 'in', res = 500)
+png(here('figs/priorsumcomp.png'), height = 4.5, width = 6, family = 'serif', units = 'in', res = 500)
 print(p1)
 dev.off()
 
@@ -142,7 +142,7 @@ lims <- list(
 
 p <- optex(apagrd, fwdatcmp, apasumdat, rnkmetsum = c(1, 16), met = 'nse', lims = lims)
 
-png(here('figs/optex.png'), height = 8.7, width = 9, family = 'serif', units = 'in', res = 500)
+png(here('figs/optex.png'), height = 8, width = 9, family = 'serif', units = 'in', res = 500)
 print(p)
 dev.off()
 

### 2023-08-31 R/funcs.R

@@ -156,6 +156,14 @@ priorcomp <- function(dat, met, topbot = 3){
 # Fwoxy apa comparison to EBASE for different priors, rmse summary
 priorsumcomp <- function(dat, met = 'r2'){
 
+  
+  metsel <- tibble(
+    met = c('r2', 'rmse', 'mape', 'mae', 'nse'),
+    lbspr = c('italic(R)^2', 'RMSE', 'MAPE', 'mae', 'NSE')
+  ) %>% 
+    filter(met == !!met)
+  ylab <- parse(text = paste0('`%`', '~change~', metsel$lbspr))
+  
   toplo <- dat %>% 
     unnest(ests) %>% 
     select(amean, asd, rmean, rsd, ndays, ind, var, !!met) %>% 
@@ -175,50 +183,62 @@ priorsumcomp <- function(dat, met = 'r2'){
       .by = c('ndays', 'prior', 'val')
     ) %>% 
     mutate(
-      grandmed = median(medv), 
-      .by = c('ndays', 'val')
-    ) %>% 
-    mutate(
-           medv = 100 * (medv - grandmed) / ((grandmed + medv) / 2),
-           param = gsub('mean$|sd$', '', prior),
-           param = factor(param, levels = c('a', 'r'), labels = c('a', 'R')),
-           prior = gsub('^a|^r|^b', '', prior), 
-           prior = factor(prior, levels = c('mean', 'sd'), labels = c('mu', 'sigma')), 
-           ndays = factor(ndays, levels = c('1 day', '7 days', '30 days'), labels = c('1~day', '7~days', '30~days'))
+      upr = medv + (iqr / 2), 
+      lwr = medv - (iqr / 2),
+      param = gsub('mean$|sd$', '', prior),
+      param = factor(param, levels = c('a', 'r'), labels = c('a', 'R')),
+      prior = gsub('^a|^r|^b', '', prior), 
+      prior = factor(prior, levels = c('mean', 'sd'), labels = c('mu', 'sigma')), 
+      ndays = factor(ndays, levels = c('1 day', '7 days', '30 days'), labels = c('1~day', '7~days', '30~days'))
     )
   
-  metsel <- tibble(
-    met = c('r2', 'rmse', 'mape', 'mae', 'nse'),
-    lbspr = c('italic(R)^2', 'RMSE', 'MAPE', 'mae', 'NSE')
-  ) %>% 
-    filter(met == !!met)
-  ylab <- parse(text = paste0('`%`', '~change~', metsel$lbspr))
[…]

### 2023-08-31 manu-draft.qmd
update sensitivity analysis results text

@@
 Functions are available for viewing model results, including a plot of the modeled dissolved oxygen with the observed, a time series plot of the primary metabolic estimates (P, R, and D, as areal rates with units $\text{mmol}/\text{m}^2/\text{d}$), and a time series plot of the credible intervals for the $a$, $R$, and $b$ parameters.
-An additional option allows the user to specify the model optimization period, where the time period can vary from a minimum of one day to the maximum of the entire length of the time series.
-This differs from the BASE approach where the optimization period is set at one day.
+An additional option allows the user to specify the model optimization period, where the time period can vary from a minimum of 1 day to the maximum of the entire length of the time series.
+This differs from the BASE approach where the optimization period is set at 1 day.
 Sensitivity of the EBASE method to different optimization periods is described in @sec-cmpmod.
@@
 First, results were compared to the known values after changing characteristics of their prior distributions.
-Second, results were compared using a model period of optimization for one, seven, and 30 days to assess how results varied based on the amount of data used for the Bayesian estimation.
+Second, results were compared using a model period of optimization for 1, 7, and 30 days to assess how results varied based on the amount of data used for the Bayesian estimation.
 Details of these analyses are as follows.
@@
 The optimization period used for a time series determines how much data are used to estimate the $a$ and $R$ parameters.
-Unlike BASE, the EBASE software allows flexibility in the period of time used for optimization, with a minimum being one day and a maximum being the length of the time series.
+Unlike BASE, the EBASE software allows flexibility in the period of time used for optimization, with a minimum being 1 day and a maximum being the length of the time series.
 For a chosen optimization period, single estimates for $a$ and $R$ are returned by the model ($b$ is also returned, but was fixed for the sensitivity analysis).
-A short period of optimization (e.g., one day) may return unstable parameter estimates if forcing factors (e.g., wind, temperature) that affect estuarine biogeochemical rates occur at time scales longer the chosen period of optimization for EBASE.
+A short period of optimization (e.g., 1 day) may return unstable parameter estimates if forcing factors (e.g., wind, temperature) that affect estuarine biogeochemical rates occur at time scales longer the chosen period of optimization for EBASE.
 As such, potentially longer optimization periods may return more stable and robust estimates given that more data are used for parameter estimation, although at the risk of overgeneralizing metabolic rates if variation occurs at time scales less than the optimization period.
-Results from optimization periods of one, seven, and thirty days were compared for the synthetic one-year time series.
+Results from optimization periods of 1, 7, and 30 days were compared for the synthetic one-year time series.
 Each optimization period was also evaluated for every unique combination of prior distributions, totaling 48 (16 times three optimization periods) unique analyses for the assessment of changing priors of all parameters.
@@
 Changing the prior distributions affected the ability of EBASE to return the known parameters from the synthetic time series, with results varying by optimization period.
-@fig-priorcomp shows a summary of the comparisons of the synthetic data with EBASE for each unique combination of prior distributions and the three optimization periods of one, seven, and thirty days.
+@fig-priorcomp shows a summary of the comparisons of the synthetic data with EBASE for each unique combination of prior distributions and the three optimization periods of 1, 7, and 30 days.
 The results are shown using Nash-Sutcliffe Efficiency (NSE) values [@Nash70; @Moriasi07]:
@@
 where $Y_i$ is the estimate from EBASE or the synthetic times series for the optimization period $i$, evaluated for each metabolic parameter, each unique combination of priors, and each optimization period.
-The value $n$ is the number of optimization periods in the one-year time series, i.e., $n=365$ for one day, $n \approx 52$ for seven days, and $n \approx 12$ for 30 days.
+The value $n$ is the number of optimization periods in the one-year time series, i.e., $n=365$ for 1 day, $n \approx 52$ for 7 days, and $n \approx 12$ for 30 days.
 All metabolic estimates were averaged within the number of days for each optimization period prior to calculating NSE values, if appropriate.
@@
 The top three ranked models were identical across optimization periods and median NSE values across parameters for the top models increased with optimization period (1 day = `r optmod[['1']]`, 7 days = `r optmod[['7']]`, 30 days = `r optmod[['30']]`).
-The NSE values for all parameters were greater than zero for the top three models in the 30 day optimization period, whereas the $a$ parameter had NSE less than zero for the top model in the 7 day optimization period and $R$ and $a$ had negative NSE values for the top model in the one day optimization period.
+The NSE values for all parameters were greater than zero for the top three models in the 30 day optimization period, whereas the $a$ parameter had NSE less than zero for the top model in the 7 day optimization period and $R$ and $a$ had negative NSE values for the top model in the 1 day optimization period.
 About half of the parameters regardless of optimization period or priors had NSE \< 0 (`r perneg`).
@@
 Median NSE values across all models, parameters, and each optimization period for the different subsets of the prior combinations are shown.
-For example, the left-most bars in top left panel with the x-axis value for the $a$ parameter shows the median NSE values for all models using a one-day optimization period where the $\mu$ value for the $a$ prior distribution was low or high.
+For example, the left-most bars in top left panel with the x-axis value for the $a$ parameter shows the median NSE values for all models using a 1-day optimization period where the $\mu$ value for the $a$ prior distribution was low or high.
 The lines indicate the inter-quartile range of NSE values for each subset.
@@
 The patterns suggest that EBASE results are more sensitive to model optimization period as shown by higher median NSE values with longer optimization times (note the different y-axis scales, also confirmed in @fig-priorplot).
-Similarly, median values were much more variable for the one day optimization period, as shown by a mix of both positive and negative values in the interquartile ranges.
+Similarly, median values were much more variable for the 1 day optimization period, as shown by a mix of both positive and negative values in the interquartile ranges.
 Lower median NSE values were observed for higher values for $\mu$, where higher median values for were observed for higher values of $\sigma$, regardless of optimization period.
 
-The actual model results for the best and worst performing combinations of prior distributions for the one day, seven day, and thirty day optimization periods are shown in @fig-optex (all results are averaged at the time-step of the optimization period for visual comparability).
+The actual model results for the best and worst performing combinations of prior distributions for the 1 day, 7 day, and 30 day optimization periods are shown in @fig-optex (all results are averaged at the time-step of the optimization period for visual comparability).
 The best performing models were those with the highest median NSE values across all parameter estimates in @fig-priorcomp.
-The best performing models for each optimization period (@fig-optex a, c, e) showed similar results where the parameter estimates were similar to the synthetic time series, although more variation in the EBASE results were of course observed with the one day optimization period.
+The best performing models for each optimization period (@fig-optex a, c, e) showed similar results where the parameter estimates were similar to the synthetic time series, although more variation in the EBASE results were of course observed with the 1 day optimization period.
 The worst performing model for each optimization period (@fig-optex b, d, f) demonstrates the negative NSE values that were obtained (i.e., EBASE values much lower or higher than the synthetic values).
@@
 The latter comparison provided an assessment of metabolic rates that are expected to be minimally influenced by tidal advection, as a fundamental assumption of metabolic models using *in situ* dissolved oxygen time series from fixed monitoring stations.
-The prior distributions for EBASE were uninformed using the same large $\sigma$ values from @sec-cmpmod and moderate $\mu$ values ($\mu$ = 0.2 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$ for $a$, $\mu$ = 20 $\text{mmol}/\text{m}^3/\text{d}$ for $R$, $\mu$ = 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ for $b$).
-A one day model optimization period was used for EBASE for comparability with the Odum method.
+The prior distributions for EBASE used the default values justified in @sec-ebasemod (equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]).
+A 1 day model optimization period was used for EBASE for comparability with the Odum method that returns daily estimates.
 
@@
 As such, using one week of data for each optimization period is likely appropriate for this location.
-For other locations, using one day is not recommended given the issues stated above (i.e., unstable results), whereas an optimization period longer than a month may miss important events in the time series that can affect the metabolic estimates and will likely produce parameters that are overly generalized.
+For other locations, using 1 day is not recommended given the issues stated above (i.e., unstable results), whereas an optimization period longer than a month may miss important events in the time series that can affect the metabolic estimates and will likely produce parameters that are overly generalized.
 
@@
 Results are compared for unique combinations of prior distributions and optimization period.
-The means ($\mu$) and standard deviations ($\sigma$) are evaluated as low (L) or high (H) values (see text for details) and the optimization period for EBASE is one day, seven days, or thirty days.
+The means ($\mu$) and standard deviations ($\sigma$) are evaluated as low (L) or high (H) values (see text for details) and the optimization period for EBASE is 1 day, 7 days, or 30 days.
 The model predictions of dissolved oxygen (DO), gross primary production (P), ecosystem respiration (R), gas exchange (D), and the $a$ parameter are evaluated based on Nash-Sutcliffe Efficiency (NSE) values between the EBASE results and synthetic values.
@@
 Each result on the x-axis represents the median values for which the combination prior applies.
-For example, the bars in the upper left panel that are farthest to the left summarize the results from @fig-priorcomp where $\mu$ is high or low for the $a$ parameter using a one day optimization period.
+For example, the bars in the upper left panel that are farthest to the left summarize the results from @fig-priorcomp where $\mu$ is high or low for the $a$ parameter using a 1 day optimization period.
 Bars are the medians and lines the interquartile ranges of NSE values for each result ($n$ = 8 for each bar) .](figs/priorsumcomp.png){#fig-priorsumcomp}
@@
 ![Comparison of EBASE results for the best and worst performing models to a synthetic time series created with known metabolic parameters.
-The best and worst performing models for the one day (a, b), seven day (c, d), and thirty day (e, f) optimization periods are based on the EBASE results with prior distributions that produced the highest and lowest median Nash-Sutcliffe Efficency values across all parameter estimates (@fig-priorcomp).
+The best and worst performing models for the 1 day (a, b), 7 day (c, d), and 30 day (e, f) optimization periods are based on the EBASE results with prior distributions that produced the highest and lowest median Nash-Sutcliffe Efficency values across all parameter estimates (@fig-priorcomp).
 The results are averaged at the time step of the model optimization period.](figs/optex.png){#fig-optex fig-align="center"}

### 2023-08-31 manu-draft.qmd
more edits from RN, mostly for sensitivity analysis

@@
 
-where $Y_i$ is the estimate from EBASE or the synthetic times series for the optimization period $i$, evaluated for each metabolic parameter, each unique combination of priors, and each optimization period.
-The value $n$ is the number of optimization periods in the one-year time series, i.e., $n=365$ for 1 day, $n \approx 52$ for 7 days, and $n \approx 12$ for 30 days.
-All metabolic estimates were averaged within the number of days for each optimization period prior to calculating NSE values, if appropriate.
+where $Y_i$ is the estimate from EBASE or the synthetic times series for the optimization period $i$, evaluated for each of five model outputs ($D$, $DO$, $R$, $a$, and $P$), each of 16 unique combinations of priors, and each of 3 optimization periods.
+The value $n$ is the number of optimization periods in the one-year time series, i.e., $n=365$ for comparability among the optimization periods.
 The NSE value is conceptually similar to the coefficient of determination, but varies from $-\infty$ to 1.
@@
 Several conclusions can be made from the results in @fig-priorcomp following the patterns of the NSE values.
-Overall, most models regardless of optimization period or priors were able to reproduce the dissolved oxygen time series.
+Overall, most models regardless of optimization period or priors were able to show some skill at reproducing the dissolved oxygen time series.
 The only models where the NSE values were less than zero for dissolved oxygen were those where $\sigma$ of the $a$ and $R$ parameters was low and $mu$ for one of the $a$ and $R$ parameters was high.
-However, substantial variation in NSE values was observed among all other parameters.
+However, substantial variation in NSE values was observed among all other model outputs, with some very low values of NSE occurring while the $DO$ simulation was fairly good, showing the potential for the model to produce the right answer for the wrong reasons.
 Not considering dissolved oxygen, gas exchange ($D$) was estimated with the highest NSE, with a median value across all optimization periods and priors slightly larger than zero (median NSE = `r prmall[['D']]`).
+This result makes sense, given that the $b$ parameter was fixed at the actual value used to create the synthetic time series.
 Conversely, the $a$ parameter was reproduced the most poorly, with a median value less than zero across all optimization periods and priors (median NSE = `r prmall[['a']]`).
-By optimization period, median NSE values across all parameters increased as the number of days used for model optimization increased (1 day = `r optnse[['1']]`, 7 days = `r optnse[['7']]`, 30 days = `r optnse[['30']]`).
-The top three ranked models were identical across optimization periods and median NSE values across parameters for the top models increased with optimization period (1 day = `r optmod[['1']]`, 7 days = `r optmod[['7']]`, 30 days = `r optmod[['30']]`).
-The NSE values for all parameters were greater than zero for the top three models in the 30 day optimization period, whereas the $a$ parameter had NSE less than zero for the top model in the 7 day optimization period and $R$ and $a$ had negative NSE values for the top model in the 1 day optimization period.
-About half of the parameters regardless of optimization period or priors had NSE \< 0 (`r perneg`).
-As such, the prior combinations that produced the highest NSE values across all parameters were those where all prior distributions had low $\mu$ and $\sigma$ values for both $a$ and $R$, although the second and third best performing models had high $\sigma$ for $R$.
 
-@fig-priorsumcomp provides an alternative summary of how the prior distributions influenced the ability of EBASE to reproduce the synthetic parameters.
+By optimization period, median NSE values across all model outputs increased as the number of days used for model optimization increased (1 day = `r optnse[['1']]`, 7 days = `r optnse[['7']]`, 30 days = `r optnse[['30']]`).
+The top three ranked models were identical across optimization periods and median NSE values across model outputs for the top models increased with optimization period (1 day = `r optmod[['1']]`, 7 days = `r optmod[['7']]`, 30 days = `r optmod[['30']]`).
+The NSE values for all model outputs were greater than zero for the top three models in the 30 day optimization period, whereas the $a$ parameter had NSE less than zero for the top model in the 7 day optimization period and $R$ and $a$ had negative NSE values for the top model in the 1 day optimization period.
+About half of the model outputs regardless of optimization period or priors had NSE \< 0 (`r perneg`).
+As such, the prior combinations that produced the highest NSE values across all parameters were those where all prior distributions had low $\mu$ and $\sigma$ values for both $a$ and $R$ for the 1 and 7 day optimization periods, whereas the same combination was observed for the 30 day optimization period with the exception that $\sigma$ was high for $R$.
+Notably, many of the top models for each optimization period had at least one high value for $\sigma$ for $a$ or $R$, suggesting that model withs sufficient flexibility to search a larger parameter space can produce metabolic estimates most similar to actual values.
+
+@fig-priorsumcomp provides an alternative summary of how the prior distributions influenced the ability of EBASE to reproduce the synthetic results.
 Median NSE values across all models, parameters, and each optimization period for the different subsets of the prior combinations are shown.
@@
 As such, each set of points denoted by the x-axis parameters in each panel include median NSE values for all models by optimization period, but subset by the low/high $\mu$ or $\sigma$ values for each parameter.
-Dissolved oxygen was generally modeled with high accuracy for all priors and it was excluded to not skew the assessment.
+Dissolved oxygen was generally modeled with high skill for all priors and it was excluded to not skew the assessment.
 The patterns suggest that EBASE results are more sensitive to model optimization period as shown by higher median NSE values with longer optimization times (note the different y-axis scales, also confirmed in @fig-priorplot).
@@
 
-The actual model results for the best and worst performing combinations of prior distributions for the 1 day, 7 day, and 30 day optimization periods are shown in @fig-optex (all results are averaged at the time-step of the optimization period for visual comparability).
-The best performing models were those with the highest median NSE values across all parameter estimates in @fig-priorcomp.
-The best performing models for each optimization period (@fig-optex a, c, e) showed similar results where the parameter estimates were similar to the synthetic time series, although more variation in the EBASE results were of course observed with the 1 day optimization period.
+The actual model results for the best and worst performing combinations of prior distributions for the three optimization periods are shown in @fig-optex (all results are averaged at the time-step of the optimization period for visual comparability).
+The best performing models were those with the highest median NSE values across all model outputs in @fig-priorcomp.
+The best performing models for each optimization period (@fig-optex a, c, e) showed similar results where the model outputs were similar to the synthetic time series, although more variation in the EBASE results were of course observed with the 1 day optimization period.
 The worst performing model for each optimization period (@fig-optex b, d, f) demonstrates the negative NSE values that were obtained (i.e., EBASE values much lower or higher than the synthetic values).
-The worst performing models were caused by constraining EBASE to a small parameter space, such as a high $\mu$ and low $\sigma$ for $a$ as in the bottom row for subplot b, d, and f.
+The worst performing models were caused by constraining EBASE to a small parameter space, such as a high $\mu$ and low $\sigma$ for $a$ as in the bottom row for subplots (b), (d), and (f).
+In such cases, the $DO$ simulation (fourth row) can still be quite good as a result of compensating errors in $a$ and $R$.
+The EBASE-derived rates of photosynthesis and respiration are both much higher than in the synthetic data for subplots (b) and (d), errors that compensate well enough to produce fairly accurate recoveries of the $DO$ time series.
+Further, the worst model for the 30 day optimization period in subplot (f) had the same priors as the 7 day optimization period, but was able to better reproduce the synthetic results, suggesting that poor estimates from inadequate priors can in part be mitigated using longer optimization periods.
 
 As a final evaluation of EBASE to return known metabolic parameters, a second synthetic time series was created that added estimated effects of noise to the original synthetic time series.
-The objective of the analysis is to develop an expectation of how metabolic estimates may change using in situ data with noise that may substantially mask a biological signal in dissolved oxygen measurements.
+The objective of the analysis is to develop an expectation of how metabolic estimates may change using *in situ* data with noise that may substantially mask a biological signal in dissolved oxygen measurements.
 The noise estimates were created using a weighted regression analysis [@Beck15; @Beck21] on actual observed dissolved oxygen time series for Apalachicola Bay using data from 2021.
-This approach models dissolved oxygen as a function of time, salinity, and season using a moving window approach with weights of the independent variables appropriate for each time period within the windows.
+This approach models dissolved oxygen as a function of time (day), hour, and water depth using a moving window approach with weights of the independent variables appropriate for each time period within the windows.
 The results of this analysis return a predicted and detided dissolved oxygen time series.
-The instrument noise was estimated as the difference between the observed and predicted time series and the tidal noise was estimated as the difference between the predicted and detided time series.
-Both the instrument and tidal noise were added to the original synthetic time series above (@fig-synapanos).
-EBASE was then used with unconstrained priors (large $\sigma$) on the synthetic time series with noise, then compared with results from the original synthetic time series.
+Two separate time series of noise, representing residual (any other random process not related to metabolism or tidal advection) or tidal noise, were dervied from the observed, predicted and detided time series.
+The residual noise was estimated as the difference between the observed and predicted time series and the tidal noise was estimated as the difference between the predicted and detided time series.
+Both the residual and tidal noise were added to the original synthetic time series above (@fig-synapanos).
+EBASE was then used on the synthetic time series with noise, then compared with results from the original synthetic time series.
+The default priors justified in @sec-ebasemod (equations [-@eq-aprior] and [-@eq-rprior]) were used with a 7 day optimization period.
+The prior for the $b$ parameter was fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ following similar logic used for the sensitivity analysis described above.
 
-@fig-synapanoscmp shows the comparison of the EBASE results for the synthetic time series with instrument and tidal noise to those from the original synthetic time series.
-Results for both time series produced similar dissolved oxygen estimates (top-left panel, @fig-synapanoscmp), although notable differences were observed in the remaining parameters.
+@fig-synapanoscmp shows the comparison of the EBASE results for the synthetic time series with noise (residual and tidal) to those from the original synthetic time series without noise.
+Only results for $a$ and $R$ are shown Results for both time series produced similar dissolved oxygen estimates (top-left panel, @fig-synapanoscmp), although notable differences were observed in the remaining parameters.
 In particular, gross primary production (P) and ecosystem respiration (R) were generally higher and more variable for results from the noisy time series, with differences in the former attributed to higher estimates in the $a$ parameter.
@@
 ![A subset of the synthetic time series and the synthetic time series with noise for evaluating the ability of EBASE to recover known metabolic parameters.
-The original synthetic time series and the same time series with instrument and tidal noise is shown in (a) and the estimated instrument and tidal noise is shown in (b).](figs/synapanos.png){#fig-synapanos fig-align="center"}
+The original synthetic time series and the same time series with residual and tidal noise is shown in (a) and the estimated residual and tidal noise added to the synthetic time series is shown in (b).](figs/synapanos.png){#fig-synapanos fig-align="center"}
 

### 2023-08-31 R/figs.R

@@ -184,7 +184,7 @@ toplo2 <- toplo %>%
   pivot_longer(-DateTimeStamp) %>% 
   mutate(
     name = case_when(
-      name == 'obsnoise' ~ 'Instrument', 
+      name == 'obsnoise' ~ 'Residual', 
       name == 'tidnoise' ~ 'Tidal'
     ),
     value = value / 32 * 1000

### 2023-09-01 manu-draft.qmd
finished RN edits

@@
 
-@fig-synapanoscmp shows the comparison of the EBASE results for the synthetic time series with noise (residual and tidal) to those from the original synthetic time series without noise.
-Only results for $a$ and $R$ are shown Results for both time series produced similar dissolved oxygen estimates (top-left panel, @fig-synapanoscmp), although notable differences were observed in the remaining parameters.
-In particular, gross primary production (P) and ecosystem respiration (R) were generally higher and more variable for results from the noisy time series, with differences in the former attributed to higher estimates in the $a$ parameter.
-As for the previous results, the $b$ parameter was estimated with low precision for both input time series, although results were centered around the true value of 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$.
-Gas exchange (D) was similar between the input time series, suggesting that lack of precision in the $b$ parameter had a minimal influence regardless of noise.
+@fig-synapanoscmp shows the comparison of the results recovered from EBASE for the synthetic time series with noise (residual and tidal) and for the synthetic time series without noise.
+EBASE was able to recover reasonable estimates for both $a$ and $R$, although the results from the noisy time series were slightly biased towards higher values as shown by the blue regression lines on the right side of @fig-synapanoscmp.
+The higher values for $a$ will produce higher estimates for $P$, and combined with $R$, will produce inflated values for metabolic estimates compared to the results for the synthetic time series without noise.
+Although the results for the latter were not biased towards higher values, both estimates provided by EBASE had a relatively large range of variablity around the known values.
+As noted above, a longer optimization period could reduce this variability.
 
@@
 For a better comparison with the EBASE methods, the gas exchange parameterization in the Odum method was modified to use the @Wanninkhof14 approach and the $b$ parameter was fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ [as available in the WtRegDO R package, @Beck21].
-An additional and fundamental difference between EBASE and the Odum method is that the former is likelihood based with metabolic rates estimated by fitting the model to the data, whereas the latter is strictly arithmetic based primarily on integration of the diel dissolved oxygen curve.
-Thus, potentially different results are expected between the approaches based on differences in formulation of the mass balance equations for metabolism, as well as alternative statistical approaches to estimating the parameters.
-Each comparison also evaluated differences in the metabolic estimates using the observed dissolved oxygen time series and a detided time series using weighted regression [@Beck15; @Beck21].
+An additional and fundamental difference between EBASE and the Odum method is that the former is likelihood-based with metabolic rates estimated by fitting the model to the data, whereas the latter is strictly arithmetic-based primarily on integration of the diel dissolved oxygen curve.
+Thus, potentially different results are expected based on the alternative statistical approaches to estimate the metabolic results.
+Each comparison also evaluated differences in the results using the observed dissolved oxygen time series and a detided time series using weighted regression [@Beck15; @Beck21].
 The latter comparison provided an assessment of metabolic rates that are expected to be minimally influenced by tidal advection, as a fundamental assumption of metabolic models using *in situ* dissolved oxygen time series from fixed monitoring stations.
 The prior distributions for EBASE used the default values justified in @sec-ebasemod (equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]).
+Compared to the previous analyses, the prior distribution for the $b$ parameter was not fixed given that its estimate should vary as a function of the data using real observations.
 A 1 day model optimization period was used for EBASE for comparability with the Odum method that returns daily estimates.
 
-The common set of metabolic parameters returned by the EBASE and Odum methods are net ecosystem metabolism (NEM), gross primary production (P), ecosystem respiration (R), and gas exchange (D).
-Pairwise comparisons of the daily estimates returned by each method were evaluated using simple summary statistics, including Pearson correlation coefficients ($\rho$) and root mean square errors (RMSE) of the linear regression fit comparing the same metabolic parameter between methods.
+Pairwise comparisons of the daily estimates of $NEM$, $P$, and $R$ returned by each method were evaluated using simple summary statistics, including Pearson correlation coefficients ($\rho$) and root mean square differences (RMSD) of the linear regression fit comparing the same flux type between methods.
 Results were assessed separately for estimates using the observed or detided dissolved oxygen time series.
-
 @fig-apacmpfig and `r run_reference(id = "tbl-apacmptab")` provide a comparison of the metabolic estimates from the EBASE and Odum methods.
 Correlations for all metabolic estimates, regardless of the input time series, were positive and significantly correlated, as expected.
-However, correlations were generally stronger for the metabolic results based on the detided dissolved oxygen time series.
-The only exception for the correlation estimates was gas exchange (D), which had a slightly weaker correlation using the detided results.
-@fig-apacmpfig shows that the Odum results for gas exchange were more variable than those for EBASE using the detided dissolved oxygen, where the latter often estimated gas exhange at much smaller values compared to the former.
-The RMSE values of each regression comparing the metabolic estimates between methods were generally lower in the detided results, excluding D which was higher with the detided dissolved oxygen relative to the observed.
+However, correlations were generally stronger (and RMSDs smaller) for the metabolic results based on the detided dissolved oxygen time series.
+The observed time series clearly produced a much larger range of values for results from both the EBASE and Odum methods.
 
 The comparisons in @fig-apacmpfig and `r run_reference(id = "tbl-apacmptab")` suggest that tidal effects on continuous monitoring data can severely violate assumptions required for methods of estimating ecosystem metabolism [as described in detail in @Beck15].
-These effects are easily identified by stronger correlations and lower RMSE using the detided dissolved oxygen.
+These effects are easily identified by stronger correlations and lower RMSD using the detided dissolved oxygen.
 Further, effects of tidal advection can be seen with estimates from the Odum method as negative values for P and R, i.e., "anomalous" values.
@@
 The success of these results was predicated on the requirements of EBASE to address several critical needs that existing methods for estimating metabolism do not address for coastal applications.
-These needs were addressed by 1) using a Bayesian framework that estimates the best fit of the model to the data with posterior probabilities describing the likelihood of the fit, and 2) modifying the underlying mass-balance model for dissolved oxygen with several components more appropriate for estuarine settings.
-First, the Bayesian framework enables hypothesis testing of how a model can be fit to data by using prior knowledge in the model fitting process, which results in posterior probabilities describing the certainty of the parameter estimates [@Hilborn13].
+These needs were addressed orimarily using a Bayesian framework that estimates the best fit of the model to the data with posterior probabilities describing the likelihood of the fit.
+This framework enables hypothesis testing of how a model can be fit to data by using prior knowledge in the model fitting process, which results in posterior probabilities describing the certainty of the parameter estimates [@Hilborn13].
 The advantages of the Bayesian approach have led to widespread adoption over frequentist approaches in many scientific applications, especially in the coastal and marine community [e.g., @myers2001; @borsuk2001; @brown2017], yet these methods have not been rigorously explored for estimating metabolic rates in estuaries.
 Our results demonstrated that even using uninformed priors, EBASE can reproduce known metabolic parameters and rates with reasonable certainty.
-However, an exception was the ability of the model to estimate stable values for the $b$ parameter, although the resulting estimate of gas exchange was similar to the known values.
-This result could be explained by the dominance of the relative processes that manifest signals in the dissolved oxygen time series.
-That is, estuarine dissolved oxygen at Apalachicola Bay may be dominated by production and respiration, and less so from gas exchange, such that the available data provided insufficient information for EBASE to estimate $b$.
-This result aligns with others that have suggested gas exchange may be a relatively small component of the metabolic signal in shallow estuaries of the Gulf of Mexico [@Murrell18].
+However, initial analyses suggested an inability of the model to estimate stable values for the $b$ parameter that influence gas exchange.
+This parameter was fixed for the sensitivity analyses and we were unable to evaluate the ability of EBASE to recover it from the synthetic time series.
+Future work should focus on understanding the sensitivity of EBASE to estimating $b$ where the dominance of the relative processes that manifest signals in the dissolved oxygen time series could vary.
+For example, estuarine dissolved oxygen at Apalachicola Bay may be dominated by production and respiration, and less so from gas exchange, such that the available data may provide insufficient information for EBASE to estimate $b$.
+This aligns with others that have suggested gas exchange may be a relatively small component of the metabolic signal in shallow estuaries of the Gulf of Mexico [@Murrell18].
 In such cases, more constrained prior distributions for $b$ could be appropriate or EBASE could be tested in settings with expected larger contributions of gas exchange expressed by dissolved oxygen.
 
-The second advantage of EBASE that makes the model more appropriate for estuarine settings included several modifications to the underlying mass-balance equation for dissolved oxygen (@eq-ebase).
-As described in @sec-basemodif, EBASE is an estuarine adaptation of BASE [@Grace15] that was developed for stream applications and many assumptions of the underlying model of the latter were inappropriate for the former.
-Specifically, EBASE was modified to address three primary concerns of using BASE in an estuarine setting.
-First, the estimation of gross primary production as a function of PAR assumed a linear relationship such that the increase of production per unit increase of PAR was consistent for any level of the latter.
-The BASE model includes an exponential parameter, $p$, that defines a non-linear relationship between P and PAR (although it should be noted that BASE provides an option to set $p = 1$).
-@Holtgrieve10 describes a simple linear photosynthesis-irradiance curve as appropriate for systems that are never light-saturated and this assumption likely holds true for estuaries that are commonly more turbid than open-ocean or freshwater locations.
-Second, EBASE removed the temperature-dependency of respiration and assumed that respiration did not change with water temperature, consistent with applications of the Odum method to other shallow estuaries [@Beck15; @Murrell18].
-This modification addresses recent work that suggests respiration in streams may be reduced during the night with lower temperatures as the respiration of labile autochthonous organic matter may contribute to changes in the dissolved oxygen time series [@sadro2014; @schindler2017].
-There is currently no justification for this assumption in estuaries.
-Finally, the gas exchange in EBASE used the parameterization in @Wanninkhof14 that is primarily wind-based and developed for the open-ocean, which is clearly more appropriate than the prior formulation in BASE that includes a temperature and depth dependency for streams.
-Overall, these changes also reduced the required inputs and many monitoring locations nationwide include the requisite data for applying EBASE (i.e., dissolved oxygen, water temperature, salinity, PAR, and wind speed, as collected by the NERRS).
-
 Our results also confirmed that the effects of tidal advection on metabolic estimates can lead to biases in understanding gross primary production and ecosystem respiration.
-This result has previously been shown using the Odum open water method [@kemp1980; @caffrey2003; @Beck15] and confirmed herein, whereby results using observed time series can produce over-estimates of metabolic rates with more variation than those without tidal influence.
-Our assessment of the synthetic time series with noise and use of the detided dissolved oyxgen time series from the observed data at Apalachicola Bay to estimate metabolism both demonstrated this bias (Figures [-@fig-synapanoscmp] and [-@fig-apacmpfig]).
+This result has previously been shown using the Odum open water method [@kemp1980; @caffrey2003; @Beck15] and confirmed herein, whereby results using observed time series can produce over- and under-estimates of metabolic rates with more variation than those without tidal influence (Figures [-@fig-synapanoscmp] and [-@fig-apacmpfig]).
 As a result, detiding dissolved oxygen data is recommended prior to estimating metabolism, regardless of whether EBASE or an alternative method is used.
 In addition to inaccurate and imprecise metabolic estimates, tidal influences are also most likely observed as negative (or anomalous) values in the production and respiration rates using the Odum method [@caffrey2003].
-Tidal effects may be less obvious using EBASE since negative values cannot be obtained based on constraints of the prior distributions, although as noted above, larger and more variable estimates are returned.
+Tidal effects may be less obvious using EBASE since negative values cannot be obtained based on constraints of the prior distributions, although as noted above, more variable estimates are returned.
 A potential future approach for developing EBASE could be an explicit formulation of tidal advection in the model using a synoptic tidal height time series in the Bayesian framework.
@@
 
-EBASE provides several computational advantages over BASE implemented in the R package that was created for the purposes herein [@Beck23].
+There were several considerations during the development of EBASE that makes the model appropriate for estuarine settings.
+As described in @sec-basemodif, EBASE is an estuarine adaptation of BASE [@Grace15] that was developed for stream applications and many assumptions of the underlying model of the latter were inappropriate for the former.
+The most critical change was the formulation for gas exchange in EBASE using the parameterization in @Wanninkhof14 that is primarily wind-based and developed for the open-ocean, which is clearly more appropriate than the prior formulation in BASE that includes a temperature and depth dependency for streams.
+EBASE also removed the temperature-dependency of respiration and assumed that respiration did not change with water temperature, consistent with applications of the Odum method to other shallow estuaries [@Beck15; @Murrell18].
+Overall, these changes also reduced the required inputs and many monitoring locations nationwide include the requisite data for applying EBASE (i.e., dissolved oxygen, water temperature, salinity, PAR, and wind speed, as collected by the NERRS).
+
+EBASE also provides several computational advantages over BASE implemented in the R package that was created for the purposes herein [@Beck23].
 Most importantly, EBASE was developed to allow different model optimization lengths, whereas the BASE method was developed to estimate results for a single day.
 Allowing EBASE to incorporate more than a day of observations in the optimization can likely produce more accurate and precise estimates for the metabolic parameters.
-Physical and biological characteristics of an ecosystem that affect the metabolic rates are likely acting at time scales longer than a day, e.g., wind patterns that can affect gas exchange as quantified by the $b$ parameter may persist for several days.
+Physical and biological characteristics of an ecosystem that affect the metabolic rates are likely acting at time scales longer than a day; e.g., wind patterns that can affect gas exchange as quantified by the $b$ parameter (such as fetch and waves) as well as ecological and biogeochemical characteristics influencing the $a$ and $R$ parameters (such as community composition and nutrient availability) may persist for several days.
 Exposing the model to additional observations may produce more stable results as the parameters for the best fit are estimated by the Bayesian routine as a function of the data that are evaluated each optimization period.
 However, we cannot provide a precise recommendation on the number of days to use for the optimization period as the exact length of time that ecosystem characteristics can affect metabolic rates may not be known and likely varies by location.
-Our example using the synthetic time series from Apalachicola Bay suggests that results were similar for the 7 and 30 day optimization period.
-As such, using one week of data for each optimization period is likely appropriate for this location.
-For other locations, using 1 day is not recommended given the issues stated above (i.e., unstable results), whereas an optimization period longer than a month may miss important events in the time series that can affect the metabolic estimates and will likely produce parameters that are overly generalized.
+Our example using the synthetic time series from Apalachicola Bay suggested that the most precise results were obtained using the 30 day optimization period, although results are likely to vary across sytems.
+For most applications, very short (e.g., 1 day) or very long (e.g., \> 30 days) are not recommended, such that the former may produce unstable results and the latter may miss important events in the time series producing parameters that are overly generalized.
+However, an advantage of the Bayesian approach is the ability to evaluate different optimization periods and, therefore, the amount of data on which to estimate key parameters.
+Hypotheses can be generated on the amount of time over which critical metabolic parameters may vary across systems.
 
@@
 Like most models, explicit choices are needed to apply EBASE to time series data.
-First, the use of unconstrained prior distributions was shown to reproduce known metabolic parameters with reasonable certainty, although the $b$ parameter was still relatively unstable and unconstrained priors may not provide useful estimates in estuarine settings outside of the location evaluated herein.
-This may require further constraining the priors, although doing so may produce biased results in the absence of additional data to justify a constraint.
+First, the use of unconstrained prior distributions was shown to reproduce known metabolic parameters with reasonable certainty, although initial assessments suggested the $b$ parameter was relatively unstable and unconstrained priors may not provide useful estimates.
+Parameters that are poorly estimated may require constrained priors, although doing so may produce biased results in the absence of additional data to justify a constraint.
 Field-based measurements or alternative metabolic models could be used for informing priors.
 As an example, initial application of BASE to estuarine data prior to developing EBASE demonstrated that unstable and unrealistic metabolic rates were returned with uninformed priors.
-The Odum method was used on the same data to develop an appropriate expectation of the range of values for the reaeration coefficient, $K$, which was then used to inform the prior distribution using the BASE method.
+The Odum method was used on the same data to develop an expectation of the range of values for the reaeration coefficient, $K$, which was then used to create an informed prior distribution for the same parameter using the BASE method.
 More stable and realistic metabolic estimates were then returned by BASE, demonstrating the value of the Bayesian approach that can incorporate prior knowledge.
-A similar approach could be applied to create more accurate estimates from EBASE depending on location.
+A similar approach using alternative metabolic models for informing priors could be applied to create more accurate estimates from EBASE.
 An additional choice for using EBASE is the model optimization period, although some guidance is provided above.
@@
 
-The development of EBASE represents a new approach for estimating metabolic rates in estuarine settings that leverages existing methods by adapting theoretical components of the underlying mass-balance model using a Bayesian framework.
-Our results have demonstrated that unknown metabolic parameters can be estimated with reasonable certainty using uninformed prior distributions and EBASE could be a valuable approach to improve the understanding of critical ecosystem processes as the method can be readily applied to other settings.
+The development of EBASE represents a new approach for estimating metabolic rates in estuarine settings that leverages existing methods in a Bayesian framework.
+Our results have demonstrated that unknown metabolic parameters can be estimated with reasonable certainty and EBASE could be a valuable approach to improve the understanding of critical ecosystem processes as the method can be readily applied to other settings.
 A notable limitation of EBASE was an inability to precisely recover the $b$ parameter from a synthetic time series, although this result may be explained by a minimal contribution of gas exchange as expressed in the dissolved oxygen time series.
@@
 We thank Darren Giling for initial conversations on applying BASE to estuarine data.
-This research was funded by award 1924559 from the National Science Foundation to co-PIs Ray Najjar, Maria Herrmann, and Kathleen Hill.
+This research was funded by award 1924559 from the National Science Foundation to co-PIs Raymond Najjar, Maria Herrmann, and Kathleen Hill.
 

### 2023-09-01 manu-draft.qmd
added tassone ref

@@
 A unified model to explain the large variability of metabolic rates within and across estuaries is clearly lacking as a critical tool for understanding metabolic processes.
-Bayesian techniques have not been fully explored in estuarine applications despite their potential to address the above [although see @ciavatta2008].
+Bayesian techniques have only been explored in limited estuarine applications despite their potential to address the above \[ @ciavatta2008; @tassone2019\].
 Bayesian techniques allow for incorporation of prior knowledge about model parameters and robust estimation of parameter probability distributions [@Holtgrieve10] using highly efficient computational algorithms [@Grace15; @winslow2016].

### 2023-09-11 manu-draft.qmd
added JMA edits

@@
 Sensitivity of the EBASE method to different optimization periods is described in @sec-cmpmod.
+Finally, the EBASE software can accommodate missing observations in the input data by interpolating gaps prior to estimating the metabolic parameters.
+Results obtained from interpolated data are automatically removed from the output, eliminating the need to pre- and post-process the data when using EBASE.
 All results provided herein were generated using the EBASE R package.
@@
 Overall, most models regardless of optimization period or priors were able to show some skill at reproducing the dissolved oxygen time series.
-The only models where the NSE values were less than zero for dissolved oxygen were those where $\sigma$ of the $a$ and $R$ parameters was low and $mu$ for one of the $a$ and $R$ parameters was high.
+The only models where the NSE values were less than zero for dissolved oxygen were those where $\sigma$ of the $a$ and $R$ parameters was low and $\mu$ for one of the $a$ and $R$ parameters was high.
 However, substantial variation in NSE values was observed among all other model outputs, with some very low values of NSE occurring while the $DO$ simulation was fairly good, showing the potential for the model to produce the right answer for the wrong reasons.
@@
 As such, the prior combinations that produced the highest NSE values across all parameters were those where all prior distributions had low $\mu$ and $\sigma$ values for both $a$ and $R$ for the 1 and 7 day optimization periods, whereas the same combination was observed for the 30 day optimization period with the exception that $\sigma$ was high for $R$.
-Notably, many of the top models for each optimization period had at least one high value for $\sigma$ for $a$ or $R$, suggesting that model withs sufficient flexibility to search a larger parameter space can produce metabolic estimates most similar to actual values.
+Notably, many of the top models for each optimization period had at least one high value for $\sigma$ for $a$ or $R$, suggesting that models with sufficient flexibility to search a larger parameter space can produce metabolic estimates most similar to actual values.
 
@@
 The success of these results was predicated on the requirements of EBASE to address several critical needs that existing methods for estimating metabolism do not address for coastal applications.
-These needs were addressed orimarily using a Bayesian framework that estimates the best fit of the model to the data with posterior probabilities describing the likelihood of the fit.
+These needs were addressed primarily using a Bayesian framework that estimates the best fit of the model to the data with posterior probabilities describing the likelihood of the fit.
 This framework enables hypothesis testing of how a model can be fit to data by using prior knowledge in the model fitting process, which results in posterior probabilities describing the certainty of the parameter estimates [@Hilborn13].

### 2023-10-13 manu-draft.qmd
need equatags for flextable equations, use flextable 0.9.1 for table caption workaround

@@
 All priors are normal distributions and truncated to positive values.
-$b$ is also truncated to a maximum of 0.502 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$.](figs/priorplot.png){#fig-priorplot fig-align="center"}
+$b$ is also truncated to a maximum of 0.502 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$.
+Probabilities have been normalized to the maximum probability.](figs/priorplot.png){#fig-priorplot fig-align="center"}
 

### 2023-10-13 manu-draft.qmd
update figure 2 for rn comments

@@
 A synthetic oxygen time series was then generated using the inputs from Apalachicola Bay ([@fig-synapa]) and @eq-ebase as a forward calculating model with estimated values for $a$, $R$, and $b$.
-All metabolic rates for production, respiration, and gas exchange ([@fig-synapa]b and c) were also estimated with the synthetic time series.
+All metabolic rates for production, respiration, and gas exchange ([@fig-synapa]b) were also estimated with the synthetic time series.
 The time series was estimated using the observed data described above, the separate metabolic components and their parameters in @eq-ebase, and approximate metabolic rates from the Odum open-water method as implemented in the WtRegDO R package [@Beck21] as an approximate estimate for back-calculation of the EBASE parameters.
@@
 ![Synthetic time series for one year of continuous data at Apalachicola Bay.
-Synthetic input to EBASE was the (a) dissolved oxygen time series, with results compared to the (b) synthetic $a$ parameter and (c) synthetic metabolic estimates for gross primary production ($P$), the negative of ecosystem respiration ($-R$), and gas exchange ($-D$).](figs/synapa.png){#fig-synapa}
+Synthetic data were created using the (a) $a$ parameter to estimate (b) metabolic estimates for gross primary production ($P$), the negative of ecosystem respiration ($-R$), and gas exchange ($-D$) and the (c) dissolved oxygen time series (including saturation, $C_{sat}$).](figs/synapa.png){#fig-synapa}
 

### 2023-10-13 R/figs.R

@@ -55,29 +55,49 @@ dev.off()
 
 # simulated apa data --------------------------------------------------------------------------
 
-p1 <- ggplot(fwdatcmp, aes(x = DateTimeStamp, y = DO_obs)) + 
+p1 <- ggplot(fwdatcmp, aes(x = DateTimeStamp, y = a)) + 
   geom_line(linewidth = 0.2) + 
   scale_x_datetime(breaks = "1 month", labels = date_format("%b")) +
   theme_minimal() + 
-  labs(
-    x = NULL, 
-    y = expression(paste(O [2], ' (mmol ', m^-3, ')')),
-    title = '(a) Synthetic dissolved oxygen'
-  )
-
-p2 <- ggplot(fwdatcmp, aes(x = DateTimeStamp, y = a)) + 
-  geom_line(linewidth = 0.2) + 
-  scale_x_datetime(breaks = "1 month", labels = date_format("%b")) +
-  theme_minimal() + 
+  coord_cartesian(ylim = c(0, NA)) +
   labs(
     x = NULL, 
     y = expression(paste(italic(a), ' (mmol ', m^-2, d^-1, ') / (W ', m^{-2}, ')')), 
-    title = expression(paste('(b) Synthetic ', italic(a), ' parameter'))
+    title = expression(paste('(a) Synthetic ', italic(a), ' parameter'))
   )
 
-p3 <- ebase_plot(fwdatcmp, instantaneous = F) +
+p2 <- ebase_plot(fwdatcmp, instantaneous = F) +
   scale_x_date(breaks = "1 month", labels = date_format("%b")) +
-  labs(title = '(c) Synthetic metabolic estimates')
+  labs(title = '(b) Synthetic metabolic estimates')
+
+toplo3 <- fwdat %>% 
+  mutate(
+    DateTimeStamp = dmy_hms(datet, tz = 'America/Jamaica')
+  ) %>% 
+  select(
+    DateTimeStamp, 
+    DO_obs = `oxy,mmol/m3`, 
+    DO_sat = `oxysu,mmmol/m3`
+  ) %>% 
+  pivot_longer(DO_obs:DO_sat)
+
+p3 <- ggplot(toplo3, aes(x = DateTimeStamp, y = value, color = name)) + 
+  geom_line(linewidth = 0.2) + 
+  scale_color_manual(
+    values = c('black', 'grey'), 
+    labels = c(expression(italic('C')), expression(italic(C[sat])))
+  ) +
[…]

### 2023-10-16 manu-draft.qmd
remove priorsumcomp fig

@@
 
-@fig-priorsumcomp provides an alternative summary of how the prior distributions influenced the ability of EBASE to reproduce the synthetic results.
-Median NSE values across all models, parameters, and each optimization period for the different subsets of the prior combinations are shown.
-For example, the left-most bars in top left panel with the x-axis value for the $a$ parameter shows the median NSE values for all models using a 1-day optimization period where the $\mu$ value for the $a$ prior distribution was low or high.
-The lines indicate the inter-quartile range of NSE values for each subset.
-As such, each set of points denoted by the x-axis parameters in each panel include median NSE values for all models by optimization period, but subset by the low/high $\mu$ or $\sigma$ values for each parameter.
-Dissolved oxygen was generally modeled with high skill for all priors and it was excluded to not skew the assessment.
-The patterns suggest that EBASE results are more sensitive to model optimization period as shown by higher median NSE values with longer optimization times (note the different y-axis scales, also confirmed in @fig-priorplot).
-Similarly, median values were much more variable for the 1 day optimization period, as shown by a mix of both positive and negative values in the interquartile ranges.
-Lower median NSE values were observed for higher values for $\mu$, where higher median values for were observed for higher values of $\sigma$, regardless of optimization period.
-
 The actual model results for the best and worst performing combinations of prior distributions for the three optimization periods are shown in @fig-optex (all results are averaged at the time-step of the optimization period for visual comparability).
@@
 
-![Comparison of EBASE results to a synthetic time series created with known model parameters.
-($a$, $R$, and $b$).
+![Comparison of EBASE results to a synthetic time series created with known model parameters ($a$, $R$, and $b$).
 Results are compared for unique combinations of prior distributions and optimization period.
 The means ($\mu$) and standard deviations ($\sigma$) are evaluated as low (L) or high (H) values (see text for details) and the optimization period for EBASE is 1 day, 7 days, or 30 days.
-The model predictions of dissolved oxygen (DO), gross primary production (P), ecosystem respiration (R), gas exchange (D), and the $a$ parameter are evaluated based on Nash-Sutcliffe Efficiency (NSE) values between the EBASE results and synthetic values.
+The model predictions of dissolved oxygen (DO), gross primary production (P), ecosystem respiration (R), air-water oxygen flux (D), and light efficiency ($a$) are evaluated based on Nash-Sutcliffe Efficiency (NSE) values between the EBASE results and synthetic values.
 The $b$ parameter is a fixed constant and was not evaluated.
 Rankings of each combination as the median NSE value across all five model outputs in a row are shown on the right, with the top three as black/bold and bottom three as black/italic.](figs/priorcomp.png){#fig-priorcomp}
-
-{{< pagebreak >}}
-
-![Summaries of Nash-Sutcliffe Efficiency (NSE) values across all the comparisons of EBASE and the five synthetic model outputs separated by different combinations of prior values (note the different y-axis scales).
-Each result on the x-axis represents the median values for which the combination prior applies.
-For example, the bars in the upper left panel that are farthest to the left summarize the results from @fig-priorcomp where $\mu$ is high or low for the $a$ parameter using a 1 day optimization period.
-Bars are the medians and lines the interquartile ranges of NSE values for each result ($n$ = 8 for each bar) .](figs/priorsumcomp.png){#fig-priorsumcomp}
 

### 2023-10-16 R/figs.R

@@ -121,18 +121,6 @@ png(here('figs/priorcomp.png'), height = 4, width = 6, family = 'serif', units =
 print(p1)
 dev.off()
 
-# Fwoxy apa comparison met sum --------------------------------------------
-
-fl <- paste0(tempdir(), '/apasumdat.RData')
-download.file('https://github.com/fawda123/BASEmetab_script/raw/master/data/apasumdat.RData', destfile = fl)
-load(file = fl)
-
-p1 <- priorsumcomp(apasumdat, met = 'nse')
-
-png(here('figs/priorsumcomp.png'), height = 4.5, width = 6, family = 'serif', units = 'in', res = 500)
-print(p1)
-dev.off()
-
 # best/worst fwoxy apa comp -------------------------------------------------------------------
 
 # grid ests

### 2023-10-16 manu-draft.qmd
figures updated with rn comments

@@
 
-![A subset of the synthetic time series and the synthetic time series with noise for evaluating the ability of EBASE to recover known metabolic parameters.
-The original synthetic time series and the same time series with residual and tidal noise is shown in (a) and the estimated residual and tidal noise added to the synthetic time series is shown in (b).](figs/synapanos.png){#fig-synapanos fig-align="center"}
+![A subset of the synthetic time series (Figure [-@fig-synapa]c) and the synthetic time series with noise for evaluating the ability of EBASE to recover known metabolic parameters.
+The (a) estimated residual and tidal noise added to the (b) synthetic time series are shown.](figs/synapanos.png){#fig-synapanos fig-align="center"}
 
@@
 The left plots show the time series and the right plots show a 1:1 comparison for each result.
-See @fig-synapanos for an example of the dissolved oxygen for both input time series.](figs/synapanoscmp.png){#fig-synapanoscmp fig-align="center"}
+See @fig-synapanos for subsets of both dissolved oxygen time series.](figs/synapanoscmp.png){#fig-synapanoscmp fig-align="center"}
 
@@
 Results for (a) gross primary production ($P$), (b) ecosystem respiration ($R$), and (c) net ecosystem metabolism ($NEM$) are compared using the observed and detided dissolved oxygen time series as input to each model.
-All values are areal as mmol O$_2$ m$^2$ d$^{-1}$.
 Black lines show the 1:1 comparison and the red lines show the linear regression fit.](figs/apacmpfig.png){#fig-apacmpfig fig-align="center"}

### 2023-10-16 R/figs.R

@@ -178,16 +178,6 @@ tomod <- fwdatinp %>%
 toplo <- tomod %>% 
   filter(month(DateTimeStamp) == 6) 
 toplo1 <- toplo %>% 
-  select(DateTimeStamp, DO_obs, DO_nos) %>% 
-  pivot_longer(-DateTimeStamp) %>% 
-  mutate(
-    name = case_when(
-      name == 'DO_obs' ~ 'Synthetic', 
-      name == 'DO_nos' ~ 'Synethic + noise'
-    ), 
-    value = value / 32 * 1000
-  )
-toplo2 <- toplo %>% 
   select(DateTimeStamp, tidnoise, obsnoise) %>%
   pivot_longer(-DateTimeStamp) %>% 
   mutate(
@@ -197,19 +187,30 @@ toplo2 <- toplo %>%
     ),
     value = value / 32 * 1000
   )
+toplo2 <- toplo %>% 
+  select(DateTimeStamp, DO_obs, DO_nos) %>% 
+  pivot_longer(-DateTimeStamp) %>% 
+  mutate(
+    name = case_when(
+      name == 'DO_obs' ~ 'Synthetic', 
+      name == 'DO_nos' ~ 'Synethic + noise'
+    ), 
+    value = value / 32 * 1000
+  )
 
 p1 <- ggplot(toplo1, aes(x = DateTimeStamp, y = value, color = name)) + 
-  geom_line() + 
-  labs(
-    color = NULL, 
-    title = '(a) Time series'
-  )
-p2 <- ggplot(toplo2, aes(x = DateTimeStamp, y = value, color = name)) + 
   geom_line() + 
   scale_color_manual(values = c('darkgrey', 'black')) +
   labs(
     color = NULL,
-    title = '(b) Noise'
+    title = '(1) Noise'
+  )
+
+p2 <- ggplot(toplo2, aes(x = DateTimeStamp, y = value, color = name)) + 
+  geom_line() + 
+  labs(
+    color = NULL, 
+    title = '(b) Time series'
   )
 
 p <- p1 + p2 + plot_layout(ncol = 1) & scale_x_datetime(expand = c(0,0)) & theme_minimal() & labs(x = NULL, y = expression(paste(O [2], ' (mmol ', m^-3, ')'))) & theme(panel.grid.minor = element_blank(), legend.position = 'top')

### 2023-10-16 R/funcs.R

@@ -420,7 +420,7 @@ optex <- function(apagrd, fwdatcmp, apasumdat, rnkmetsum, met, lims = NULL){
                               'R~(mmol~m^{2}~d^{-1})', 
                               'D~(mmol~m^{2}~d^{-1})',
                               'O[2]~(mmol~m^{-3})',
-                              'italic(a)~(mmol~m^{-3}~d^{-1})/(W~m^{-2})')
+                              'italic(a)~(mmol~d^{-1}~W^{-1})')
       )
     ) %>% 
     select(-grp)
@@ -444,7 +444,7 @@ optex <- function(apagrd, fwdatcmp, apasumdat, rnkmetsum, met, lims = NULL){
             legend.title = element_blank(), 
             legend.text = element_text(size = 14),
             axis.text.x = element_text(size = 12),
-            axis.title.y = element_text(size = 8.5),
+            axis.title.y = element_text(size = 12),
             axis.text.y = element_text(size = 10),
             strip.text.x = element_text(size = rel(1.4)), 
             panel.grid.minor = element_blank(), 
@@ -551,8 +551,8 @@ syncomp_plo <- function(resobs, resnos, fwdatcmp){
     mutate(
       var = factor(var, 
                    levels = c('a', 'R'), 
-                   labels = c('italic(a)~(mmol~m^{-3}~d^{-1})/(W~m^{-2})',
-                              'R~(mmol~m^{2}~d^{-1})'
+                   labels = c('italic(a)~(mmol~m^{-2}~d^{-1})/(W~m^{-2})',
+                              'R~(mmol~m^{-2}~d^{-1})'
                    )
       ), 
       name = factor(name, levels = c('Synthetic', 'EBASE recovered', 'EBASE recovered (+ noise)'))
@@ -665,9 +665,9 @@ apacmp_plo <- function(dat, alph = 0.5, col = 'black'){
     )
   
   subfigs <- list(
-    P = '(a) P', 
-    R = '(b) R', 
-    NEM = '(c) NEM'
+    P = '(a)~P~(mmol~m^{-2}~d^{-1})', 
+    R = '(b)~R~(mmol~m^{-2}~d^{-1})', 
+    NEM = '(c)~NEM~(mmol~m^{-2}~d^{-1})'
   )
   
   lvs <- levels(toplo$var)
@@ -693,7 +693,7 @@ apacmp_plo <- function(dat, alph = 0.5, col = 'black'){
[…]

### 2023-10-16 manu-draft.qmd
ebase parameter table removed

@@
 
-load(file = here('tabs/ebaseiotab.RData'))
 load(file = here('tabs/apacmptab.RData'))
@@
 The $b$ parameter is estimated from the data and has units of $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$.
-As such, the three parameters estimated by EBASE are $a$, $R$, and $b$ using the required input data that includes time series of dissolved oxygen, water temperature, salinity, PAR, wind speed, and water column depth (`r run_reference(id = "tbl-ebaseiotab")`).
+As such, the three parameters estimated by EBASE are $a$, $R$, and $b$ using the required input data that includes time series of dissolved oxygen, water temperature, salinity, PAR, wind speed, and water column depth.
 
@@
 The estimated dissolved oxygen concentration is also returned, which can be compared to the observed values as one measure of model performance.
-All parameter inputs and outputs for EBASE are shown in `r run_reference(id = "tbl-ebaseiotab")`.
 Areal rates for $P$, $R$, and $D$ ($\text{mmol}/\text{m}^2/\text{d}$) are returned for comparability of estimates across locations where water column depth may vary.
@@
 
-<!-- {{< pagebreak >}} -->
-
 ## Tables {.unnumbered}
-
-```{r}
-#| label: tbl-ebaseiotab
-#| tbl-cap: "EBASE model parameters showing the required input, parameters derived from the model input, and model output.
-The model notation for production, respiration, and gas exchange shows both the simple parameter name and the full model component used to estimate the parameter."
-ebaseiotab
-```
-
-{{< pagebreak >}}
 

### 2023-10-16 R/tabs.R

@@ -5,65 +5,6 @@ library(here)
 
 source(file = here('R/funcs.R'))
 
-# EBASE i/o -----------------------------------------------------------------------------------
-
-totab <- tibble(
-  Type = c('Input', 'Input', 'Input', 'Input', 'Input', 'Input',
-           'Input-derived', 'Input-derived', 'Input-derived',
-           'Output', 'Output', 'Output', 'Output', 'Output', 'Output'),
-  Description = c('Dissolved oxygen', 'Water temperature', 'Salinity', 'Total photosynthetically active radiation', 'Wind speed', 'Water column depth',
-                  'Wind speed at 10 meter height, squared', 'Schmidt number (from water temperature and salinity)', 'Dissolved oxygen at saturation (from water temperature and salinity)',
-                  'Dissolved oxygen (modelled)', 'Production', 'Respiration', 'Gas exchange', 'Light efficiency', 'Sensitivity of gas transfer to wind speed'),
-  `Model notation` =  c('C_d', '-', '-', 'PAR', '-', 'H', 
-                        'U102', 'Schmidt number', 'C_sat', 
-                        'C_mod', 'P', 'R',  'D', 'a', 'b'),
-  Units = c('mg/L', 'C', 'psu', 'W/m2', 'm/s', 'm', 
-            'm2', 'unitless', 'mmol/m3',
-            'mmol/m3', 'mmol O2 m2/d', 'mmol O2 m2/d', 'mmol O2 m2/d', '(mmol/m2/d)/(W/m2)', '(cm/hr)/(m2/s2)')
-) %>% 
-  flextable::as_grouped_data(groups = 'Type')
-
-eqsz <- fp_text_default(font.size = 9)
-# col2wid <- 1
-# col3wid <- 2
-# col14wid <- 6.5 - (col2wid + col3wid)
-ebaseiotab <- flextable(totab) %>% 
-  flextable::compose(i = 2, j = 3, value = as_paragraph(as_equation('C', props = eqsz))) %>% 
-  flextable::compose(i = 5, j = 3, value = as_paragraph(as_equation('PAR', props = eqsz))) %>% 
-  flextable::compose(i = 7, j = 3, value = as_paragraph(as_equation('H', props = eqsz))) %>% 
[…]

### 2023-10-16 manu-draft.qmd
apacmptab now kable

@@
 library(here)
-library(officer)
-library(officedown)
-library(flextable)
 library(dplyr)
@@
 Results were assessed separately for estimates using the observed or detided dissolved oxygen time series.
-@fig-apacmpfig and `r run_reference(id = "tbl-apacmptab")` provide a comparison of the metabolic estimates from the EBASE and Odum methods.
+@fig-apacmpfig and @tbl-apacmptab provide a comparison of the metabolic estimates from the EBASE and Odum methods.
 Correlations for all metabolic estimates, regardless of the input time series, were positive and significantly correlated, as expected.
@@
 
-The comparisons in @fig-apacmpfig and `r run_reference(id = "tbl-apacmptab")` suggest that tidal effects on continuous monitoring data can severely violate assumptions required for methods of estimating ecosystem metabolism [as described in detail in @Beck15].
+The comparisons in @fig-apacmpfig and @tbl-apacmptab suggest that tidal effects on continuous monitoring data can severely violate assumptions required for methods of estimating ecosystem metabolism [as described in detail in @Beck15].
 These effects are easily identified by stronger correlations and lower RMSD using the detided dissolved oxygen.

### 2023-10-16 R/tabs.R

@@ -1,4 +1,3 @@
-library(flextable)
 library(tidyverse)
 library(lubridate)
 library(here)
@@ -94,15 +93,8 @@ totab <- grd %>%
   ) 
 
 apacmptab <- totab %>%
-  flextable() %>% 
-  fontsize(part = 'all', size = 12) %>% 
-  font(part = 'all', fontname = 'Times New Roman') %>% 
-  padding(padding = 1, part = 'all') %>% 
-  width(width = 6.5 / ncol_keys(.)) %>% 
-  flextable::compose(part = 'header', j = 'RMSD', value = as_paragraph(as_equation("\\text{RMSD}~(\\text{mmol}~\\text{O}_2/\\text{m}^2/\\text{d})"))) %>% 
-  flextable::compose(part = 'header', j = 'Corr.', value = as_paragraph(as_equation("\\rho"))) %>% 
-  flextable::align(align = 'center', j = 2:ncol_keys(.), part = 'all') %>% 
-  font(part = 'footer', fontname = 'Times New Roman') %>% 
-  flextable::align(align = 'right', part = 'footer')
+  knitr::kable(
+    col.names = c('Parameter', '$\\rho$', 'RMSD (mmol O$^2$/m$^2$/d)')
+  )
 
 save(apacmptab, file = here('tabs/apacmptab.RData'))

### 2023-10-16 manu-draft.qmd
intro revision from rn comments

@@
     number-sections: true
-title: "Fitting metabolic models to dissolved oxygen data: The Estuarine BAyesian Single-station Estimation method"
+title: "Fitting metabolic models to dissolved oxygen data: The Estuarine BAyesian Single-station Estimation method (EBASE)"
 bibliography: refs.bib
@@
 Continuous monitoring data for dissolved oxygen have been used to estimate the contribution of gross primary production and ecosystem respiration to net ecosystem metabolism.
-Most models for estimating metabolic processes make several assumptions on how biological and physical factors affect the expression of metabolism in dissolved oxygen data, which can lead to poorly constrained errors that influence the interpretation of results.
+Most methods for estimating metabolic processes make several assumptions on how biological and physical factors affect the expression of metabolism in dissolved oxygen data, which can lead to poorly constrained errors that influence the interpretation of results.
 In estuaries, these methods can inform our understanding of spatial and temporal patterns in trophic state, yet very few models have been developed that address challenges of estimating metabolism in these dynamic systems.
@@
 This approach can also vary the time period over which critical model parameters are estimated.
-We demonstrate that EBASE can reproduce known metabolic parameters from a synthetic time series created with continuous monitoring data from the Apalachicola estuarine reserve and provide a quantitative assessment of how the prior distributions affect the model results.
+We demonstrate that EBASE can reproduce known metabolic parameters from a synthetic time series created with continuous monitoring data from the Apalachicola Bay (southeastern United States) and provide a quantitative assessment of how the prior distributions affect the model results.
 The approach is also bench-marked with more conventional methods for estimating metabolism, with critical differences attributed to formulation of the metabolic mass-balance equations and the underlying statistical framework (i.e., Bayesian vs simple integration).
@@
 For example, an assessment in Randers Fjord (Denmark) showed agreement in both the sign and magnitude among four different approaches [@gazeau2005].
-However, gross primary production (P) and ecosystem respiration (R) estimates from incubations were consistently lower than those from open-water methods, consistent with the estuarine respiration synthesis of @hopkinson2005.
+However, gross primary production and ecosystem respiration estimates from incubations were consistently lower than those from open-water methods, consistent with the estuarine respiration synthesis of @hopkinson2005.
 Even within one technique, there are numerous choices to be made that alter the outcome.
-For example, in the commonly used ${}^{14}$C technique for measuring primary production, @cloern2014 noted that investigators have made divergent choices with regard to filtering mesozooplankton grazers, measuring ${}^{14}$C in dissolved organic carbon, accounting for changing spectral light quality with depth, and choosing the euphotic zone depth, the incubation time, and the incubation location (*in situ* vs.
+For example, in the commonly used ${}^{14}$C technique for measuring primary production, @cloern2014 noted that investigators have made divergent choices with regard to filtering mesozooplankton grazers, measuring ${}^{14}$C in dissolved organic carbon, changing spectral light quality with depth, the euphotic zone depth, the incubation time, and the incubation location (*in situ* vs.
 on-deck).
-These challenges are further compounded in estuaries where high temporal and spatial variability have been noted in global syntheses of P and R.
-@cloern2014 found that annual phytoplankton primary production can vary 10-fold within an estuary and up to 5-fold from year-to-year.
-@hopkinson2005 focused on cross-system variations and found benthic and pelagic respiration rates varying by more than factors of 30 and 40, respectively, ranges that are broadly consistent with whole system respiration rates based on the open-water method [@hoellein2013].
+These challenges are further compounded in estuaries, where high temporal and spatial variability have been noted in global syntheses of photosynthesis and respiration.
+@cloern2014 found that annual phytoplankton primary production can vary 10-fold within an estuary and up to 5-fold from year to year.
+@hopkinson2005 focused on cross-system variations and found benthic and pelagic respiration rates varying by more than factors of 30 and 40, respectively, ranges that are broadly consistent with whole-system respiration rates based on the open-water method [@hoellein2013].
 
 Advances in sensor technology for continuous collection of dissolved oxygen time series suggest that the open-water method [@Odum56] could be leveraged for applications across multiple locations.
-This approach exploits the diel cycle of dissolved oxygen and allows the simultaneous determination of whole-system gross primary production (net primary production $+$ autotrophic respiration), ecosystem respiration (heterotrophic $+$ autotrophic respiration), and NEM (P $-$ R).
+This method exploits the diel cycle of dissolved oxygen and allows the simultaneous determination of whole-system gross primary production (P $=$ net primary production $+$ autotrophic respiration), ecosystem respiration (R $=$ heterotrophic $+$ autotrophic respiration), and NEM (P $-$ R).
 A global synthesis of warm-season, open-water metabolism studies [@hoellein2013] was based on a total of 47 estimates, almost all of which were derived from a single study of U.S.
@@
 In particular, tidal advection can violate assumptions of the method that the measurements are from the same parcel of water with a continuous metabolic history [@kemp1980].
-Striking evidence of this challenge is the large number of physically impossible negative P and R estimates in the open-water analysis of 28 3-year time series by @caffrey2003.
+Striking evidence of this limitation is the large number of physically impossible negative P and R estimates in the open-water analysis of 28 3-year time series by @caffrey2003.
 The computed rates were negative 23% of the time, on average, and as high as 69% for individual sites.
-Estimating gas exchange using the open-water method in estuaries has also been challenging.
-Most approaches estimate gas exchange using empirical formula dependent on wind speed and temperature or held at a constant value [e.g., @caffrey2004; @russell2007; @nagel2009; @Beck15; @Murrell18].
-Gas exchange is a first-order process in the estuarine dissolved oxygen budget and, therefore, inaccurate estimations based on empirical formulae or fixed values can lead to poorly constrained errors that influence interpretation of results.
+
+Another limitation of the open-water method is the estimation of the air-water oxygen flux, which is an important term in the oxygen budget.
+Most approaches estimate this flux using an empirically derived gas transfer velocity that is dependent on wind speed and temperature or held at a constant value [e.g., @caffrey2004; @russell2007; @nagel2009; @Beck15; @Murrell18].
+Particularly challenging is that the exchange velocity in estuaries also depends on water depth and tides [@Ho16] and possibly additional factors, such as fetch and the concentration of total suspended solids [@Borges11].
+The importance of the air--water gas flux in the oxygen budget opens up the possibility that the oxygen data themselves may be used to determine the gas transfer velocity, an approach that has been applied in streams [e.g., @riley2013].
 
 A unified model to explain the large variability of metabolic rates within and across estuaries is clearly lacking as a critical tool for understanding metabolic processes.
-Bayesian techniques have only been explored in limited estuarine applications despite their potential to address the above \[ @ciavatta2008; @tassone2019\].
+Bayesian techniques have only been explored in limited estuarine applications despite their potential to address the above [@ciavatta2008; @tassone2019].
 Bayesian techniques allow for incorporation of prior knowledge about model parameters and robust estimation of parameter probability distributions [@Holtgrieve10] using highly efficient computational algorithms [@Grace15; @winslow2016].
@@
 
-The goal of this paper is to describe the development and application of a new Bayesian method for simultaneously determining gross primary production, ecosystem respiration, net ecosystem metabolism, and the gas exchange (D) in estuaries from high-resolution dissolved oxygen time series.
+The goal of this paper is to describe the development and application of a new Bayesian method for simultaneously determining gross primary production and ecosystem respiration in estuaries from high-resolution dissolved oxygen time series.
 We call this method EBASE (Estuarine BAyesian Single-station Estimation), which builds and improves on the BASE method [@Grace15] using an approach more appropriate for estuarine time series.
@@
 These comparisons combine results using real observations and those based on detided inputs to demonstrate effects of tidal advection on metabolic estimates.
+We also attempt to constrain the gas transfer velocity using this approach but are unable, at least with the data sets used in this study.
 EBASE is available as an open-source software package created with the R statistical programming language [@Beck23; @RCT23] and the results herein provide practical suggestions for use of the software on novel datasets.

### 2023-10-16 manu-draft.qmd
methods revision from rn comments done

@@
 
-The Estuarine BAyesian Single station Estimation (EBASE) method provides a simple approach to quantify metabolic parameters from continuous dissolved oxygen time series.
-The EBASE mass balance equation is:
+EBASE uses a mass balance equation for dissolved oxygen, assuming a well-mixed water column of depth $H$ and no lateral transport processes:
 
 $$
-H\frac{dC}{dt} = aPAR - R - k_w\left(C_{Sat} - C \right )
+H\frac{dC}{dt} = P - R - D
 $$ {#eq-ebase}
 
-The parameter $k_w$ is the gas transfer velocity defined as:
+where the terms on the right side of the equation are gross primary production ($P$), ecosystem respiration ($R$), and the net upward diffusive gas flux at the air-water interface ($D$), respectively, processes that result in the change in dissolved oxygen concentration ($dC$) per unit time ($dt$).
+The equation has dimensions of moles per unit area of time.
+$P$ is modeled as $aPAR$, where $PAR$ is the surface photosynthetically active radiation and $a$ is the light efficiency; $R$ is constant; and $D$ is modeled in the standard manner [e.g, @sarmiento2013] as $k_w\left(C_{sat} - C\right)$, where $k_w$ is the gas transfer velocity and $C_{sat}$ is the saturation (or equilibrium) concentration, calculated from salinity and temperature [@Garcia92].
+The gas transfer velocity is modeled using the formulation in @Wanninkhof14:
 
@@
 
-For @eq-ebase, the terms on the right side of the equation are gross primary production ($P$, $aPAR$), ecosystem respiration ($R$), and gas exchange ($D$, remainder), respectively, processes that result in the change in dissolved oxygen ($dC$) per unit time ($dt$); $H$ is the water depth.
-Production is modeled linearly as a function of surface irradiance (PAR, $\text{W}/\text{m}^2$) and $a$ (light efficiency, $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$).
-Respiration is modeled with the single parameter $R$ ($\text{mmol}/\text{m}^2/\text{d}$).
-Gas exchange is modeled using the formulation in @Wanninkhof14 as a wind-based parameterization for the gas transfer velocity [@Ho16] multiplied by the oxygen concentration gradient as the difference between the concentration measured in the surface water layer, $C$ ($\text{mmol}/\text{m}^3$) , and the saturation (or equilibrium) concentration, $C_{Sat}$ ($\text{mmol}/\text{m}^3$) calculated from salinity (psu) and temperature ($^\circ$C) [@Millero81; @Garcia92].
-The gas transfer velocity, $k_w$ in @eq-kw, is modeled using wind speed squared at 10 meters above the water ($U_{10}^2$, $\text{m}^2/\text{s}^2$) and the Schmidt number ($Sc$), which is the ratio of the kinematic viscosity of water to the molecular diffusivity of oxygen in water [$S_c$ is computed from water temperature and salinity using the polynomial fit in @Wanninkhof14].
-The $b$ parameter is estimated from the data and has units of $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$.
+where $b$ is constant, $U_{10}$ is the wind speed at 10 m above the water, and $Sc$ is the Schmidt number, which is the ratio of the kinematic viscosity of water to the molecular diffusivity of oxygen in water [$Sc$ is computed from water temperature and salinity using the polynomial fit in @Wanninkhof14].
+The $b$ parameter is fairly well known for open-ocean conditions, but may be different in estuaries, where the transfer velocity is influenced by factors other than wind speed and $Sc$.
 As such, the three parameters estimated by EBASE are $a$, $R$, and $b$ using the required input data that includes time series of dissolved oxygen, water temperature, salinity, PAR, wind speed, and water column depth.
@@
 The gas transfer velocity, $k_w$ (@eq-kw), increases as the water column in the vicinity of the air--water interface becomes more turbulent.
-As such, $k_w$ is usually modeled as a function of measurable variables that can potentially predict turbulent mixing of the surface water, e.g., wind speed, current velocity, and water depth [@Ho16]; fetch and suspended solids have also been considered for estuarine systems [@Borges11].
-$k_w$ also increases with temperature because the film of water at the air-sea interface becomes thinner as the viscosity decreases and the random motion of gas molecules increases.
+As such, $k_w$ is usually modeled as a function of measurable variables that can potentially predict turbulent mixing of the surface water, such as wind speed.
+$k_w$ also increases with temperature because the film of water at the air-water interface becomes thinner as the viscosity decreases and the random motion of gas molecules increases.
 This dependence of $k_w$ on temperature is usually modeled with the Schmidt number as in @eq-ebase, which combines information on the viscosity of the fluid and the diffusivity of the gas.
@@
 
-where $\bar{T}$ is the average water temperature over each 24-hour period in the oxygen time series and $K$ is the reaeration coefficient estimated from the data.
+where $T$ is the instantaneous water temperature, $\bar{T}$ is the average water temperature over each 24-hour period in the oxygen time series (both in $^{\circ}$C), and $K$ is the reaeration coefficient estimated from the data.
 @eq-gasexbase indicates that $k_w$ increases with temperature and depth.
@@
 An expectation is that $k_w$ would increase as water depth decreases for a given current speed because turbulence generated at the sediment--water interface by currents would more easily reach the surface [@Ho16].
-Based on these differences and the lack of availability of other possible drivers of the transfer velocity in coastal systems (e.g., currents, fetch, suspended solids, and wave height), the wind-based parameterization for gas exchange was used for EBASE [@Wanninkhof14].
-However, this parameterization is largely based on open-ocean data and so we allow $b$, which was esimated by @Wanninkhof14 to be 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$, to be a free parameter that may reflect spatial and temporal variability in factors other than wind and temperature that may influence transfer velocity.
+Based on these differences and the lack of availability of other possible drivers of the transfer velocity in coastal systems, the well-established wind- and temperature-based parameterization for gas exchange of @Wanninkhof14 was used for EBASE, but allowing $b$, which was esimated by @Wanninkhof14 to be 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$, to be a free parameter that may reflect spatial and temporal variability in factors other than wind and temperature that may influence transfer velocity.
 
@@
 
+For ease of application, an R package was developed to implement EBASE [@Beck23; @RCT23], similar to the R package available for BASE (<https://github.com/dgiling/BASEmetab>).
+Changes in model implementation were included in the package as additional changes to BASE.
+First, users can specify the model optimization period, where the time period can vary from a minimum of 1 day to the maximum of the entire length of the time series.
+This differs from the BASE approach where the optimization period is set at 1 day.
+Sensitivity of the EBASE method to different optimization periods is described in @sec-cmpmod.
+Second, the EBASE software can accommodate missing observations in the input data by interpolating gaps prior to estimating the metabolic parameters.
+Results obtained from interpolated data are automatically removed from the output, eliminating the need to pre- and post-process the data when using EBASE.
+Again, this feature was not included in the original BASE method and a complete time series was required for use.
+A detailed web page (<https://fawda123.github.io/EBASE/>) describes how to use the package, with explanations of the functions for viewing model results, including a plot of the modeled dissolved oxygen with the observed, a time series plot of the primary metabolic estimates (P, R, and D, as areal rates with units $\text{mmol}/\text{m}^2/\text{d}$), and a time series plot of the credible intervals for the $a$, $R$, and $b$ parameters.
+All results provided herein were generated using the EBASE R package.
+
 ### EBASE model estimation {#sec-ebasemod}
 
-The remaining three parameters $a$ (light efficiency, $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$), $R$ (respiration, $\text{mmol}/\text{m}^2/\text{d}$), and $b$ ($\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$) are estimated by likelihood given the observed data and prior distributions for the parameters.
-The JAGS software [as for BASE, @Plummer03] is used with the EBASE R package [described below, @Beck23; @RCT23] to estimate the unknown parameters using Markov Chain Monte Carlo (MCMC) simulations in a Bayesian framework.
-At each time step, the change in oxygen concentration between time steps is calculated from the equation using model inputs and parameter guesses based on the prior distributions, and then a finite difference approximation is used to estimate modeled oxygen concentration.
-Estimations proceed using a forward-in-time integration where the dissolved oxygen at each time step is informed by parameter estimates from the preceding time steps.
-The estimated dissolved oxygen concentration is also returned, which can be compared to the observed values as one measure of model performance.
-Areal rates for $P$, $R$, and $D$ ($\text{mmol}/\text{m}^2/\text{d}$) are returned for comparability of estimates across locations where water column depth may vary.
+The remaining three parameters $a$ , $R$, and $b$ are estimated by likelihood given the observed data and prior distributions for the parameters.
+The "Just Another Gibbs Sampler" (JAGS) software [as for BASE, @Plummer03] is used with the EBASE R package (described above) to estimate probability distributions of the unknown parameters using Markov Chain Monte Carlo (MCMC) simulations in a Bayesian framework applied to a finite-difference form of @eq-ebase.
+A JAGS model file is included with the package that implements @eq-ebase, with options to supply a custom model file as needed.
 
@@
 Examples of the prior distributions in Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior] are shown in @fig-priorplot.
-
-For ease of application, an R package was developed to implement the EBASE methods described above [@Beck23].
-A detailed web page (<https://fawda123.github.io/EBASE/>) describes the required data inputs for estimating the metabolic parameters using the core R functions that pass inputs to the JAGS software.
-A JAGS model file is included with the package that implements @eq-ebase, with options to supply a custom model file as needed.
-Users can also specify alternative prior distributions from those in Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior] for each parameter.
-Functions are available for viewing model results, including a plot of the modeled dissolved oxygen with the observed, a time series plot of the primary metabolic estimates (P, R, and D, as areal rates with units $\text{mmol}/\text{m}^2/\text{d}$), and a time series plot of the credible intervals for the $a$, $R$, and $b$ parameters.
-An additional option allows the user to specify the model optimization period, where the time period can vary from a minimum of 1 day to the maximum of the entire length of the time series.
-This differs from the BASE approach where the optimization period is set at 1 day.
-Sensitivity of the EBASE method to different optimization periods is described in @sec-cmpmod.
-Finally, the EBASE software can accommodate missing observations in the input data by interpolating gaps prior to estimating the metabolic parameters.
-Results obtained from interpolated data are automatically removed from the output, eliminating the need to pre- and post-process the data when using EBASE.
-All results provided herein were generated using the EBASE R package.
+Functions in the EBASE R package allow alternative prior distributions from those above.
 

### 2023-10-16 manu-draft.qmd
working on results from RN comments

@@
 
-A preliminary assessment of the ability of EBASE to produce reasonable parameter estimates for $a$, $R$, and $b$ (@eq-ebase) was conducted using a synthetic time series with actual data as inputs.
-These assessments evaluated if 1) EBASE estimates similar values for the known parameters in the synthetic time series, and 2) how the estimated values change with different model configurations.
-The synthetic time series used inputs from one year of continuous data from Apalachicola Bay, Florida.
-The input data were observations from the year 2021 at the Cat Point water quality monitoring station (29.7021$^{\circ}$ N, --84.8802$^{\circ}$ W) and included 15-minute observations of water temperature and salinity.
-Continuous meteorological observations from the nearby East Bay station (29.7694$^{\circ}$ N, --84.8815$^{\circ}$ W, \~10 km north of Cat Point) were combined with the water quality data at the same time step and included air temperature, wind speed, and PAR.
-Missing observations in the water quality and meteorological data (`r wqmis`% and `r mtmis`% of all observations in each dataset, respectively) were filled using autoregressive modelling of the actual parameters versus time to create a complete dataset from January 1st to December 31st.
+A preliminary assessment of the ability of EBASE to produce reasonable parameter estimates for $a$, $R$, and $b$ (@eq-ebase) was conducted using a synthetic time series of dissolved oxygen concentration created from observed water temperature, salinity, air temperature, $PAR$, and wind speed measured every 15 minutes.
+These assessments evaluated if 1) EBASE estimates similar values for the known parameters used to create the synthetic time series, and 2) how the estimated values change with different model configurations.
+The synthetic time series used inputs from one year (2021) of continuous data from Apalachicola Bay, Florida.
+Water temperature and salinity were from the Cat Point water quality monitoring station (29.7021$^{\circ}$ N, --84.8802$^{\circ}$ W) and the meteorological observations were from the nearby East Bay station (29.7694$^{\circ}$ N, --84.8815$^{\circ}$ W, \~10 km north of Cat Point).
+Missing observations in the water and meteorological data (`r wqmis`% and `r mtmis`% of all observations in each dataset, respectively) were filled using autoregressive modelling of the actual data versus time to create a complete dataset from January 1st to December 31st.
 
-A synthetic oxygen time series was then generated using the inputs from Apalachicola Bay ([@fig-synapa]) and @eq-ebase as a forward calculating model with estimated values for $a$, $R$, and $b$.
-All metabolic rates for production, respiration, and gas exchange ([@fig-synapa]b) were also estimated with the synthetic time series.
-The time series was estimated using the observed data described above, the separate metabolic components and their parameters in @eq-ebase, and approximate metabolic rates from the Odum open-water method as implemented in the WtRegDO R package [@Beck21] as an approximate estimate for back-calculation of the EBASE parameters.
-In short, $P$, $R$, and $D$ as $\text{mmol}/\text{m}^2/\text{d}$ were estimated using the Odum method and the separate synthetic components for EBASE were created by estimating the corresponding parameters, such that $P /PAR = a$, $R$ as is, and $D = k_w\left(C_{sat} - C \right)$, where $k_w$ includes the $b$ parameter.
-Daily estimates for each were created as the daily average for each parameter fit with a polynomial spline to create a smoothed output.
-The synthetic dissolved oxygen time series was then created as the integrated sum of the separate components for $P$, $R$, and $D$ divided by water column depth.
-The only parameter that could not be created as a continuous time series was $b$, which was set to 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ [@Wanninkhof14].
-Accordingly, the synthetic dissolved oxygen time series and the observed water quality and meteorological data were used with EBASE to determine if reasonable estimates for the known parameters and metabolic rates were recovered.
+The synthetic oxygen time series was generated using @eq-ebase as a forward calculating model with specified daily values for $a$, $R$, and $b$.
+The $b$ parameter was fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ [@Wanninkhof14].
+Daily values of $a$ and $R$ were estimatd by 1) applying the Odum open-water method as implemented in the WtRegDO R package \[@Beck21\], 2) computing monthly averages of the resulting daily estimates of $P$ and $R$, 3) computing the monthly average of the parameter $a$ by dividing the monthly average of $P$ by the monthly average of $PAR$, and 4) by fitting a spline function to the monthly averages of $R$ and $a$.
+
+The resulting synthetic time series of light efficiency, dissolved oxygen budget terms ($P$, $D$, and $R$), and dissolved oxygen concentration are shown in @fig-synapa.
+The light efficiency has a complex structure, with maxima in the summer, fall, and winter, and a strong minimum in spring (Figure [-@fig-synapa]a).
+The oxygen budget terms (Figure [-@fig-synapa]b) reveal that (1) photosynthesis has a structure similar to that of light efficiency, but noisier, reflecting the synoptic-scale variability of $PAR$; (2) respiration has the expected structure of being greatest during the summer [@caffrey2004], stimulated by greater autochthonous organic matter variability and high temperature; and (3) the air–water gas flux mimic productivity on short time scales, with outgassing during high productivity and ingassing during low productivity.
+The synthetic dissolved oxygen time series shows high variability at multiple time scales (Figure [-@fig-synapa]c).
+There is a prominent diurnal cycle that is superimposed on subseasonal variability, presumably driven by radiation and winds.
+
+#### Recovery of model parameters
+
+#### Sensitivity to priors
 

### 2023-10-17 manu-draft.qmd
more edits from rn comments

@@
 Water temperature and salinity were from the Cat Point water quality monitoring station (29.7021$^{\circ}$ N, --84.8802$^{\circ}$ W) and the meteorological observations were from the nearby East Bay station (29.7694$^{\circ}$ N, --84.8815$^{\circ}$ W, \~10 km north of Cat Point).
-Missing observations in the water and meteorological data (`r wqmis`% and `r mtmis`% of all observations in each dataset, respectively) were filled using autoregressive modelling of the actual data versus time to create a complete dataset from January 1st to December 31st.
+Missing observations in the water and meteorological data (`r wqmis`% and `r mtmis`% of all observations in each dataset, respectively) were filled using autoregressive modelling of the actual data versus time to create a complete dataset from January 1st to December 31st [@akaike1969].
 
@@
 The $b$ parameter was fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ [@Wanninkhof14].
-Daily values of $a$ and $R$ were estimatd by 1) applying the Odum open-water method as implemented in the WtRegDO R package \[@Beck21\], 2) computing monthly averages of the resulting daily estimates of $P$ and $R$, 3) computing the monthly average of the parameter $a$ by dividing the monthly average of $P$ by the monthly average of $PAR$, and 4) by fitting a spline function to the monthly averages of $R$ and $a$.
+Daily values of $a$ and $R$ were estimatd by 1) applying the Odum open-water method as implemented in the WtRegDO R package [@Beck21], 2) computing monthly averages of the resulting daily estimates of $P$ and $R$, 3) computing the monthly average of the parameter $a$ by dividing the monthly average of $P$ by the monthly average of $PAR$, and 4) by fitting a spline function to the monthly averages of $R$ and $a$.
 
@@
 The light efficiency has a complex structure, with maxima in the summer, fall, and winter, and a strong minimum in spring (Figure [-@fig-synapa]a).
-The oxygen budget terms (Figure [-@fig-synapa]b) reveal that (1) photosynthesis has a structure similar to that of light efficiency, but noisier, reflecting the synoptic-scale variability of $PAR$; (2) respiration has the expected structure of being greatest during the summer [@caffrey2004], stimulated by greater autochthonous organic matter variability and high temperature; and (3) the air–water gas flux mimic productivity on short time scales, with outgassing during high productivity and ingassing during low productivity.
+The oxygen budget terms (Figure [-@fig-synapa]b) reveal that (1) photosynthesis has a structure similar to that of light efficiency, but noisier, reflecting the synoptic-scale variability of $PAR$; (2) respiration has the expected structure of being greatest during the summer [@caffrey2004], stimulated by greater autochthonous organic matter variability and high temperature; and (3) the air--water gas flux mimic productivity on short time scales, with outgassing during high productivity and ingassing during low productivity.
 The synthetic dissolved oxygen time series shows high variability at multiple time scales (Figure [-@fig-synapa]c).
@@
 The minimum and maximum values for the standard deviations for $R$ and $a$ were chosen as 10% and 1000% of the default mean values to test a range from high precision to completely uninformed values.
-The mean and standard deviation for the $b$ parameter were not tested and the prior value was set as the fixed constant used in the synthetic time series (0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$).
+The mean and standard deviation for the $b$ parameter were not tested and the prior value was set as the fixed constant used in the synthetic time series, 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$.
 There is minimal prior knowledge on expectations for the $b$ parameter and initial assessments with the synthetic time series demonstrated that the parameter was poorly constrained, primarily because the fixed parameter could not be estimated using observed data (i.e., wind, water temperature) that affected the actual realization of $b$.
@@
 A short period of optimization (e.g., 1 day) may return unstable parameter estimates if forcing factors (e.g., wind, temperature) that affect estuarine biogeochemical rates occur at time scales longer the chosen period of optimization for EBASE.
-As such, potentially longer optimization periods may return more stable and robust estimates given that more data are used for parameter estimation, although at the risk of overgeneralizing metabolic rates if variation occurs at time scales less than the optimization period.
+As such, potentially longer optimization periods may return more stable and robust estimates given that more data are used for parameter estimation, although at the risk of overgeneralizing metabolic rates if biogeochemical variation occurs at time scales less than the optimization period.
 Results from optimization periods of 1, 7, and 30 days were compared for the synthetic one-year time series.
@@
 
-# gasex nse
+# all parameter nse
 prmall <- apasumdat %>% 
   tidyr::unnest('ests') %>% 
-  filter(var != 'DO_mod') %>% 
   na.omit %>% 
@@
 Several conclusions can be made from the results in @fig-priorcomp following the patterns of the NSE values.
-Overall, most models regardless of optimization period or priors were able to show some skill at reproducing the dissolved oxygen time series.
+Overall, most models regardless of optimization period or priors were able to show some skill at reproducing the dissolved oxygen time series, which had a median NSE across all optimization periods and priors of `r prmall[['DO_mod']]`.
 The only models where the NSE values were less than zero for dissolved oxygen were those where $\sigma$ of the $a$ and $R$ parameters was low and $\mu$ for one of the $a$ and $R$ parameters was high.
 However, substantial variation in NSE values was observed among all other model outputs, with some very low values of NSE occurring while the $DO$ simulation was fairly good, showing the potential for the model to produce the right answer for the wrong reasons.
-Not considering dissolved oxygen, gas exchange ($D$) was estimated with the highest NSE, with a median value across all optimization periods and priors slightly larger than zero (median NSE = `r prmall[['D']]`).
-This result makes sense, given that the $b$ parameter was fixed at the actual value used to create the synthetic time series.
-Conversely, the $a$ parameter was reproduced the most poorly, with a median value less than zero across all optimization periods and priors (median NSE = `r prmall[['a']]`).
+After dissolved oxygen, gas exchange ($D$) was estimated with the highest median NSE of `r prmall[['D']]`.
+This result makes sense, given that differences in $D$ across all optimization periods and priors are due solely to difference in $DO$.
+Conversely, the $a$ parameter was reproduced the most poorly, with a median NSE of `r prmall[['a']]`.
 
@@
 
+#### Impact of noise on parameter recovery
+
 As a final evaluation of EBASE to return known metabolic parameters, a second synthetic time series was created that added estimated effects of noise to the original synthetic time series.
 The objective of the analysis is to develop an expectation of how metabolic estimates may change using *in situ* data with noise that may substantially mask a biological signal in dissolved oxygen measurements.
-The noise estimates were created using a weighted regression analysis [@Beck15; @Beck21] on actual observed dissolved oxygen time series for Apalachicola Bay using data from 2021.
+The noise estimates were created using a weighted regression analysis [@Beck15; @Beck21] on observed dissolved oxygen time series for Apalachicola Bay in 2021.
 This approach models dissolved oxygen as a function of time (day), hour, and water depth using a moving window approach with weights of the independent variables appropriate for each time period within the windows.
 The results of this analysis return a predicted and detided dissolved oxygen time series.
-Two separate time series of noise, representing residual (any other random process not related to metabolism or tidal advection) or tidal noise, were dervied from the observed, predicted and detided time series.
-The residual noise was estimated as the difference between the observed and predicted time series and the tidal noise was estimated as the difference between the predicted and detided time series.
-Both the residual and tidal noise were added to the original synthetic time series above (@fig-synapanos).
+Two separate time series of noise, representing residual (any other random process not related to metabolism or tidal advection) or tidal noise, were derived from the observed, predicted and detided time series.
+The residual noise was estimated as the difference between the observed and predicted time series and the tidal noise was estimated as the difference between the predicted and detided time series (Figure [-@fig-synapanos]a).
+Both the residual and tidal noise were added to the original synthetic time series in Figure [-@fig-synapa]c and are shown in Figure [-@fig-synapanos]b.
 EBASE was then used on the synthetic time series with noise, then compared with results from the original synthetic time series.
-The default priors justified in @sec-ebasemod (equations [-@eq-aprior] and [-@eq-rprior]) were used with a 7 day optimization period.
+The default priors justified in @sec-ebasemod (Equations [-@eq-aprior] and [-@eq-rprior]) were used with a 7 day optimization period.
 The prior for the $b$ parameter was fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ following similar logic used for the sensitivity analysis described above.
@@
 The latter comparison provided an assessment of metabolic rates that are expected to be minimally influenced by tidal advection, as a fundamental assumption of metabolic models using *in situ* dissolved oxygen time series from fixed monitoring stations.
-The prior distributions for EBASE used the default values justified in @sec-ebasemod (equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]).
+The prior distributions for EBASE used the default values justified in @sec-ebasemod (Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]).
 Compared to the previous analyses, the prior distribution for the $b$ parameter was not fixed given that its estimate should vary as a function of the data using real observations.
@@
 Pairwise comparisons of the daily estimates of $NEM$, $P$, and $R$ returned by each method were evaluated using simple summary statistics, including Pearson correlation coefficients ($\rho$) and root mean square differences (RMSD) of the linear regression fit comparing the same flux type between methods.
-Results were assessed separately for estimates using the observed or detided dissolved oxygen time series.
 @fig-apacmpfig and @tbl-apacmptab provide a comparison of the metabolic estimates from the EBASE and Odum methods.
@@
 A potential future approach for developing EBASE could be an explicit formulation of tidal advection in the model using a synoptic tidal height time series in the Bayesian framework.
-This has the advantage of simultaneously detiding and estimating metabolic rates using the same model, whereas using weighted regression prior to EBASE, as used herein, requires additional analysis and assumptions.
+Such an approach would have the advantage of simultaneously detiding and estimating metabolic rates using the same model, whereas using weighted regression prior to EBASE, as used herein, requires additional analysis and assumptions.
 However, field measurements [e.g., control volume approaches, @falter2008; @nidzieko2014] may be required to provide reasonable constraints on the prior distributions for any parameters that quantify advective influences on dissolved oxygen.
 
-There were several considerations during the development of EBASE that makes the model appropriate for estuarine settings.
-As described in @sec-basemodif, EBASE is an estuarine adaptation of BASE [@Grace15] that was developed for stream applications and many assumptions of the underlying model of the latter were inappropriate for the former.
-The most critical change was the formulation for gas exchange in EBASE using the parameterization in @Wanninkhof14 that is primarily wind-based and developed for the open-ocean, which is clearly more appropriate than the prior formulation in BASE that includes a temperature and depth dependency for streams.
-EBASE also removed the temperature-dependency of respiration and assumed that respiration did not change with water temperature, consistent with applications of the Odum method to other shallow estuaries [@Beck15; @Murrell18].
+As described in @sec-basemodif, EBASE is an estuarine adaptation of BASE [@Grace15], which was developed for stream applications and some assumptions of the underlying model of the latter were inappropriate for the former.
+The most critical change made to BASE was adopting the gas transfer velocity parameterization in @Wanninkhof14, which is primarily wind-based and more appropriate than the depth-dependent formulation in BASE.
+EBASE also removed the temperature dependency of respiration, consistent with applications of the Odum method to other shallow estuaries [@Beck15; @Murrell18].
 Overall, these changes also reduced the required inputs and many monitoring locations nationwide include the requisite data for applying EBASE (i.e., dissolved oxygen, water temperature, salinity, PAR, and wind speed, as collected by the NERRS).
@@
 EBASE also provides several computational advantages over BASE implemented in the R package that was created for the purposes herein [@Beck23].
-Most importantly, EBASE was developed to allow different model optimization lengths, whereas the BASE method was developed to estimate results for a single day.
+Most importantly, EBASE was developed to allow different model optimization periods, whereas BASE was developed to estimate results for a single day.
 Allowing EBASE to incorporate more than a day of observations in the optimization can likely produce more accurate and precise estimates for the metabolic parameters.
-Physical and biological characteristics of an ecosystem that affect the metabolic rates are likely acting at time scales longer than a day; e.g., wind patterns that can affect gas exchange as quantified by the $b$ parameter (such as fetch and waves) as well as ecological and biogeochemical characteristics influencing the $a$ and $R$ parameters (such as community composition and nutrient availability) may persist for several days.
+Physical and biological characteristics of an ecosystem that affect the metabolic rates are likely acting at time scales longer than a day, e.g., ecological and biogeochemical characteristics influencing the $a$ and $R$ parameters (such as community composition and nutrient availability) may persist for several days.
 Exposing the model to additional observations may produce more stable results as the parameters for the best fit are estimated by the Bayesian routine as a function of the data that are evaluated each optimization period.
 However, we cannot provide a precise recommendation on the number of days to use for the optimization period as the exact length of time that ecosystem characteristics can affect metabolic rates may not be known and likely varies by location.
-Our example using the synthetic time series from Apalachicola Bay suggested that the most precise results were obtained using the 30 day optimization period, although results are likely to vary across sytems.
+Our example using the synthetic time series from Apalachicola Bay suggested that the most precise results were obtained using the 30 day optimization period, although results are likely to vary across systems.
 For most applications, very short (e.g., 1 day) or very long (e.g., \> 30 days) are not recommended, such that the former may produce unstable results and the latter may miss important events in the time series producing parameters that are overly generalized.
@@
 The EBASE R package can also accommodate missing observations in the input data.
-This allows the estimation of metabolic rates for time series that cover relatively long periods of time (e.g., more than year), when data gaps are likely to occur for several reasons (e.g., equipment malfunction, routine maintenance, etc.).
-The R package accommodates data gaps using linear interpolation for any required input used by EBASE.
-This allows the Bayesian routine to provide continuous estimates across the time series.
+This feature allows the estimation of metabolic rates for time series that cover relatively long periods of time (e.g., more than year), when data gaps are likely to occur for several reasons (e.g., equipment malfunction and routine maintenance).
+The R package accommodates data gaps using linear interpolation for any required input used by EBASE, allowing the Bayesian routine to provide continuous estimates across the time series.
 The results are then automatically post-processed to remove any metabolic estimates that exceed a threshold for the number of interpolated observations.
-The analyst can determine the appropriate length of time to use for excluding results based, although the default setting for EBASE is 12 hours, where results for an entire optimization period are removed if any interpolated period exceeded 12 hours within that period.
+The analyst can determine the appropriate length of time to use for excluding results, although the default setting for EBASE is 12 hours, where results for an entire optimization period are removed if any interpolated period exceeded 12 hours within that period.
 Although this does not apply to the synthetic time series, some of the estimates using the 2021 observed data for Apalachicola Bay were excluded from the analysis.
@@
 Our results have demonstrated that unknown metabolic parameters can be estimated with reasonable certainty and EBASE could be a valuable approach to improve the understanding of critical ecosystem processes as the method can be readily applied to other settings.
-A notable limitation of EBASE was an inability to precisely recover the $b$ parameter from a synthetic time series, although this result may be explained by a minimal contribution of gas exchange as expressed in the dissolved oxygen time series.
+A notable limitation of EBASE was an inability to constrain the gas transfer velocity from a synthetic time series, although this result may be explained by a minimal contribution of gas exchange as expressed in the dissolved oxygen time series.
 Effects of tidal advection on dissolved oxygen measurements were also observed, demonstrating a violation of the assumption that the measured data represent a water sample having a continuous metabolic history.

### 2023-10-17 R/figs.R
default ebase prior plot comparison

@@ -109,6 +109,12 @@ png(here('figs/synapa.png'), height = 7, width = 7, family = 'serif', units = 'i
 print(p)
 dev.off()
 
+# parameter recovery with defaults ------------------------------------------------------------
+
+fl <- paste0(tempdir(), '/apadef.RData')
+download.file('https://github.com/fawda123/BASEmetab_script/raw/master/data/apadef.RData', destfile = fl)
+load(file = fl)
+
 # Fwoxy apa comparison ------------------------------------------------------------------------
 
 fl <- paste0(tempdir(), '/apasumdat.RData')

### 2023-10-17 R/funcs.R

@@ -359,6 +359,81 @@ cmpplo <- function(res, fwdatcmp, subttl, ylbs = TRUE){
   
 }
 
+# compare synthetic with default parameters
+defplo <- function(apadef, fwdatcmp){
+  
+  res <- apadef %>% 
+    enframe() %>% 
+    mutate(
+      value = purrr::map(value, credible_prep)
+    ) %>% 
+    unnest('value')
+  
+  toplo <- fwdatcmp %>% 
+    pivot_longer(!all_of(c('Date', 'DateTimeStamp')), names_to = 'var', values_to = 'Synthetic') %>% 
+    filter(var %in% c('a', 'R', 'b')) %>% 
+    summarise(
+      Synthetic = mean(Synthetic, na.rm = T), 
+      .by = c('Date', 'var')
+    ) %>% 
+    inner_join(res, ., by  = c('Date', 'var')) %>% 
+    mutate(
+      var = factor(var, 
+                   levels = c('a', 'R', 'b'), 
+                   labels = c(
+                     'italic(a)~(mmol~d^{-1}~W^{-1})',
+                     'R~(mmol~m^{2}~d^{-1})', 
+                     'b~(cm~hr^{-1})/(m^2~s^{-2})'
+                   )
+      ), 
+      name = factor(name, 
+                    levels = c('opt7', 'opt30'), 
+                    labels = c('(a)~7~day', '(b)~30~day')
+      )
+    ) %>% 
+    reframe(
+      mean = na.omit(mean), 
+      lo = na.omit(lo), 
+      hi = na.omit(hi), 
+      Synthetic = mean(Synthetic, na.rm = T),
+      Date = min(Date),
+      .by = c('grp', 'name', 'var')
+    )
+  
+  brks <- seq.Date(min(toplo$Date), max(toplo$Date), by = '3 months')
+  
+  p <- ggplot(toplo, aes(x = Date)) + 
+    # geom_line(aes(y = mean, color = 'EBASE')) + 
+    geom_point(aes(y = mean, color = 'EBASE'), show.legend = T) + 
+    geom_line(aes(y = Synthetic, color = 'Synthetic')) +
+    geom_errorbar(aes(ymin = lo, ymax = hi, color = 'EBASE'), width = 0, show.legend = F) +
+    facet_grid(var ~ name, scales = 'free_y', labeller = label_parsed, switch = 'y') +
+    scale_x_date(date_labels = '%b', breaks = brks) + 
+    theme_minimal() +
+    theme(
+      strip.placement = 'outside', 
[…]

### 2023-10-18 R/figs.R
save figs.R

@@ -115,6 +115,12 @@ fl <- paste0(tempdir(), '/apadef.RData')
 download.file('https://github.com/fawda123/BASEmetab_script/raw/master/data/apadef.RData', destfile = fl)
 load(file = fl)
 
+p <- defplo(apadef, fwdatcmp)
+
+png(here('figs/defplo.png'), height = 7, width = 7, family = 'serif', units = 'in', res = 500)
+print(p)
+dev.off()
+
 # Fwoxy apa comparison ------------------------------------------------------------------------
 
 fl <- paste0(tempdir(), '/apasumdat.RData')

### 2023-10-19 R/figs.R
plots with ndays results where 7 or 30 have last incomplete group removed

@@ -115,6 +115,10 @@ fl <- paste0(tempdir(), '/apadef.RData')
 download.file('https://github.com/fawda123/BASEmetab_script/raw/master/data/apadef.RData', destfile = fl)
 load(file = fl)
 
+# final groups have smaller sample size
+apadef <- apadef %>% 
+  lapply(function(x) filter(x, grp != max(grp)))
+
 p <- defplo(apadef, fwdatcmp)
 
 png(here('figs/defplo.png'), height = 7, width = 7, family = 'serif', units = 'in', res = 500)
@@ -159,6 +163,23 @@ lims <- list(
   R = c(0, 600), 
   a = c(0, 11)
 )
+
+# # view incomplete groups in apagrd, the last one
+# lapply(apagrd$out, function(x) table(x[[1]]$grp))
+# remove incomplete groups
+apagrd <- apagrd %>% 
+  mutate(
+    out = purrr::pmap(list(ndays, out), function(ndays, out){
+
+      # filter incomplete groups for 7, 30 day opt
+      if(ndays %in% c(7, 30))
+        out[[1]] <- out[[1]] %>% 
+          filter(grp != max(grp))
+      
+      return(out)
+      
+    })
+  )
 
 p <- optex(apagrd, fwdatcmp, apasumdat, rnkmetsum = c(1, 16), met = 'nse', lims = lims)
 
@@ -240,6 +261,10 @@ load(file = fl)
 fl <- paste0(tempdir(), '/resnos.RData')
 download.file('https://github.com/fawda123/BASEmetab_script/raw/master/data/resnos.RData', destfile = fl)
 load(file = fl)
+
+# remove incomplete groups from resobs, resnos
+resobs <- resobs %>% filter(grp != max(grp))
+resnos <- resnos %>% filter(grp != max(grp))
 
 p <- syncomp_plo(resobs, resnos, fwdatcmp) 
   

### 2023-10-19 manu-draft.qmd
default results and text added for RN comments

@@
 
+The synthetic time series was used as input to EBASE to predict metabolic parameters with the default prior distributions above.
+Two optimization periods of 7 and 30 days were evaluated.
+This analysis provided a demonstration that the default priors are not entirely unreasonable based on the ability to reproduce the known parameters in the synthetic time series [@fig-defplo].
+Overall, EBASE produced reasonable results for $a$ and $R$, with higher accuracy using the 30 day optimization period (Figure [-@fig-defplo]b) compared to the 7 day optimization period (Figure [-@fig-defplo]a).
+However, the model was unable to return the $b$ parameter with acceptable accuracy, suggesting the data were insufficient to constrain this parameter.
+Further, many of the 95% credible intervals from the posterior distributions of each parameter did not overlap the known values from the synthetic time series.
+Estimates for each optimization period may be inaccurate despite overall similarities comparing all estimates to the known values for the whole year.
+
 #### Sensitivity to priors
@@
 Results from optimization periods of 1, 7, and 30 days were compared for the synthetic one-year time series.
+For the models with 7 and 30 day optimization periods, the final period in the annual time series was excluded from the results due to incomplete number of observations compared to earlier periods (i.e., the total number of days in the time series was not evenly divided by 7 or 30, resulting in a remainder for the last period).
 Each optimization period was also evaluated for every unique combination of prior distributions, totaling 48 (16 times three optimization periods) unique analyses for the assessment of changing priors of all parameters.
@@
 By optimization period, median NSE values across all model outputs increased as the number of days used for model optimization increased (1 day = `r optnse[['1']]`, 7 days = `r optnse[['7']]`, 30 days = `r optnse[['30']]`).
-The top three ranked models were identical across optimization periods and median NSE values across model outputs for the top models increased with optimization period (1 day = `r optmod[['1']]`, 7 days = `r optmod[['7']]`, 30 days = `r optmod[['30']]`).
+The top three ranked models were identical for the 1 and 7 day optimization period, whereas the top three models for the 30 day optimization appeared less dependent on the choice of priors.
+Median NSE values across model outputs for the top models increased with optimization period (1 day = `r optmod[['1']]`, 7 days = `r optmod[['7']]`, 30 days = `r optmod[['30']]`).
 The NSE values for all model outputs were greater than zero for the top three models in the 30 day optimization period, whereas the $a$ parameter had NSE less than zero for the top model in the 7 day optimization period and $R$ and $a$ had negative NSE values for the top model in the 1 day optimization period.
-About half of the model outputs regardless of optimization period or priors had NSE \< 0 (`r perneg`).
-As such, the prior combinations that produced the highest NSE values across all parameters were those where all prior distributions had low $\mu$ and $\sigma$ values for both $a$ and $R$ for the 1 and 7 day optimization periods, whereas the same combination was observed for the 30 day optimization period with the exception that $\sigma$ was high for $R$.
-Notably, many of the top models for each optimization period had at least one high value for $\sigma$ for $a$ or $R$, suggesting that models with sufficient flexibility to search a larger parameter space can produce metabolic estimates most similar to actual values.
+Less than half of the model outputs regardless of optimization period or priors had NSE \< 0 (`r perneg`).
+As such, the prior combinations that produced the highest NSE values across all parameters were those where all prior distributions had low $\mu$ and $\sigma$ values for both $a$ and $R$ for the 1 and 7 day optimization periods, whereas the combination of priors did not have a notable effect on the output for the 30 day optimization period.
+Notably, many of the top models for the shorter optimization periods had at least one high value for $\sigma$ for $a$ or $R$, suggesting that models with sufficient flexibility to search a larger parameter space can produce metabolic estimates most similar to actual values.
 
@@
 The best performing models for each optimization period (@fig-optex a, c, e) showed similar results where the model outputs were similar to the synthetic time series, although more variation in the EBASE results were of course observed with the 1 day optimization period.
-The worst performing model for each optimization period (@fig-optex b, d, f) demonstrates the negative NSE values that were obtained (i.e., EBASE values much lower or higher than the synthetic values).
-The worst performing models were caused by constraining EBASE to a small parameter space, such as a high $\mu$ and low $\sigma$ for $a$ as in the bottom row for subplots (b), (d), and (f).
+The worst performing models for the 1 and 7 day optimization periods (@fig-optex b, d) demonstrate the negative NSE values that were obtained (i.e., EBASE values much lower or higher than the synthetic values).
+The worst performing models were caused by constraining EBASE to a small parameter space, such as a high $\mu$ and low $\sigma$ for $a$ as in the bottom row for subplots (b) and (d).
 In such cases, the $DO$ simulation (fourth row) can still be quite good as a result of compensating errors in $a$ and $R$.
 The EBASE-derived rates of photosynthesis and respiration are both much higher than in the synthetic data for subplots (b) and (d), errors that compensate well enough to produce fairly accurate recoveries of the $DO$ time series.
-Further, the worst model for the 30 day optimization period in subplot (f) had the same priors as the 7 day optimization period, but was able to better reproduce the synthetic results, suggesting that poor estimates from inadequate priors can in part be mitigated using longer optimization periods.
+Further, resutls for the worst model for the 30 day optimization period in subplot (f) are visually indistinguishable from those of the best model in subplot (e), suggesting that poor estimates from inadequate priors can in part be mitigated using longer optimization periods.
 
@@
 The prior distributions for EBASE used the default values justified in @sec-ebasemod (Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]).
-Compared to the previous analyses, the prior distribution for the $b$ parameter was not fixed given that its estimate should vary as a function of the data using real observations.
+Similar to the previous analyses, the prior distribution for the $b$ parameter was fixed given the difficulty in recovery.
 A 1 day model optimization period was used for EBASE for comparability with the Odum method that returns daily estimates.
@@
 
+![Comparison of EBASE results with default prior distributions to a synthetic time series created with known model parameters ($a$, $R$, and $b$).
+Results are compared for EBASE parameters estimated with a (a) 7 day and (b) 30 day optimization period.
+95% credible intervals from the posterior distributions of the parameter estimates from EBASE are also shown.
+The results are averaged at the time step of the model optimization period.](figs/defplo.png){#fig-defplo}
+
+{{< pagebreak >}}
+
 ![Comparison of EBASE results to a synthetic time series created with known model parameters ($a$, $R$, and $b$).

### 2023-10-19 manu-draft.qmd
fix apadef plot figure caption
minor text edit

@@
 Two optimization periods of 7 and 30 days were evaluated.
-This analysis provided a demonstration that the default priors are not entirely unreasonable based on the ability to reproduce the known parameters in the synthetic time series [@fig-defplo].
+This analysis provided a demonstration that the default priors are not entirely unreasonable based on the ability to reproduce the known parameters in the synthetic time series ([@fig-defplo]).
 Overall, EBASE produced reasonable results for $a$ and $R$, with higher accuracy using the 30 day optimization period (Figure [-@fig-defplo]b) compared to the 7 day optimization period (Figure [-@fig-defplo]a).
@@
 
-![Comparison of EBASE results with default prior distributions to a synthetic time series created with known model parameters ($a$, $R$, and $b$).
+![Comparison of EBASE results using default prior distributions to a synthetic time series created with known model parameters ($a$, $R$, and $b$).
 Results are compared for EBASE parameters estimated with a (a) 7 day and (b) 30 day optimization period.

### 2023-11-19 manu-draft.qmd
formatting for submission

@@
     reference-doc: my_styles.docx
-    number-sections: true
+    number-sections: false
 title: "Fitting metabolic models to dissolved oxygen data: The Estuarine BAyesian Single-station Estimation method (EBASE)"
@@
 ```
+
+*Running head*: EBASE metabolic model
+
+*Key words*: Bayesian, dissolved oxygen, ecosystem metabolism, estuary, gas exchange, open source
+
 ## Abstract {.unnumbered}
@@
 Similarly, estuarine primary production consumes dissolved inorganic carbon delivered from land, forming particulate organic carbon that may be buried in estuarine sediments [e.g., @hu2006].
-Estuarine metabolic processes are fundamentally important, as illustrated by the dissolved inorganic carbon budget of eastern North American estuaries [@najjar2018]: net primary production (11.7 $\pm$ 3.0 Tg C yr$^{-1}$, best estimate $\pm$ 2 standard errors) and heterotrophic respiration (13.2 $\pm$ 3.1 Tg C yr$^{–1}$) far exceed inputs from upland sources (5.6 $\pm$ 1.2 Tg C yr$^{-1}$) and outputs to the atmosphere (4.2 $\pm$ 1.7 Tg C yr$^{–1}$) and ocean (2.9 $\pm$ 2.3 Tg C yr$^{–1}$).
+Estuarine metabolic processes are fundamentally important, as illustrated by the dissolved inorganic carbon budget of eastern North American estuaries [@najjar2018]; net primary production and heterotrophic respiration far exceed inputs from upland sources and outputs to the atmosphere and ocean.
 As such, small changes in either net primary production or heterotrophic respiration can have a large impact on the difference between the two, net ecosystem metabolism (NEM, sometimes also referred to as net ecosystem production), which ultimately governs the form and quantity of estuarine materials that are lost to the atmosphere, the sediments, and the ocean.
@@
 A similar synthesis was done by @yvon-durocher2012.
-These studies evaluated the open-water method as applied to continuous monitoring data available for several decades across the National Estuarine Research Reserve System (NERRS).
+These studies evaluated the open-water method as applied to continuous monitoring data available for several decades across the National Estuarine Research Reserve System (NERRS), a network of 30 estuaries across the U.S.
 Thus, the approach can be broadly applied where data are available, although it is not without limitations.
@@
 
-Another limitation of the open-water method is the estimation of the air-water oxygen flux, which is an important term in the oxygen budget.
+Another limitation of the open-water method is the estimation of the air--water oxygen flux, which is an important term in the oxygen budget.
 Most approaches estimate this flux using an empirically derived gas transfer velocity that is dependent on wind speed and temperature or held at a constant value [e.g., @caffrey2004; @russell2007; @nagel2009; @Beck15; @Murrell18].
@@
 The rationale and changes made to BASE are described in detail, followed by a demonstration of how EBASE can estimate known metabolic parameters with reasonable certainty from a synthetic time series.
-Lastly, EBASE is compared to metabolic estimates using the Odum open-water method to identify key differences related to both the theoretical and statistical differences associated with each method.
+Lastly, metabolic estimates from EBASE are compared with those of the Odum open-water method to identify key differences related to both the theoretical and statistical differences associated with each method.
 These comparisons combine results using real observations and those based on detided inputs to demonstrate effects of tidal advection on metabolic estimates.
@@
 
-where the terms on the right side of the equation are gross primary production ($P$), ecosystem respiration ($R$), and the net upward diffusive gas flux at the air-water interface ($D$), respectively, processes that result in the change in dissolved oxygen concentration ($dC$) per unit time ($dt$).
-The equation has dimensions of moles per unit area of time.
+where the terms on the right side of the equation are gross primary production ($P$), ecosystem respiration ($R$), and the net upward diffusive gas flux at the air--water interface ($D$), respectively, processes that result in the change in dissolved oxygen concentration ($dC$) per unit time ($dt$).
+The equation has dimensions of moles per unit area and time.
 $P$ is modeled as $aPAR$, where $PAR$ is the surface photosynthetically active radiation and $a$ is the light efficiency; $R$ is constant; and $D$ is modeled in the standard manner [e.g, @sarmiento2013] as $k_w\left(C_{sat} - C\right)$, where $k_w$ is the gas transfer velocity and $C_{sat}$ is the saturation (or equilibrium) concentration, calculated from salinity and temperature [@Garcia92].
@@
 The $b$ parameter is fairly well known for open-ocean conditions, but may be different in estuaries, where the transfer velocity is influenced by factors other than wind speed and $Sc$.
-As such, the three parameters estimated by EBASE are $a$, $R$, and $b$ using the required input data that includes time series of dissolved oxygen, water temperature, salinity, PAR, wind speed, and water column depth.
+As such, the three parameters estimated by EBASE are $a$, $R$, and $b$ using the required input data that include time series of dissolved oxygen, water temperature, salinity, PAR, wind speed, and water column depth.
 
@@
 As such, $k_w$ is usually modeled as a function of measurable variables that can potentially predict turbulent mixing of the surface water, such as wind speed.
-$k_w$ also increases with temperature because the film of water at the air-water interface becomes thinner as the viscosity decreases and the random motion of gas molecules increases.
+$k_w$ also increases with temperature because the film of water at the air--water interface becomes thinner as the viscosity decreases and the random motion of gas molecules increases.
 This dependence of $k_w$ on temperature is usually modeled with the Schmidt number as in @eq-ebase, which combines information on the viscosity of the fluid and the diffusivity of the gas.
@@
 An expectation is that $k_w$ would increase as water depth decreases for a given current speed because turbulence generated at the sediment--water interface by currents would more easily reach the surface [@Ho16].
-Based on these differences and the lack of availability of other possible drivers of the transfer velocity in coastal systems, the well-established wind- and temperature-based parameterization for gas exchange of @Wanninkhof14 was used for EBASE, but allowing $b$, which was esimated by @Wanninkhof14 to be 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$, to be a free parameter that may reflect spatial and temporal variability in factors other than wind and temperature that may influence transfer velocity.
+Based on these differences and the lack of availability of other possible drivers of the transfer velocity in coastal systems, the well-established wind- and temperature-based parameterization for gas exchange of @Wanninkhof14 was used for EBASE, but allowing $b$, which was estimated by @Wanninkhof14 to be 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$, to be a free parameter that may reflect spatial and temporal variability in factors other than wind and temperature that may influence transfer velocity.
 
@@
 For ease of application, an R package was developed to implement EBASE [@Beck23; @RCT23], similar to the R package available for BASE (<https://github.com/dgiling/BASEmetab>).
-Changes in model implementation were included in the package as additional changes to BASE.
+EBASE differs from BASE in model implementation in two ways.
 First, users can specify the model optimization period, where the time period can vary from a minimum of 1 day to the maximum of the entire length of the time series.
@@
 Again, this feature was not included in the original BASE method and a complete time series was required for use.
-A detailed web page (<https://fawda123.github.io/EBASE/>) describes how to use the package, with explanations of the functions for viewing model results, including a plot of the modeled dissolved oxygen with the observed, a time series plot of the primary metabolic estimates (P, R, and D, as areal rates with units $\text{mmol}/\text{m}^2/\text{d}$), and a time series plot of the credible intervals for the $a$, $R$, and $b$ parameters.
+A detailed web page (<https://fawda123.github.io/EBASE/>) describes how to use the package, with explanations of the functions for viewing model results, including a plot of the modeled dissolved oxygen with the observed, a time series plot of the oxygen budget terms (P, R, and D), and a time series plot of the credible intervals for the $a$, $R$, and $b$ parameters.
 All results provided herein were generated using the EBASE R package.
@@
 
-The remaining three parameters $a$ , $R$, and $b$ are estimated by likelihood given the observed data and prior distributions for the parameters.
+The parameters $a$, $R$, and $b$ are estimated by likelihood given the observed data and prior distributions for the parameters.
 The "Just Another Gibbs Sampler" (JAGS) software [as for BASE, @Plummer03] is used with the EBASE R package (described above) to estimate probability distributions of the unknown parameters using Markov Chain Monte Carlo (MCMC) simulations in a Bayesian framework applied to a finite-difference form of @eq-ebase.
@@
 Reasonable uninformed prior distributions for EBASE were chosen with mean values using approximate estimates from the literature [@caffrey2004; @Wanninkhof14] and standard deviations that were sufficiently large to allow the Bayesian routine to search an unconstrained parameter space.
-Specifically, the prior mean for $R$ was chosen as 300 $\text{mmol}/\text{m}^2/\text{d}$ (\~10 $\text{g O}_2/\text{m}^2/\text{d}$) using a national synthesis of metabolic rates for US estuaries [Table 2 in @caffrey2004].
-Although @caffrey2004 focused on small, shallow estuaries that are part of the NERRS, this dataset is the largest, longest, and most consistently collected in the US and it is reasonable to expect others may apply EBASE to these data.
+
+Specifically, the prior mean for $R$ was chosen as 300 $\text{mmol}/\text{m}^2/\text{d}$ (\~10 $\text{g O}_2/\text{m}^2/\text{d}$) using a synthesis based on NERRS data [Table 2 in @caffrey2004].
+Although this synthesis focused on small, shallow estuaries that are part of the NERRS, this dataset is the largest, longest, and most consistently collected in the US and it is reasonable to expect others may apply EBASE to these data.
 The prior mean for $a$ was chosen assuming that $P$ and $R$ are comparable at long time scales, such that $aPAR$ = $R$, or $a$ = $\left(300~\text{mmol}/\text{m}^2/\text{d}\right)/PAR$.
@@
 Lastly, the prior mean for $b$ was chosen as 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ following @Wanninkhof14.
+
 The prior standard deviation for $R$ was based on the standard deviation of the mean values of $R$ at the NERRS sites in @caffrey2004.
@@
 The standard deviation for $a$ was set proportionally to 2 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$.
-For the prior standad deviation of $b$, we considered using the 20% error estimated for $b$ by @Wanninkhof14.
+For the prior standard deviation of $b$, we considered using the 20% error estimated for $b$ by @Wanninkhof14.
 However, we chose a larger value, 50%, because the wind speed-based gas transfer velocity formulation of @Wanninkhof14 was mainly developed and calibrated for open-ocean applications, where the complications of surfactants [@frew1997] and tidal currents [@Ho16], as well as fetch and suspended matter [@Borges11], are expected to be minimal.
-Together, the default priors for $a$, $R$, and $b$ are:
+
+Together, the default prior distributions for $a$, $R$, and $b$ (@fig-priorplot) are:
 
@@
 Initial development of EBASE showed that using an undefined upper limit of $b$ led to unstable and unreasonable estimates for all parameters.
-Examples of the prior distributions in Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior] are shown in @fig-priorplot.
 Functions in the EBASE R package allow alternative prior distributions from those above.
@@
 A preliminary assessment of the ability of EBASE to produce reasonable parameter estimates for $a$, $R$, and $b$ (@eq-ebase) was conducted using a synthetic time series of dissolved oxygen concentration created from observed water temperature, salinity, air temperature, $PAR$, and wind speed measured every 15 minutes.
-These assessments evaluated if 1) EBASE estimates similar values for the known parameters used to create the synthetic time series, and 2) how the estimated values change with different model configurations.
-The synthetic time series used inputs from one year (2021) of continuous data from Apalachicola Bay, Florida.
-Water temperature and salinity were from the Cat Point water quality monitoring station (29.7021$^{\circ}$ N, --84.8802$^{\circ}$ W) and the meteorological observations were from the nearby East Bay station (29.7694$^{\circ}$ N, --84.8815$^{\circ}$ W, \~10 km north of Cat Point).
-Missing observations in the water and meteorological data (`r wqmis`% and `r mtmis`% of all observations in each dataset, respectively) were filled using autoregressive modelling of the actual data versus time to create a complete dataset from January 1st to December 31st [@akaike1969].
+This assessment evaluated if (1) EBASE estimates similar values for the known parameters used to create the synthetic time series, and (2) how the estimated values change with different model configurations.
+The synthetic time series used inputs from one year (2021) of continuous data from the NERRS reserve in Apalachicola Bay, Florida.
+Water temperature and salinity were from the Cat Point water quality monitoring station (29.7021$^{\circ}$ N, 84.8802$^{\circ}$ W) and the meteorological observations were from the nearby East Bay station (29.7694$^{\circ}$ N, 84.8815$^{\circ}$ W, \~10 km north of Cat Point).
+Missing observations in the water and meteorological data (`r wqmis`% and `r mtmis`% of all observations in each dataset, respectively) were filled using autoregressive modelling [@akaike1969] of the actual data versus time to create a complete dataset from January 1st to December 31st.
 
@@
 The $b$ parameter was fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ [@Wanninkhof14].
-Daily values of $a$ and $R$ were estimatd by 1) applying the Odum open-water method as implemented in the WtRegDO R package [@Beck21], 2) computing monthly averages of the resulting daily estimates of $P$ and $R$, 3) computing the monthly average of the parameter $a$ by dividing the monthly average of $P$ by the monthly average of $PAR$, and 4) by fitting a spline function to the monthly averages of $R$ and $a$.
+Daily values of $a$ and $R$ were estimatd by (1) applying the Odum open-water method as implemented in the WtRegDO R package [@Beck21], (2) computing monthly averages of the resulting daily estimates of $P$ and $R$, (3) computing the monthly average of the parameter $a$ by dividing the monthly average of $P$ by the monthly average of $PAR$, and (4) by fitting a spline function to the monthly averages of $R$ and $a$.
 
@@
 
+## References {.unnumbered}
+
+::: {#refs}
+:::
+
 ## Acknowledgments {.unnumbered}
@@
 
-## Figures {.unnumbered}
+## Figure Legends {.unnumbered}
 
@@
 ![Synthetic time series for one year of continuous data at Apalachicola Bay.
-Synthetic data were created using the (a) $a$ parameter to estimate (b) metabolic estimates for gross primary production ($P$), the negative of ecosystem respiration ($-R$), and gas exchange ($-D$) and the (c) dissolved oxygen time series (including saturation, $C_{sat}$).](figs/synapa.png){#fig-synapa}
+Synthetic data were created using the (a) $a$ parameter to estimate (b) metabolic estimates for gross primary production ($P$), ecosystem respiration ($R$), and gas exchange ($D$) and the (c) dissolved oxygen time series (including saturation, $C_{sat}$).](figs/synapa.png){#fig-synapa}
 
@@
 The means ($\mu$) and standard deviations ($\sigma$) are evaluated as low (L) or high (H) values (see text for details) and the optimization period for EBASE is 1 day, 7 days, or 30 days.
-The model predictions of dissolved oxygen (DO), gross primary production (P), ecosystem respiration (R), air-water oxygen flux (D), and light efficiency ($a$) are evaluated based on Nash-Sutcliffe Efficiency (NSE) values between the EBASE results and synthetic values.
+The model predictions of dissolved oxygen (DO), gross primary production (P), ecosystem respiration (R), air--water oxygen flux (D), and light efficiency ($a$) are evaluated based on Nash-Sutcliffe Efficiency (NSE) values between the EBASE results and synthetic values.
 The $b$ parameter is a fixed constant and was not evaluated.
@@
 ![Comparison of EBASE results for the best and worst performing models to a synthetic time series created with known metabolic parameters.
-The best and worst performing models for the 1 day (a, b), 7 day (c, d), and 30 day (e, f) optimization periods are based on the EBASE results with prior distributions that produced the highest and lowest median Nash-Sutcliffe Efficency values across all parameter estimates (@fig-priorcomp).
+The best and worst performing models for the 1 day (a, b), 7 day (c, d), and 30 day (e, f) optimization periods are based on the EBASE results with prior distributions that produced the highest and lowest median Nash-Sutcliffe Efficiency values across all parameter estimates (@fig-priorcomp).
 The results are averaged at the time step of the model optimization period.](figs/optex.png){#fig-optex fig-align="center"}
@@
 ```
-
-{{< pagebreak >}}
-
-## References {.unnumbered}

### 2023-11-20 manu-draft.qmd
update abstract from rgn comments

@@
 
-*Key words*: Bayesian, dissolved oxygen, ecosystem metabolism, estuary, gas exchange, open source
+*Key words*: Bayesian, dissolved oxygen, ecosystem metabolism, estuary, gas exchange, open-source
 
@@
 
-Continuous monitoring data for dissolved oxygen have been used to estimate the contribution of gross primary production and ecosystem respiration to net ecosystem metabolism.
-Most methods for estimating metabolic processes make several assumptions on how biological and physical factors affect the expression of metabolism in dissolved oxygen data, which can lead to poorly constrained errors that influence the interpretation of results.
-In estuaries, these methods can inform our understanding of spatial and temporal patterns in trophic state, yet very few models have been developed that address challenges of estimating metabolism in these dynamic systems.
-This paper describes the development, assessment, and interpretation of results from a new approach for estimating metabolic rates in estuaries, the Estuarine BAyesian Single-station Estimation (EBASE) method.
-EBASE provides several advantages over existing models by using an integration formula appropriate for estuarine systems and using a Bayesian framework where critical model parameters are fit as a function of the data as informed by a set of prior distributions.
-This approach can also vary the time period over which critical model parameters are estimated.
-We demonstrate that EBASE can reproduce known metabolic parameters from a synthetic time series created with continuous monitoring data from the Apalachicola Bay (southeastern United States) and provide a quantitative assessment of how the prior distributions affect the model results.
-The approach is also bench-marked with more conventional methods for estimating metabolism, with critical differences attributed to formulation of the metabolic mass-balance equations and the underlying statistical framework (i.e., Bayesian vs simple integration).
-The EBASE approach is available using open source software and can be readily applied to multiple years of long-term monitoring data that are available in many estuaries.
+Continuous measurements of dissolved oxygen (DO) are useful for quantifying ecosystem metabolism, which is critical for understanding estuarine biogeochemistry and ecology, but current methods applied to these data may lead to unphysical results and poorly constrained errors.
+Here we present a new approach for estimating estuarine metabolism: EBASE (Estuarine BAyesian Single-station Estimation).
+EBASE applies a Bayesian framework to a simple process-based model and DO observations, allowing the estimation of critical model parameters, specifically light efficiency and respiration, as informed by a set of prior distributions.
+EBASE improves upon the stream-based model from which it was derived by accommodating missing DO data and allowing the user to set the time period over which parameters are estimated.
+We demonstrate that EBASE can recover known metabolic parameters from a synthetic time series, even in the presence of noise (e.g., due to tidal advection) and when prior distributions are uninformed.
+Optimization periods of 7 days and 30 days are more preferable than 1 day.
+A comparison with the more-conventional method of Odum reveals the ability of EBASE to avoid unphysical results (such as negative photosynthesis and respiration) and improves when the DO data are detided.
+EBASE is available using open-source software and can be readily applied to multiple years of long-term monitoring data that are available in many estuaries.
 Overall, EBASE provides an accessible method to parameterize a simple metabolic model appropriate for estuarine systems and will provide additional understanding of processes that influence ecosystem status and condition.
-
-*Key words*: Bayesian, dissolved oxygen, ecosystem metabolism, estuary, gas exchange, open source
 
@@
 This differs from the BASE approach where the optimization period is set at 1 day.
-Sensitivity of the EBASE method to different optimization periods is described in @sec-cmpmod.
+Sensitivity of the EBASE method to different optimization periods is described below.
 Second, the EBASE software can accommodate missing observations in the input data by interpolating gaps prior to estimating the metabolic parameters.
@@
 EBASE was then used on the synthetic time series with noise, then compared with results from the original synthetic time series.
-The default priors justified in @sec-ebasemod (Equations [-@eq-aprior] and [-@eq-rprior]) were used with a 7 day optimization period.
+The default priors (Equations [-@eq-aprior] and [-@eq-rprior]) were used with a 7 day optimization period.
 The prior for the $b$ parameter was fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ following similar logic used for the sensitivity analysis described above.
@@
 The latter comparison provided an assessment of metabolic rates that are expected to be minimally influenced by tidal advection, as a fundamental assumption of metabolic models using *in situ* dissolved oxygen time series from fixed monitoring stations.
-The prior distributions for EBASE used the default values justified in @sec-ebasemod (Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]).
+The prior distributions for EBASE used the default values (Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]).
 Similar to the previous analyses, the prior distribution for the $b$ parameter was fixed given the difficulty in recovery.
@@
 
-As described in @sec-basemodif, EBASE is an estuarine adaptation of BASE [@Grace15], which was developed for stream applications and some assumptions of the underlying model of the latter were inappropriate for the former.
+As described in the materials and procedures section, EBASE is an estuarine adaptation of BASE [@Grace15], which was developed for stream applications and some assumptions of the underlying model of the latter were inappropriate for the former.
 The most critical change made to BASE was adopting the gas transfer velocity parameterization in @Wanninkhof14, which is primarily wind-based and more appropriate than the depth-dependent formulation in BASE.

### 2023-11-20 manu-draft.qmd
fix ref for wtregdo

@@
 The $b$ parameter was fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ [@Wanninkhof14].
-Daily values of $a$ and $R$ were estimatd by (1) applying the Odum open-water method as implemented in the WtRegDO R package [@Beck21], (2) computing monthly averages of the resulting daily estimates of $P$ and $R$, (3) computing the monthly average of the parameter $a$ by dividing the monthly average of $P$ by the monthly average of $PAR$, and (4) by fitting a spline function to the monthly averages of $R$ and $a$.
+Daily values of $a$ and $R$ were estimatd by (1) applying the Odum open-water method as implemented in the WtRegDO R package [@Beck23a], (2) computing monthly averages of the resulting daily estimates of $P$ and $R$, (3) computing the monthly average of the parameter $a$ by dividing the monthly average of $P$ by the monthly average of $PAR$, and (4) by fitting a spline function to the monthly averages of $R$ and $a$.
 
@@
 The objective of the analysis is to develop an expectation of how metabolic estimates may change using *in situ* data with noise that may substantially mask a biological signal in dissolved oxygen measurements.
-The noise estimates were created using a weighted regression analysis [@Beck15; @Beck21] on observed dissolved oxygen time series for Apalachicola Bay in 2021.
+The noise estimates were created using a weighted regression analysis [@Beck15; @Beck23a] on observed dissolved oxygen time series for Apalachicola Bay in 2021.
 This approach models dissolved oxygen as a function of time (day), hour, and water depth using a moving window approach with weights of the independent variables appropriate for each time period within the windows.
@@
 The Odum method used an approach described in @Murrell18 created for estuarine application that uses a gas-exchange approach from @Thebault08.
-For a better comparison with the EBASE methods, the gas exchange parameterization in the Odum method was modified to use the @Wanninkhof14 approach and the $b$ parameter was fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ [as available in the WtRegDO R package, @Beck21].
+For a better comparison with the EBASE methods, the gas exchange parameterization in the Odum method was modified to use the @Wanninkhof14 approach and the $b$ parameter was fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ [as available in the WtRegDO R package, @Beck23a].
 An additional and fundamental difference between EBASE and the Odum method is that the former is likelihood-based with metabolic rates estimated by fitting the model to the data, whereas the latter is strictly arithmetic-based primarily on integration of the diel dissolved oxygen curve.
 Thus, potentially different results are expected based on the alternative statistical approaches to estimate the metabolic results.
-Each comparison also evaluated differences in the results using the observed dissolved oxygen time series and a detided time series using weighted regression [@Beck15; @Beck21].
+Each comparison also evaluated differences in the results using the observed dissolved oxygen time series and a detided time series using weighted regression [@Beck15; @Beck23a].
 The latter comparison provided an assessment of metabolic rates that are expected to be minimally influenced by tidal advection, as a fundamental assumption of metabolic models using *in situ* dissolved oxygen time series from fixed monitoring stations.

### 2023-11-20 manu-draft.qmd
submitted

@@
       correspondence: false
-  - Maria Hermann:
+  - Maria Herrmann:
       email: mxh367@psu.edu

### 2024-02-22 revision/review_response.qmd
add review response start file

---
format: 
  docx:
    reference-doc: response_styles.docx
bibliography: ../refs.bib

execute: 
  echo: false
  warning: false
  message: false
---

```{r echo = F, cache = F, eval = F}
#| eval: false
spelling::spell_check_files(here::here('revision/review-response.qmd'))
```

### Reviewer 1

This manuscript describes an advancement in the computation of aquatic estuarine metabolism (i.e., primary production and reparation) using open-water dissolved oxygen (DO) time series. The authors adapted a Bayesian model originally developed for non-tidal, lotic systems (i.e., BASE) to 1) now work with time series that have missing dissolved oxygen observations and 2) can accommodate longer time periods for metabolism estimates. The latter could produce non- real estimates of metabolism in earlier methods while the former is achieved through linear interpolation during pre-processing and estimate removal by a user defined threshold during post-processing. The authors describe in sufficient detail how to operationalize the model and provide a comparison with a historically more standard arithmetic-based model that has been used in earlier studies of estuarine metabolism. Furthermore, the authors provide an R package “EBASE” to allow others to estimate estuarine metabolism using this method advancement; a package I am sure the coastal community will find useful. Overall, this paper was well written and the advancement significant. This paper should be considered for publication in L&O Methods following incorporation of the comments below.

*Response: text goes here*

There were two studies cited in the introduction (Line 98) as having used Bayesian techniques to estimate aquatic metabolism in estuaries. Please elaborate on how the method described in this study is consistent with and/or different from those in the cited studies.

*Response: text goes here*

The authors describe pitfalls in their model, particularly when it comes to estimating the gas transfer velocity. Additionally, the authors provide evidence that sometimes the right answer (i.e., modeled DO time series, lines 349-352) can be achieved through incorrect processes. Similar issues have been described and addressed in the stream metabolism literature which was not cited or discussed in this study, please see Appling et al. 2018 - https://doi.org/10.1002/2017JG004140). Consider adding to the discussion how EBASE might incorporate these advancements to address these issues.

*Response: text goes here*

Detiding (i.e., removing the influence of tidal cycles on time series data) the DO time series substantially improved model estimates and therefore was recommended as a step prior to applying the model. However, it was not clear how the authors detided their time series (they used Weighted Regression but it is unclear how this method removes tidal influence from the predicted time series – see lines 400-402). Please add additional information on the detiding approach and which variables were detided (i.e., was only DO detided? Wouldn’t all water-related input variables, such as water temperature and salinity, need to be detided?).

*Response: text goes here*

How would your results be impacted by using different observed data collection frequencies such as every 5-min, 15-min, 1 hour?

*Response: text goes here*

In my use of Bayesian models for estimating metabolism, with each model run you can get different metabolism estimates. This is due to the prior distribution space that the model can operate in. Is this something you have noticed with your model? If so, could you add a sentence somewhere explaining that a user should expect different modeled estimates with each model run?

*Response: text goes here*

Line 32-36: Estuaries are also significantly influenced by their freshwater end-members, consider adding ‘river’ to ‘…interface between land, river, ocean, and atmosphere’. Similarly consider incorporating rivers into ‘…transported from land to rivers to ocean…’.

*Response: text goes here*

Line 47: Consider adding how materials are lost to the atmosphere, the sediments, and the ocean. For example, “…that are lost to the atmosphere (i.e., outgassing), the sediments (i.e., burial), and the ocean (i.e., advection).”.

*Response: text goes here*

Line 48-50: Consider adding the underwater eddy covariance method to the list of methods for estimating aquatic metabolism. See Berg et al. 2022 (https://doi.org/10.1146/annurev-marine- 042121-012329)

*Response: text goes here*

Line 80: As written, it seems ‘assumptions’ should be singular, not plural (…can violate an assumption of the method…).

*Response: text goes here*

Line 82: I appreciate that the sign of P and R estimates in the literature has been applied non- uniformly. However, I believe the convention is that P is positive (DO is produced) while R is negative (DO is consumed). Could you change the physically impossible statement to state negative P and positive R? Or add a sentence about why both values are described with a positive sign – this is somewhat described on line 236.

*Response: text goes here*

Line 96: Add the word statistical between ‘Bayesian’ and ‘techniques’ (i.e., Bayesian statistical techniques…)

*Response: text goes here*

Line 98: These two cited studies – Ciavatta et al. 2008; Tassone and Bukaveckas 2019 - have used Bayesian approaches to estimate estuarine metabolism; what makes them limited? How is the approach described in this manuscript different from these two earlier studies? Have those earlier two studies informed the model selection/development described in this manuscript?

*Response: text goes here*

Line 106: Curious why your method builds off BASE when there has been an updated Bayesian technique for stream metabolism that more accurately estimates uncertainty and reduces equifinality - please see Appling et al. (2018) cited above. Additionally, one of the sources cited above (Tassone and Bukaveckas 2019) used the updated Bayesian technique to estimate single- station estuarine metabolism.

*Response: text goes here*

Line 126: Presumably, EBASE produces daily metabolic estimates. What constitutes a day in EBASE? Is it from midnight to midnight the following day or from 4 am to 4 am?

*Response: text goes here*

Line 129-131: I normally see water column depth represented as Z not H. Should clarify if the values of P, R, and D are positive, negative or both.

*Response: text goes here*

Line 135: Consider making C = Cobs if it represents the observed DO concentration. If that is not what it represents, then will need to be described in text.

*Response: text goes here*

Line 181: Specifying the model optimization period is an advancement of the BASE model but is not described in text. What does ‘optimization’ mean? What are the implications of using 1 day, 7 days, 30 days, or the entire time series length on the output? If you use 7 days, do you only get weekly estimates of metabolism?

*Response: text goes here*

Line 185: On line 546 it is stated that linear interpolation is used, that should be described here. Furthermore, the threshold discussion on line 547-551 should be mentioned here. Lastly, on line 550-551, a 12 hour threshold should be further discussed. Why 12- hours? Why not 6 considering anything longer than 6 and you would be missing half of a semidiurnal tidal cycle.

*Response: text goes here*

Line 188: With your improvements to the original BASE method, could EBASE be used to estimate metabolism in non-tidal, lotic ecosystems?

*Response: text goes here*

Line 248-252: How was the NERRS data accessed? Also be sure to cite the NERRS system-wide monitoring program, see link on NERRS data citation here - https://cdmo.baruch.sc.edu/data/citation.cfm

*Response: text goes here*

Line 267: Consider rephrasing – “…has a structure similar to, yet noisier than, that of light efficiency…”.

*Response: text goes here*

Line 270-271: Wouldn't one expect the opposite? During periods of high productivity, ingassing should occur given the relatively high(er) rates of primary production. Then during periods of low productivity (i.e., outside of the growing season) that outgassing would dominate as respiration of allochthonous material continues to occur while primarily production would be light limited (assuming the example dataset used is in a temperate environment).

*Response: text goes here*

Line 276-278: As pointed to earlier in the manuscript, it is unclear what is meant by the optimization period. Why are the results in Figure 3 averaged by the optimization period? Why not show the daily estimates? I liked seeing the fit of the EBASE estimates on the synthetic time series however, I would like to see the fit of a linear regression of the synthetic (observed) estimates vs. the EBASE (predicted) estimates.

*Response: text goes here*

Line 313: This explanation of optimization needs to come much earlier in the text and be expanded upon.

*Response: text goes here*

Line 324-327: I'm not sure if I follow correctly - are you seeking daily metabolic estimates? If so, are the 7 & 30 day optimizations using a forward looking moving window approach that advances by one-day (similar to a rolling average)?

*Response: text goes here*

Line 473-477: It is stated that “Bayesian…methods have not been rigorously explored for estimating metabolic rates in estuaries” however, I point back to the two cited studies in the introduction (line 98) that did use Bayesian techniques to estimate estuarine metabolism. Consider rewording this sentence and/or expand on why those earlier studies are not considered rigorous.

*Response: text goes here*

Line 581: Maybe I missed it but how many MCMC simulations were used? How would a user change the number of MCMC simulations in the EBASE package? Can you provide a recommendation for the number of simulations to use or at least how many were used in this analysis?

*Response: text goes here*

Figures 3 & 5: These figures have a gray background while the rest are white. Please make the background color of the figures consistent across figures.

*Response: text goes here*

Figure 7: Consider adding points to the lines in the left column plots. You may need to make the thickness of the lines smaller.

*Response: text goes here*

### Reviewer 2

Beck et al. developed a new method for estimating metabolism in estuaries using continuous dissolved oxygen data. Metabolism is a key component of ecosystems, yet it is inherently difficult to estimate, especially in estuaries due to tidal currents. Their method is based on a stream Bayesian approach, and the authors expanded it by adjusting the time interval over which model parameters are estimated. The longer intervals allow for estimation over tidal cycles, which acts as noise in the time series, and allows for improved model estimation. The authors also compared their model to a more conventional approach (i.e., Odum model) and show the model’s improved estimation, specifically avoiding impossible results, such as negative production.

The model is written in the programming language R, making it widely available to the public. The online documentation and vignettes are clear and show the users key features and examples of the package. This is a nice bonus to go along with a well-written manuscript and expand the impact of the products.

I enjoyed reading the manuscript. It is technical where it needs to be, but also provides general context for the problem and the rationale for the technique. I was tripped up during a few of the sections, and I provide some suggestions that might improve the flow of the article. I’m also wondering if this model approach is only applicable to estuaries, or if it could be useful for other aquatic ecosystems? Overall, the authors did a great job. Nice work!

*Response: text goes here*

#### General comments

Table of model parameters. There are several parameters, most are relatively simple (single letter), and they appear on figures. I think it would be valuable to have a table. This could include, name, acronym, units, and any other relevant information (source/equation/etc.). I can also see organizing this table by inputs and outputs (L146) to help future users identify their input data and expected outputs.

*Response: text goes here*

Is EBASE only applicable to estuaries? It seems that the main change from BASE is the inclusion of a wider temporal window to estimate model parameters and changes in the b parameterization. Can these modifications also improve upon metabolism estimates in other systems (lakes, streams, etc.)? While I see the goal of improving metabolism in estuaries, there may be wider applicability to rivers, lakes, ocean, or other coastal areas. If so, mentioning what kinds of systems (big lakes, deep lakes, big rivers, complex hydrodynamics, etc.?) would be most appropriate for the model and potential get improved estimates over other metabolism approaches. This may expand the impact of this paper.

*Response: text goes here*

Model parameter b.

I’m less familiar with this term, and the authors could do a little bit more to explain what it is. On L143, they say that it is “fairly well known for open for open-ocean conditions”, but they never actually define the term. Looking at Eq 2, it seems to relate to “gas transfer efficiency” given that it is a coefficient on the wind-speed/viscosity term. That being said, I’m also a little confused on how Eq 3 is integrated into the model framework. Eq. 3 seems to relate more closely with streams as water depth is included as a scaler for kw. So somehow the two models (one more typical for lentic systems, the other more typical for lotic systems) are integrated together. In theory this makes sense in an estuary as both wind and flow likely affect gas transfer.

Diving into eq 3, higher water levels usually indicate higher flows and thus higher gas transfer velocity (Kw). I’m not exactly sure how temperature plays into this equation, and if it is appropriate for an estuary. It seems to suggest warm periods within a day have higher exchange rates, due to differences in kinetics at the water surface. However warmer temperatures may also reflect increased stratification and less vertical mixing (small lakes). There may be something to clean from changes in temperature that drive changes in turbulence, but simply using warmer = more exchange seems like an oversimplification. Eq 3 also includes a term related to water depth (H). In flowing waters this equation suggests that Kw increases with water level, likely a result of high-water levels = higher velocities = higher turbulence = higher kw. In an estuary, I’m not sure this relationship truly holds as water depth is not as good of a surrogate for velocity. Highest water level may be at slack tide and thus minimal water movements. It could be that change in water depth (assuming the tidal currents are strong relative to the river inputs) would be a better term to reflect instantaneous discharge. Or possibly flow could be included as a possible input parameter. But I realize this term is not typically available.

All told, perhaps the issue with poorly constraining b is related to assumptions in the equations used to model gas exchange in an estuary. The authors do a good job explaining that the model parameter b was difficult to constrain in the example dataset. This is understandable, and my comments are mainly my mental exploration of ways that might improve that aspect of the model. I’m guessing the authors have thought extensively about this problem, and I’m mainly offering ideas. Perhaps the authors could include some other ideas for why the b parameter is difficult to constrain? Or explore additional datasets to understand when/where it can be estimated more accurately? For example, if the model was applied to an artificial dataset from a stream, lake, or other type of estuary does it perform better? If so, what is different about the system, which might help with tuning the model? Would any of these help understand the parameterization/limitation of the current model, and or identify systems it is most appropriate?

*Response: text goes here*

#### Specific comments

L26: which open-source software? If only one (i.e., R), then can list it here.

*Response: text goes here*

L40: Organic matter produced in the estuary has other outcomes. Re-respired, consumed and incorporated into biota, exported to the ocean, etc.

*Response: text goes here*

L83: which computed rates were negative? P, R, or both? L128: mean depth?

*Response: text goes here*

L133: Thanks for providing the units. I am curious about area and time. I’m assuming that area is surface water area and time is the interval of the dissolved oxygen data. If so, perhaps expand on these so the author knows what the temporal and horizontal spatial scales are.

*Response: text goes here*

L134: “R is constant”. I believe R is allowed to vary across days, but R is constant within a day. Please specify.

*Response: text goes here*

L143: what is b? and for what time interval is b constant? This ends up being an output (Figure 3) and it is confusing to say that it is constant. So maybe it is constant within a day? Is this a standard parameter

*Response: text goes here*

for gas exchange? As with the other parameters, providing some type of ‘name’ would be helpful, especially if provided in a table.

*Response: text goes here*

L164: The equation implies that Kw would increase as water level (H) increases. It seems the authors have this statement inversed or talking about measured reaeration (K). I believe this relates to streams where increasing water levels relate to higher water velocities, and thus higher turbulence and higher gas transfer velocity (kw). See General Comments.

*Response: text goes here*

L189: Thanks for putting together the web page. I’m curious if the data in this manuscript are included within the package itself, so it is possible to regenerate the results entirely? I’m also wondering how stable this website is and if any online information that is described in the manuscript will change? I’m not that familiar with these online reference materials, but I’m thinking of a static version (tag) that will stay with the paper so a reader 10 years from now could access the exact same materials that go along with this paper.

*Response: text goes here*

L210: Thank you for providing the units for these parameters. I am wondering if that authors considered using negative exponents to ensure the division symbols are properly addressed. For example, 300 mmol/m2/d. While I understand this to be mmol m-2 d-1, days could be mistakenly assumed to be in the numerator of the equation. While unlikely in this situation, the subsequent equations (L218) get even more complicated “(mmol/m2/d)/(W/m2)”, so it could be worth ensuring all the units are clear. Further, this unit has m2 twice, so they could be removed from the final unit (mmol d-1 W-1).

*Response: text goes here*

L237: The upper limit of 0.502 for b. This makes sense to constrain this parameter. I also noticed that many of the estimates of b are close to this value rather than near the synthetic value for the example dataset (Figure 3). I’m wondering if anything within the model could be improved to handle this issue? See general comments.

*Response: text goes here*

L272: I’m curious about the model estimates for kw. The authors show the results for flux (D), but this will vary with concentrations and k. Wondering if there is anything else to glean from those modelled results?

*Response: text goes here*

L301: The units for R are listed as (mmol/m2/d)/(W/m2). Perhaps the denominator is a typo?

*Response: text goes here*

L541: Wondering if it would be helpful to mention somewhere in this paragraph, “it depends on the question/objective of the project”? For example, if we are interested in annual summations and comparisons among systems, it may be fine to use a longer optimization period. But if we are interested in daily rates, timing of events in relation to storms, floods, etc. we may need to use a shorter optimization period. So the researcher’s question should be considered.

*Response: text goes here*

### Reviewer 3

Beck et al. present a new software package to estimate ecosystem metabolism from time series of dissolved oxygen and accompanying physical variables in estuarine systems. Accurately estimating metabolic rates are critical to understanding impacts on estuaries and their role in material cycling. The work is highly novel as no such statistical approach is currently available for estuaries. The software adapts a package built for estimating metabolism in stream ecosystems using a Bayesian framework, which is highly suited to this task due to the ability to place prior distributions on physical parameters with known physical limitations. I found the paper to be very clear and the choices made in model development, setting of priors, and model validation to be robust and well documented. The improvements made to the model, such as the option to vary the optimization period, are valuable and were demonstrated to be important to accurately estimating rates (and mirror suggestions for dealing with inaccurate K in streams; e.g. Appling et al. 2018 https://dx.doi.org/10.1002/2017JG004140)

In particular, the focus on the importance of accurately capturing the physical variables is a strength of the paper. It is made clear from the outcomes of the synthetic dataset analysis that it is better to use a literature estimate of the parameter b, a constant in the equation that governs air-water exchange, then to let the model estimate inaccurate values for b. The discussion around this issue was appropriate, but I wonder if some more very clear advice could be provided to a potential user that wants to use this package to answer questions about ecosystem metabolism in estuaries but does not have the aim of further refining the modelling approaches. Could they use the estimate from Wanninkhof (2014) as was used in subsequent analyses here? Would an appropriate sensitivity analysis be to vary this estimate by 20% (as mentioned L224) to check if their P and R results are highly sensitive to this parameter? Or is further testing in a range of estuaries required before this can be applied?

*Response: text goes here*

#### Specific comments:

L14: The first use of “unphysical” in the abstract was a little unclear without further detail. Consider a short phrase like “may lead to estimates that are physically impossible” be used to define this first.

*Response: text goes here*

L122: The sentence “We also attempt to constrain the gas transfer velocity using this approach but are unable, at least with the data sets used in this study.“ was unclear. Consider replacing “this approach” with a more precise statement (I assume “this approach” is the Bayesian framework, rather than a reference to the previous sentence on the test with the detided inputs?). I’m also not sure if “constrain” is the right word, as the issue was accurately estimating this parameter.

*Response: text goes here*

L159 & L514: The air-water exchange isn’t depth dependent in BASE. The inputs (and metabolic results) in base are all volumetric and reaeration coefficient (K) is in per day. Depth is not provided to the model but can be used to convert the metabolic estimates afterwards.

*Response: text goes here*

L258: typo in “estimatd”

*Response: text goes here*

L269: Should it be “availability” rather than “variability”? of course, a more variable pool of organic matter could also be correct.

*Response: text goes here*

L352: The “right answer for wrong reasons” is interesting. It could be clearer here that even though DO is estimated accurately, the P and R estimates will be inaccurate in this scenario.-7

*Response: text goes here*

### 2024-02-22 revision/review_response.qmd
started revisions, readded input/output table

@@
 
-*Response: text goes here*
+*Response: Corrected.*
 
@@
 
-*Response: text goes here*
+*Response: Added.*
 
@@
 
-*Response: text goes here*
+*Response: Yes, C represents observed dissolved oxygen concentration (as noted on line 132).
+We have retained C instead of using Cobs for simplicity in the model equations.
+Further, the addition of the model input/output table as requested by another review helps clarify data.*
 
@@
 
-*Response: text goes here*
+*Response: The NERRS data were accessed through the CDMO, citation was added.*
 
@@
 
-*Response: text goes here*
+*Response: Rephrased.*
 
@@
 
-*Response: text goes here*
+*Response: Figures 3 & 5 were changed to white background.*
 
@@
 
-*Response: text goes here*
+*Response: Points were added, line thickness reduced in the left column plots.*
 
@@
 
-*Response: text goes here*
+*Response: We have added a new table showing model inputs/outputs, parameter names, acronyms, and units.*
 
@@
 So maybe it is constant within a day?
-Is this a standard parameter
-
-*Response: text goes here*
-
-for gas exchange? As with the other parameters, providing some type of ‘name’ would be helpful, especially if provided in a table.
+Is this a standard parameter for gas exchange?
+As with the other parameters, providing some type of ‘name’ would be helpful, especially if provided in a table.
 
@@
 
-*Response: text goes here*
+*Response: Yes, this was a typo, corrected.*
 

### 2024-02-22 manu-draft.qmd
started revisions, readded input/output table

@@
 Thus, the approach can be broadly applied where data are available, although it is not without limitations.
-In particular, tidal advection can violate assumptions of the method that the measurements are from the same parcel of water with a continuous metabolic history [@kemp1980].
+In particular, tidal advection can violate an assumption of the method that the measurements are from the same parcel of water with a continuous metabolic history [@kemp1980].
 Striking evidence of this limitation is the large number of physically impossible negative P and R estimates in the open-water analysis of 28 3-year time series by @caffrey2003.
@@
 A unified model to explain the large variability of metabolic rates within and across estuaries is clearly lacking as a critical tool for understanding metabolic processes.
-Bayesian techniques have only been explored in limited estuarine applications despite their potential to address the above [@ciavatta2008; @tassone2019].
+Bayesian statistical techniques have only been explored in limited estuarine applications despite their potential to address the above [@ciavatta2008; @tassone2019].
 Bayesian techniques allow for incorporation of prior knowledge about model parameters and robust estimation of parameter probability distributions [@Holtgrieve10] using highly efficient computational algorithms [@Grace15; @winslow2016].
@@
 The $b$ parameter is fairly well known for open-ocean conditions, but may be different in estuaries, where the transfer velocity is influenced by factors other than wind speed and $Sc$.
-As such, the three parameters estimated by EBASE are $a$, $R$, and $b$ using the required input data that include time series of dissolved oxygen, water temperature, salinity, PAR, wind speed, and water column depth.
+As such, the three parameters estimated by EBASE are $a$, $R$, and $b$ using the required input data that include time series of dissolved oxygen, water temperature, salinity, PAR, wind speed, and water column depth (@tbl-ebaseiotab).
 
@@
 A JAGS model file is included with the package that implements @eq-ebase, with options to supply a custom model file as needed.
+All parameter inputs and outputs for EBASE are shown in @tbl-ebaseiotab.
 
@@
 This assessment evaluated if (1) EBASE estimates similar values for the known parameters used to create the synthetic time series, and (2) how the estimated values change with different model configurations.
-The synthetic time series used inputs from one year (2021) of continuous data from the NERRS reserve in Apalachicola Bay, Florida.
+The synthetic time series used inputs from one year (2021) of continuous data from the NERRS reserve in Apalachicola Bay, Florida [@nerr2023].
 Water temperature and salinity were from the Cat Point water quality monitoring station (29.7021$^{\circ}$ N, 84.8802$^{\circ}$ W) and the meteorological observations were from the nearby East Bay station (29.7694$^{\circ}$ N, 84.8815$^{\circ}$ W, \~10 km north of Cat Point).
@@
 The light efficiency has a complex structure, with maxima in the summer, fall, and winter, and a strong minimum in spring (Figure [-@fig-synapa]a).
-The oxygen budget terms (Figure [-@fig-synapa]b) reveal that (1) photosynthesis has a structure similar to that of light efficiency, but noisier, reflecting the synoptic-scale variability of $PAR$; (2) respiration has the expected structure of being greatest during the summer [@caffrey2004], stimulated by greater autochthonous organic matter variability and high temperature; and (3) the air--water gas flux mimic productivity on short time scales, with outgassing during high productivity and ingassing during low productivity.
+The oxygen budget terms (Figure [-@fig-synapa]b) reveal that (1) photosynthesis has a structure similar to, yet noisier than, that of light efficiency, reflecting the synoptic-scale variability of $PAR$; (2) respiration has the expected structure of being greatest during the summer [@caffrey2004], stimulated by greater autochthonous organic matter variability and high temperature; and (3) the air--water gas flux mimic productivity on short time scales, with outgassing during high productivity and ingassing during low productivity.
 The synthetic dissolved oxygen time series shows high variability at multiple time scales (Figure [-@fig-synapa]c).
@@
 Low and high values for the means and standard deviations for $R$ were assessed following reasonable ranges from @caffrey2004 using logic similar to that for choosing the default priors.
-The minimum and maximum mean values for $R$ were evaluated as 138 and 1009 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$, which we rounded to $1/2$ and 3 times the mean (150 and 900 $\text{mmol}/\text{m}^2/\text{d}$) and proportionally for $a$ as 2.0 and 12 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$.
+The minimum and maximum mean values for $R$ were evaluated as 138 and 1009 $\text{mmol}/\text{m}^2/\text{d}$, which we rounded to $1/2$ and 3 times the mean (150 and 900 $\text{mmol}/\text{m}^2/\text{d}$) and proportionally for $a$ as 2.0 and 12 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$.
 The minimum and maximum values for the standard deviations for $R$ and $a$ were chosen as 10% and 1000% of the default mean values to test a range from high precision to completely uninformed values.
@@
 ```{r}
+#| label: tbl-ebaseiotab
+#| tbl-cap: "EBASE model parameters showing the required input, parameters derived from the model input, and model output."
+ebaseiotab
+```
+
+{{< pagebreak >}}
+
+```{r}
 #| label: tbl-apacmptab

### 2024-02-22 R/funcs.R

@@ -421,7 +421,7 @@ defplo <- function(apadef, fwdatcmp){
       axis.text.y = element_text(size = 10),
       strip.text = element_text(size = rel(1.4)), 
       panel.grid.minor = element_blank(), 
-      panel.background = element_rect(fill = 'gray97', color = NA),
+      # panel.background = element_rect(fill = 'gray97', color = NA),
       plot.margin = unit(c(0,0,0,0), "cm")
     ) + 
     guides(color = guide_legend(override.aes = list(linetype = c(NA, 1), shape = c(16, NA)))) +
@@ -523,7 +523,7 @@ optex <- function(apagrd, fwdatcmp, apasumdat, rnkmetsum, met, lims = NULL){
             axis.text.y = element_text(size = 10),
             strip.text.x = element_text(size = rel(1.4)), 
             panel.grid.minor = element_blank(), 
-            panel.background = element_rect(fill = 'gray97', color = NA),
+            # panel.background = element_rect(fill = 'gray97', color = NA),
             plot.margin = unit(c(0,0,0,0), "cm")
           ) + 
           labs(
@@ -632,10 +632,12 @@ syncomp_plo <- function(resobs, resnos, fwdatcmp){
       ), 
       name = factor(name, levels = c('Synthetic', 'EBASE recovered', 'EBASE recovered (+ noise)'))
     )
-  
+
   p1 <- ggplot(toplo1, aes(x = DateTimeStamp, y = val, color = name)) + 
-    geom_line(linewidth = 0.8) +
+    geom_line(linewidth = 0.5) +
+    geom_point(data = subset(toplo1, name != 'Synthetic')) + 
     facet_wrap(~var, scales = 'free', ncol = 1, strip.position = 'left', labeller = label_parsed) + 
+    guides(colour = guide_legend(override.aes = list(pch = c(NA, 16, 16)))) +
     theme_minimal(base_size = 12) + 
     theme(
       strip.placement = 'outside', 

### 2024-02-22 R/tabs.R

@@ -1,8 +1,33 @@
+library(flextable)
 library(tidyverse)
 library(lubridate)
 library(here)
 
 source(file = here('R/funcs.R'))
+
+# EBASE i/o -----------------------------------------------------------------------------------
+
+totab <- tibble(
+  Type = c('Input', 'Input', 'Input', 'Input', 'Input', 'Input',
+           'Input-derived', 'Input-derived', 'Input-derived',
+           'Output', 'Output', 'Output', 'Output', 'Output', 'Output'),
+  Description = c('Dissolved oxygen (observed)', 'Water temperature', 'Salinity', 'Total photosynthetically active radiation', 'Wind speed', 'Water column depth',
+                  'Wind speed at 10 meter height, squared', 'Schmidt number (from water temperature and salinity)', 'Dissolved oxygen at saturation (from water temperature and salinity)',
+                  'Dissolved oxygen (modelled)', 'Production', 'Respiration', 'Gas exchange', 'Light efficiency', 'Sensitivity of gas transfer to wind speed'),
+  `Model notation` =  c('C', '-', '-', 'PAR', '-', 'H', 
+                        'U$_{10}^2$', 'Sc', 'C$_{sat}$', 
+                        'C_mod', 'P', 'R',  'D', 'a', 'b'),
+  Units = c('mmol m$^{-3}$', '$^\\circ$C', 'psu', 'W m$^{-2}$', 'm s$^{-1}$', 'm', 
+            'm$^2$ s$^{-2}$', 'unitless', 'mmol m$^{-3}$',
+            'mmol m$^{-3}$', 'mmol m$^2$ d$^{-1}$', 'mmol m$^2$ d$^{-1}$', 'mmol m$^2$ d$^{-1}$', '(mmol m$^2$ d$^{-1}$)/(W m$^{-2}$)', '(cm hr$^{-1}$)/(m$^2$ s$^{-2}$)')
+  ) |> 
+  mutate(
+    Type = ifelse(duplicated(Type), '', Type)
+  )
+
+ebaseiotab <- knitr::kable(totab)
+
+save(ebaseiotab, file = here('tabs/ebaseiotab.RData'))
 
 # comparison of EBASE to other methods with actual apa data -----------------------------------
 

### 2024-02-22 R/tabs.R
save file

@@ -16,7 +16,7 @@ totab <- tibble(
                   'Dissolved oxygen (modelled)', 'Production', 'Respiration', 'Gas exchange', 'Light efficiency', 'Sensitivity of gas transfer to wind speed'),
   `Model notation` =  c('C', '-', '-', 'PAR', '-', 'H', 
                         'U$_{10}^2$', 'Sc', 'C$_{sat}$', 
-                        'C_mod', 'P', 'R',  'D', 'a', 'b'),
+                        'C$_{mod}$', 'P', 'R',  'D', 'a', 'b'),
   Units = c('mmol m$^{-3}$', '$^\\circ$C', 'psu', 'W m$^{-2}$', 'm s$^{-1}$', 'm', 
             'm$^2$ s$^{-2}$', 'unitless', 'mmol m$^{-3}$',
             'mmol m$^{-3}$', 'mmol m$^2$ d$^{-1}$', 'mmol m$^2$ d$^{-1}$', 'mmol m$^2$ d$^{-1}$', '(mmol m$^2$ d$^{-1}$)/(W m$^{-2}$)', '(cm hr$^{-1}$)/(m$^2$ s$^{-2}$)')

### 2024-02-27 revision/review_response.qmd
more revisions

@@
 ```
+
+*We thank all three reviewers for providing constructive comments that have helped to improve the manuscript.
+Our responses to each of these comments are provided below.
+Note that line number refer to those in the original manuscript.*
 
@@
 
-*Response: text goes here*
+*Response: Added.*
 
-Line 48-50: Consider adding the underwater eddy covariance method to the list of methods for estimating aquatic metabolism. See Berg et al. 2022 (https://doi.org/10.1146/annurev-marine- 042121-012329)
+Line 48-50: Consider adding the underwater eddy covariance method to the list of methods for estimating aquatic metabolism. See Berg et al. 2022 (https://doi.org/10.1146/annurev-marine-042121-012329)
 
-*Response: text goes here*
+*Response: Added to the list, including the citation, "...use of inert gases, and aquatic eddy covariance (Kemp and Testa 2011; Staehr et al. 2011; Berg et al. 2022).*
 
@@
 
-*Response: text goes here*
+*Response: Yes, we agree the sign has not been consistently described in the literature.
+We have added clarifying text at the end of the paragraph: "Note that R is often expressed as a negative value indicating consumption of dissolved oxygen.
+Here and throughout, we express correct values of R as positive values given the metabolic model described below."*
 
@@
 
-*Response: text goes here*
+*Response: The "metabolic" day in EBASE is midnight to midnight the following day.
+Clarifying text was added to line 182.
+In our experience, very similar estimates are obtained whether the day is defined midnight to midnight or some other period.
+This is especially true if the optimization period is greater than one day, as suggested in model application, i.e., lines 317 - 320, 523-525.*
 
@@
 
-*Response: text goes here*
+*Response: The water column depth notation was changed to Z (including the documentation and code in EBASE).
+Clarification was added for P, R, and D on line 132: "Both $P$ and $R$ are expressed as positive values, whereas D can be positive (outgassing) or negative (ingassing)."*
 
@@
 We have retained C instead of using Cobs for simplicity in the model equations.
-Further, the addition of the model input/output table as requested by another review helps clarify data.*
+Further, the addition of the model input/output table as requested by another reviewer helps clarify the input data.*
 
@@
 
-*Response: text goes here*
+*Response: This was clarified in the text.*
+
+*Line 181:  "First, users can specify the model optimization period as the amount of time over which the unknown parameters are estimated from the Bayesian prior distributions (i.e., $a$, $R$, and $b$, see below)."*
+
+*Line 184: "Note that only one estimate for each of $a$, $R$, and $b$ is returned for each optimization period as defined by the model, although $P$ and $D$ can vary continuously with each timestep given the dependence on other inputs that can vary across the time series (e.g., $PAR$ and wind)."*
 
@@
 
-*Response: text goes here*
+*Response: 12 hours was chosen as a default value to cover approximately half of a diurnal cycle.
+Users may want to consider smaller gaps if tidal cycles are having a large influence on the observed time series.
+Our default value was chosen under the assumption that tidal effects, although they may be present, are likely not the dominant signal in the data.
+If the latter is expected, detiding is strongly encouraged.
+We have added text to this effect.*
+
+*Line 186: "All gaps are linearly interpolated to allow the model estimation to proceed continuously and any results for an optimization period with gaps larger than 12 hours are removed, eliminating the need to pre- and post-process the data when using EBASE.
+The user can vary the criteria for removing results (e.g., 6 hours as a more conservative criteria)."*
+
+*Line 555: "More conservative gaps can be chosen if, for example, semidiurnal tidal signals operating at time scales less than 12 hours are expected to influence the results.
+However, detiding is strongly encouraged in these cases and the default gap size of 12 hours that describes the diurnal oxygen signal is likely appropriate after detiding."*
+
+*Also note that data for an entire optimization period are removed if any gaps greater than the chosen gap size are present, providing an additional safety measure that the results are not based on inaccurately interpolated inputs.*
 

### 2024-02-27 manu-draft.qmd
more revisions

@@
 ```
-
 *Running head*: EBASE metabolic model
@@
 
-Estuaries play a key role in the global cycling of elements because of high rates of biogeochemical activity at the interface between land, ocean, and atmosphere.
-As materials are transported from land to ocean, estuaries profoundly transform or filter these materials through various processes, chief among which are primary production and respiration, collectively referred to as metabolism [@schubel1984].
+Estuaries play a key role in the global cycling of elements because of high rates of biogeochemical activity at the interface between land, river, ocean, and atmosphere.
+As materials are transported from land to rivers to ocean, estuaries profoundly transform or filter these materials through various processes, chief among which are primary production and respiration, collectively referred to as metabolism [@schubel1984].
 For example, estuarine respiration of organic carbon delivered from land produces carbon dioxide, which may be lost to the atmosphere via outgassing [e.g, @cai2011].
@@
 Estuarine metabolic processes are fundamentally important, as illustrated by the dissolved inorganic carbon budget of eastern North American estuaries [@najjar2018]; net primary production and heterotrophic respiration far exceed inputs from upland sources and outputs to the atmosphere and ocean.
-As such, small changes in either net primary production or heterotrophic respiration can have a large impact on the difference between the two, net ecosystem metabolism (NEM, sometimes also referred to as net ecosystem production), which ultimately governs the form and quantity of estuarine materials that are lost to the atmosphere, the sediments, and the ocean.
+As such, small changes in either net primary production or heterotrophic respiration can have a large impact on the difference between the two, net ecosystem metabolism (NEM, sometimes also referred to as net ecosystem production), which ultimately governs the form and quantity of estuarine materials that are lost to the atmosphere (i.e., outgassing), the sediments (i.e., burial), and the ocean (i.e., advection).
 
-Methods for estimating metabolism in aquatic environments are numerous and include open-water techniques, bottle-based incubations, ecosystem budgets, oxygen isotopes, and use of inert gases [@kemp2011; @staehr2011].
+Methods for estimating metabolism in aquatic environments are numerous and include open-water techniques, bottle-based incubations, ecosystem budgets, oxygen isotopes, use of inert gases, and aquatic eddy covariance [@kemp2011; @staehr2011; @berg2022].
 Based on the variety of assumptions and limitations of each method, comparisons of metabolic rates within and between ecosystems have been challenging.
@@
 In particular, tidal advection can violate an assumption of the method that the measurements are from the same parcel of water with a continuous metabolic history [@kemp1980].
-Striking evidence of this limitation is the large number of physically impossible negative P and R estimates in the open-water analysis of 28 3-year time series by @caffrey2003.
+Striking evidence of this limitation is the large number of physically impossible negative P and R estimates in the open-water analysis of 28 3-year time series by [@caffrey2003].
 The computed rates were negative 23% of the time, on average, and as high as 69% for individual sites.
+Note that R is often expressed as a negative value indicating consumption of dissolved oxygen.
+Here and throughout, we express correct values of R as positive values given the metabolic model described below.
 
@@
 
-EBASE uses a mass balance equation for dissolved oxygen, assuming a well-mixed water column of depth $H$ and no lateral transport processes:
+EBASE uses a mass balance equation for dissolved oxygen, assuming a well-mixed water column of depth $Z$ and no lateral transport processes:
 
 $$
-H\frac{dC}{dt} = P - R - D
+Z\frac{dC}{dt} = P - R - D
 $$ {#eq-ebase}
@@
 where the terms on the right side of the equation are gross primary production ($P$), ecosystem respiration ($R$), and the net upward diffusive gas flux at the air--water interface ($D$), respectively, processes that result in the change in dissolved oxygen concentration ($dC$) per unit time ($dt$).
+Both $P$ and $R$ are expressed as positive values, whereas $D$ can be positive (outgassing) or negative (ingassing).
 The equation has dimensions of moles per unit area and time.
@@
 $$
-\frac{k_w}{H} = K \cdot 1.024^{\left(T - \bar{T}\right)}
+\frac{k_w}{Z} = K \cdot 1.024^{\left(T - \bar{T}\right)}
 $$ {#eq-gasexbase}
@@
 EBASE differs from BASE in model implementation in two ways.
-First, users can specify the model optimization period, where the time period can vary from a minimum of 1 day to the maximum of the entire length of the time series.
+First, users can specify the model optimization period as the amount of time over which the unknown parameters are estimated from the Bayesian prior distributions (i.e., $a$, $R$, and $b$, see below).
+The time period can vary from a minimum of 1 day (defined as midnight to midnight the following day) to the maximum of the entire length of the time series.
 This differs from the BASE approach where the optimization period is set at 1 day.
 Sensitivity of the EBASE method to different optimization periods is described below.
+Note that only one estimate for each of $a$, $R$, and $b$ is returned for each optimization period as defined by the model, although $P$ and $D$ can vary continuously with each timestep given the dependence on other inputs that can vary across the time series (e.g., $PAR$ and wind).
 Second, the EBASE software can accommodate missing observations in the input data by interpolating gaps prior to estimating the metabolic parameters.
-Results obtained from interpolated data are automatically removed from the output, eliminating the need to pre- and post-process the data when using EBASE.
+All gaps are linearly interpolated to allow the model estimation to proceed continuously and any results for an optimization period with gaps larger than 12 hours are removed, eliminating the need to pre- and post-process the data when using EBASE.
+The user can vary the criteria for removing results (e.g., 6 hours as a more conservative criteria).
 Again, this feature was not included in the original BASE method and a complete time series was required for use.
@@
 This feature allows EBASE to be applied more easily to other data sets because an analyst does not need to manually screen data that include missing values prior to analysis.
+More conservative gaps can be chosen if, for example, semidiurnal tidal signals operating at time scales less than 12 hours are expected to influence the results.
+However, detiding is strongly encouraged in these cases and the default gap size of 12 hours that describes the diurnal oxygen signal is likely appropriate after detiding.
 

### 2024-02-27 R/tabs.R

@@ -14,7 +14,7 @@ totab <- tibble(
   Description = c('Dissolved oxygen (observed)', 'Water temperature', 'Salinity', 'Total photosynthetically active radiation', 'Wind speed', 'Water column depth',
                   'Wind speed at 10 meter height, squared', 'Schmidt number (from water temperature and salinity)', 'Dissolved oxygen at saturation (from water temperature and salinity)',
                   'Dissolved oxygen (modelled)', 'Production', 'Respiration', 'Gas exchange', 'Light efficiency', 'Sensitivity of gas transfer to wind speed'),
-  `Model notation` =  c('C', '-', '-', 'PAR', '-', 'H', 
+  `Model notation` =  c('C', '-', '-', 'PAR', '-', 'Z', 
                         'U$_{10}^2$', 'Sc', 'C$_{sat}$', 
                         'C$_{mod}$', 'P', 'R',  'D', 'a', 'b'),
   Units = c('mmol m$^{-3}$', '$^\\circ$C', 'psu', 'W m$^{-2}$', 'm s$^{-1}$', 'm', 

### 2024-02-28 revision/review_response.qmd
more revisions

@@
 bibliography: ../refs.bib
+csl: ../limnology-and-oceanography.csl
 
@@
 
-*We thank all three reviewers for providing constructive comments that have helped to improve the manuscript.
+*We thank all three reviewers for providing thorough and constructive comments that have helped to improve the manuscript.
 Our responses to each of these comments are provided below.
-Note that line number refer to those in the original manuscript.*
+Note that line numbers refer to those in the original manuscript.*
 
@@
 
-*Response: text goes here*
+*Response: The detiding approach used herein is thoroughly described in another L&O paper cited in the main text [@Beck15].
+Describing this approach in detail is beyond the scope of the paper, although we have added some text for clarification.
+Also, the method has only been applied to dissolved oxygen data, so we have not used it for any of the remaining parameters.
+Clarification for both points has been added to line 399: "The results of this analysis return a predicted and detided dissolved oxygen time series, where the detided results are simply the predictions at average tidal height [see @Beck15 for details].
+This method was only applied to dissolved oxygen since it has not been evaluated with any other parameters."*
 
@@
 
-*Response: text goes here*
+*Response: The resolution of "continuous" monitoring data has not been evaluated in this study, although our guess is that more frequent sampling may produce more robust estimates.
+However, this may not matter given the time scales at which metabolism changes (i.e., > 1 day).
+In our experience, estimates from the Odum open-water method are nearly identical using 15-minute or hourly samples.
+Not sure the same would apply for EBASE given the fundamental differences between the approaches.
+We have added a point in the discussion on line 541 that speaks to this question: "These hypotheses may also consider frequency of continuous monitoring (e.g., 15 minute vs hourly) and how this may influence the results, as this is currently unknown for EBASE."*
 
@@
 
-*Response: text goes here*
+*Response: Surprisingly, we have not encountered this issue with EBASE.
+Obviously the results depend on the prior distributions, which was the purpose of our sensitivity analysis, although we have not noticed slightly different results for different runs with the same priors.
+It is possible these differences are present, but small enough not to notice.
+We would hope that a user of this approach would be familiar enough with Bayesian and MCMC methods to expect these minor differences.*
 
@@
 
-*Line 181:  "First, users can specify the model optimization period as the amount of time over which the unknown parameters are estimated from the Bayesian prior distributions (i.e., $a$, $R$, and $b$, see below)."*
-
-*Line 184: "Note that only one estimate for each of $a$, $R$, and $b$ is returned for each optimization period as defined by the model, although $P$ and $D$ can vary continuously with each timestep given the dependence on other inputs that can vary across the time series (e.g., $PAR$ and wind)."*
+*Line 181: "First, users can specify the model optimization period as the amount of time over which the unknown parameters are estimated from the Bayesian prior distributions (i.e., $a$, $R$, and $b$).
+The optimization period determines how much data are used to estimate each parameter.
+Only one estimate for each of $a$, $R$, and $b$ is returned for each optimization period as defined by the model, although $P$ and $D$ can vary continuously given the dependence on other inputs that can vary at each time step (e.g., $PAR$ and wind).
+The optimization period can vary from a minimum of 1 day (defined as midnight to midnight the following day) to a maximum of the entire length of the time series.
+This differs from the BASE approach where the optimization period is set at 1 day.
+Sensitivity of the EBASE method to different optimization periods is described below."*
 
@@
 
-*Response: text goes here*
+*Response: In theory, EBASE could be used in non-tidal, lotic systems, since the primary difference with BASE is the use of a wind- and temperature-based gas exchange parameterization appropriate for open-water systems [@Wanninkhof14].
+However, additional comparisons are needed (e.g., Odum vs EBASE in a lake) and our thorough justification for the default prior distributions are based on the estuarine literature.
+This is still a valuable poin and we have added text to line 519:*
+
+*"Further, EBASE could be applied to other non-tidal, lotic systems, although additional comparisons with existing methods are needed and the default uninformed prior distributions should be chosen carefully given the justification herein based primarily on estuarine literature [e.g., @caffrey2004]."*
 
@@
 
-*Response: text goes here*
+*Response: Yes, this is correct and can even be seen in Figure 2, e.g., late September to early October shows P > R with negative D (ingassing), then P < R with positive D (outgassing).
+We have corrected this error in the text.*
 
@@
 
-*Response: text goes here*
+*Response: Please see the response to line 181.
+In particular, our revision notes that only one value each for $a$, $R$, and $b$ is returned for each optimization period.
+The vertical lines for each point in Figure 3 are credible intervals showing uncertainy around each individual estimate.
+The statement in the caption that "the results are averaged at the time step of the model optimization period" is incorrect and removed.
+This really only applies to the synthetic results in Figure 5 to show a more clear comparison with the recovered values from EBASE.*
 
@@
 
-*Response: text goes here*
+*Response: Agreed, please see the response to line 181. This section was also slightly shortened in response to the changes on line 181.*
 
@@
 
-*Response: text goes here*
+*Response: The text was edited for clarity: "Results from optimization periods of 1, 7, and 30 days were compared to the synthetic one-year time series ($b$ was fixed for the sensitivity analysis) based on daily averages for both the synthetic and EBASE results.
+For the latter, parameters that are only returned as individual values for each optimization period ($a$, $R$, and $b$) were assigned as daily values for comparability of summary statistics where $n$ is the same across analyses."*
 
@@
 
-*Response: text goes here*
+*Response: This detail was omitted and we have now included it on line 199 with the description of the MCMC approach (default 10,000 simulations).*
 
@@
 
-*Response: text goes here*
+*Response: Yes, it is possible with caveats.
+Please see our response to reviewer on line 188.
+Specifically, additional comparisons are needed and the default uninformed priors based on estuarine literature must be reconsidered.*
 
@@
 
-*Response: text goes here*
+*Response: Fixed.*
 
@@
 
-*Response: text goes here*
+*Response: Changed to "availability", this seems more correct.*
 
@@
 *Response: text goes here*
+
+## References

### 2024-02-28 manu-draft.qmd
more revisions

@@
 EBASE differs from BASE in model implementation in two ways.
-First, users can specify the model optimization period as the amount of time over which the unknown parameters are estimated from the Bayesian prior distributions (i.e., $a$, $R$, and $b$, see below).
-The time period can vary from a minimum of 1 day (defined as midnight to midnight the following day) to the maximum of the entire length of the time series.
+First, users can specify the model optimization period as the amount of time over which the unknown parameters are estimated from the Bayesian prior distributions (i.e., $a$, $R$, and $b$).
+The optimization period determines how much data are used to estimate each parameter.
+Only one estimate for each of $a$, $R$, and $b$ is returned for each optimization period as defined by the model, although $P$ and $D$ can vary continuously given the dependence on other inputs that can vary at each time step (e.g., $PAR$ and wind).
+The optimization period can vary from a minimum of 1 day (defined as midnight to midnight the following day) to a maximum of the entire length of the time series.
 This differs from the BASE approach where the optimization period is set at 1 day.
 Sensitivity of the EBASE method to different optimization periods is described below.
-Note that only one estimate for each of $a$, $R$, and $b$ is returned for each optimization period as defined by the model, although $P$ and $D$ can vary continuously with each timestep given the dependence on other inputs that can vary across the time series (e.g., $PAR$ and wind).
 Second, the EBASE software can accommodate missing observations in the input data by interpolating gaps prior to estimating the metabolic parameters.
@@
 Again, this feature was not included in the original BASE method and a complete time series was required for use.
-A detailed web page (<https://fawda123.github.io/EBASE/>) describes how to use the package, with explanations of the functions for viewing model results, including a plot of the modeled dissolved oxygen with the observed, a time series plot of the oxygen budget terms (P, R, and D), and a time series plot of the credible intervals for the $a$, $R$, and $b$ parameters.
+A detailed web page ([https://fawda123.github.io/EBASE/](#0){.uri}) describes how to use the package, with explanations of the functions for viewing model results, including a plot of the modeled dissolved oxygen with the observed, a time series plot of the oxygen budget terms (P, R, and D), and a time series plot of the credible intervals for the $a$, $R$, and $b$ parameters.
 All results provided herein were generated using the EBASE R package.
@@
 The parameters $a$, $R$, and $b$ are estimated by likelihood given the observed data and prior distributions for the parameters.
-The "Just Another Gibbs Sampler" (JAGS) software [as for BASE, @Plummer03] is used with the EBASE R package (described above) to estimate probability distributions of the unknown parameters using Markov Chain Monte Carlo (MCMC) simulations in a Bayesian framework applied to a finite-difference form of @eq-ebase.
+The "Just Another Gibbs Sampler" (JAGS) software [as for BASE, @Plummer03] is used with the EBASE R package (described above) to estimate probability distributions of the unknown parameters using Markov Chain Monte Carlo (MCMC) simulations (default 10,000) in a Bayesian framework applied to a finite-difference form of @eq-ebase.
 A JAGS model file is included with the package that implements @eq-ebase, with options to supply a custom model file as needed.
@@
 The $b$ parameter was fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ [@Wanninkhof14].
-Daily values of $a$ and $R$ were estimatd by (1) applying the Odum open-water method as implemented in the WtRegDO R package [@Beck23a], (2) computing monthly averages of the resulting daily estimates of $P$ and $R$, (3) computing the monthly average of the parameter $a$ by dividing the monthly average of $P$ by the monthly average of $PAR$, and (4) by fitting a spline function to the monthly averages of $R$ and $a$.
+Daily values of $a$ and $R$ were estimated by (1) applying the Odum open-water method as implemented in the WtRegDO R package [@Beck23a], (2) computing monthly averages of the resulting daily estimates of $P$ and $R$, (3) computing the monthly average of the parameter $a$ by dividing the monthly average of $P$ by the monthly average of $PAR$, and (4) by fitting a spline function to the monthly averages of $R$ and $a$.
 
@@
 The light efficiency has a complex structure, with maxima in the summer, fall, and winter, and a strong minimum in spring (Figure [-@fig-synapa]a).
-The oxygen budget terms (Figure [-@fig-synapa]b) reveal that (1) photosynthesis has a structure similar to, yet noisier than, that of light efficiency, reflecting the synoptic-scale variability of $PAR$; (2) respiration has the expected structure of being greatest during the summer [@caffrey2004], stimulated by greater autochthonous organic matter variability and high temperature; and (3) the air--water gas flux mimic productivity on short time scales, with outgassing during high productivity and ingassing during low productivity.
+The oxygen budget terms (Figure [-@fig-synapa]b) reveal that (1) photosynthesis has a structure similar to, yet noisier than, that of light efficiency, reflecting the synoptic-scale variability of $PAR$; (2) respiration has the expected structure of being greatest during the summer [@caffrey2004], stimulated by greater autochthonous organic matter availability and high temperature; and (3) the air--water gas flux mimic productivity on short time scales, with ingassing during high productivity and outgassing during low productivity.
 The synthetic dissolved oxygen time series shows high variability at multiple time scales (Figure [-@fig-synapa]c).
@@
 
-The optimization period used for a time series determines how much data are used to estimate the $a$ and $R$ parameters.
-Unlike BASE, the EBASE software allows flexibility in the period of time used for optimization, with a minimum being 1 day and a maximum being the length of the time series.
-For a chosen optimization period, single estimates for $a$ and $R$ are returned by the model ($b$ is also returned, but was fixed for the sensitivity analysis).
+The optimization period used for a time series determines how much data are used to estimate the metabolic parameters.
 A short period of optimization (e.g., 1 day) may return unstable parameter estimates if forcing factors (e.g., wind, temperature) that affect estuarine biogeochemical rates occur at time scales longer the chosen period of optimization for EBASE.
 As such, potentially longer optimization periods may return more stable and robust estimates given that more data are used for parameter estimation, although at the risk of overgeneralizing metabolic rates if biogeochemical variation occurs at time scales less than the optimization period.
-Results from optimization periods of 1, 7, and 30 days were compared for the synthetic one-year time series.
+Results from optimization periods of 1, 7, and 30 days were compared to the synthetic one-year time series (\$b\$ was fixed for the sensitivity analysis) based on daily averages for both the synthetic and EBASE results.
+For the latter, parameters that are only returned as individual values for each optimization period (\$a\$, \$R\$, and \$b\$) were assigned as daily values for comparability of summary statistics where \$n\$ is the same across analyses.
 For the models with 7 and 30 day optimization periods, the final period in the annual time series was excluded from the results due to incomplete number of observations compared to earlier periods (i.e., the total number of days in the time series was not evenly divided by 7 or 30, resulting in a remainder for the last period).
@@
 This approach models dissolved oxygen as a function of time (day), hour, and water depth using a moving window approach with weights of the independent variables appropriate for each time period within the windows.
-The results of this analysis return a predicted and detided dissolved oxygen time series.
+The results of this analysis return a predicted and detided dissolved oxygen time series, where the detided results are simply the predictions at average tidal height [see @Beck15 for details].
+This method was only applied to dissolved oxygen since it has not been evaluated with any other parameters.
 Two separate time series of noise, representing residual (any other random process not related to metabolism or tidal advection) or tidal noise, were derived from the observed, predicted and detided time series.
@@
 Overall, these changes also reduced the required inputs and many monitoring locations nationwide include the requisite data for applying EBASE (i.e., dissolved oxygen, water temperature, salinity, PAR, and wind speed, as collected by the NERRS).
+Further, EBASE could be applied to other non-tidal, lotic systems, although additional comparisons with existing methods are needed and the default uninformed prior distributions should be chosen carefully given the justification herein based primarily on estuarine literature [e.g., @caffrey2004].
 
@@
 Hypotheses can be generated on the amount of time over which critical metabolic parameters may vary across systems.
+These hypotheses may also consider frequency of continuous monitoring (e.g., 15 minute vs hourly) and how this may influence the results, as this is currently unknown for EBASE.
 
@@
 Results are compared for EBASE parameters estimated with a (a) 7 day and (b) 30 day optimization period.
-95% credible intervals from the posterior distributions of the parameter estimates from EBASE are also shown.
-The results are averaged at the time step of the model optimization period.](figs/defplo.png){#fig-defplo}
+95% credible intervals from the posterior distributions of the parameter estimates from EBASE are also shown.](figs/defplo.png){#fig-defplo}
 

### 2024-02-29 revision/review_response.qmd
more revisions

@@
 
-*Response: text goes here*
+*Response: Thank you for providing thoughtful comments on our manuscript.  Please see our responses below.*
 
@@
 *Response: Yes, it is possible with caveats.
-Please see our response to reviewer on line 188.
+Please see our response to reviewer one on line 188.
 Specifically, additional comparisons are needed and the default uninformed priors based on estuarine literature must be reconsidered.*
@@
 In theory this makes sense in an estuary as both wind and flow likely affect gas transfer.
+
+*Response: Apologies for this confusion.
+Eq.
+3 is not used in EBASE, rather it is discussed as the gas transfer used in BASE that is inappropriate for estuaries, justifying our use of the gas exchange model in @Wanninkhof14.
+Line 166 was revised to make this more clear: "Based on these differences and the lack of availability of other possible drivers of the transfer velocity in coastal systems, Equation 3 is not used herein.
+Instead, the well-established wind- and temperature-based parameterization for gas exchange of @Wanninkhof14 was used for EBASE (Equation 2)..."*
 
@@
 
-*Response: text goes here*
+*Response: Added R in parentheses.*
 
@@
 
-*Response: text goes here*
+*Response: Sentence was revised as follows: "...forming particulate organic carbon with multiple fates, e.g., buried in estuarine sediments, incorporated into biota, transferred to the ocean, etc.
+[@hu2006; @middelburg2007]."*
 
-L83: which computed rates were negative? P, R, or both? L128: mean depth?
+L83: which computed rates were negative? P, R, or both?
 
-*Response: text goes here*
+*Response: Revised as "The computed rates were negative 23% of the time for both P and R..."*
+
+L128: mean depth?
+
+*Response: The software can handle a single value (e.g., mean depth) or a vector of tidal height depths equal in length to the input time series.
+This is expressed in the documentation: https://fawda123.github.io/EBASE/reference/ebase.html.
+This is also expressed in the new table showing model inputs/outputs as requested above.*
 
@@
 
-*Response: text goes here*
+*Response: See the new table for inputs/outputs.*
 
@@
 
-*Response: text goes here*
+*Response: We have specified that the unknown metabolic parameters ($a$, $R$, and $b$) are constant at the unit of the model optimization period chosen by the user, although P and D vary continuously given their dependence on measured input parameters (e.g., PAR, wind) that can change at each time step.
+Please see our response to reviewer 1 on line 181.*
 
@@
 
-*Response: text goes here*
+*Response: Please see our response to the previous comment and the new input/output table.
+It's an ambiguous parameter, but really just an indication of how sensitive the gas transfer is to wind speed.
+The sentence was revised as "The $b$ parameter is fairly well known for open-ocean conditions primarily as an expression of the sensitivity of gas exchange to wind..."*
 
@@
 
-*Response: text goes here*
+*Response: This equation is *
 
@@
 
-*Response: text goes here*
+*Response: It's the exception and not the norm that published software remains the same after a descriptive manuscript is published.
+For this reason, we have added a version of EBASE that is appropriate for this paper and included the version number in the text.
+......* EDIT
 

### 2024-02-29 manu-draft.qmd
more revisions

@@
 A comparison with the more-conventional method of Odum reveals the ability of EBASE to avoid unphysical results (such as negative photosynthesis and respiration) and improves when the DO data are detided.
-EBASE is available using open-source software and can be readily applied to multiple years of long-term monitoring data that are available in many estuaries.
+EBASE is available using open-source software (R) and can be readily applied to multiple years of long-term monitoring data that are available in many estuaries.
 Overall, EBASE provides an accessible method to parameterize a simple metabolic model appropriate for estuarine systems and will provide additional understanding of processes that influence ecosystem status and condition.
@@
 For example, estuarine respiration of organic carbon delivered from land produces carbon dioxide, which may be lost to the atmosphere via outgassing [e.g, @cai2011].
-Similarly, estuarine primary production consumes dissolved inorganic carbon delivered from land, forming particulate organic carbon that may be buried in estuarine sediments [e.g., @hu2006].
+Similarly, estuarine primary production consumes dissolved inorganic carbon delivered from land, forming particulate organic carbon with multiple fates, e.g., buried in estuarine sediments, incorporated into biota, transferred to the ocean, etc.
+[@hu2006; @middelburg2007].
 Estuarine metabolic processes are fundamentally important, as illustrated by the dissolved inorganic carbon budget of eastern North American estuaries [@najjar2018]; net primary production and heterotrophic respiration far exceed inputs from upland sources and outputs to the atmosphere and ocean.
@@
 Striking evidence of this limitation is the large number of physically impossible negative P and R estimates in the open-water analysis of 28 3-year time series by [@caffrey2003].
-The computed rates were negative 23% of the time, on average, and as high as 69% for individual sites.
+The computed rates were negative 23% of the time for both P and R, on average, and as high as 69% for individual sites.
 Note that R is often expressed as a negative value indicating consumption of dissolved oxygen.
@@
 where $b$ is constant, $U_{10}$ is the wind speed at 10 m above the water, and $Sc$ is the Schmidt number, which is the ratio of the kinematic viscosity of water to the molecular diffusivity of oxygen in water [$Sc$ is computed from water temperature and salinity using the polynomial fit in @Wanninkhof14].
-The $b$ parameter is fairly well known for open-ocean conditions, but may be different in estuaries, where the transfer velocity is influenced by factors other than wind speed and $Sc$.
+The $b$ parameter is fairly well known for open-ocean conditions primarily as an expression of the sensitivity of gas exchange to wind, but may be different in estuaries, where the transfer velocity is influenced by factors other than wind speed and $Sc$.
 As such, the three parameters estimated by EBASE are $a$, $R$, and $b$ using the required input data that include time series of dissolved oxygen, water temperature, salinity, PAR, wind speed, and water column depth (@tbl-ebaseiotab).
@@
 An expectation is that $k_w$ would increase as water depth decreases for a given current speed because turbulence generated at the sediment--water interface by currents would more easily reach the surface [@Ho16].
-Based on these differences and the lack of availability of other possible drivers of the transfer velocity in coastal systems, the well-established wind- and temperature-based parameterization for gas exchange of @Wanninkhof14 was used for EBASE, but allowing $b$, which was estimated by @Wanninkhof14 to be 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$, to be a free parameter that may reflect spatial and temporal variability in factors other than wind and temperature that may influence transfer velocity.
+Based on these differences and the lack of availability of other possible drivers of the transfer velocity in coastal systems, @eq-gasexbase is not used herein.
+Instead, the well-established wind- and temperature-based parameterization for gas exchange of @Wanninkhof14 was used for EBASE [@eq-kw], but allowing $b$, which was estimated by @Wanninkhof14 to be 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$, to be a free parameter that may reflect spatial and temporal variability in factors other than wind and temperature that may influence transfer velocity.
 

### 2024-02-29 R/tabs.R

@@ -11,7 +11,7 @@ totab <- tibble(
   Type = c('Input', 'Input', 'Input', 'Input', 'Input', 'Input',
            'Input-derived', 'Input-derived', 'Input-derived',
            'Output', 'Output', 'Output', 'Output', 'Output', 'Output'),
-  Description = c('Dissolved oxygen (observed)', 'Water temperature', 'Salinity', 'Total photosynthetically active radiation', 'Wind speed', 'Water column depth',
+  Description = c('Dissolved oxygen (observed)', 'Water temperature', 'Salinity', 'Total photosynthetically active radiation', 'Wind speed', 'Water column depth (mean or tidal height vector)',
                   'Wind speed at 10 meter height, squared', 'Schmidt number (from water temperature and salinity)', 'Dissolved oxygen at saturation (from water temperature and salinity)',
                   'Dissolved oxygen (modelled)', 'Production', 'Respiration', 'Gas exchange', 'Light efficiency', 'Sensitivity of gas transfer to wind speed'),
   `Model notation` =  c('C', '-', '-', 'PAR', '-', 'Z', 

### 2024-03-01 manu-draft.qmd
more revisions

@@
 ```{=tex}
-\newcommand{\auni}{\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)}
-\newcommand{\runi}{\text{mmol}/\text{m}^2/\text{d}}
-\newcommand{\buni}{\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)}
+\newcommand{\auni}{\left(\text{mmol}~\text{m}^{-2}~\text{d}^{-1}\right)/\left(\text{W}~\text{m}^{-2}\right)}
+\newcommand{\runi}{\text{mmol}~\text{m}^{-2}~\text{d}^{-1}}
+\newcommand{\buni}{\left(\text{cm}~\text{hr}^{-1}\right)/\left(\text{m}^2~\text{s}^{-2}\right)}
 ```
@@
 
-Continuous measurements of dissolved oxygen (DO) are useful for quantifying ecosystem metabolism, which is critical for understanding estuarine biogeochemistry and ecology, but current methods applied to these data may lead to unphysical results and poorly constrained errors.
+Continuous measurements of dissolved oxygen (DO) are useful for quantifying ecosystem metabolism, which is critical for understanding estuarine biogeochemistry and ecology, but current methods applied to these data may lead to estimates that are physically impossible and poorly constrained errors.
 Here we present a new approach for estimating estuarine metabolism: EBASE (Estuarine BAyesian Single-station Estimation).
@@
 These comparisons combine results using real observations and those based on detided inputs to demonstrate effects of tidal advection on metabolic estimates.
-We also attempt to constrain the gas transfer velocity using this approach but are unable, at least with the data sets used in this study.
+We also attempt to constrain (i.e., accurately estimate) the gas transfer velocity using EBASE but are unable, at least with the data sets used in this study.
 EBASE is available as an open-source software package created with the R statistical programming language [@Beck23; @RCT23] and the results herein provide practical suggestions for use of the software on novel datasets.
@@
 Both $P$ and $R$ are expressed as positive values, whereas $D$ can be positive (outgassing) or negative (ingassing).
-The equation has dimensions of moles per unit area and time.
+The equation has dimensions of moles per unit area and time (e.g., $\text{mmol}~\text{m}^{-2}~\text{d}^{-1}$).
 $P$ is modeled as $aPAR$, where $PAR$ is the surface photosynthetically active radiation and $a$ is the light efficiency; $R$ is constant; and $D$ is modeled in the standard manner [e.g, @sarmiento2013] as $k_w\left(C_{sat} - C\right)$, where $k_w$ is the gas transfer velocity and $C_{sat}$ is the saturation (or equilibrium) concentration, calculated from salinity and temperature [@Garcia92].
@@
 Based on these differences and the lack of availability of other possible drivers of the transfer velocity in coastal systems, @eq-gasexbase is not used herein.
-Instead, the well-established wind- and temperature-based parameterization for gas exchange of @Wanninkhof14 was used for EBASE [@eq-kw], but allowing $b$, which was estimated by @Wanninkhof14 to be 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$, to be a free parameter that may reflect spatial and temporal variability in factors other than wind and temperature that may influence transfer velocity.
+Instead, the well-established wind- and temperature-based parameterization for gas exchange of @Wanninkhof14 was used for EBASE [@eq-kw], but allowing $b$, which was estimated by @Wanninkhof14 to be 0.251 $\left(\text{cm}~\text{hr}^{-1}\right)/\left(\text{m}^2~\text{s}^{-2}\right)$, to be a free parameter that may reflect spatial and temporal variability in factors other than wind and temperature that may influence transfer velocity.
 
@@
 A detailed web page ([https://fawda123.github.io/EBASE/](#0){.uri}) describes how to use the package, with explanations of the functions for viewing model results, including a plot of the modeled dissolved oxygen with the observed, a time series plot of the oxygen budget terms (P, R, and D), and a time series plot of the credible intervals for the $a$, $R$, and $b$ parameters.
-All results provided herein were generated using the EBASE R package.
+All results provided herein were generated using the EBASE R package (specific materials for this manuscript can be found at <https://github.com/fawda123/ebase-manu>).
 
@@
 
-Specifically, the prior mean for $R$ was chosen as 300 $\text{mmol}/\text{m}^2/\text{d}$ (\~10 $\text{g O}_2/\text{m}^2/\text{d}$) using a synthesis based on NERRS data [Table 2 in @caffrey2004].
+Specifically, the prior mean for $R$ was chosen as 300 $\text{mmol}~\text{m}^{-2}~\text{d}^{-1}$ (\~10 $\text{g O}_2~\text{m}^{-2}~\text{d}^{-1}$) using a synthesis based on NERRS data [Table 2 in @caffrey2004].
 Although this synthesis focused on small, shallow estuaries that are part of the NERRS, this dataset is the largest, longest, and most consistently collected in the US and it is reasonable to expect others may apply EBASE to these data.
-The prior mean for $a$ was chosen assuming that $P$ and $R$ are comparable at long time scales, such that $aPAR$ = $R$, or $a$ = $\left(300~\text{mmol}/\text{m}^2/\text{d}\right)/PAR$.
-Using the global mean surface $PAR$ of 80 $W/m^2$, which is based on a global mean surface total shortwave radiation of 200 $\text{W}/\text{m}^2$ [@kiehl1997] and a $PAR$ fraction of total shortwave radiation of 0.4 [@jacovides2004], the prior mean for $a$ is reasonably set at 4 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$.
-Lastly, the prior mean for $b$ was chosen as 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ following @Wanninkhof14.
+The prior mean for $a$ was chosen assuming that $P$ and $R$ are comparable at long time scales, such that $aPAR$ = $R$, or $a$ = $\left(300~\text{mmol}~\text{m}^{-2}~\text{d}^{-1}\right)/PAR$.
+Using the global mean surface $PAR$ of 80 $\text{W}~m^{-2}$, which is based on a global mean surface total shortwave radiation of 200 $\text{W}~\text{m}^{-2}$ [@kiehl1997] and a $PAR$ fraction of total shortwave radiation of 0.4 [@jacovides2004], the prior mean for $a$ is reasonably set at 4 $\left(\text{mmol}~\text{m}^{-2}~\text{d}^{-1}\right)/\left(\text{W}~\text{m}^{-2}\right)$.
+Lastly, the prior mean for $b$ was chosen as 0.251 $\left(\text{cm}~\text{hr}^{-1}\right)/\left(\text{m}^2~\text{s}^{-2}\right)$ following @Wanninkhof14.
 
 The prior standard deviation for $R$ was based on the standard deviation of the mean values of $R$ at the NERRS sites in @caffrey2004.
-The standard deviation of $R$ was 171 $\text{mmol}/\text{m}^2/\text{d}$, which was rounded to 50% of the mean, or 150 $\text{mmol}/\text{m}^2/\text{d}$.
-The standard deviation for $a$ was set proportionally to 2 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$.
+The standard deviation of $R$ was 171 $\text{mmol}~\text{m}^{-2}~\text{d}^{-1}$, which was rounded to 50% of the mean, or 150 $\text{mmol}~\text{m}^{-2}~\text{d}^{-1}$.
+The standard deviation for $a$ was set proportionally to 2 $\left(\text{mmol}~\text{m}^{-2}~\text{d}^{-1}\right)/\left(\text{W}~\text{m}^{-2}\right)$.
 For the prior standard deviation of $b$, we considered using the 20% error estimated for $b$ by @Wanninkhof14.
@@
 $$
-a \sim N(4, 2)I(0,\infty) \;\; \left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)
+a \sim N(4, 2)I(0,\infty) \;\; \left(\text{mmol}~\text{m}^{-2}~\text{d}^{-1}\right)/\left(\text{W}~\text{m}^{-2}\right)
 $$ {#eq-aprior}
@@
 $$
-R \sim N(300, 150)I(0,\infty) \;\; \text{mmol}/\text{m}^2/\text{d}
+R \sim N(300, 150)I(0,\infty) \;\; \text{mmol}~\text{m}^{-2}~\text{d}^{-1}
 $$ {#eq-rprior}
@@
 $$
-b \sim N(0.251, 0.125)I(0,0.502) \;\; \left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)
+b \sim N(0.251, 0.125)I(0,0.502) \;\; \left(\text{cm}~\text{hr}^{-1}\right)/\left(\text{m}^2~\text{s}^{-2}\right)
 $$ {#eq-bprior}
@@
 All prior distributions were constrained to positive values based on known physical constraints and requirements of the model formula in @eq-ebase.
-Additionally, the prior distribution for $b$ was constrained to an upper limit of 0.502 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ (\~twice the default mean).
+Additionally, the prior distribution for $b$ was constrained to an upper limit of 0.502 $\left(\text{cm}~\text{hr}^{-1}\right)/\left(\text{m}^2~\text{s}^{-2}\right)$ (\~twice the default mean).
 Initial development of EBASE showed that using an undefined upper limit of $b$ led to unstable and unreasonable estimates for all parameters.
@@
 The synthetic oxygen time series was generated using @eq-ebase as a forward calculating model with specified daily values for $a$, $R$, and $b$.
-The $b$ parameter was fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ [@Wanninkhof14].
+The $b$ parameter was fixed at 0.251 $\left(\text{cm}~\text{hr}^{-1}\right)/\left(\text{m}^2~\text{s}^{-2}\right)$ [@Wanninkhof14].
 Daily values of $a$ and $R$ were estimated by (1) applying the Odum open-water method as implemented in the WtRegDO R package [@Beck23a], (2) computing monthly averages of the resulting daily estimates of $P$ and $R$, (3) computing the monthly average of the parameter $a$ by dividing the monthly average of $P$ by the monthly average of $PAR$, and (4) by fitting a spline function to the monthly averages of $R$ and $a$.
@@
 Low and high values for the means and standard deviations for $R$ were assessed following reasonable ranges from @caffrey2004 using logic similar to that for choosing the default priors.
-The minimum and maximum mean values for $R$ were evaluated as 138 and 1009 $\text{mmol}/\text{m}^2/\text{d}$, which we rounded to $1/2$ and 3 times the mean (150 and 900 $\text{mmol}/\text{m}^2/\text{d}$) and proportionally for $a$ as 2.0 and 12 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$.
+The minimum and maximum mean values for $R$ were evaluated as 138 and 1009 $\text{mmol}~\text{m}^{-2}~\text{d}^{-1}$, which we rounded to $1/2$ and 3 times the mean (150 and 900 $\text{mmol}~\text{m}^{-2}~\text{d}^{-1}$) and proportionally for $a$ as 2.0 and 12 $\left(\text{mmol}~\text{m}^{-2}~\text{d}^{-1}\right)/\left(\text{W}~\text{m}^{-2}\right)$.
 The minimum and maximum values for the standard deviations for $R$ and $a$ were chosen as 10% and 1000% of the default mean values to test a range from high precision to completely uninformed values.
-The mean and standard deviation for the $b$ parameter were not tested and the prior value was set as the fixed constant used in the synthetic time series, 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$.
+The mean and standard deviation for the $b$ parameter were not tested and the prior value was set as the fixed constant used in the synthetic time series, 0.251 $\left(\text{cm}~\text{hr}^{-1}\right)/\left(\text{m}^2~\text{s}^{-2}\right)$.
 There is minimal prior knowledge on expectations for the $b$ parameter and initial assessments with the synthetic time series demonstrated that the parameter was poorly constrained, primarily because the fixed parameter could not be estimated using observed data (i.e., wind, water temperature) that affected the actual realization of $b$.
@@
 The only models where the NSE values were less than zero for dissolved oxygen were those where $\sigma$ of the $a$ and $R$ parameters was low and $\mu$ for one of the $a$ and $R$ parameters was high.
-However, substantial variation in NSE values was observed among all other model outputs, with some very low values of NSE occurring while the $DO$ simulation was fairly good, showing the potential for the model to produce the right answer for the wrong reasons.
+However, substantial variation in NSE values was observed among all other model outputs, with some very low values of NSE occurring while the $DO$ simulation was fairly good, showing the potential for the model to produce the right answer (accurate $DO$) for the wrong reasons (inaccurate $P$, $R$, or $D$), since $DO$ is simply a mass balance of $P$, $R$, and $D$ ([@eq-ebase]).
 After dissolved oxygen, gas exchange ($D$) was estimated with the highest median NSE of `r prmall[['D']]`.
@@
 The default priors (Equations [-@eq-aprior] and [-@eq-rprior]) were used with a 7 day optimization period.
-The prior for the $b$ parameter was fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ following similar logic used for the sensitivity analysis described above.
+The prior for the $b$ parameter was fixed at 0.251 $\left(\text{cm}~\text{hr}^{-1}\right)/\left(\text{m}^2~\text{s}^{-2}\right)$ following similar logic used for the sensitivity analysis described above.
 
@@
 The Odum method used an approach described in @Murrell18 created for estuarine application that uses a gas-exchange approach from @Thebault08.
-For a better comparison with the EBASE methods, the gas exchange parameterization in the Odum method was modified to use the @Wanninkhof14 approach and the $b$ parameter was fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ [as available in the WtRegDO R package, @Beck23a].
+For a better comparison with the EBASE methods, the gas exchange parameterization in the Odum method was modified to use the @Wanninkhof14 approach and the $b$ parameter was fixed at 0.251 $\left(\text{cm}~\text{hr}^{-1}\right)/\left(\text{m}^2~\text{s}^{-2}\right)$ [as available in the WtRegDO R package, @Beck23a].
 An additional and fundamental difference between EBASE and the Odum method is that the former is likelihood-based with metabolic rates estimated by fitting the model to the data, whereas the latter is strictly arithmetic-based primarily on integration of the diel dissolved oxygen curve.
@@
 Hypotheses can be generated on the amount of time over which critical metabolic parameters may vary across systems.
+These hypotheses may relate to the length of time over which potential factors affect metabolism, e.g., storm events may affect metabolism at short time scales and a shorter optimization period could be used, whereas longer optimization periods may be more desirable if interested in long-term, climate-related effects on metabolism.
 These hypotheses may also consider frequency of continuous monitoring (e.g., 15 minute vs hourly) and how this may influence the results, as this is currently unknown for EBASE.
@@
 All priors are normal distributions and truncated to positive values.
-$b$ is also truncated to a maximum of 0.502 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$.
+$b$ is also truncated to a maximum of 0.502 $\left(\text{cm}~\text{hr}^{-1}\right)/\left(\text{m}^2~\text{s}^{-2}\right)$.
 Probabilities have been normalized to the maximum probability.](figs/priorplot.png){#fig-priorplot fig-align="center"}

### 2024-03-04 revision/review_response.qmd
remove depth-dependency justification

@@
 
-```{r echo = F, cache = F, eval = F}
+```{r}
+library(dplyr)
+library(tidyr)
+library(lubridate)
+library(ggplot2)
+library(readr)
+```
+
+```{r}
 #| eval: false
@@
 *Response: The resolution of "continuous" monitoring data has not been evaluated in this study, although our guess is that more frequent sampling may produce more robust estimates.
-However, this may not matter given the time scales at which metabolism changes (i.e., > 1 day).
+However, this may not matter given the time scales at which metabolism changes (i.e., \> 1 day).
 In our experience, estimates from the Odum open-water method are nearly identical using 15-minute or hourly samples.
@@
 
-*Response: Yes, this is correct and can even be seen in Figure 2, e.g., late September to early October shows P > R with negative D (ingassing), then P < R with positive D (outgassing).
+*Response: Yes, this is correct and can even be seen in Figure 2, e.g., late September to early October shows P \> R with negative D (ingassing), then P \< R with positive D (outgassing).
 We have corrected this error in the text.*
@@
 In particular, our revision notes that only one value each for $a$, $R$, and $b$ is returned for each optimization period.
-The vertical lines for each point in Figure 3 are credible intervals showing uncertainy around each individual estimate.
+The vertical lines for each point in Figure 3 are credible intervals showing uncertainty around each individual estimate.
 The statement in the caption that "the results are averaged at the time step of the model optimization period" is incorrect and removed.
@@
 Eq.
-3 is not used in EBASE, rather it is discussed as the gas transfer used in BASE that is inappropriate for estuaries, justifying our use of the gas exchange model in @Wanninkhof14.
-Line 166 was revised to make this more clear: "Based on these differences and the lack of availability of other possible drivers of the transfer velocity in coastal systems, Equation 3 is not used herein.
+3 is not used in EBASE, rather it is discussed as the gas transfer used in BASE that is inappropriate for our application, justifying use of the gas exchange model in @Wanninkhof14.
+Line 166 was revised to make this more clear: "Based on this difference and the lack of availability of other possible drivers of the transfer velocity in coastal systems, Equation 3 is not used herein.
 Instead, the well-established wind- and temperature-based parameterization for gas exchange of @Wanninkhof14 was used for EBASE (Equation 2)..."*
@@
 But I realize this term is not typically available.
+
+*Response: We have noted an error in our description of Equation 3 that includes the depth-dependency.
+As noted by Reviewer 3, @Grace15 do not include a depth-dependency in this equation.
+We have since removed the dependency and revised the text accordingly, emphasizing the importance of wind for the transfer velocity in open-water systems.
+Regarding temperature, we feel justified in its inclusion for gas exchange given the previously noted relationship with the Schmidt number, a known factor to be important in open-water systems.*
 
@@
 
-*Response: See the new table for inputs/outputs.*
+*Response: Please see the new table for inputs/outputs.
+Although equation 1 can apply to different units of space and time, we have added those used herein in parentheses: "The equation has dimensions of moles per unit area and time (e.g., $\text{mmol}~\text{m}^{-2}~\text{d}^{-1}$)."*
 
@@
 
-*Response: This equation is *
+*Response: The depth-dependency for kw described in [@Grace15] was incorrect and an error on our part.
+We have removed $H$ (now $Z$) from the equation and updated the text accordingly, emphasizing the wind-dependence on the transfer velocity as a key factor influencing our choice to use the method from @Wanninkhof14: "However, open-water systems like estuaries are more susceptible to wind-based influences on gas exchange, particularly for $k_w$ [@Wanninkhof14; @vandam2019].
+Based on this difference and the lack of availability of other possible drivers of the transfer velocity in coastal systems, Equation 3 is not used herein.
+Instead, the well-established wind- and temperature-based parameterization for gas exchange of @Wanninkhof14 was used for EBASE..."*
 
@@
 For this reason, we have added a version of EBASE that is appropriate for this paper and included the version number in the text.
-......* EDIT
+EBASE is also now on CRAN, the Centralized R Archive Network, where a majority of publicly available R packages are hosted, in addition to the GitHub repository.
+We have also added text on line 193 indicating where source materials for this manuscript (i.e., to reproduce results) can be found: "All results provided herein were generated using the EBASE R package (specific materials for this manuscript can be found at <https://github.com/fawda123/ebase-manu>)."* EDIT
 
@@
 
-*Response: text goes here*
+*Response: Point well taken, we have not been consistent in the exponent notation as the text uses positive with fractions and the figures use negative with no fractions.
+We have corrected all instances in the text to use negative exponents.
+However, we have retained the units for $a$ since this makes the relationship between $P$, $a$, and $PAR$ more clear, i.e., the units for PAR ($W~m^{-2}$) when multiplied by the units for $a$ ($\left(\text{mmol}~\text{m}^{-2}~\text{d}^{-1}\right)/\left(W~m^{-2}\right)$) produce the units for P ($\text{mmol}~\text{m}^{-2}~\text{d}^{-1}$).*
 
@@
 
-*Response: text goes here*
+*Response: Yes, we noticed this behavior as well.
+To us, this is a reflection of the instability in the parameter estimation.
+Earlier iterations of model development did not have this constraint and the $b$ parameter was often estimated at unreasonably high values, often causing the other parameters ($a$ and $R$) to also be unreasonable.
+We previously described some rationale for why this parameters was so poorly constrained and offer some potential solutions for future research (i.e., lines 482-490, 502-509.*
 
@@
 
-*Response: text goes here*
+*Response: Perhaps, although our focus was on the primary metabolic parameters returned by EBASE given the goal of assessing parameter recovery (of which kw is included).
+For sake of interest, below is a plot of D and kw for the synthetic data (data available here: <https://raw.githubusercontent.com/fawda123/BASEmetab_script/master/data/apafwoxy2.csv>).
+Note that kw is directly proportional to wind, whereas D is mostly a function of kw and the difference between C and Csat.*
+
+```{r}
+#| fig-width: 8
+#| fig-height: 4
+
+fwdat <- read_csv(file = url('https://raw.githubusercontent.com/fawda123/BASEmetab_script/master/data/apafwoxy2.csv'))
+
+fwdatcmp <- fwdat %>% 
+  mutate(
+    DateTimeStamp = dmy_hms(datet, tz = 'America/Jamaica'),
+    Date = as.Date(DateTimeStamp),
+    D = -1 * `gasex,mmol/m2/d`,
+    kw = `kw,m/s`
+  ) %>% 
+  select(Date, D, kw) %>% 
+  pivot_longer(cols = c(D, kw), names_to = 'var', values_to = 'val') %>% 
+  summarise(
+    val = mean(val, na.rm = TRUE),
+    .by = c('Date', 'var')
+  ) %>% 
+  mutate(
+    var = factor(var, levels = c('D', 'kw'), labels = c('D (mmol m-2 d-1)', 'kw (m s-1)'))
+  )
+
+ggplot(fwdatcmp, aes(x = Date, y = val, color = var)) +
+  geom_line() + 
+  geom_point() + 
+  theme_minimal() + 
+  facet_wrap(~var, scales = 'free_y', ncol = 1, strip.position = 'left') +
+  theme(
+    strip.placement = 'outside', 
+    legend.position = 'none'
+  ) +
+  labs(
+    x = NULL,
+    y = NULL
+  )
+```
 
@@
 
-*Response: text goes here*
+*Response: Good suggestion, we have added this text to line 541 (also note the last sentence that was added in response to a general comment from reviewer 1 regarding how frequency of continuous sampling may affect the results): "These hypotheses may relate to the length of time over which potential factors affect metabolism, e.g., storm events may affect metabolism at short time scales and a shorter optimization period could be used, whereas longer optimization periods may be more desirable if interested in long-term, climate-related effects on metabolism.
+These hypotheses may also consider frequency of continuous monitoring (e.g., 15 minute vs hourly) and how this may influence the results, as this is currently unknown for EBASE."*
 
@@
 
-*Response: text goes here*
+*Response: Text was modified as suggested.*
 
@@
 
-*Response: text goes here*
+*Response: The sentence was modified to clarify the meaning.
+We have retained "constrain" given that it's used throughout the paper, but we provide an explicit meaning in parentheses: "We also attempt to constrain (i.e., accurately estimate) the gas transfer velocity using EBASE but are unable...".*
 
@@
 
-*Response: text goes here*
+*Response: Thank you for pointing out this error.  We have removed this from the text and revised accordingly.  Please see our response to reviewer 2 for line 164.*
 
@@
 
-L352: The “right answer for wrong reasons” is interesting. It could be clearer here that even though DO is estimated accurately, the P and R estimates will be inaccurate in this scenario.-7
+L352: The “right answer for wrong reasons” is interesting. It could be clearer here that even though DO is estimated accurately, the P and R estimates will be inaccurate in this scenario.
 
-*Response: text goes here*
+*Response: Sentence was revised as follows: "...the potential for the model to produce the right answer (accurate $DO$) for the wrong reasons (inaccurate $P$, $R$, or $D$), since $DO$ is simply a mass balance of $P$, $R$, and $D$ (Equation 1)".*
 

### 2024-03-04 manu-draft.qmd
remove depth-dependency justification

@@
 $$
-\frac{k_w}{Z} = K \cdot 1.024^{\left(T - \bar{T}\right)}
+k_w = K \cdot 1.024^{\left(T - \bar{T}\right)}
 $$ {#eq-gasexbase}
@@
 where $T$ is the instantaneous water temperature, $\bar{T}$ is the average water temperature over each 24-hour period in the oxygen time series (both in $^{\circ}$C), and $K$ is the reaeration coefficient estimated from the data.
-@eq-gasexbase indicates that $k_w$ increases with temperature and depth.
-The temperature dependence is consistent with the known decrease of $k_w$ with the Schmidt number, but the depth dependence is difficult to justify.
-An expectation is that $k_w$ would increase as water depth decreases for a given current speed because turbulence generated at the sediment--water interface by currents would more easily reach the surface [@Ho16].
-Based on these differences and the lack of availability of other possible drivers of the transfer velocity in coastal systems, @eq-gasexbase is not used herein.
+@eq-gasexbase indicates that $k_w$ increases with temperature, consistent with the known decrease of $k_w$ with the Schmidt number.
+However, open-water systems like estuaries are more susceptible to wind-based influences on gas exchange, particularly for $k_w$ [@Wanninkhof14; @vandam2019].
+Based on this difference and the lack of availability of other possible drivers of the transfer velocity in coastal systems, @eq-gasexbase is not used herein.
 Instead, the well-established wind- and temperature-based parameterization for gas exchange of @Wanninkhof14 was used for EBASE [@eq-kw], but allowing $b$, which was estimated by @Wanninkhof14 to be 0.251 $\left(\text{cm}~\text{hr}^{-1}\right)/\left(\text{m}^2~\text{s}^{-2}\right)$, to be a free parameter that may reflect spatial and temporal variability in factors other than wind and temperature that may influence transfer velocity.
@@
 As described in the materials and procedures section, EBASE is an estuarine adaptation of BASE [@Grace15], which was developed for stream applications and some assumptions of the underlying model of the latter were inappropriate for the former.
-The most critical change made to BASE was adopting the gas transfer velocity parameterization in @Wanninkhof14, which is primarily wind-based and more appropriate than the depth-dependent formulation in BASE.
+The most critical change made to BASE was adopting the gas transfer velocity parameterization in @Wanninkhof14, which is primarily wind-based and more appropriate than the formulation in BASE.
 EBASE also removed the temperature dependency of respiration, consistent with applications of the Odum method to other shallow estuaries [@Beck15; @Murrell18].

### 2024-03-06 revision/review_response.qmd
revisions done but need to proof and add cran

@@
 
-*Response: text goes here*
+*Response: Thank you for providing these thorough comments on our manuscript.  Please see our responses below.*
 
@@
 
-*Response: text goes here*
+*Response: We agree this clarification is needed.
+The first study [@ciavatta2008] actually did not use a Bayesian approach, rather it provided simultaneous estimation of the metabolic parameters using a conventional empirical model framework.
+This reference to a Bayesian approach was in error and we have removed this citation from the introduction.
+@tassone2019 does provide a good example for comparison with our work and we elaborate on this in the revised text on line 105: "@tassone2019 adapted a Bayesian approach developed for streams [@appling2018] in an estuarine setting to estimate metabolic rates in the James River.
+Although this approach demonstrated the utility of such an approach in estuaries, much of the core model structure developed for streams was unchanged and may not have been fully descriptive of fundamental processes influencing metabolism in open-water environments."*
+
+*We then continue the justification for our use of BASE, adding the following:*
+
+*Line 108: "The underlying model components can also easily be updated given the relatively simple format of the approach."*
+
+*Line 115: "We focus on BASE given the relative simplicity and, therefore, ease of application compared to other stream-based Bayesian approaches [e.g., @appling2018]."*
+
+*The streamMetabolizer package is a great contribution to the field, although our implicit goal of this work was to modify a readily accessible open source approach such as BASE for our needs.
+The framework was easily extendable for our chosen gas-exchange method [@Wanninkhof14] and modification of the software functionality (i.e., user-chosen optimization period and dealing with data gaps).*
 
@@
 
-*Response: text goes here*
+*Response: Thank you for pointing out this reference.
+We have included an explicit mention of equifinality and a citation for @appling2018 on line 352.
+Also please see our addition to the discussion on line 490.
+This is described in full in response to the general comments from Reviewer 3.*
 
@@
 
-*Response: text goes here*
+*Response: Please see our response to your general comment above.*
 
@@
 
-*Response: text goes here*
+*Response: Please see our response to your general comment above.
+Although @tassone2019 does provide another useful example, we still feel that these methods have not been thoroughly explored in estuaries.*
 
@@
 We have since removed the dependency and revised the text accordingly, emphasizing the importance of wind for the transfer velocity in open-water systems.
-Regarding temperature, we feel justified in its inclusion for gas exchange given the previously noted relationship with the Schmidt number, a known factor to be important in open-water systems.*
+Regarding temperature, we feel justified in its inclusion for gas exchange given the previously noted relationship with the Schmidt number, a known factor to be important in open-water systems.
+Also see our response to your comment on line 164.*
 
@@
 
-*Response: text goes here*
+*Response: Yes, we agree this issue is perplexing and worth exploring further.
+We have mentioned in the original draft how this could be explored further, e.g., evaluation in other locations where gas exchange may be a more dominant component of the metabolic model (line 490).
+We have also added a brief description of @appling2018 that confront this issue in stream metabolic models.
+Please see our response to the general comments from Reviewer 3.*
 
@@
 
-*Response: text goes here*
+*Response: Thank you for these suggestions.
+We agree an explicit description of how the approach can be applied in light of uncertainties in the $b$ parameter is needed.
+Some discussion is already included (e.g., line 488).
+We have also added text describing how @appling2018 addressed the issue, although this suggestions would be further model development.
+Revisions were made to the first paragraph of the discussion.*
+
+*Line 490: "For application, users of EBASE may simply use a fixed value for $b$ (e.g., 0.251 $\left(\text{cm}~\text{hr}^{-1}\right)/\left(\text{m}^2~\text{s}^{-2}\right)$) or perform a simple sensitivity analysis by varying $b$ +/- a small percentage to determine how $P$ and $R$ may change.
+Further model development may consider alternative approaches for dealing with equifinality, i.e., when the same modeled DO time series can be returned with different estimates of $P$, $R$, and $D$.
+@appling2018 provide a useful example for a stream metabolic model that addresses equifinality by allowing model parameters to vary with key drivers at sub-daily scales (e.g., gas exchange versus discharge) and avoiding assumptions that model residuals are uncorrelated."*
 
@@
 
-*Response: Sentence was revised as follows: "...the potential for the model to produce the right answer (accurate $DO$) for the wrong reasons (inaccurate $P$, $R$, or $D$), since $DO$ is simply a mass balance of $P$, $R$, and $D$ (Equation 1)".*
+*Response: Sentence was revised as follows: "...the potential for the model to produce the right answer (accurate $DO$) for the wrong reasons (inaccurate $P$, $R$, or $D$), since $DO$ is simply a mass balance of $P$, $R$, and $D$ (Equation 1).
+This issue, or equifinality, has been encountered in stream metabolic models [@appling2018]."*
 

### 2024-03-06 manu-draft.qmd
revisions done but need to proof and add cran

@@
 A unified model to explain the large variability of metabolic rates within and across estuaries is clearly lacking as a critical tool for understanding metabolic processes.
-Bayesian statistical techniques have only been explored in limited estuarine applications despite their potential to address the above [@ciavatta2008; @tassone2019].
+Bayesian statistical techniques have only been explored in limited estuarine applications despite their potential to address the above [e.g., @tassone2019].
 Bayesian techniques allow for incorporation of prior knowledge about model parameters and robust estimation of parameter probability distributions [@Holtgrieve10] using highly efficient computational algorithms [@Grace15; @winslow2016].
 These techniques are very powerful because they not only estimate P and R as standard open-water methods do, but they retrieve estimates and their uncertainties of parameters related to gas exchange, P, and R (e.g., the initial slope of the photosynthesis--irradiance curve), which reveals quantitative information about the mechanisms of metabolism and gas exchange.
-A potentially useful application developed for streams is the BAyesian Single-station Estimation (BASE) method [@Grace15].
+@tassone2019 adapted a Bayesian approach developed for streams [@appling2018] in an estuarine setting to estimate metabolic rates in the James River.
+Although this approach demonstrated the utility of such an approach in estuaries, much of the core model structure developed for streams was unchanged and may not have been fully descriptive of fundamental processes influencing metabolism in open-water environments.
+Another potentially useful application developed for streams is the BAyesian Single-station Estimation (BASE) method [@Grace15].
 BASE has been demonstrated to accurately determine metabolic parameters and their uncertainty using continuous monitoring data that are similar to those available at many estuarine locations.
+The underlying model components can also easily be updated given the relatively simple format of the approach.
 As such, modification of BASE to include model characterizations more appropriate for estuaries, while maintaining the fundamental Bayesian approach, could prove useful.
@@
 We call this method EBASE (Estuarine BAyesian Single-station Estimation), which builds and improves on the BASE method [@Grace15] using an approach more appropriate for estuarine time series.
+We focus on BASE given the relative simplicity and, therefore, ease of application compared to other stream-based Bayesian approaches [e.g., @appling2018].
 The rationale and changes made to BASE are described in detail, followed by a demonstration of how EBASE can estimate known metabolic parameters with reasonable certainty from a synthetic time series.
@@
 We also attempt to constrain (i.e., accurately estimate) the gas transfer velocity using EBASE but are unable, at least with the data sets used in this study.
-EBASE is available as an open-source software package created with the R statistical programming language [@Beck23; @RCT23] and the results herein provide practical suggestions for use of the software on novel datasets.
+EBASE is available as an open-source software package created with the R statistical programming language \[@Beck23; @RCT23\] and the results herein provide practical suggestions for use of the software on novel datasets.
 
@@
 However, substantial variation in NSE values was observed among all other model outputs, with some very low values of NSE occurring while the $DO$ simulation was fairly good, showing the potential for the model to produce the right answer (accurate $DO$) for the wrong reasons (inaccurate $P$, $R$, or $D$), since $DO$ is simply a mass balance of $P$, $R$, and $D$ ([@eq-ebase]).
+This issue, or equifinality, has been encountered in stream metabolic models [@appling2018].
 After dissolved oxygen, gas exchange ($D$) was estimated with the highest median NSE of `r prmall[['D']]`.
@@
 The advantages of the Bayesian approach have led to widespread adoption over frequentist approaches in many scientific applications, especially in the coastal and marine community [e.g., @myers2001; @borsuk2001; @brown2017], yet these methods have not been rigorously explored for estimating metabolic rates in estuaries.
+
 Our results demonstrated that even using uninformed priors, EBASE can reproduce known metabolic parameters and rates with reasonable certainty.
@@
 In such cases, more constrained prior distributions for $b$ could be appropriate or EBASE could be tested in settings with expected larger contributions of gas exchange expressed by dissolved oxygen.
+For application, users of EBASE may simply use a fixed value for $b$ (e.g., 0.251 $\left(\text{cm}~\text{hr}^{-1}\right)/\left(\text{m}^2~\text{s}^{-2}\right)$) or perform a simple sensitivity analysis by varying $b$ +/- a small percentage to determine how $P$ and $R$ may change.
+Further model development may consider alternative approaches for dealing with equifinality, i.e., when the same modeled DO time series can be returned with different estimates of $P$, $R$, and $D$.
+@appling2018 provide a useful example for a stream metabolic model that addresses equifinality by allowing model parameters to vary with key drivers at sub-daily scales (e.g., gas exchange versus discharge) and avoiding assumptions that model residuals are uncorrelated.
 

### 2024-03-07 revision/review_response.qmd
add JA comments

@@
 However, additional comparisons are needed (e.g., Odum vs EBASE in a lake) and our thorough justification for the default prior distributions are based on the estuarine literature.
-This is still a valuable poin and we have added text to line 519:*
+This is still a valuable point and we have added text to line 519:*
 
@@
 Earlier iterations of model development did not have this constraint and the $b$ parameter was often estimated at unreasonably high values, often causing the other parameters ($a$ and $R$) to also be unreasonable.
-We previously described some rationale for why this parameters was so poorly constrained and offer some potential solutions for future research (i.e., lines 482-490, 502-509.*
+We previously described some rationale for why this parameters was so poorly constrained and offer some potential solutions for future research (i.e., lines 482-490, 502-509.)*
 
@@
 *Response: Thank you for these suggestions.
-We agree an explicit description of how the approach can be applied in light of uncertainties in the $b$ parameter is needed.
+We agree an explicit description of how the approach can be applied regarding uncertainties in the $b$ parameter is needed.
 Some discussion is already included (e.g., line 488).
-We have also added text describing how @appling2018 addressed the issue, although this suggestions would be further model development.
+We have also added text describing how @appling2018 addressed the issue, although this addition describes further model development rather than application.
 Revisions were made to the first paragraph of the discussion.*

### 2024-03-07 manu-draft.qmd
add JA comments

@@
 Bayesian statistical techniques have only been explored in limited estuarine applications despite their potential to address the above [e.g., @tassone2019].
-Bayesian techniques allow for incorporation of prior knowledge about model parameters and robust estimation of parameter probability distributions [@Holtgrieve10] using highly efficient computational algorithms [@Grace15; @winslow2016].
+Bayesian models allow for incorporation of prior knowledge about model parameters and robust estimation of parameter probability distributions [@Holtgrieve10] using highly efficient computational algorithms [@Grace15; @winslow2016].
 These techniques are very powerful because they not only estimate P and R as standard open-water methods do, but they retrieve estimates and their uncertainties of parameters related to gas exchange, P, and R (e.g., the initial slope of the photosynthesis--irradiance curve), which reveals quantitative information about the mechanisms of metabolism and gas exchange.
@@
 Lastly, metabolic estimates from EBASE are compared with those of the Odum open-water method to identify key differences related to both the theoretical and statistical differences associated with each method.
-These comparisons combine results using real observations and those based on detided inputs to demonstrate effects of tidal advection on metabolic estimates.
+These comparisons combine results using real observations and those based on detided dissolved oxygen inputs to demonstrate effects of tidal advection on metabolic estimates.
 We also attempt to constrain (i.e., accurately estimate) the gas transfer velocity using EBASE but are unable, at least with the data sets used in this study.
-EBASE is available as an open-source software package created with the R statistical programming language \[@Beck23; @RCT23\] and the results herein provide practical suggestions for use of the software on novel datasets.
+EBASE is available as an open-source software package created with the R statistical programming language [@Beck23; @RCT23] and the results herein provide practical suggestions for use of the software on novel datasets.
 
@@
 The optimization period used for a time series determines how much data are used to estimate the metabolic parameters.
-A short period of optimization (e.g., 1 day) may return unstable parameter estimates if forcing factors (e.g., wind, temperature) that affect estuarine biogeochemical rates occur at time scales longer the chosen period of optimization for EBASE.
+A short period of optimization (e.g., 1 day) may return unstable parameter estimates if forcing factors (e.g., wind, temperature) that affect estuarine biogeochemical rates occur at time scales longer than the chosen period of optimization for EBASE.
 As such, potentially longer optimization periods may return more stable and robust estimates given that more data are used for parameter estimation, although at the risk of overgeneralizing metabolic rates if biogeochemical variation occurs at time scales less than the optimization period.
-Results from optimization periods of 1, 7, and 30 days were compared to the synthetic one-year time series (\$b\$ was fixed for the sensitivity analysis) based on daily averages for both the synthetic and EBASE results.
-For the latter, parameters that are only returned as individual values for each optimization period (\$a\$, \$R\$, and \$b\$) were assigned as daily values for comparability of summary statistics where \$n\$ is the same across analyses.
+Results from optimization periods of 1, 7, and 30 days were compared to the synthetic one-year time series ($b$ was fixed for the sensitivity analysis) based on daily averages for both the synthetic and EBASE results.
+For the latter, parameters that are only returned as individual values for each optimization period ($a$, $R$, and $b$) were assigned as daily values for comparability of summary statistics where $n$ is the same across analyses.
 For the models with 7 and 30 day optimization periods, the final period in the annual time series was excluded from the results due to incomplete number of observations compared to earlier periods (i.e., the total number of days in the time series was not evenly divided by 7 or 30, resulting in a remainder for the last period).
@@
 The EBASE-derived rates of photosynthesis and respiration are both much higher than in the synthetic data for subplots (b) and (d), errors that compensate well enough to produce fairly accurate recoveries of the $DO$ time series.
-Further, resutls for the worst model for the 30 day optimization period in subplot (f) are visually indistinguishable from those of the best model in subplot (e), suggesting that poor estimates from inadequate priors can in part be mitigated using longer optimization periods.
+Further, results for the worst model for the 30 day optimization period in subplot (f) are visually indistinguishable from those of the best model in subplot (e), suggesting that poor estimates from inadequate priors can in part be mitigated using longer optimization periods.
 
@@
 We thank Darren Giling for initial conversations on applying BASE to estuarine data.
+We thank three anonymous reviewers that provided helpful feedback that improved the manuscript.
 This research was funded by award 1924559 from the National Science Foundation to co-PIs Raymond Najjar, Maria Herrmann, and Kathleen Hill.

### 2024-03-09 revision/review_response.qmd
add CRAN links and version to text and reviewer respones

@@
 *Response: It's the exception and not the norm that published software remains the same after a descriptive manuscript is published.
-For this reason, we have added a version of EBASE that is appropriate for this paper and included the version number in the text.
-EBASE is also now on CRAN, the Centralized R Archive Network, where a majority of publicly available R packages are hosted, in addition to the GitHub repository.
-We have also added text on line 193 indicating where source materials for this manuscript (i.e., to reproduce results) can be found: "All results provided herein were generated using the EBASE R package (specific materials for this manuscript can be found at <https://github.com/fawda123/ebase-manu>)."* EDIT
+For this reason, we have added a version of EBASE that is appropriate for this paper (v1.0.0) and included the version number in the text.
+EBASE is also now on [CRAN](https://cran.r-project.org/web/packages/EBASE/index.html), the Centralized R Archive Network, where a majority of publicly available R packages are hosted, in addition to the [GitHub repository](https://github.com/fawda123/EBASE).
+We have also added text on line 193 indicating where source materials for this manuscript (i.e., to reproduce results) can be found: "All results provided herein were generated using the EBASE R package (specific materials for this manuscript can be found at <https://github.com/fawda123/ebase-manu>)."*
 

### 2024-03-09 manu-draft.qmd
add CRAN links and version to text and reviewer respones

@@
 We also attempt to constrain (i.e., accurately estimate) the gas transfer velocity using EBASE but are unable, at least with the data sets used in this study.
-EBASE is available as an open-source software package created with the R statistical programming language [@Beck23; @RCT23] and the results herein provide practical suggestions for use of the software on novel datasets.
+EBASE is available as an open-source software package created with the R statistical programming language [<https://cran.r-project.org/web/packages/EBASE/index.html>, @Beck23; @RCT23] and the results herein provide practical suggestions for use of the software on novel datasets.
 
@@
 
-For ease of application, an R package was developed to implement EBASE [@Beck23; @RCT23], similar to the R package available for BASE (<https://github.com/dgiling/BASEmetab>).
+For ease of application, an R package was developed to implement EBASE [<https://cran.r-project.org/web/packages/EBASE/index.html>, @Beck23; @RCT23], similar to the R package available for BASE (<https://github.com/dgiling/BASEmetab>).
 EBASE differs from BASE in model implementation in two ways.
@@
 A detailed web page ([https://fawda123.github.io/EBASE/](#0){.uri}) describes how to use the package, with explanations of the functions for viewing model results, including a plot of the modeled dissolved oxygen with the observed, a time series plot of the oxygen budget terms (P, R, and D), and a time series plot of the credible intervals for the $a$, $R$, and $b$ parameters.
-All results provided herein were generated using the EBASE R package (specific materials for this manuscript can be found at <https://github.com/fawda123/ebase-manu>).
+All results provided herein were generated using the EBASE R package ([v1.0.0](https://cran.r-project.org/web/packages/EBASE/index.html), specific materials for this manuscript can be found at <https://github.com/fawda123/ebase-manu>).
 

### 2024-03-12 revision/review_response.qmd
MH comments added

@@
 
-*Response: We have noted an error in our description of Equation 3 that includes the depth-dependency.
+*Response: We have noted lack of clarity in our description of Equation 3 that includes the depth-dependency.
 As noted by Reviewer 3, @Grace15 do not include a depth-dependency in this equation.
-We have since removed the dependency and revised the text accordingly, emphasizing the importance of wind for the transfer velocity in open-water systems.
+We have since revised the equation and text accordingly, emphasizing the importance of wind for the transfer velocity in open-water systems.
 Regarding temperature, we feel justified in its inclusion for gas exchange given the previously noted relationship with the Schmidt number, a known factor to be important in open-water systems.
@@
 
-*Response: The depth-dependency for kw described in [@Grace15] was incorrect and an error on our part.
-We have removed $H$ (now $Z$) from the equation and updated the text accordingly, emphasizing the wind-dependence on the transfer velocity as a key factor influencing our choice to use the method from @Wanninkhof14: "However, open-water systems like estuaries are more susceptible to wind-based influences on gas exchange, particularly for $k_w$ [@Wanninkhof14; @vandam2019].
-Based on this difference and the lack of availability of other possible drivers of the transfer velocity in coastal systems, Equation 3 is not used herein.
-Instead, the well-established wind- and temperature-based parameterization for gas exchange of @Wanninkhof14 was used for EBASE..."*
+*Response: This was an inaccurate statement on our part.  We have updated the equation and text accordingly:*
+
+*Line 141: "Equation 1 is similar to the oxygen mass balance equation in BASE [@Grace15], with the main difference being the treatment of gas exchange.
+While both assume that gas exchange is proportional to the concentration gradient across the air–water interface, the treatments of the proportionality differ.
+@Grace15 used the following for the volumetric gas flux ($Dvol$) in BASE:*
+
+$$
+Dvol = [\, -K \cdot 1.024^{\left(T - \bar{T} \right)}\left(C_{sat} - C \right) ]\,
+$$
+
+*where $T$ is the instantaneous water temperature, $\bar{T}$ is the average water temperature over each 24-hour period in the oxygen time series (both in $^{\circ}$C), $C$ is the measured dissolved oxygen, and $C_{sat}$ is the saturation concentration.
+$K$ is the reaeration coefficient, which was estimated from the data.
+In EBASE, gas flux per unit surface area of the air-water interface is proportional to the concentration gradient, with the proportionality taken from the well-established wind- and temperature-based parameterization for gas exchange of @Wanninkhof14 (see Equation 2).
+EBASE also allows $b$ to be a free parameter that may reflect spatial and temporal variability in factors other than wind and temperature that may influence transfer velocity.
+@Wanninkhof14 estimated a fixed $b$ parameter as 0.251 $\left(\text{cm}~\text{hr}^{-1}\right)/\left(\text{m}^2~\text{s}^{-2}\right)$."*
 

### 2024-03-12 manu-draft.qmd
MH comments added

@@
 @eq-ebase is similar to the oxygen mass balance equation in BASE [@Grace15], with the main difference being the treatment of gas exchange.
-The gas transfer velocity, $k_w$ (@eq-kw), increases as the water column in the vicinity of the air--water interface becomes more turbulent.
-As such, $k_w$ is usually modeled as a function of measurable variables that can potentially predict turbulent mixing of the surface water, such as wind speed.
-$k_w$ also increases with temperature because the film of water at the air--water interface becomes thinner as the viscosity decreases and the random motion of gas molecules increases.
-This dependence of $k_w$ on temperature is usually modeled with the Schmidt number as in @eq-ebase, which combines information on the viscosity of the fluid and the diffusivity of the gas.
-
-@Grace15 used the following for the gas transfer in BASE:
+While both assume that gas exchange is proportional to the concentration gradient across the air–water interface, the treatments of the proportionality differ.
+@Grace15 used the following for the volumetric gas flux ($Dvol$) in BASE:
 
 $$
-k_w = K \cdot 1.024^{\left(T - \bar{T}\right)}
+Dvol = [\, -K \cdot 1.024^{\left(T - \bar{T} \right)}\left(C_{sat} - C \right) ]\,
 $$ {#eq-gasexbase}
 
-where $T$ is the instantaneous water temperature, $\bar{T}$ is the average water temperature over each 24-hour period in the oxygen time series (both in $^{\circ}$C), and $K$ is the reaeration coefficient estimated from the data.
-@eq-gasexbase indicates that $k_w$ increases with temperature, consistent with the known decrease of $k_w$ with the Schmidt number.
-However, open-water systems like estuaries are more susceptible to wind-based influences on gas exchange, particularly for $k_w$ [@Wanninkhof14; @vandam2019].
-Based on this difference and the lack of availability of other possible drivers of the transfer velocity in coastal systems, @eq-gasexbase is not used herein.
-Instead, the well-established wind- and temperature-based parameterization for gas exchange of @Wanninkhof14 was used for EBASE [@eq-kw], but allowing $b$, which was estimated by @Wanninkhof14 to be 0.251 $\left(\text{cm}~\text{hr}^{-1}\right)/\left(\text{m}^2~\text{s}^{-2}\right)$, to be a free parameter that may reflect spatial and temporal variability in factors other than wind and temperature that may influence transfer velocity.
+where $T$ is the instantaneous water temperature, $\bar{T}$ is the average water temperature over each 24-hour period in the oxygen time series (both in $^{\circ}$C), $C$ is the measured dissolved oxygen, and $C_{sat}$ is the saturation concentration.
+$K$ is the reaeration coefficient, which was estimated from the data.
+In EBASE, gas flux per unit surface area of the air-water interface is proportional to the concentration gradient, with the proportionality taken from the well-established wind- and temperature-based parameterization for gas exchange of @Wanninkhof14 (see @eq-kw).
+EBASE also allows $b$ to be a free parameter that may reflect spatial and temporal variability in factors other than wind and temperature that may influence transfer velocity.
+@Wanninkhof14 estimated a fixed $b$ parameter as 0.251 $\left(\text{cm}~\text{hr}^{-1}\right)/\left(\text{m}^2~\text{s}^{-2}\right)$.
 

### 2024-03-12 manu-draft.qmd
resubmitted to journal

@@
 We also attempt to constrain (i.e., accurately estimate) the gas transfer velocity using EBASE but are unable, at least with the data sets used in this study.
-EBASE is available as an open-source software package created with the R statistical programming language [<https://cran.r-project.org/web/packages/EBASE/index.html>, @Beck23; @RCT23] and the results herein provide practical suggestions for use of the software on novel datasets.
+EBASE is available as an open-source software package created with the R statistical programming language [<https://cran.r-project.org/web/packages/EBASE/index.html>, @Beck24; @RCT23] and the results herein provide practical suggestions for use of the software on novel datasets.
 
@@
 
-For ease of application, an R package was developed to implement EBASE [<https://cran.r-project.org/web/packages/EBASE/index.html>, @Beck23; @RCT23], similar to the R package available for BASE (<https://github.com/dgiling/BASEmetab>).
+For ease of application, an R package was developed to implement EBASE [<https://cran.r-project.org/web/packages/EBASE/index.html>, @Beck24; @RCT23], similar to the R package available for BASE (<https://github.com/dgiling/BASEmetab>).
 EBASE differs from BASE in model implementation in two ways.
@@
 
-EBASE also provides several computational advantages over BASE implemented in the R package that was created for the purposes herein [@Beck23].
+EBASE also provides several computational advantages over BASE implemented in the R package that was created for the purposes herein [@Beck24].
 Most importantly, EBASE was developed to allow different model optimization periods, whereas BASE was developed to estimate results for a single day.

### 2024-03-21 manu-draft.qmd
SE edits

@@
 $$
-Z\frac{dC}{dt} = P - R - D
+Z\frac{dC}{dt} = P - R + D
 $$ {#eq-ebase}
@@
 where the terms on the right side of the equation are gross primary production ($P$), ecosystem respiration ($R$), and the net upward diffusive gas flux at the air--water interface ($D$), respectively, processes that result in the change in dissolved oxygen concentration ($dC$) per unit time ($dt$).
-Both $P$ and $R$ are expressed as positive values, whereas $D$ can be positive (outgassing) or negative (ingassing).
+Both $P$ and $R$ are expressed as positive values, whereas $D$ can be positive (ingassing) or negative (outgassing).
 The equation has dimensions of moles per unit area and time (e.g., $\text{mmol}~\text{m}^{-2}~\text{d}^{-1}$).
@@
 The light efficiency has a complex structure, with maxima in the summer, fall, and winter, and a strong minimum in spring (Figure [-@fig-synapa]a).
-The oxygen budget terms (Figure [-@fig-synapa]b) reveal that (1) photosynthesis has a structure similar to, yet noisier than, that of light efficiency, reflecting the synoptic-scale variability of $PAR$; (2) respiration has the expected structure of being greatest during the summer [@caffrey2004], stimulated by greater autochthonous organic matter availability and high temperature; and (3) the air--water gas flux mimic productivity on short time scales, with ingassing during high productivity and outgassing during low productivity.
+The oxygen budget terms (Figure [-@fig-synapa]b) reveal that (1) photosynthesis has a structure similar to, yet noisier than, that of light efficiency, reflecting the synoptic-scale variability of $PAR$; (2) respiration has the expected structure of being greatest during the summer [@caffrey2004], stimulated by greater autochthonous organic matter availability and high temperature; and (3) the air--water gas flux mimic productivity on short time scales, with outgassing during high productivity and ingassing during low productivity.
 The synthetic dissolved oxygen time series shows high variability at multiple time scales (Figure [-@fig-synapa]c).

### 2024-03-21 manu-draft.qmd
change EBASE version

@@
 A detailed web page ([https://fawda123.github.io/EBASE/](#0){.uri}) describes how to use the package, with explanations of the functions for viewing model results, including a plot of the modeled dissolved oxygen with the observed, a time series plot of the oxygen budget terms (P, R, and D), and a time series plot of the credible intervals for the $a$, $R$, and $b$ parameters.
-All results provided herein were generated using the EBASE R package ([v1.0.0](https://cran.r-project.org/web/packages/EBASE/index.html), specific materials for this manuscript can be found at <https://github.com/fawda123/ebase-manu>).
+All results provided herein were generated using the EBASE R package ([v1.0.1](https://cran.r-project.org/web/packages/EBASE/index.html), specific materials for this manuscript can be found at <https://github.com/fawda123/ebase-manu>).
 
